### 2008-10-31 individual.tex
added individual.tex and .pdf
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@230 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%&LaTeX

\documentclass[a4paper,12pt,man]{apa} % nobf, doc, apa

\usepackage[]{amsmath, amsfonts, amstext, amsthm} 
\usepackage{amssymb}
\usepackage[]{graphics} 
\usepackage{graphicx} 
\usepackage{epsfig}
\usepackage{epstopdf}

\newcommand{\citep}{\cite} 
\newcommand{\citet}{\citeA}
\newcommand{\mat}{\mathbf} 
\newcommand{\vc}{\mathbf}

\newcommand{\pkg}{\texttt}
\newcommand{\code}{\texttt}

\renewcommand{\baselinestretch}{2}

\title{A Framework for Discrete Change}

\author{Ingmar Visser \& Maarten Speekenbrink}

\date{\today}

\affiliation{Department of Psychology, University of Amsterdam\\
 Correspondence concerning this article should be addressed to:
 \\
 Ingmar Visser \\
 Department of Psychology, University of Amsterdam \\
 Roetersstraat 15 \\
 1018 WB Amsterdam \\
 phone: +31 (20) 5256723 \\
 fax: +31 (20) 6390279 \\
 email: i.visser@uva.nl}

\shorttitle{Modelling Discrete Change}

\abstract{A class of models is developed for measuring and detecting
discrete change in learning and development.  The basic model for
detecting such change is the latent or hidden Markov model.
Traditionally, these models were restricted to categorical, mostly
binary, observed variables, placing severe restrictions on possible
measurement models.  In this paper, the basic model is extended to
include arbitrary distributions for the observed variables, including
multi-variate distributions.  Moreover, there is optional support to
include time-varying predictors.  In effect, this model consists of
mixtures of general linear models with Markovian depencies over time
to model the change process.  In addition, transition parameters can
be made to depend on covariates as well, such that the switching
regime between states depends on characteristics of the individual or
the experimental situation.  The model is illustrated with an example
of participants' learning in the weather prediction task.}

\begin{document}
\maketitle

% intro
Discrete change frequently occurs in learning and development: in
learning concepts, in performance on Piagetian tasks, in
discrimination learning and in conditioning.  This chapter is
concerned with detecting the time points of change in (individual)
time series.  We present a framework of dependent mixture models that can
be used to differentiate between gradual and discrete learning events
in individual time series data.  Before presenting the model in formal
terms and providing some illustrations, we first review some examples
in which discrete change is found.

% phenomena of discrete change in development/learning
Piagetian developmental theory (REFERENCE??)  assumes step-wise
changes in the strategies that children apply in all kinds of tasks
such as the conservation learning and the balance scale task.  Van der
Maas et al.  (1992) developed a catastrophe model to describe phase
transitions in learning and developmental processes.  They applied the
catastrophe model to learning in the conservation of liquid task
(REFERENCE??)  in which children have to judge relative volumes of
liquid in glasses of different heights and widths.  Young children
tend to ignore the width dimension and hence always choose the glass
with the highest level of liquid (REFERENCE??).  Van der Maas et al.
(1992) showed that there is a sudden transition to a new strategy in
which the children also take the width of the glasses into account
when judging the volume of liquids.

% balance scale rules
Jansen and Van der Maas (2001) applied the catastrophe model to
development of strategies on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 (Siegler, 1981).  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 (Siegler, 1981).

% hysteresis on the balance scale
Jansen and Van der Maas (2001) found clear evidence for stage-wise
transitions between Rule 1 and Rule 2 by testing criteria that were
derived from the catastrophe model.  In particular, they found bimodal
test scores and inaccessibility.  The latter means that there are no
in-between strategies: children apply either Rule 1 or Rule 2 and
there is no in-between option.  Jansen and Van der Maas also found
evidence for hysteresis: the phenomenon that switching between
strategies is assymetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different trials.  In particular, if the
distance dimension in the balance scale problems is made more salient
by increasing the distance difference between weights on either side
of the balance scale, children may switch from Rule 1 to Rule 2.  If
subsequently the distance difference is decreased again, children may
switch back to using Rule 1.  Hysteresis is the phenomenon that this
switch back occurs at a different value of the control variable, in
this case the distance difference.

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior (Gallistel et al 2004).  In particular,
in their study, evidence was found for sudden onset of learning: at
the start of the learning experiment, the pigeons did not learn
anything and performance was stable; after a number of trials,
learning kicks in and there are large increases in performance.  The
interest here is in modeling the distribution of onset times: that is,
the trials at which learning suddenly takes off.  A similar interest
in process onset times is found in addiction research.  For example,
(REFERENCE) study the age at which children start using alcohol and
how this related to eventual outcomes in terms of addiction.

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  This kind of learning is referred to as all-or-none
learning or concept identification learning.  Raijmakers et al (2001)
found evidence for different strategies applied by children when faced
with such a learning task.  Schmittmann et al (2006) reanalyzed their
data using hidden Markov models to show that both strategies are
characterized by sudden transitions in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example, Visser,
Raijmakers, \& Van der Maas (2008) analyzed data from a single
participant in an experimental task that manipulates the trade-off
between speed and accuracy.  The data consisted of three time series
with each around 150 repeated measurements of both reaction time and
accuracy.  Below we provide examples of analyzing time series from
single participants from two experiments; one from the Iowa Gambling
Task and one from the weather prediction task.  The interest in these
tasks is to show that participants develop different strategies over
time in responding to the stimuli and that the transition from one
strategy to the next is a discrete event.  Before providing these
illustrations, below we give a formalization of dependent mixture
models and a brief overview of the DepmixS4 package that was developed
to specify and fit such models.

\section{Dependent Mixture Models}

% outline
In this section we describe a class of models which are especially
suitable for describing and testing discrete change in (individual)
time series data.  The dependent mixture model is similar to, but
slightly different from two other types of models that are in use for
modelling discrete change: the hidden and the latent Markov models.

% history of Markov models/LMM applications
Markov models have been used extensively in the social sciences; for
example, in analyzing language learning \cite{Miller1952,Miller1963},
in the analysis of paired associate learning \cite{Wickens1982}.  In
these models, the focus is on survey type data: a few repeated
measurements taken in a large sample.  \citeA{Langeheine1990}
discusses latent Markov models and their use in sociology ands
political science (see also McCutcheon, 1987).  Latent transition
models, for example, been used in studying development of math skills
\citep{Collins1992} and in medical applications \citep{Reboussin1998};
\citet{Kaplan2008} provides an overview of such models, that are
called stage-sequential models in the developmental psychology
literature.

\nocite{McCutcheon1987}

% hidden Markov models
Hidden Markov models (HMM) tend to be used in the analysis of long
univariate and individual timeseries.  For example, HMMs are the model
of choice in speech recognition applications \cite{Rabiner1989}.  In
biology, HMMs are used to analyze DNA sequences and in econometric
science, to analyze changes in stock market prices and commodities
\cite{Kim1994}.

% depmix model
The dependent mixture model that we propose here spans the range from
latent Markov models for few repeated measurements with many
participants to hidden Markov models for individual time series.  In
addition, the dependent mixture model includes multivariate responses.
The dependent mixture model consists of the following elements:
\begin{enumerate}
	\item $S$ is a collection of discrete states
	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix
	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density 
\end{enumerate}

% state space
The state space, which is a set of discrete states, captures the
different states that the learning or developmental process under
consideration can be in.  In the balance scale example mentioned
above, children are applying one of two possible strategies in
responding to the items.  The states are characterized by their
corresponding observation densities.  Using for example Rule 1 in the
balance scale task leads to correct answers on some items and
incorrect answers on others.  A different strategy may lead to correct
answers on some items and to guessing behavior on other items.  In
analyzing categorization learning data, in which participants learn to
categorize a set of objects, a typical initial state is that
participants are guessing because at the start of the task they have
no knowledge of which features are important in categorization.

% transitions/matrix
The transition matrix $\mat{A}$ describes the transitions between
states over repeated measures or trials.  This matrix summarizes the
probabilities of transitioning from one state to another which
represents learning or development.  The transition model contains the
Markov assumption:
$$Pr(S_{t}|S_{t-1}, \ldots, S_{1}) = Pr(S_{t}|S_{t-1}),$$
which means that the current state (at time $t$) only depends on the
previous state $S_{t-1}$, and not on earlier states.

% observation densities
The observation densities $\mat{B}$ form the measurement part of the 
model; these describe the distributions of the observations 
conditional on the current state. Hence, these distributions 
characterize the state, and in our examples, these characterize the 
strategy that a participant is using at a given measurement occasion.

% log-likelihood
The log-likelihood of DMMs is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
Pr(\vc{O}_{t}|\vc{O}_{1}, 
	\ldots, \vc{O}_{t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

Note that so far no assumptions have been made about the response
distributions $b_{j}$, hence these can be arbitrary univariate or
multivariate distributions.

\section{DepmixS4}

% outline/abstract: this has redundant information, mostly discussed
% in detail below
\pkg{depmixS4} implements a general framework for defining and fitting
dependent mixture models in the R programming language \citep{R2008}.
This includes standard Markov models, latent/hidden Markov models, and
latent class and finite mixture distribution models.  The models can
be fitted on mixed multivariate data with multinomial and/or gaussian
distributions.  Parameters can be estimated subject to general linear
constraints.  Parameter estimation is done through an EM algorithm or
by a direct optimization approach with gradients using the
\code{Rdonlp2} optimization routine when contraints are imposed on the
parameters.

% design goals
The \pkg{depmixS4} package was motivated by the fact that Markovian
models are used commonly in the social sciences, but no comprehensive
package was available for fitting such models.  Common programs for
Markovian models include Panmark \citep{Pol1996}, and for latent class
models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
number of important features.  There are currently some packages in R
that handle hidden Markov models but they lack a number of features
that we needed in our research.  In particular, \pkg{depmixS4} was
designed to meet the following goals:
\begin{enumerate}

	\item to be able to fit transition models with covariates, i.e.,
	to have time-dependent transition matrices

	\item to be able to include covariates in the prior or initial
	state probabilities of models

	\item to allow for easy extensibility, in particular, to be able
	to add new response distributions, both univariate and
	multivariate, and similarly to be able to allow for the addition
	of other transition models, e.g., continuous time observation
	models

\end{enumerate}

Although \pkg{depmixS4} is designed to deal with longitudinal or time
series data, for say $T>100$, it can also handle the limit case with
$T=1$ in analyzing cross-sectional data.  In those cases, there are no
time dependencies between observed data, and the model reduces to a
finite mixture model, or a latent class model.  Although there are
other specialized (R) packages to deal with mixture data, one specific
feature that we needed which is not available in other packages is the
possibility to include covariates on the prior probabilities of class
membership.

\subsection{Reponse distributions and parameters}

% this needs some more on the design, interface to glm and similar
% other models etc. 
The package is built using S4 classes (object oriented classes in R)
to allow easy extensibility (REFERENCE to S4 methodology).

Each row of the transition matrix and the initial state probabilities:
\begin{itemize}
	\item is modeled as a multinomial distribution
	\item uses the logistic link function to include covariates
	\item can have time-dependent covariates
\end{itemize}

% response distributions include the possibility to have direct
% effects of covariates or predictors as well as indirect effects
% through the use of covariates on the transition matrix and the
% initial state probabilities
Current options for the response models are models from the
generalized linear modeling framework:

\begin{itemize}
	\item normal distribution; continuous, gaussian data
	\item binomial logistic; binary data
	\item Poisson distribution; count data
	\item multinomial logistic; multiple choice data
\end{itemize}

All response models have the option of including covariates
Other link functions may be used; eg the probit for binary data.

% possibility of multivariate responses: local independence or true
% multivariate distributions, eg multivariate normal

\subsection{Parameter Estimation}

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  The EM
algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization (although it is usually
faster than direct optimization at the start, so possibly a
combination of EM and direct optimization is fastest).  Second,
applying constraints to parameters can be problematic; in particular,
EM can lead to wrong parameter estimates when applying constraints.
Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
\cite{Tamura2007,Spellucci2002}, because it handles general linear
(in)equality constraints, and optionally also non-linear constraints.

\section{Illustrations}

% \subsection{Toy data: parameter retrieval} ????

Two illustrations are provided below of models that analyze single
participant time series data from two common experimental paradigms.
In both of these, participants learn different strategies through
trial and error.

\subsection{Iowa gambling task}

% what is the IGT? 
The Iowa gambling task (IGT) is an experimental paradigm designed to
mimic real-life decision-making situations (Bechara, Damasio, Damasio
\& Anderson, 1994), in the way that it factors uncertainty, reward and
punishment (Dunn, Dalgleish, \& Lawrence, 2006).  The task requires
the selection of cards from four decks.  Each deck is characterized by
a certain amount of gain (delivered on each draw), frequency of loss,
and amount of loss.  Two decks (A and B) yield consistently high
rewards, but also high, probabilistic penalties and are both (equally)
disadvantageous in the long run.  The other two decks (C and D) yield
consistently smaller rewards, but also low, probabilistic penalties
and are both (equally) advantageous in the long run.  It is assumed
that the ventromedial prefrontal cortex (VMPFC) is active in the IGT
as VMPC patients show impaired task performance.  Their preference for
the decks with immediate high rewards indicates ``myopia for the
future''.

% what is the HDT? developmental trends and relevance?
Crone \& van der Molen (2004) designed a developmentally appropriate
analogue of the IGT, the Hungry Donkey Task (HDT), with a similar win
and loss schedule although the abolute amounts were redcuced by a
factor of 25.  The HDT is a pro-social game inviting the player to
assist a hungry donkey to collect as many apples as possible, by
opening one of four doors.  Again, doors A and B are characterized by
a high constant gain (10 apples), whereas doors C and D deliver a low
constant gain (2 apples).  At doors A and C, a loss of 50 apples (A)
or 10 apples (C) is delivered in 50\% of the trials.  For doors B and
D, frequency of loss is only 10\%.  The median loss of doors B and D
is 10 and 2, respectively.  Crone and van der Molen administered the
HDT to children from four age groups (6-9, 10-12, 13-15, and 18-25
year-olds) and concluded that children also fail to consider future
consequences.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset (Huizenga, Crone, \& Jansen, 2007)
indicated that participants might solve the task by sequentially
considering the three dimensions (constant gain, frequency of loss,
and amount of loss) in order to choose a door.  Most youngest children
in the dataset seem to focus on the dominant dimension in the task,
frequency of loss, resulting in equal preference for doors B and D.
Older participants seem to use a two-dimensional rule where
participants first focus on the frequency of loss and then consider
amount of loss, resulting in a preference for door D. A third very
small subgroup seems to use an integrative rule where participants
combine all three dimensions in the appropriate way.  Participants
using the integrative rule pick cards from doors C and D, which are
advantageous in the long run.

% problematic aspects of this analysis
Typical analyses of these data use the last 60 trials in a series of
200 trials. A silent assumption that is made in these analyses is that
behavior has stabilized after 140 trials of learning; this could very 
well be wrong and it is highly likely that there are individual
differences in this learning process.

% proposed analysis here
Single participant choices analyzed.

\includegraphics[width=7cm]{graphs/igtdata4.pdf}

\begin{itemize}
	\item N=200
	\item 4 choice data, displayed in blocks of 20 trials
	\item optimal strategy is choosing C or D
\end{itemize}

Results are the 4-state model (best by AIC/BIC), model predicted 
probabilities are in the figure ???. States are characterized by 
different types of behavior, shifting from B/D strategy to C/D 
(optimal) strategy.

\includegraphics[width=7cm]{graphs/igtmodels4.pdf}

\subsection{Weather prediction task}

Still to add models and data.

\section{Discussion}

\begin{itemize}
	\item depmixS4 can be downloaded from: http://r-forge.r-project.org/depmix/
	\item It is feasible to fit hidden Markov models in moderate length time series
	\item Many applications in experimental psychology
	\item Future developments: 
	\begin{enumerate}
		\item richer measurement models, eg factor models, AR models etc
		\item richer transition models, eg continuous time measurement occasions
		\item explicit state durations
		\item identifiability of models
		\item model selection
	\end{enumerate}
	\item standard errors of parameters
\end{itemize}

References

Bechara, A., Damasio, A. R., Damasio, H., \& Anderson, S. W.(1994).
Insensitivity to future consequences following damage to human
prefrontal cortex.  Cognition, 50(1Ð3), 7Ð15.

Crone, E. A.,  \& van der Molen, M. W. (2004).  Developmental changes in
real life decision making: Performance on a gambling task previously
shown to depend on the ventromedial prefrontal cortex.  Developmental
Neuropsychology, 25(3), 251-279.

Dunn, B. D., Dalgleish, T.,  \& Lawrence, A. D. (2006).  The somatic
marker hypothesis: A critical evaluation.  Neuroscience and
Biobehavioral Reviews, 30(2), 239-271.

Huizenga, H. M., Crone, E. A., \& Jansen, B. R. J. (2007).
Decision-making in healthy children, adolescents and adults explained
by the use of increasingly complex proportional reasoning rules.
Developmental Science, 10(6), 814-825.

Siegler, R. S. (1981).  Developmental sequences within and between
concepts.  Monographs of the Society for Research in Child
Development, 46(2, Serial No.  189).

Visser, Raijmakers, \& Van der Maas (2008). Dynamics book chapter.

\section*{Author note}

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
Economic Learning and Social Evolution (ELSE).  Thanks to Hilde
Huizenga, Brenda Jansen and Anna van Duijvenvoorde for the Iowa
Gambling task data.

\bibliography{all,ingmar}

\end{document}

### 2008-11-03 papers/invidividual/individual.tex
added papers directory which is not complete
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@232 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2008-10-31 individual.tex
@@
 
-\subsection{Reponse distributions and parameters}
+\subsection{Response distributions and parameters}
 
@@
 The package is built using S4 classes (object oriented classes in R)
-to allow easy extensibility (REFERENCE to S4 methodology).
+to allow easy extensibility (Chambers, 1998).
 
@@
 
-Still to add models and data.
+%Still to add models and data.
+
+The Weather Prediction Task (WPT, Knowlton, Squire \& Gluck, 1994) 
+is a probabilistic categorization task, in which participants learn to 
+predict the state of the weather (sunny, or rainy) on the basis 
+of four ``tarot'' cards (cards with abstract geometrical patterns). 
+Each cue pattern is associated with a particular probability distribution 
+over the states of the weather. In order to perform in the task, 
+participants must predict the weather in accordance with these
+conditional probabilities.
+
+%The WPT has been popular in neuropsychological research, 
+%particularly because amnesic patients perform this task rather
+%well, despite not being able to remember actually many aspects
+%of the task (or in some cases, even performing the task at all).
+%This has led to the conclusion that probabilistic category learning
+%depends on implicit memory, which is separate from explicit
+%memory. While this conclusion is debatable (Speekenbrink, Channon 
+%\& Shanks, 2008), the finding of relatively unimpaired performance
+%by amnesic individuals remains striking.
+
+There are different accounts of probabilistic category learning. 
+According to instance or exemplar learning theories, participants
+learn by storing each encountered cue-outcome pairing. When presented
+with a cue pattern, these exemplars are retrieved from memory, and weighted
+according to their similarity to the probe cue pattern, to form a 
+classification. According to associative theories, participants gradually 
+learn by gradually associating the individual cues (or cue patterns in 
+configural learning) to the outcomes. In rule-learning, participants are 
+taken to extract rules by which to categorize the different cue patterns. 
+Gluck, Shohamy and Myers (2002) proposed a number of such rules (or strategies). A main
+difference between these is whether responses are based on the 
+presence/absence of a single cue, or whether responses are based on 
+cue patterns. Gluck et al. formulated all strategies in a deterministic and
+optimal manner (e.g., the multi-cue strategy corresponded to giving the optimal
+response to each cue pattern). Meeter et al. allowed for probabilistic 
+responding (a small probability of giving the non-optimal response).
+
+Alternative non-strategy based analyses of the WPT (Lagnado et al, Speekenbrink et al) 
+have estimated response strategies by variations of logistic regression.
+
+%associative, rule-based.  
+Here, we analyze the behavior of a single individual performing the
+WPT for 200 trials. We let each state be characterized by a Generalized
+Linear Model with a Bernoulli response and logistic link function. We are
+interested in whether a DMM can recover a strategy model in line with Gluck et al. 
+As we fit the data to a single subject, we must place some constraints.
+Specifically, we constrain the state transitions to be in a ``left-right'' format
+(states can only proceed to the immediately adjacent state and never back) and 
+the initial state
+
+A single state model (usual GLM) 
+We started with a 3-state model,
+
+\begin{itemize}
+	\item N=200
+	\item 4 choice data, displayed in blocks of 20 trials
+	\item optimal strategy is choosing C or D
+\end{itemize}
 
@@
 
+Chambers, J. M. (1998). Programming with Data: A Guide to the S Lan-
+guage. New York: Springer-Verlag.
+
 Crone, E. A.,  \& van der Molen, M. W. (2004).  Developmental changes in
@@
 
+Gluck, M. A., Shohamy, D., \& Myers, C. (2002). How do people solve the 
+weather prediction task?: Individual variability in strategies for 
+probabilistic category learning. Learning \& Memory, 9, 408-418.
+
 Huizenga, H. M., Crone, E. A., \& Jansen, B. R. J. (2007).
@@
 Developmental Science, 10(6), 814-825.
+
+Knowlton, B. J., Squire, L. R., \& Gluck, M. A. (1994). 
+Probabilistic classification learning in amnesia. 
+Learning \& Memory, 1 , 106-120.
 
@@
 Huizenga, Brenda Jansen and Anna van Duijvenvoorde for the Iowa
-Gambling task data.
+Gambling task data and David Lagnado for the Weather Prediction Task
+data.
 

### 2008-11-03 papers/jss/dmm4.tex
added papers directory (lots of fighting with svn ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%&LaTeX

\documentclass[a4paper,12pt,doc]{apaIngmar} % nobf, doc, apa, man

\usepackage[]{amsmath, amsfonts, amstext, amsthm}

\usepackage{amssymb}

\usepackage[]{graphics}

\usepackage{graphicx}

\usepackage{epsfig}

\usepackage{epstopdf}

\newcommand{\citep}{\cite}

\newcommand{\citet}{\citeA}

\newcommand{\mat}{\mathbf}

\newcommand{\vc}{\mathbf}

\title{DMM4: Dependent Mixture Models in S4 Classes}

\author{Ingmar Visser}
\date{\today}

\affiliation{Department of Psychology, University of Amsterdam\\

 Correspondence concerning this article should be addressed to:
 \\
 Ingmar Visser \\
 Department of Psychology, University of Amsterdam \\
 Roetersstraat 15 \\
 1018 WB Amsterdam \\
 phone: +31 (20) 5256735 \\
 fax: +31 (20) 6390279 \\
 email: i.visser@uva.nl}

\renewcommand{\baselinestretch}{1.5}

\shorttitle{DMM4: Dependent Mixture Models in S4}

\abstract{DMM4 implements mixture models and latent Markov models for
mixed indicator from various distributions.  Optimization can be
subjected to general linear constraints and is done using nlm, optim
or NPSOL. Covariates can be regressed on model parameters through
identity or logit link functions.}

\begin{document}

\bibliographystyle{apacite}

\maketitle

\newpage

In this paper an outline is provided for a program that computes
parameter estimates for the latent/hidden Markov mode (LMM) and for
latent transition analysis (LTA), both univariate and multivariate.
The difference between the two types of model is not a principled one
but rather of degree, being N is small and T large in LMM and vice
versa in LTA. Even so, when T is indeed large as is usually the case
in the hidden Markov model literature, the transition matrix is taken
to be stationary.  The basis for the current approach is to use the
hidden Markov model (HMM) estimation equations since these are
suitable for long time series which is not the case for the standard
EM algorithm used for latent class analysis (LCA) and LTA which soon
runs into the problem of sparse contingency tables and underflow
problems during computation.  The approach is illustrated using a
number of simulated and real data examples.

\section{Introduction}

The model that is considered in this paper is the multivariate (or
multi indicator) latent Markov model.  The main contribution of this
paper is the extension of estimation algorithms to allow for arbitrary
length time series of observations.  In traditional estimation methods
this is impossible due to underflow problems.  Moreover, the
estimation procedure that I present is not based on the complete
contingency table, which is impossible or impractical in the case of
long time series.  In particular, one might want to model a single
time series.  The usual approach to latent Markov models is then faced
with a huge contingency table, $2^{T}$ cells for a binary variable
measured at $T$ time points, in which only a single cell has a one and
the others are zero.  Howzthat for sparsity?

The model can be seen as an extension of several other models, in
particular, the hidden Markov model (HMM), the latent class model
(LCM), the latent transition model (LTM) and also the simple Markov
model (MM).  Markov models (and their subsequent refinements) have a
long history in psychology, mainly in the psychology of learning
\citep{Wickens1982}.  These models have in common that they are all
used to describe one or more sequences of responses $O_{1}, \ldots,
O_{T}$, where each of the $O_{t}$'s may be multivariate.  In this
paper mixed categorical and continuous observations are considered.

The model concerned in this paper is the combination of the latent
class model and the hidden Markov model.  There is quite some work
already on this model, but it is usually limited to small numbers of
repeated meausurements, in which case this model is usually called the
latent transition model: it is the repeated measurements extension of
the latent class model.  It is customary in those models to have
time-varying (transition) parameters.  In the time series literature
on the other hand, where hidden Markov models are frequently used,
transition parameters are taken to be time-homogeneous.  This
assumption is necessary when analyzing single time-series since at
each time $t$ only a single transition is observed.  The multi-variate
latent Markov model is defined by the following elements:
\begin{enumerate}
	\item a set $\vc{S}$ of latent states $S_{i},\, i=1, \ldots , n$,

	\item a matrix $\mat{A}$ of transition probabilities $a_{ij}$ for
	the transition from state $S_{i}$ to state $S_{j}$,

	\item a set $\vc{B}$ of observation functions $b_{j}(\cdot)$ that
	provide the conditional probabilities associated with latent state
	$S_{j}$,

	\item a vector $\pmb{\pi}$ of latent state initial probabilities
	$\pi_{i}$.
\end{enumerate}
When transitions are added to the latent class model, it is more
appropriate to refer to the classes as states.  The word class is
rather more associated with a stable trait-like attribute whereas a
state can change over time.

The data has the general form $O_{1}^{1}, \ldots, O_{1}^{m},
O_{2}^{1}, \ldots, O_{2}^{m}, \ldots, O_{T}^{1}, \ldots, O_{T}^{m}$
for an $m$-variate time series of length $T$ and independent
realizations of such series.  It is important to note that the latent
Markov model is commonly associated with data of this type, but the
estimation procedures that are used are not suitable for long time
series due to underflow problems.  Conversely the hidden Markov model
is only used for univariate time series and multiple dependent time
series or multivariate time series have not been considered in
previous literature on hidden Markov models.  In the next section, the
likelihood and estimation procedure for this model is described, given
data of the above form, for arbitrary $m$ and $T$.

\section{Computing likelihood, gradients and Hessian}

Single-indicator latent class models.  Estimation logic is different:
the starting point is not a complete contingency table, which is then
weeded.  The starting point is a stochastic automaton which produces
time series.  In time series analysis, one does not expect a complete
contingency table; in the contingency table logic, a single time
series produces an enormous table with only a single filled cell.

For computing the likelihood and its partial derivatives, I follow the
strategy used by \cite{Lystig2002}.  They developed a method to
compute exact gradients and information matrix by reformulating the
forward procedure of the Baum-Welch algorithm for estimating hidden
Markov model parameters.  Scores and information are computed at a
single pass through the data, using the intermediate variables from
the forward recursion.  Their article only discusses univariate time
series.  Our aim in this section is to generalize their results to
multiple dependent and independent time series.  With the gradients
available, it is possible to use general purpose optimization routines
that allow box constraints and arbitrary (non-) linear constraints.

\subsection{Likelihood}

The loglikelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
change the forward algorithm in such a way as to allow computing the
gradients of the loglikelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
	Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1}),
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$.  Note that for a
simple Markov chain these probabilities reduce to
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, \vc{O}_{t-1}) =
Pr(\vc{O}_{t}|\vc{O}_{t-1})$.  The log-likelihood can now be expressed
as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
	\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j}
	b_{j}(\vc{O}_{1}) \label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots,
	\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times
	(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$.  Combining
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and
equation~(\ref{eq:condLogl}) gives the following expression for the
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
\end{equation}

The above forward recursion can readily be generalized to mixture
models, in which it is assumed that the data are realizations of a
number of different LMMs and the goal is to assign posterior
probabilities to sequences of observations.  This situation occurs,
for example, in learning data where different learning strategies may
lead to different answer patterns.  From an observed sequence of
responses, it may not be immediately clear from which learning process
they stem.  Hence, it is interesting to consider a mixture of latent
Markov models which incorporate restrictions that are consistent with
each of the learning strategies.

\subsection{Likelihood for mixtures}

The easiest way to compute the likelihood of a mixture of LMMs is by
realizing that such a mixture is itself an LMM with a number of
constraints on the initial parameters and the transition matrix.  For
example, consider a mixture of two LMMs, in which one component has
two states and the other three states.  Later in this paper I will
present an example of such a mixture when analyzing discrimination
learning data.  The assumption is that children and adults use
different strategies in such a task which can be translated into
different latent Markov models.  In this example the first component
has two states and the following parameters:
\begin{gather*}
\mat{A} =\begin{pmatrix} 
1 & 0\\ \alpha & 1-\alpha
\end{pmatrix} \quad \text{and} \quad
\pmb{\pi}=\begin{pmatrix} \pi_{1}\\ \pi_{2} \end{pmatrix}.
\end{gather*}
The response parameters $b$ are all to freely estimated. The second 
component has three  states and the following parameters:
\begin{gather*}
\mat{A} =\begin{pmatrix} 
1 & 0 & 0\\ 0 & g & 1-g \\ \alpha & \beta & \gamma
\end{pmatrix} \quad \text{and} \quad
\pmb{\pi}=\begin{pmatrix} \pi_{1}\\\pi_{2}\\\pi_{3}  \end{pmatrix}.
\end{gather*}
By combing the transition matrices and the initial vectors the 
following model results:
\begin{gather*}
\mat{A} =\begin{pmatrix} 
1 & 0 & 0 & 0 & 0\\
\alpha & 1-\alpha & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0
\\0 & 0 & 0 & g & 1-g 
\\ 0 & 0 & \alpha & \beta & \gamma
\end{pmatrix} \quad \text{and} \quad
\pmb{\pi}=\begin{pmatrix} \pi_{1}^{1}\\ \pi_{2}^{1}  \\ 
\pi_{1}^{2}\\\pi_{2}^{2}\\\pi_{3}^{2} \end{pmatrix},
\end{gather*}
whith the constraint that the $\pi$'s sum to unity.  It can easily be
seen that computing the log-likelihood of such a model leads to many
unneccessary computations because the transition matrix $\mat{A}$
contains blocks of zeroes that have to be multiplied $T$ times for
each time series under consideration.  Therefor, below the
log-likelihood is computed as a real mixture of the components above
with mixture parameters $p_{k}$.

To compute the likelihood of a mixture of $K$ models, define the
forward recursion variables as follows (these variables now have an
extra index $k$ indicating that observation and transition
probabilities are from latent model $k$):
\begin{align}
\begin{split}
\phi_{1}(j_{k}) &=  Pr(\vc{O}_{1}, 
S_{1}=j_{k})=p_{k}\pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}).
\end{split}\label{eq:fwd1mix} \\
\begin{split}
\phi_{t}(j_{k})   &=   Pr(\vc{O}_{t}, S_{t}=j_{k}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
			&= \left[ \sum_{k=1}^{K} \sum_{i=1}^{n_{k}} \phi_{t-1}(i_{k}) 
			a_{ij_{k}}b_{j_{k}}(\vc{O}_{t}) \right] \times (\Phi_{t-1})^{-1},
\end{split}\label{eq:fwdtmix} 
\end{align}
where $\Phi_{t} =  \sum_{k=1}^{K}\sum_{i=1}^{n_{k}} 
\phi_{t}(j_{k})$.  Note that the double sum over $k$ and $n_{k}$ is 
simply an enumeration of all the states of the model. Now, because 
$a_{ij_{k}}=0$ whenever $S_{i}$ is not part of component $k$, the sum 
over $k$ can be dropped and hence equation~\ref{eq:fwdtmix} reduces 
to:
\begin{equation}
	\phi_{t}(j_{k}) = \left[ \sum_{i=1}^{n_{k}} \phi_{t-1}(i_{k}) 
			a_{ij_{k}}b_{j_{k}}(\vc{O}_{t}) \right] \times (\Phi_{t-1})^{-1}
\end{equation}

Writing this out in matrices and vectors and rearranging provides 
useful expressions for implementational  purposes, for each $k$ we 
get:
\begin{align}
	\vc{\phi}_{1} &=  p_{k} \times \pmb{\pi} \otimes 
\vc{b}(\vc{O}_{1}),\\
	\vc{\phi}_{t} &= \left[ \mat{A}^{t} \times \pmb{\phi}_{t-1} \right ] 
		\otimes \vc{b} (\vc{O}_{t}) \times (\Phi_{t-1})^{-1},
\end{align}
where $^{t}$ denotes matrix tranpose, $\times$ denotes matrix/vector 
multiplication and $\otimes$ denotes the Hadamard product of 
vectors/matrices.

\paragraph{Computational considerations} 
From equations~(\ref{eq:fwd1}--\ref{eq:fwdt}), it can be seen that
computing the forward variables, and hence the log-likelihood, takes
$O(Tn^{2})$ computations, for an $n$-state model and a time series of
length $T$.  Consider the mixture of two components, one with two
states and the other with three states, as in the example above.
Using equations~(\ref{eq:fwd1}--\ref{eq:fwdt}) to compute the
log-likelihood of this model one needs $O(Tn^{2})=O(T\times 25)$
computations whereas with the mixture
equationsequations~(\ref{eq:fwd1mix}--\ref{eq:fwdtmix}), one only
needs $\sum_{n_{i}} O(n_{i}^{2}T)$ computations, in this case $O(T
\times 13)$.  So, it can be seen that in this easy example the
computational cost is almost halved.

Note that in above equations $b_{j}(\vc{O}_{t})$ factors into
$\prod_{i=1}^{m} b_{j} (O_{t}^{i})$, as is the case in the latent
class model.

%\subsection{Posterior probabilities}

%Viterbi algorithm?

% %Using the equations above for the likelihood of single and mixture 
% models, arriving at the posterior probabilities of membership of the 
% components can be easily accomplished.

% %The posterior probability is the probability that data $\vc{O}$ are 
% generated by mixture component $k$, i.e., $Pr(C=k|\vc{O})$, where $C$ 
% denotes the mixture component. This probability can be expressed as:
%\begin{eqnarray}
%	Pr(C=k|\vc{O}) &=& \frac{Pr(C=k) Pr(\vc{O}|C=k)}
%	{\sum_{k=1}^{K} Pr(C=k)Pr(\vc{O}|C=k)} \\
%	&=& \frac {p_{k} l_{TMC=k}}
%	{\sum}
%\end{eqnarray}

\subsection{Gradients}

\newcommand{\fpp}{\frac{\partial} {\partial \lambda_{1}}}

See equations 10--12 in \cite{Lystig2002} for the score recursion
functions of the hidden Markov model for a univariate time series.
Here the corresponding score recursion for the multivariate mixture
case are provided.  The $t=1$ components of this score recursion are
defined as (for an arbitrary parameter $\lambda_{1}$):
\begin{align}
\psi_{1}(j_{k};\lambda_{1}) &:=  \fpp Pr(\vc{O}_{1}|S_{1}=j_{k}) \\
\begin{split} 
	&= \left[  \fpp p_{k} \right] \pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}) + 
	p_{k}\left[ \fpp \pi_{j_{k}} \right] b_{j_{k}}(\vc{O}_{1}) \\
	& \qquad  + p_{k}\pi_{j_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{1})\right],
\end{split} \label{eq:psi1}
\end{align}
and for $t>1$ the definition is:
\begin{align}
	\psi_{t}(j_{k};\lambda_{1}) & = \frac{\fpp Pr(\vc{O}_{1}, \ldots,
	\vc{O}_{t}, S_{t}=j_{k})} {Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}
	\\
	\begin{split} 
		& = \sum_{i=1}^{n_{k}} \Bigg\{
		\psi_{t-1}(i;\lambda_{1})a_{ij_{k}} b_{j_{k}}(\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right]
			b_{j_{k}} (\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
			(\vc{O}_{t}) \right] \Bigg\} 
			\times (\Phi_{t-1})^{-1}.
		\end{split} 
	\label{eq:psit}
\end{align}

Writing this out in matrix notation (and dropping the $k$-subscripts 
for components) we get for $t=1$ and for each $k$:
\begin{eqnarray}
	\pmb{\psi}_{1}(\lambda_{1}) = \vc{\fpp} p_{k} \times \pmb{\pi} 
	\otimes \vc{b}(O_{1})+ p_{k} \times \vc{\fpp}\pmb{\pi} \otimes
	\vc{b}(O_{1}) + p_{k} \times \pmb{\pi} \otimes \fpp \vc{b}(O_{1}),
\end{eqnarray}
and for $\pmb{\psi}_{t}$ we get: 
\begin{equation}
	\begin{split} 
		\pmb{\psi}_{t}(\lambda_{1}) &= \left[ \mat{A}^{t} \times
		\pmb{\psi_{t-1}}(\lambda_{1}) \right ] \otimes \vc{b}(O_{t}) +
		\left [ \left[ \fpp \mat{A} \right ]^{t} \times
		\pmb{\phi}_{t-1} \right ] \otimes \vc{b}(O_{t}) \\
		& \qquad + \left [ \mat{A}^{t} \times \pmb{\phi}_{t-1} \right]
		\otimes \fpp \vc{b}(O_{t}).
	\end{split}
\end{equation}

Using above equations, \cite{Lystig2002} derive the following equation 
for the partial derivative of the likelihood:
\begin{equation}
	\fpp l_{T}=
	\frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}},
\end{equation}
where $\Psi_{t}=\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} 
\psi_{t}(j_{k};\lambda_{1})$. 
Starting from the equation from the logarithm of the likelihood, this 
is easily seen to be correct: 
\begin{eqnarray}
	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1},
	\ldots, \vc{O}_{T})^{-1} \fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T})
	\\
	&=& Pr(\vc{O}_{1}, \ldots, \vc{O}_{T})^{-1} \Psi_{T}(\lambda_{1})
	Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1}) \\
	&=& \frac{Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1})}{Pr(\vc{O}_{1},
	\ldots, \vc{O}_{T})} \Psi_{T} (\lambda_{1}) \\
	&=&  \frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}}.
\end{eqnarray}

Having provided the global computations above we now need expressions
for $\fpp p_{k}$, $\fpp \pi_{i}$, $\fpp a_{ij}$, and $\fpp
b_{i}(\vc{O}_{t})$, which are dependent on the particular
$\lambda_{1}$ under consideration.  This generates special cases for
each combination of $\lambda_{1}$ and each of the components of the
gradient computation.  These cases are worked out below for each of
the choices $\lambda_{1} \in \pmb{\lambda}=\{
\vc{p},\pmb{\pi},\mat{A},\vc{b} \}$.

%%p-k cases
\paragraph{Case 1: $\fpp p_{k}$}

\begin{align} 
        \fpp p_{k} = \left\{ \begin{array}{ll} 
            1 & \mbox{if $\lambda_{1}=p_{k}$} \\ 
            0 & \mbox{otherwise} 
            \end{array} \right. 
\end{align}

%%pipi-k cases

\paragraph{Case 2: $\fpp \pi_{i}$ }

\begin{align} 
        \fpp \pi_{i}^{k} = \left\{ \begin{array}{ll} 
            1 & \mbox{if $\lambda_{1}=\pi_{i}^{k}$} \\ 
            0 & \mbox{otherwise} 
            \end{array} \right. 
\end{align}

\paragraph{Case 3: $\fpp a_{ij}$ }

\begin{align} 
        \fpp a_{ij}^{k}= \left\{ \begin{array}{ll} 
            1 & \mbox{if $\lambda_{1}=a_{ij}^{k}$} \\ 
            0 & \mbox{otherwise} 
            \end{array} \right. 
\end{align}

The cases below for $\fpp b_{j}(\vc{O}_{t})$ are slightly more
interesting.  First, remember that in the case of an $m$-variate time
series of observations, $\fpp b_{j}(\vc{O}_{t})=\fpp b_{j}(O_{t}^{1},
\ldots, O_{t}^{m})$, of which we need partial derivatives for
arbitrary parameters $\lambda_{1}$.  First note that because of local
independence we can write:
\begin{equation}
	\fpp [b_{j}(O_{t}^{1}, \ldots, O_{t}^{m})]= \frac{\partial} 
{\partial \lambda_{1} } [b_{j}(O_{t}^{1})] \times  
[b_{j}(O_{t}^{2})], \ldots,  [b_{j}(O_{t}^{m})].  
\end{equation}
Applying the chain rule for products we get:
\begin{equation}
	\fpp [b_{j}(O_{t}^{1}, \ldots, O_{t}^{m})] =
	\sum_{l=1}^{m} \left[ \prod_{i=1, \ldots, \hat{l}, \ldots, m} 
b_{j}(O_{t}^{i}) \right] \times
	\fpp  [b_{j}(O_{t}^{l})],
	\label{partialProd}
\end{equation}
where $\hat{l}$ means that that term is left out of the product.
These latter terms, $\frac{\partial} {\partial \lambda_{1} }
[b_{j}(O_{t}^{k})]$, are easy to compute given either multinomial or
gaussian observation densities.  As a consequence, this provides
gradients for the multivariate time series model for mixed categorical
and gaussian observations, e.g.\ correct/incorrect responses and
reaction times, and a categorical latent state or class.  Hence, we
need the partial derivatives $\fpp b_{j}({O}_{t}^{l})$ instead of
$\fpp b_{j}(\vc{O}_{t})$ (if $\lambda_{1} \in \{ \vc{p},
\pmb{\pi},\mat{A} \} $, then $\fpp b_{j}=0$).

\paragraph{Case 4: $\fpp b_{j}({O}_{t}^{l})$ and $\lambda_{1} \in
\vc{b}$.} These derivatives depend on the response function of the
item, i.e.\ whether the item is multinomial or gaussian (other
itemtypes are easy to add).  For the derivatives of the response
functions there are two cases depending on the type of the response
function.  For a gaussian item we have the following:
\begin{align} 
	\fpp b_{j_{k}}(O_{t}^{l})= \left\{
	\begin{array}{ll} 
		b_{j_{k}}(O_{t}^{l}) \times \left[
		\frac{O_{t}-\mu}{\sigma^{2}} \right] & \mbox{if $\lambda_{1} =
		\mu_{j_{k}}^{l} $ } 
		\\
		b_{j_{k}}(O_{t}^{l}) \times \left[
		\frac{(O_{t}-\mu)^{2}}{\sigma^{3}} - \frac{1}{\sigma} \right]
		& \mbox{if $\lambda_{1} = \sigma_{j_{k}}^{l} $} 
		\\
		0 & \mbox{otherwise} \end{array} \right.
\end{align}

For an $c$-category item $l$ with response probabilities $p_{1}, 
\ldots, p_{c}$ we have the following:
\begin{align} 
	\fpp b_{j_{k}}(O_{t}^{l})= \left\{ 
	\begin{array}{ll} 
		1 & \mbox{if $\lambda_{1} = p_{i}$ \& $O_{t}=i$}  \\ 
		0 & \mbox{otherwise} 
	\end{array} 
	\right\}
\end{align}

%\newpage
%\section*{Author note}
%We would like to thank many people.

\bibliography{all}

\end{document}

### 2008-11-03 papers/jss/hmmlta.tex
added papers directory (lots of fighting with svn ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2008-11-03 papers/jss/dmm4.tex
@@
 
-\documentclass[a4paper,12pt,doc]{apaIngmar} % nobf, doc, apa, man
+\documentclass[a4paper,man,nobf]{apa}
+ \usepackage[]{amsmath, amsfonts, amstext, amsthm}
+ \usepackage[oztex, final]{graphics}
+ \usepackage[]{epsfig}
 
-\usepackage[]{amsmath, amsfonts, amstext, amsthm}
+\usepackage[round]{natbib}
 
-\usepackage{amssymb}
+\title{A comprehensive approach to fitting multivariate latent Markov 
+models and latent transition models: An implementation and 
+applications}
 
-\usepackage[]{graphics}
+\author{Ingmar Visser}
+\affiliation{Developmental Processes Research Group, Department of 
+Psychology, University of Amsterdam}
+\date{\today}
 
-\usepackage{graphicx}
+% \acknowledgements{Author notes etc.}   
+ \rightheader{Depedent Mixtures}                    % defaults to title
+\shorttitle{Depedent Mixtures}                      % defaults to title
+% \leftheader{Authors}
 
-\usepackage{epsfig}
+%\batchmode
 
-\usepackage{epstopdf}
+%\newcommand{\citep}{\cite}
+%\newcommand{\citet}{\citeA}
 
-\newcommand{\citep}{\cite}
-
-\newcommand{\citet}{\citeA}
-
+\newcommand{\vc}{\mathbf}
 \newcommand{\mat}{\mathbf}
 
-\newcommand{\vc}{\mathbf}
+\abstract{In this article, an outline is presented for a program that 
+computes parameter estimates with standard errors for the 
+multivariate hidden Markov model. The model itself can be seen as an 
+extension and generalization of both the hidden or latent Markov 
+model and latent class models. The parameter procedure is different 
+from the traditional approach with these models. This approach allows 
+modeling of arbitrarily long time series efficiently and without 
+exponentially growing computational effort. Moreover, because 
+analytical gradients and Hessian are available, models can easily be 
+fitted subject to arbitrary linear and non-linear constraints. 
+Multi-group and mixture analyses are also possible. A number of 
+illustrative examples are presented: 1) mixture analysis of 
+(Markovian) learning data , 2) analysis of sleep stages from EEG, and 
+3) intra- and inter-individual differences in personality structure.}
 
-\title{DMM4: Dependent Mixture Models in S4 Classes}
-
-\author{Ingmar Visser}
-\date{\today}
-
-\affiliation{Department of Psychology, University of Amsterdam\\
-
- Correspondence concerning this article should be addressed to:
- \\
- Ingmar Visser \\
- Department of Psychology, University of Amsterdam \\
- Roetersstraat 15 \\
- 1018 WB Amsterdam \\
- phone: +31 (20) 5256735 \\
- fax: +31 (20) 6390279 \\
- email: i.visser@uva.nl}
-
-\renewcommand{\baselinestretch}{1.5}
-
-\shorttitle{DMM4: Dependent Mixture Models in S4}
-
-\abstract{DMM4 implements mixture models and latent Markov models for
-mixed indicator from various distributions.  Optimization can be
-subjected to general linear constraints and is done using nlm, optim
-or NPSOL. Covariates can be regressed on model parameters through
-identity or logit link functions.}
+%\keywords{latent Markov model; latent class model; latent transition model; mixtures; gradients; 
+%%%% Hessian; estimation}
 
@@
 
-\bibliographystyle{apacite}
+%\bibliographystyle{plainnat}
 
 \maketitle
-
-\newpage
 
@@
 multi indicator) latent Markov model. The main contribution of this 
-paper is the extension of estimation algorithms to allow for arbitrary
-length time series of observations.  In traditional estimation methods
-this is impossible due to underflow problems.  Moreover, the
-estimation procedure that I present is not based on the complete
-contingency table, which is impossible or impractical in the case of
-long time series.  In particular, one might want to model a single
-time series.  The usual approach to latent Markov models is then faced
-with a huge contingency table, $2^{T}$ cells for a binary variable
-measured at $T$ time points, in which only a single cell has a one and
-the others are zero.  Howzthat for sparsity?
+paper is the extension of estimation algorithms to allow for 
+arbitrary length time series of observations. In traditional 
+estimation methods this is impossible due to underflow problems. 
+Moreover, the estimation procedure that I present is not based on the 
+complete contingency table, which is impossible or impractical in the 
+case of long time series. In particular, one might want to model a 
+single time series. The usual approach to latent Markov models is 
+then faced with a huge contingency table, $2^{T}$ cells for a binary 
+variable measured at $T$ time points, in which only a single cell has 
+a one and the others are zero. Howzthat for sparsity?
 
@@
 long history in psychology, mainly in the psychology of learning 
-\citep{Wickens1982}.  These models have in common that they are all
-used to describe one or more sequences of responses $O_{1}, \ldots,
-O_{T}$, where each of the $O_{t}$'s may be multivariate.  In this
-paper mixed categorical and continuous observations are considered.
+\citep{Wic82}. These models have in common that they are all used to 
+describe one or more sequences of responses $O_{1}, \ldots, O_{T}$, 
+where each of the $O_{t}$'s may be multivariate. In this paper mixed 
+categorical and continuous observations are considered.
 
@@
 already on this model, but it is usually limited to small numbers of 
-repeated meausurements, in which case this model is usually called the
-latent transition model: it is the repeated measurements extension of
-the latent class model.  It is customary in those models to have
-time-varying (transition) parameters.  In the time series literature
-on the other hand, where hidden Markov models are frequently used,
-transition parameters are taken to be time-homogeneous.  This
-assumption is necessary when analyzing single time-series since at
-each time $t$ only a single transition is observed.  The multi-variate
-latent Markov model is defined by the following elements:
+repeated meausurements, in which case this model is usually called 
+the latent transition model: it is the repeated measurements 
+extension of the latent class model. It is customary in those models 
+to have time-varying (transition) parameters. In the time series 
+literature on the other hand, where hidden Markov models are 
+frequently used, transition parameters are taken to be 
+time-homogeneous.  This assumption is necessary when analyzing single 
+time-series since at each time $t$ only a single transition is 
+observed. The multi-variate latent Markov model is defined by the 
+following elements:
 \begin{enumerate}
   	\item a set $\vc{S}$ of latent states $S_{i},\, i=1, \ldots , n$,
-
   	\item a matrix $\mat{A}$ of transition probabilities $a_{ij}$ for 
-	the transition from state $S_{i}$ to state $S_{j}$,
-
+the transition 
+	from state $S_{i}$ to state $S_{j}$,
   	\item a set $\vc{B}$ of observation functions $b_{j}(\cdot)$ that 
@@
 $S_{j}$, 
-
   	\item a vector $\pmb{\pi}$ of latent state initial probabilities 
@@
 previous literature on hidden Markov models. In the next section, the 
-likelihood and estimation procedure for this model is described, given
-data of the above form, for arbitrary $m$ and $T$.
+likelihood and estimation procedure for this model is described, 
+given data of the above form, for arbitrary $m$ and $T$.
 
@@
 
-For computing the likelihood and its partial derivatives, I follow the
-strategy used by \cite{Lystig2002}.  They developed a method to
+For computing the likelihood and its partial derivatives, I follow 
+the strategy used by \cite{Lys2002}. They developed a method to 
 compute exact gradients and information matrix by reformulating the 
@@
 multiple dependent and independent time series. With the gradients 
-available, it is possible to use general purpose optimization routines
-that allow box constraints and arbitrary (non-) linear constraints.
+available, it is possible to use general purpose optimization 
+routines that allow box constraints and arbitrary (non-) linear 
+constraints.
 
@@
 The loglikelihood of hidden Markov models is usually computed by the 
-so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
-rather by the forward part of this algorithm.  \cite{Lystig2002}
-change the forward algorithm in such a way as to allow computing the
-gradients of the loglikelihood at the same time.  They start by
-rewriting the likelihood as follows (for ease of exposition the
-dependence on the model parameters is dropped):
+so-called forward-backward algorithm \citep{Bau66,Rab89}, or rather 
+by the forward part of this algorithm.
+\cite{Lys2002} change the forward algorithm in such a way as to allow 
+computing the gradients of the loglikelihood at the same time. They 
+start by rewriting the likelihood as follows (for ease of exposition 
+the dependence on the model parameters is dropped): 
 \begin{equation}
 	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
-	Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1}),
+Pr(\vc{O}_{t}|\vc{O}_{1}, 
+	\ldots, \vc{O}_{t-1}), 
 	\label{condLike}
@@
 simple Markov chain these probabilities reduce to 
-$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, \vc{O}_{t-1}) =
-Pr(\vc{O}_{t}|\vc{O}_{t-1})$.  The log-likelihood can now be expressed
-as:
+$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
+\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
+The log-likelihood can now be expressed as:
 \begin{equation}
@@
 
-To compute the log-likelihood, \cite{Lystig2002} define the following 
+To compute the log-likelihood, \cite{Lys2002} define the following 
 (forward) recursion:
 \begin{align}
-	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j}
-	b_{j}(\vc{O}_{1}) \label{eq:fwd1} \\
+	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
+	\label{eq:fwd1} \\
 \begin{split}
@@
 lead to different answer patterns. From an observed sequence of 
-responses, it may not be immediately clear from which learning process
-they stem.  Hence, it is interesting to consider a mixture of latent
-Markov models which incorporate restrictions that are consistent with
-each of the learning strategies.
+responses, it may not be immediately clear from which learning 
+process they stem. Hence, it is interesting to consider a mixture of 
+latent Markov models which incorporate restrictions that are 
+consistent with each of the learning strategies.
 
@@
 
-\paragraph{Computational considerations} 
-From equations~(\ref{eq:fwd1}--\ref{eq:fwdt}), it can be seen that
+\paragraph{Computational considerations} From 
+equations~(\ref{eq:fwd1}--\ref{eq:fwdt}), it can be seen that 
 computing the forward variables, and hence the log-likelihood, takes 
@@
 
-% %Using the equations above for the likelihood of single and mixture 
-% models, arriving at the posterior probabilities of membership of the 
-% components can be easily accomplished.
+%Using the equations above for the likelihood of single and mixture 
+models, arriving at the posterior probabilities of membership of the 
+components can be easily accomplished.
 
-% %The posterior probability is the probability that data $\vc{O}$ are 
-% generated by mixture component $k$, i.e., $Pr(C=k|\vc{O})$, where $C$ 
-% denotes the mixture component. This probability can be expressed as:
+%Nou, nee dus!
+
+%The posterior probability is the probability that data $\vc{O}$ are 
+generated by mixture component $k$, i.e., $Pr(C=k|\vc{O})$, where $C$ 
+denotes the mixture component. This probability can be expressed as:
 %\begin{eqnarray}
@@
 
-See equations 10--12 in \cite{Lystig2002} for the score recursion
+See equations 10--12  in \cite{Lys2002} for the score recursion 
 functions of the hidden Markov model for a univariate time series. 
@@
 \psi_{t}(j_{k};\lambda_{1})  & =  \frac{\fpp Pr(\vc{O}_{1}, \ldots, 
-	\vc{O}_{t}, S_{t}=j_{k})} {Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}
-	\\
+\vc{O}_{t}, S_{t}=j_{k})}
+			{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}  \\
 \begin{split} 
-		& = \sum_{i=1}^{n_{k}} \Bigg\{
-		\psi_{t-1}(i;\lambda_{1})a_{ij_{k}} b_{j_{k}}(\vc{O}_{t}) \\
-			&\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right]
-			b_{j_{k}} (\vc{O}_{t}) \\
+			 & =  %\sum_{k=1}^{K} klopt dit?? is \fpp a_{ij} wel altijd 0 voor andere componenten??
+			 \sum_{i=1}^{n_{k}} \Bigg\{ \psi_{t-1}(i;\lambda_{1})a_{ij_{k}} 
+			 b_{j_{k}}(\vc{O}_{t}) \\ 
+			 &\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] b_{j_{k}} 
+(\vc{O}_{t}) \\
 			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
@@
 			\times (\Phi_{t-1})^{-1}.
-		\end{split} 
-	\label{eq:psit}
+\end{split} \label{eq:psit}
 \end{align}
@@
 	\pmb{\psi}_{1}(\lambda_{1}) = \vc{\fpp} p_{k} \times \pmb{\pi} 
-	\otimes \vc{b}(O_{1})+ p_{k} \times \vc{\fpp}\pmb{\pi} \otimes
-	\vc{b}(O_{1}) + p_{k} \times \pmb{\pi} \otimes \fpp \vc{b}(O_{1}),
+\otimes \vc{b}(O_{1})+
+		p_{k} \times \vc{\fpp}\pmb{\pi} \otimes \vc{b}(O_{1}) +
+		p_{k} \times \pmb{\pi} \otimes \fpp \vc{b}(O_{1}),
 \end{eqnarray}
@@
 \begin{split} 
-		\pmb{\psi}_{t}(\lambda_{1}) &= \left[ \mat{A}^{t} \times
-		\pmb{\psi_{t-1}}(\lambda_{1}) \right ] \otimes \vc{b}(O_{t}) +
-		\left [ \left[ \fpp \mat{A} \right ]^{t} \times
-		\pmb{\phi}_{t-1} \right ] \otimes \vc{b}(O_{t}) \\
-		& \qquad + \left [ \mat{A}^{t} \times \pmb{\phi}_{t-1} \right]
+	\pmb{\psi}_{t}(\lambda_{1}) 
+	&= \left[ \mat{A}^{t} \times \pmb{\psi_{t-1}}(\lambda_{1})  \right ] 
+\otimes \vc{b}(O_{t}) +
+	\left [ \left[ \fpp \mat{A} \right ]^{t} \times \pmb{\phi}_{t-1} 
+\right ] \otimes \vc{b}(O_{t}) \\
+	& \qquad + \left [ \mat{A}^{t} \times \pmb{\phi}_{t-1}  \right ] 
 \otimes \fpp \vc{b}(O_{t}).
@@
 
-Using above equations, \cite{Lystig2002} derive the following equation 
+Using above equations, \cite{Lys2002} derive the following equation 
 for the partial derivative of the likelihood:
@@
 	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1}, 
-	\ldots, \vc{O}_{T})^{-1} \fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T})
-	\\
+\ldots, \vc{O}_{T})^{-1} 
+	\fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) \\
 	&=& Pr(\vc{O}_{1}, \ldots, \vc{O}_{T})^{-1} \Psi_{T}(\lambda_{1}) 
@@
 gaussian observation densities. As a consequence, this provides 
-gradients for the multivariate time series model for mixed categorical
-and gaussian observations, e.g.\ correct/incorrect responses and
-reaction times, and a categorical latent state or class.  Hence, we
-need the partial derivatives $\fpp b_{j}({O}_{t}^{l})$ instead of
-$\fpp b_{j}(\vc{O}_{t})$ (if $\lambda_{1} \in \{ \vc{p},
-\pmb{\pi},\mat{A} \} $, then $\fpp b_{j}=0$).
+gradients for the multivariate time series model for mixed 
+categorical and gaussian observations, e.g.\ correct/incorrect 
+responses and reaction times, and a categorical latent state or 
+class.  Hence, we need the partial derivatives $\fpp 
+b_{j}({O}_{t}^{l})$ instead of $\fpp b_{j}(\vc{O}_{t})$ (if 
+$\lambda_{1} \in \{ \vc{p}, \pmb{\pi},\mat{A} \} $, then $\fpp 
+b_{j}=0$).
 
@@
 	\begin{array}{ll} 
+		b_{j_{k}}(O_{t}^{l}) \times  \left[  \frac{O_{t}-\mu}{\sigma^{2}}  
+\right] & 
+			\mbox{if $\lambda_{1} = \mu_{j_{k}}^{l} $ }  \\ 
             	b_{j_{k}}(O_{t}^{l}) \times \left[  
-		\frac{O_{t}-\mu}{\sigma^{2}} \right] & \mbox{if $\lambda_{1} =
-		\mu_{j_{k}}^{l} $ } 
-		\\
-		b_{j_{k}}(O_{t}^{l}) \times \left[
-		\frac{(O_{t}-\mu)^{2}}{\sigma^{3}} - \frac{1}{\sigma} \right]
-		& \mbox{if $\lambda_{1} = \sigma_{j_{k}}^{l} $} 
-		\\
-		0 & \mbox{otherwise} \end{array} \right.
+\frac{(O_{t}-\mu)^{2}}{\sigma^{3}} - \frac{1}{\sigma} \right] & 
+			\mbox{if $\lambda_{1} = \sigma_{j_{k}}^{l} $} \\
+		0 & \mbox{otherwise} 
+            \end{array} \right. 
 \end{align}
@@
 		0 & \mbox{otherwise} 
-	\end{array} 
-	\right\}
+            \end{array} \right. 
 \end{align}
 
-%\newpage
-%\section*{Author note}
-%We would like to thank many people.
+\subsection{Observed information}
 
-\bibliography{all}
+\newcommand{\fpt}{\frac{\partial}{\partial \lambda_{2}} }
+
+\newcommand{\fps}{\frac{\partial^{2}}{\partial\lambda_{1}\partial\lambda_{2}}}
+
+To compute the observed information, again the formulae by 
+\citet{Lys2002} are applied and generalized to the m-variate mixture 
+case. Define the information recursion as follows:
+\begin{align}
+\begin{split}
+	\omega_{1}(j_{k};\lambda_{1},\lambda_{2}) 
+	=  & \, \fps Pr(\vc{O}_{1}, S_{1}=j_{k}) = \fps p_{k} \pi_{j_{k}} 
+b_{j_{k}}(\vc{O}_{1})
+	\\  = & \, \fpp \left[ \fpt p_{k} \pi_{j_{k}} b_{j_{k}}(\vc{O}_{1}) 
+\right] 
+	\\  =  & \fpp  \bigg\{  \left[ \fpt p_{k} \right]  \pi_{j_{k}} 
+b_{j_{k}}(\vc{O}_{1}) +
+		\, p_{k} \left[ \fpt \pi_{j_{k}} \right]  b_{j_{k}}(\vc{O}_{1}) + 
+	\\  &  \, p_{k} \pi_{j_{k}} \left[ \fpt  b_{j_{k}}(\vc{O}_{1}) 
+\right] \bigg\}
+	\\  =  & \,  \left[ \fps p_{k} \right] \pi_{j_{k}} 
+b_{j_{k}}(\vc{O}_{1}) + 
+		\left[ \fpt p_{k} \right] \cdot \left[\fpp \pi_{j_{k}} \right] 
+b_{j_{k}}(\vc{O}_{1}) +
+	\\  &   \, \left[ \fpt p_{k} \right] \pi_{k} \left[ \fpp b_{j_{k}} 
+(\vc{O}_{1}) \right] +
+		\left[ \fpp p_{k} \right] \pi_{j_{k}} b_{j_{k}}(\vc{O}_{1}) + 
+	\\ &   \, p_{k} \left[ \fps \pi_{j_{k}} \right] 
+b_{j_{k}}(\vc{O}_{1}) +
+		p_{k} \left[ \fpt \pi_{j_{k}} \right] \cdot \left[ \fpp 
+b_{j_{k}}(\vc{O}_{1}) \right] +
+	\\ &   \, \left[ \fpp p_{k} \right] \pi_{j_{k}} \left[ \fpt 
+b_{j_{k}}(\vc{O}_{1}) \right] +
+		p_{k} \left[ \fpp \pi_{j_{k}} \right] \cdot \left[ \fpt 
+b_{j_{k}}(\vc{O}_{1}) \right] +
+	\\ &   \, p_{k} \pi_{j_{k}} \left[ \fps b_{j_{k}}(\vc{O}_{1}) 
+\right].
+\end{split}
+\end{align}
+
+Similarly, for $t>1$, the information recursion is defined as:
+\begin{align}
+\begin{split}
+	\omega_{t}(j_{k};\lambda_{1},\lambda_{2}) =  
+	&\frac{\fps Pr(\vc{O}_{1}, \ldots, 
+\vc{O}_{t},S_{t}=j_{k})}{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}
+	\\ = &\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} \Bigg\{ 
+	\omega_{t-1}(i;\lambda_{1},\lambda_{2}) a_{ij_{k}} 
+b_{j_{k}}(\vc{O}_{t}) +
+	\psi_{t-1}(i;\lambda_{1}) \left[ \fpt a_{ij_{k}} \right] 
+b_{j_{k}}(\vc{O}_{t}) +
+	\\  &\psi_{t-1}(i;\lambda_{1})  a_{ij_{k}}  \left[ \fpt 
+b_{j_{k}}(\vc{O}_{t}) \right] +
+	\psi_{t-1}(i;\lambda_{2}) \left[ \fpp a_{ij_{k}} \right] 
+b_{j_{k}}(\vc{O}_{t}) + 
+	\\ & \psi_{t-1}(i;\lambda_{2}) a_{ij_{k}} \left[ \fpp 
+b_{j_{k}}(\vc{O}_{t})\right]  + 
+	\phi_{t-1}(i) \left[ \fps a_{ij_{k}} \right]  b_{j_{k}}(\vc{O}_{t}) +
+	\\ & \phi_{t-1}(i) a_{ij_{k}} \left[ \fps b_{j_{k}}(\vc{O}_{t}) 
+\right]  +
+	\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] \left[ \fpt 
+b_{j_{k}}(\vc{O}_{t}) \right]  +
+	\\ & \phi_{t-1}(i) \left[ \fpt a_{ij_{k}}\right] \left[ \fpp 
+b_{j_{k}}(\vc{O}_{t}) \right]  \Bigg\} 
+	\times \frac{1}{\Phi_{t-1}}
+\end{split}
+\end{align}
+
+It is important here to provide the second order partial derivatives 
+of the terms that are involved in the multi-variate case. These are 
+the $b_{j_{k}}(\cdot)$ terms. Remember that these terms can be 
+factored and hence the derivatives can be derived from repeatedly 
+applying the chain rule for products and using equation 
+\ref{partialProd} from above for the first order derivatives: 
+\begin{align}
+\begin{split}
+	\fps b_{j_{k}} (\vc{O}_{t}) & = \fps \prod_{i=1}^{m} b_{j_{k}}^{i} 
+(O_{t}^{i}) 
+	\\ & = \fpp \left[ \fpt \prod_{i=1}^{m} b_{j_{k}}^{i} (O_{t}^{i})  
+\right] 
+	\\ & = \fpp \Bigg\{ \sum_{l=1}^{m} \left[ \prod_{i=1, \ldots, 
+\hat{l}, \ldots, m} b_{j}(O_{t}^{i}) 
+	\right] \times \fpt  [b_{j}(O_{t}^{l})] \Bigg\}  
+	\\ & = \sum_{l=1}^{m} \left[ \fpp \prod_{i=1, \ldots, \hat{l}, 
+\ldots, m} b_{j}(O_{t}^{i}) \right] 
+	\times \fpt  \left[ b_{j}(O_{t}^{l}) \right] \; + 
+	\\ &\phantom{=}\;\,  \sum_{l=1}^{m}  \left[ \prod_{i=1, \ldots, 
+\hat{l}, \ldots, m} b_{j}(O_{t}^{i}) 
+	\right] \times \fpp \fpt  \left[ b_{j}(O_{t}^{l})\right]  
+	\\ & = \sum_{k=1,k\neq	l}^{m} \sum_{l=1}^{m}  \left[ \prod_{i=1, 
+\ldots, \hat{l}, \hat{k}, \ldots, m}
+	b_{j}(O_{t}^{i}) \right] \fpt  \left[ b_{j}(O_{t}^{k})\right]  \fpt  
+\left[ b_{j}(O_{t}^{l})\right] \;+ 
+	\\ &\phantom{=}\;\, \sum_{l=1}^{m}  \left[ \prod_{i=1, \ldots, 
+\hat{l}, \ldots, m} b_{j}(O_{t}^{i}) 
+	\right] \times \fps  \left[b_{j}(O_{t}^{l}) \right] 
+\end{split}
+\end{align}
+
+In order to compute the above terms we need for arbitrary parameters 
+$\lambda_{1}$ the terms $\fps  \lambda_{1}$. Fortunately, these are 
+easy to compute as most of them are zero. In particular, with 
+$\lambda_{1} \in \{\mat{A},\vc{p},\pmb{\pi} \}$, the double 
+derivatives are zero and we only need them when 
+$\lambda_{1}\in\{\mat{B}\}$.
+These latter ones are again zero in the case of multinomial items and 
+so only the gaussian observations remain which are given here in four 
+cases:
+
+%%p-k cases
+\paragraph{Case 1: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\mu, 
+\lambda_{2}=\mu$}
+
+\begin{align} 
+        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l}) \left[ \left[ 
+\frac{O_{t}^{l}-\mu}{\sigma^{2}} \right]^{2} - \frac{1}{\sigma^{2}} 
+\right]
+\end{align}
+
+\paragraph{Case 2: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\mu, 
+\lambda_{2}=\sigma$}
+
+\begin{align} 
+        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l}) \left[ \left[ 
+\frac{(O_{t}^{l}-\mu)^{2}}{\sigma^{3}} -
+	\frac{1}{\sigma}\right]  \cdot \left[ 
+\frac{O_{t}^{l}-\mu}{\sigma^{2}} \right] + 
+	\left[ \frac{2\mu-2O_{t}^{l}}{\sigma^{3}} \right] \right]
+\end{align}
+
+\paragraph{Case 3: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\sigma, 
+\lambda_{2}=\mu$}
+
+\begin{align} 
+        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l})  \left[ \left[ 
+\frac{(O_{t}^{l}-\mu)^{2}}{\sigma^{3}} -
+	\frac{1}{\sigma}\right]  \cdot \left[ 
+\frac{O_{t}^{l}-\mu}{\sigma^{2}} \right] + 
+	\left[ \frac{2\mu-2O_{t}^{l}}{\sigma^{3}} \right] \right]
+\end{align}
+
+\paragraph{Case 4: $\fps b_{j}(O_{t}^{l}), \lambda_{1}=\sigma, 
+\lambda_{2}=\sigma$}
+
+\begin{align} 
+        \fps b_{j_{k}}(O_{t}^{l})=b_{j}(O_{t}^{l}) \left[ \left[ 
+\frac{(O_{t}^{l}-\mu)^{2}}{\sigma^{3}} -
+	\frac{1}{\sigma}\right]^{2} + \left[  \frac{1}{\sigma^{2}} - 
+\frac{3(O_{t}^{l}-\mu)^{2}}{\sigma^{4}}  \right]  \right]
+\end{align}
+
+\bibliography{psy,ingmar,hmm}
+
+% \section*{Author note}
 

### 2008-11-03 papers/vignette/depmix-intro.tex
added papers directory (lots of fighting with svn ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[a4paper]{article}

\usepackage[]{amsmath, amsfonts, amstext, amsthm}

\usepackage{amssymb}

\usepackage[]{graphics}

\usepackage{graphicx}

\usepackage{epsfig}

\usepackage{epstopdf}

%\batchmode

\usepackage[round]{natbib}

\title{depmixS4: An R-package for fitting mixture models on mixed
multivariate data with Markov dependencies}

\author{Ingmar Visser \& Maarten Speekenbrink\thanks{
Correspondence concerning this manual should be adressed to:
Ingmar Visser, Department of Psychology, University of Amsterdam,
Roetersstraat 15, 1018 WB, Amsterdam, The Netherlands}\\
Department of Psychology, University of Amsterdam\\
i.visser@uva.nl \\ 
Department of Psychology, University College London 
\\ m.speekenbrink@ucl.ac.uk}

\date{\today}

\newcommand{\vc}{\mathbf}
\newcommand{\mat}{\mathbf}

 \newcommand{\pkg}{\texttt}
 \newcommand{\code}{\texttt}

\begin{document}

\bibliographystyle{plainnat}

\maketitle

\begin{abstract}

	\pkg{depmixS4} implements a general framework for definining and
	fitting hidden Markov mixture model in the R programming language
	\citep{R2007}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach with gradients using the
	\code{Rdonlp2} optimization routine when contraints are imposed on
	the parameters.  A number of illustrative examples are included.

\end{abstract}

\newpage
\tableofcontents
\newpage

\section{Introduction}

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982},
for an overview, and \citet{Schmittmann2006} for a recent application.
In economics, latent Markov models are commonly used as regime
switching models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.
Further applications include speech recognition \citep{Rabiner1989},
EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
those latter areas of application, latent Markov models are usually
referred to as hidden Markov models.

The \pkg{depmixS4} package was motivated by the fact that Markov models
are used commonly in the social sciences, but no comprehensive package
was available for fitting such models.  Common programs for Markovian
models include Panmark \citep{Pol1996}, and for latent class models
Latent Gold \citep{Vermunt2003}.  Those programs are lacking a number
of important features, besides not being freely available.  There are
currently some packages in R that handle hidden Markov models but they
lack a number of features that we needed in our research.  In
particular, \pkg{depmixS4} was designed to meet the following goals: 
\begin{enumerate}

	\item to be able to handle parameter estimates subject to general
	linear (in)equality constraints

	\item to be able to fit transition models with covariates, i.e.,
	to have time-dependent transition matrices

	\item to be able to include covariates in the prior or initial
	state probabilities of models

	\item to allow for easy extensibility, in particular, to be able
	to add new response distributions, both univariate and
	multivariate, and similarly to be able to allow for the addition
	of other transition models, e.g., continuous time observation
	models

\end{enumerate}

Although \pkg{depmixS4} is designed to deal with
longitudinal or time series data, for say $T>100$, it can also handle
the limit case with $T=1$.  In those cases, there are no time
dependencies between observed data, and the model reduces to a finite
mixture model, or a latent class model.  Although there are other
specialized packages to deal with mixture data, one specific feature
that we needed ourselves which is to the best of our knowledge not
available in other packages is the possibility to include covariates
on the prior probabilities of class membership.  In the next section,
an outline is provided of the model and the likelihood equations.

\subsection*{Acknowledgements}

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
Economic Learning and Social Evolution (ELSE).  Han van der Maas
provided the speed-accuracy data \cite{Maas2005b} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2001}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.

% Maartje Raijmakers \cite{Raijmakers2001} provided
% the discrimination data set.

\section{Dependent mixture models}

%Only later did I find out that \citet{Leroux1992} already
%coined the term dependent mixture models in an application with hidden
%Markov mixtures of Poisson count data.

The data considered here, has the general form $O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1}, \ldots,
O_{T}^{m}$ for an $m$-variate time series of length $T$.  As an
example, consider a time series of responses generated by a single
subject in a reaction time experiment.  The data consists of three
variables, reaction time, accuracy and a covariate which is a pay-off
factor which determines the reward for speed and accuracy.  These
variables are measured on 168, 134 and 137 occasions respectively 
(in Figure~\ref{fig:speed} the first part of this series is plotted).

\begin{figure}[htbp]
  \begin{center}
	  \scalebox{0.6}{\includegraphics*{speed1.pdf}}
	  \caption{Reaction times, accuracy and pay-off values for
	  the first series of responses in dataset \texttt{speed}.}
	  \label{fig:speed}
  \end{center}
\end{figure}

The latent Markov model is commonly associated with data of this type,
albeit usually only multinomial variables are considered.  However,
common estimation procedures, such as those implemented in
\citet{Pol1996} are not suitable for long time series due to underflow
problems.  In contrast, the hidden Markov model is typically only used
for `long' univariate time series.  In the next section, the
likelihood and estimation procedure for the hidden Markov model is
described, given data of the above form.  These models are called
dependent mixture models because one of the authors (Ingmar Visser)
thought it was time for a new name for these models.

The dependent mixture model is defined by the following elements:
\begin{enumerate}

	\item a set $\vc{S}$ of latent classes or states $S_{i},\, i=1,
	\ldots , n$,

	\item matrices $\mat{A}_t$ of transition probabilities $a_{ij,t}$ for
	the transition from state $S_{i}$ to state $S_{j}$ at time $t$,

	\item a set $\vc{B}_t$ of observation functions $b_j^k(\cdot)$ that
	provide the conditional probabilities of observations $O_{t}^k$ 
	associated with latent state $S_{j}$,

	\item a vector $\pmb{\pi}$ of latent state initial probabilities
	$\pi_{i}$
\end{enumerate}
When transitions are added to the latent class model, it is more 
appropriate to refer to the classes as states. The word class is 
rather more associated with a stable trait-like attribute whereas a 
state can change over time.

\subsection{Likelihood}

The log-likelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
rather by the forward part of this algorithm.  \cite{Lystig2002}
changed the forward algorithm in such a way as to allow computing the
gradients of the log-likelihood at the same time.  They start by
rewriting the likelihood as follows (for ease of exposition the
dependence on the model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
Pr(\vc{O}_{t}|\vc{O}_{1}, 
	\ldots, \vc{O}_{t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

\subsection{Computational considerations} From
equations~(\ref{eq:fwd1}--\ref{eq:fwdt}), it can be seen that
computing the forward variables, and hence the log-likelihood, takes
$O(Tn^{2})$ computations, for an $n$-state model and a time series of
length $T$.

\subsection{Parameter estimation}

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  The EM
algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization (although it is usually
faster than direct optimization at the start, so possibly a
combination of EM and direct optimization is fastest).  Second,
applying constraints to parameters can be problematic; in particular,
EM can lead to wrong parameter estimates when applying constraints.
Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
\cite{Tamura2007,Spellucci2002}, because it handles general linear
(in)equality constraints, and optionally also non-linear constraints.

\section{Using depmixS4}

Two steps are involved in using \pkg{depmixS4} which are illustrated
below with examples:
\begin{enumerate}
	\item model specification with function \code{depmix}

	\item  model fitting with function \code{fit}
\end{enumerate}

\subsection{Example data: speed}

Throughout this manual a data set called \code{speed} is used.  It
consists of three time series with three variables: reaction time,
accuracy, and a covariate Pacc which defines the relative pay-off for
speeded and accurate responding. The participant in this experiment 
switches between fast responding at chance level and relatively 
slower responding at a high level of accuracy.

Interesting hypotheses to test are: is the switching regime symmetric?
Is there evidence for two states or does one state suffice?  Is the
guessing state actually a guessing state, i.e., is the probability
of a correct response at chance level (0.5)?

\subsection{Defining models}

A dependent mixture model is defined by the number of states, and by
the response distribution functions, and can be created with the
\code{depmix}-function as follows (see the help files for other
options):

\begin{verbatim} 
mod <- depmix(rt~1, data=speed, nstates=2)
\end{verbatim}

Above code illustrates the simplest case of a univariate time series
without covariates; besides providing the data, the only other
necessity is to specify the desired number of states. Note that the
rt's are modelled here with a Gaussian distribution as that is the
default family option in \pkg{depmixS4}. A multivariate model can be
specified by providing a list of formulae rather than a single one as
above:

\begin{verbatim} 
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
family=list(gaussian(),multinomial())
\end{verbatim}

Here it is also necessary to provide the family functions for each of
the responses.  Currently, the gaussian() and multinomial() are
implemented.

The function \code{depmix} returns an object of class \code{depmix}
which has its own summary function providing the parameter values of
the model. The object consists of three main parts: the prior model,
which specifies the initial state probabilities, the transition
models, specifying the transition probabilities for each state, and
the response models, specifying the densities for each response and
each state. See the help files for further details.

Except in simple cases, starting values can be a problem in latent
Markov models, and so in general it's best to provide them if you have
a fairly good idea of what to expect.  Providing starting values is
done through three arguments: respstart, trstart, and instart, for
response related parameters, transition parameters and prior
parameters, respectively.  The use of setpars reveals the ordering of
parameters that should be used:

\begin{verbatim}
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
family=list(gaussian(),multinomial())
setpars(mod,1:npar(mod))
\end{verbatim}

See the paragraph on parameter numbering below for further details.

\subsection{Transition matrix and initial state probabilities}

The transition matrix is parametrized as a list of multinomial
logistic models.  The initial state probabilities are similarly
parametrized as a multinomial logistic model.  Both models use a base
category parametrization, meaning that the parameter for the base
category is fixed at zero.  The default base category is the first
state.  Hence, for example, for a 3-state model, the initial state
probability model would have three parameters of which the first is
fixed at zero and the other two are freely estimated.

%See
%\citet{Agresti20??} for multinomial logistic models and various
%parameterizations.

Covariates can be specified using a one-sided formula as in the
following example:

\begin{verbatim}
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2,
transition=~Pacc)
\end{verbatim}

Note that this can be done for the initial state probabilities by
specifying prior=~X1, where X1 is the desired covariate.  The result
of this is that the transition probabilities are now dependent on the
covariate Pacc (which is an experimenter controlled variable to induce
switching between guessing and accurate responding).

\subsection{Fitting models}

Fitting models is done using the function \code{fit}.  The standard
call only requires a model object of class \code{depmix}:

\begin{verbatim} 
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
family=list(gaussian(),multinomial())
fmod <- fit(mod)
\end{verbatim}

The function returns an object of class \code{depmix.fitted} which
extends the \code{depmix} class, adding convergence information and
possibly information about constraints if these were applied.  The
function provides some online output on the progress of the
optimization, the precise form of the output depends on the
optimization method chosen.

Class \code{depmix} and \code{depmix.fitted} have \code{logLik},
\code{AIC} and \code{BIC} methods to provide fit statistics.

\subsection{Fixing and constraining parameters}

Constraining and fixing parameters is done using the \code{conpat}
argument to the \code{depmix.fit}-function, which specifies for each
parameter in the model whether it's fixed (0) or free (1 or higher).
Equality constraints can be imposed by having two parameters have the
same number in the \code{conpat} vector.  When only fixed values are
required the \code{fixed} argument can be used instead of
\code{conpat}, with zeroes for fixed parameters and other values (ones
e.g.) for non-fixed parameters.  Fitting the models subject to these
constraints is handled by the optimization routine \code{donlp2}.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete parameter vectors should be supplied,
i.e., these vectors should have length of the number of parameters of
the model, which can be obtained by calling \code{npar(object)}.
Parameters are numbered in the following order:
\begin{enumerate}
	\item  the prior model parameters

	\item  the parameters for the transition models

	\item  the response model parameters per state (and subsequently
	per response in the case of multivariate time series)

\end{enumerate}

To see the ordering of parameters use the following:
\begin{verbatim}
mod <- setpars(mod, value=1:npar(mod))mod
\end{verbatim}

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model:
\begin{verbatim}
mod <- setpars(mod,
getpars(mod,which="fixed"))mod
\end{verbatim}

\bibliography{all,ingmar}

\end{document}

### 2008-11-03 papers/vignette/toaddlater.tex
added papers directory (lots of fighting with svn ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{Mixtures of LMMs}

The above forward recursion can readily be generalized to mixture 
models, in which it is assumed that the data are realizations of a 
number of different LMMs and the goal is to assign posterior 
probabilities to sequences of observations. This situation occurs, 
for example, in learning data where different learning strategies may 
lead to different answer patterns. From an observed sequence of 
responses, it may not be immediately clear from which learning 
process they stem. Hence, it is interesting to consider a mixture of 
latent Markov models which incorporate restrictions that are 
consistent with each of the learning strategies.

To compute the likelihood of a mixture of $K$ models, define the 
forward recursion variables as follows (these variables now have an 
extra index $k$ indicating that observation and transition 
probabilities are from latent model $k$):
\begin{align}
\begin{split}
\phi_{1}(j_{k}) &=  Pr(\vc{O}_{1}, 
S_{1}=j_{k})=p_{k}\pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}).
\end{split}\label{eq:fwd1mix} \\
\begin{split}
\phi_{t}(j_{k})   &=   Pr(\vc{O}_{t}, S_{t}=j_{k}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
			&= \left[ \sum_{k=1}^{K} \sum_{i=1}^{n_{k}} \phi_{t-1}(i_{k}) 
			a_{ij_{k}}b_{j_{k}}(\vc{O}_{t}) \right] \times (\Phi_{t-1})^{-1},
\end{split}\label{eq:fwdtmix} 
\end{align}
where $\Phi_{t} = \sum_{k=1}^{K}\sum_{i=1}^{n_{k}} \phi_{t}(j_{k})$.
Note that the double sum over $k$ and $n_{k}$ is simply an enumeration
of all the states of the model.  Now, because $a_{ij_{k}}=0$ whenever
$S_{i}$ is not part of component $k$, the sum over $k$ can be dropped
and hence equation~\ref{eq:fwdtmix} reduces to:
\begin{equation}
	\phi_{t}(j_{k}) = \left[ \sum_{i=1}^{n_{k}} \phi_{t-1}(i_{k}) 
			a_{ij_{k}}b_{j_{k}}(\vc{O}_{t}) \right] \times (\Phi_{t-1})^{-1}
\end{equation}
The loglikelihood is computed by applying equation~\ref{eq:logl} on
these terms.  For multiple cases, the log-likelihood is simply the sum
over the individual log-likelihoods.

Consider a mixture of two components, one with two states
and the other with three states.  Using
equations~(\ref{eq:fwd1}--\ref{eq:fwdt}) to compute the log-likelihood
of this model one needs $O(Tn^{2})=O(T\times 25)$ computations whereas
with the mixture equations~(\ref{eq:fwd1mix}--\ref{eq:fwdtmix}),
$\sum_{n_{i}} O(n_{i}^{2}T)$ computations are needed, in this case
$O(T \times 13)$.  So, it can be seen that in this easy example the
computational cost is almost halved.

\section{Gradients}

\newcommand{\fpp}{\frac{\partial} {\partial \lambda_{1}}}

See equations 10--12  in \cite{Lystig2002} for the score recursion 
functions of the hidden Markov model for a univariate time series. 
Here the corresponding score recursion for the multivariate mixture 
case are provided. The $t=1$ components of this score recursion are 
defined as (for an arbitrary parameter $\lambda_{1}$):
\begin{align}
\psi_{1}(j_{k};\lambda_{1}) &:=  \fpp Pr(\vc{O}_{1}|S_{1}=j_{k}) \\
\begin{split} 
	&= \left[  \fpp p_{k} \right] \pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}) + 
	p_{k}\left[ \fpp \pi_{j_{k}} \right] b_{j_{k}}(\vc{O}_{1}) \\
	& \qquad  + p_{k}\pi_{j_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{1})\right],
\end{split} \label{eq:psi1}
\end{align}
and for $t>1$ the definition is:
\begin{align}
\psi_{t}(j_{k};\lambda_{1})  & =  \frac{\fpp Pr(\vc{O}_{1}, \ldots, 
\vc{O}_{t}, S_{t}=j_{k})}
			{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}  \\
\begin{split} 
	& =  
			 \sum_{i=1}^{n_{k}} \Bigg\{ \psi_{t-1}(i;\lambda_{1})a_{ij_{k}} 
			 b_{j_{k}}(\vc{O}_{t}) \\ 
			 &\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] b_{j_{k}} 
(\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
(\vc{O}_{t}) \right] \Bigg\} 
			\times (\Phi_{t-1})^{-1}.
\end{split} \label{eq:psit}
\end{align}

Using above equations, \cite{Lystig2002} derive the following equation 
for the partial derivative of the likelihood:
\begin{equation}
	\fpp l_{T}= 	
		\frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}},
\end{equation}
where $\Psi_{t}=\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} 
\psi_{t}(j_{k};\lambda_{1})$. 
Starting from the equation from the logarithm of the likelihood, this 
is easily seen to be correct: 
\begin{eqnarray*}
	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})^{-1} 
	\fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) \\
	&=&  \frac{Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1})}{Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})}  \Psi_{T} (\lambda_{1}) \\
	&=&  \frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}}.
\end{eqnarray*}

Further, to actually compute the gradients, the partial derivatives of
the parameters and observation distribution functions are neccessary,
i.e., $\fpp p_{k}$, $\fpp \pi_{i}$, $\fpp a_{ij}$, and $\fpp
\vc{b}_{i}(\vc{O}_{t})$.  Only the latter case requires some
attention.  We need the following derivatives $\fpp
\vc{b}_{j}(\vc{O}_{t})=\fpp \vc{b}_{j}(O_{t}^{1}, \ldots, O_{t}^{m})$, for
arbitrary parameters $\lambda_{1}$. To stress that $\vc{b}_{j}$ is a
vector of functions, we here used boldface. First note that because of local
independence we can write:
\begin{equation*}
	\fpp \left[ b_{j}(O_{t}^{1}, \ldots, O_{t}^{m}) \right] = \frac{\partial} 
{\partial \lambda_{1} } \left[ b_{j}(O_{t}^{1}) \right] \times  
\left[ b_{j}(O_{t}^{2}) \right], \ldots,  \left[ b_{j}(O_{t}^{m}) \right].  
\end{equation*}
Applying the chain rule for products we get:
\begin{equation}
	\fpp [b_{j}(O_{t}^{1}, \ldots, O_{t}^{m})] =
	\sum_{l=1}^{m} \left[ \prod_{i=1, \ldots, \hat{l}, \ldots, m} 
b_{j}(O_{t}^{i}) \right] \times
	\fpp  [b_{j}(O_{t}^{l})],
	\label{partialProd}
\end{equation}
where $\hat{l}$ means that that term is left out of the product. 
These latter terms, $\frac{\partial} {\partial \lambda_{1} }  
[b_{j}(O_{t}^{k})]$, are easy to compute given either multinomial or 
gaussian observation densities $b_{j}(\dot)$

\subsection{Generating data}

The \code{dmm}-class has a \code{generate} method that can be used to 
generate data according to a specified model.

\begin{verbatim}
gen<-generate(c(100,50),mod)
\end{verbatim}

\section{Multi group/case analysis}

\begin{verbatim}
conpat=rep(1,15)
conpat[1]=0
conpat[8:9]=0
conpat[14:15]=0
stv=c(1,0.9,0.1,0.1,0.9,5.5,0.2,0.5,0.5,6.4,0.25,0.9,0.1,0.5,0.5)
mod<-dmm(nstates=2,itemt=c("n",2),stval=stv,conpat=conpat)
\end{verbatim}

\code{depmix4} can handle multiple cases or multiple groups. A
multigroup model is specified using the function \code{mgdmm} as
follows:

\begin{verbatim}
mgr <- mgdmm(dmm=mod,ng=3,trans=TRUE,obser=FALSE)
mgrfree <- mgdmm(dmm=mod,ng=3,trans=FALSE)
\end{verbatim}

The \code{ng} argument specifies the number of groups, and the
\code{dmm} argument specifies the model for each group.  \code{dmm}
can be either a single model or list of models oflength \code{ng}.  If
it is a single model, each group has an identical structural model
(same fixed and constrained parameters).  Three further arguments can
be used to constrain parameters between groups, \code{trans},
\code{obser}, and \code{init} respectively.  By setting either of
these to \code{TRUE}, the corresponding transition, observation, and
initial state parameters are estimated equal between
groups\footnote{There is at this moment no way of fine-tuning this to
restrict equalities to individual parameters.  However, this can be
accomplished by manually changing the linear constraint matrix, and
the corresponding upper and lower boundaries.}.

In this example, the model from above was used and fitted on the three
observed series, and the \code{trans=TRUE} ensures that the transition
matrix parameters are constrained to be equal between the models for
these series, whereas the observation parameters are freely estimated,
i.e.\ to capture learning effects.

The loglikelihood ratio statistic can be used to test whether
constraining these transition parameters significantly reduces the
goodness-of-fit of the model.  The statistic has an approximate
$\chi^{2}$ distribution with $df=4$ because in each but the first
model, two transition matrix parameters were estimated equal to the
parameters in the first model (note that the other two transition
parameters already had to be constrained to ensure that the rows of
the transition matrices sum to 1).

\section{Mixtures of latent Markov models}

\code{depmix4} provides support for fitting mixtures of latent Markov
models using the \code{mixdmm} function; it takes a list of
\code{dmm}'s as argument, possibly together with the starting values
for the mixing proportions for each component model.  There's an
example in the helpfiles.  It fits the model to data from a
discrimination learning experiment which is provided as data set
\code{discrimination} \cite{Raijmakers2001}.

\section{Finite mixtures and latent class models}

The function \code{lca} can be used to specify latent class models
and/or finite mixture models.  It is simply a wrapper for the
\code{dmm} function, and all it does is adding appropriate numbers of
zeroes and ones to the parameter specification vectors for starting
values, fixed values and linear constraints.  When a model has class
\code{lca} the summary function does not print the transition matrix
(because it is fixed and/or not estimated).

\section{Starting values}

Although providing your own starting values is preferable,
\pkg{depmixS4} has a routine for generating starting values using the
\code{kmeans}-function from the \pkg{stats}-package.  This will
usually provide reasonable starting values, but can be way off in a
number of cases.  First, for univariate categorical time series,
\code{kmeans} does not work at all, and \pkg{depmixS4} will provide a
warning.  Second, for multivariate series with unordered categorical
items with more than 2 categories, \code{kmeans} may provide good
starting values, but they may similarly be completely off, due to the
implicit assumption in \code{kmeans} that the categories are
indicating an underlying continuum.  Starting values using
\code{kmeans} are automatically provided when a model is specified
without starting values.  The argument \code{kmst} to the
\code{fitdmm}-function can be used to control this behavior.

Starting values of the parameters, either user provided or generated,
can be further boosted by using posterior estimates using the Viterbi
algorithm \cite{Rabiner1989}.  That is, first the a posteriori latent
states are generated from the current parameter values for the data at
hand.  Next, from the a posteriori latent states, new parameter
estimates are derived.  This is done by default and can be controlled
by the \code{postst} argument.  Provided that the starting values were
close to their true values, using this procedure further pushes those
parameters in the right direction.  If however the original values
were bad, this procedure may result in bad estimates, i.e.,
optimization will lead to some non-optimal local maximum of the
loglikelihood.

### 2008-11-03 papers/individual/individual.tex
-added prelim results for WPT
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@235 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2008-11-03 papers/invidividual/individual.tex
@@
 Alternative non-strategy based analyses of the WPT (Lagnado et al, Speekenbrink et al) 
-have estimated response strategies by variations of logistic regression.
+have estimated response strategies by logistic regression, allowing the regression
+coefficients to change over time.
 
@@
 Here, we analyze the behavior of a single individual performing the
-WPT for 200 trials. We let each state be characterized by a Generalized
-Linear Model with a Bernoulli response and logistic link function. We are
-interested in whether a DMM can recover a strategy model in line with Gluck et al. 
-As we fit the data to a single subject, we must place some constraints.
-Specifically, we constrain the state transitions to be in a ``left-right'' format
-(states can only proceed to the immediately adjacent state and never back) and 
-the initial state
+WPT for 200 trials. We chose to analyse the ``average'' participant (the
+participant with performance closest to the group average) in a large
+unpublished dataset. We let each state be characterized by a GLM with a 
+Binomial distributed response and logistic link function (i.e., a logistic
+regression model). We are particularly interested in evidence for 
+strategy switching and whether a DMM can recover a strategy model 
+in line with Gluck et al. (2002).
 
-A single state model (usual GLM) 
-We started with a 3-state model,
+As we fit the data to a single subject, 
+we must place some constraints. Specifically, we constrain the state 
+transitions to be in a ``left-right'' format (states can only proceed 
+to the immediately adjacent state and never back, and must start in 
+the initial state). We fitted a single, two and three state model to 
+the data. This showed that a two state model was better than a single
+and three state model.
 
-\begin{itemize}
-	\item N=200
-	\item 4 choice data, displayed in blocks of 20 trials
-	\item optimal strategy is choosing C or D
-\end{itemize}
+\begin{table}
+\caption{Estimates for the weather prediction task}
+\label{tab:WPT}
+\begin{tabular}{lcccccccc} \hline
+ & & \multicolumn{1}{c}{1 state} & & \multicolumn{2}{c}{2 state} && \multicolumn{2}{c}{2 state (constr.)} \\ \cline{3-3} \cline{5-6} \cline{8-9}
+parameter & & $S_1$ & & $S_1$ & $S_2$ & & $S_1$ & $S_2$ \\ \hline
+(intercept) & & -0.69 & & -2.73 & 0.88 & & -1.24 & 0 \\
+cue 1 && 1.69 && 2.12 & 1.60 && 1.65 & 1.97 \\
+cue 2 && 1.12 && 0.97 & 1.63 && 0 & 1.92 \\
+cue 3 && -0.49 && 0.91 & -2.03 && 0 & -1.58 \\
+cue 4 && -1.32 && 0.69 & -3.16 && 0 & -2.67 \\ \hline
+ & & \multicolumn{1}{c}{AIC=204.47} & & \multicolumn{2}{c}{AIC=187.50} && \multicolumn{2}{c}{AIC=185.24}
+\end{tabular}
+\end{table}
+
+Investigation of the parameter estimates (see Table~\ref{tab:WPT}) indicated that the first state might be
+a single cue strategy (the regression coefficient for the first cue was of much 
+larger magnitude than that of the other cues). The second state was a multi-cue
+strategy (all cues had regression coefficients of reasonable magnitude). 
+To reduce the degrees of freedom, and improve parameter estimates, we implemented 
+constraints to force state 1 into a single cue strategy (fixing the coefficients
+of the remaining three cues to 0) and state 2 in a multi-cue strategy (forcing
+the intercept to 0). These restrictions resulted in a better AIC value of AIC=185.24 
+(df=7). Interestingly, the single cue strategy was somewhat different than 
+described by Gluck et al. Parameter estimates indicated relatively more consistent 
+predictions of ``rain'' in the absence of cue 1 ($Pr(\text{sun}) = 0.22$) and more
+inconsistent predictions of ``sun'' in the presence of cue 1 ($Pr(\text{sun}) = 0.60$). 
+The cue weights of the multi-cue strategy were in the direction of the optimal weights. 
+The Viterbi state sequence indicated that the participant used the single
+cue strategy for the first 60 trials, and then switched to the multi-cue strategy.
 

### 2008-11-06 papers/individual/individual.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@236 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 The Weather Prediction Task (WPT, Knowlton, Squire \& Gluck, 1994)
-is a probabilistic categorization task, in which participants learn to 
-predict the state of the weather (sunny, or rainy) on the basis 
-of four ``tarot'' cards (cards with abstract geometrical patterns). 
-Each cue pattern is associated with a particular probability distribution 
-over the states of the weather. In order to perform in the task, 
-participants must predict the weather in accordance with these
-conditional probabilities.
+is a probabilistic categorization task in which participants learn to
+predict (or categorize) the state of the weather (sun or rain) on the basis
+of four ``tarot'' cards (cards with abstract geometrical patterns). On a
+given trial, one, two, or three cues are present. There are a total of
+14 possible cue patterns, and each one is associated with a particular
+probability distribution over the states of the weather. In order to
+perform in the task, participants must predict the weather in accordance
+with these conditional probabilities.
 
@@
 learn by storing each encountered cue-outcome pairing. When presented
-with a cue pattern, these exemplars are retrieved from memory, and weighted
-according to their similarity to the probe cue pattern, to form a 
-classification. According to associative theories, participants gradually 
-learn by gradually associating the individual cues (or cue patterns in 
-configural learning) to the outcomes. In rule-learning, participants are 
-taken to extract rules by which to categorize the different cue patterns. 
-Gluck, Shohamy and Myers (2002) proposed a number of such rules (or strategies). A main
+with a cue pattern, these exemplars are retrieved from memory, and
+weighted according to their similarity to the probe cue pattern, to
+form a classification. According to associative theories,
+participants gradually learn by gradually associating the individual
+cues (or cue patterns in configural learning) to the outcomes. In
+rule-learning, participants are taken to extract rules by which to
+categorize the different cue patterns. Gluck, Shohamy and Myers (2002)
+proposed a number of such rules (or strategies). A main
 difference between these is whether responses are based on the
@@
 
-Alternative non-strategy based analyses of the WPT (Lagnado et al, Speekenbrink et al) 
-have estimated response strategies by logistic regression, allowing the regression
-coefficients to change over time.
+Alternative non-strategy based analyses of the WPT (Lagnado et al, Speekenbrink
+et al) have estimated response strategies by logistic regression, allowing the
+regression coefficients to change over time.
 
@@
 
-As we fit the data to a single subject, 
-we must place some constraints. Specifically, we constrain the state 
+As we fitted a DMM to the data of a single subject, it was necessary
+to place some constraints on the model. Specifically, we constrain the state
 transitions to be in a ``left-right'' format (states can only proceed
@@
 the data. This showed that a two state model was better than a single
-and three state model.
+or three state model.
 
@@
 \begin{tabular}{lcccccccc} \hline
- & & \multicolumn{1}{c}{1 state} & & \multicolumn{2}{c}{2 state} && \multicolumn{2}{c}{2 state (constr.)} \\ \cline{3-3} \cline{5-6} \cline{8-9}
+ & & \multicolumn{1}{c}{1 state} & & \multicolumn{2}{c}{2 state} &&
+ \multicolumn{2}{c}{2 state (constr.)} \\ \cline{3-3} \cline{5-6} \cline{8-9}
 parameter & & $S_1$ & & $S_1$ & $S_2$ & & $S_1$ & $S_2$ \\ \hline
@@
 cue 4 && -1.32 && 0.69 & -3.16 && 0 & -2.67 \\ \hline
- & & \multicolumn{1}{c}{AIC=204.47} & & \multicolumn{2}{c}{AIC=187.50} && \multicolumn{2}{c}{AIC=185.24}
+ & & \multicolumn{1}{c}{AIC=204.47} & & \multicolumn{2}{c}{AIC=187.50} &&
+ \multicolumn{2}{c}{AIC=185.24}
 \end{tabular}
@@
 
-Investigation of the parameter estimates (see Table~\ref{tab:WPT}) indicated that the first state might be
-a single cue strategy (the regression coefficient for the first cue was of much 
-larger magnitude than that of the other cues). The second state was a multi-cue
-strategy (all cues had regression coefficients of reasonable magnitude). 
-To reduce the degrees of freedom, and improve parameter estimates, we implemented 
-constraints to force state 1 into a single cue strategy (fixing the coefficients
-of the remaining three cues to 0) and state 2 in a multi-cue strategy (forcing
-the intercept to 0). These restrictions resulted in a better AIC value of AIC=185.24 
-(df=7). Interestingly, the single cue strategy was somewhat different than 
-described by Gluck et al. Parameter estimates indicated relatively more consistent 
-predictions of ``rain'' in the absence of cue 1 ($Pr(\text{sun}) = 0.22$) and more
-inconsistent predictions of ``sun'' in the presence of cue 1 ($Pr(\text{sun}) = 0.60$). 
-The cue weights of the multi-cue strategy were in the direction of the optimal weights. 
-The Viterbi state sequence indicated that the participant used the single
-cue strategy for the first 60 trials, and then switched to the multi-cue strategy.
+Investigation of the parameter estimates (see Table~\ref{tab:WPT}) showed that
+the regression coefficient for the first cue was of much
+larger magnitude than that of the other cues. As such, the first
+state seems representative of single cue strategy. Alternatively,
+as all regression coefficients are positive, it could indicate a ``counting''
+heuristic, where the propensity of ``sun'' responses increases when
+more cues are present (regardless of which cues they are). However, in that
+case, we would expect the regression coefficients to be of roughly identical
+magnitude. The second state seemed to represent a multi-cue strategy, as all
+cues had regression coefficients of reasonable magnitude, with differing
+directions in line with the objective cue validities.
+
+To reduce the degrees of freedom, and improve parameter estimates, we
+implemented constraints to force state 1 into a single cue strategy (fixing the
+coefficients of the remaining three cues to 0) and state 2 in a multi-cue
+strategy (forcing the intercept to 0). These restrictions resulted in a better
+AIC value of AIC=185.24 (df=7). Interestingly, the single cue strategy was
+somewhat different than described by Gluck et al. Parameter estimates indicated
+relatively more consistent predictions of ``rain'' in the absence of cue 1
+($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun'' in the
+presence of cue 1 ($Pr(\text{sun}) = 0.60$). The cue weights of the multi-cue
+strategy were in the direction of the optimal weights. The Viterbi state
+sequence indicated that the participant used the single cue strategy for the
+first 60 trials, and then switched to the multi-cue strategy.
 

### 2008-11-09 papers/individual/individual.tex
updated individual paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@237 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-% problematic aspects of this analysis
+% problematic aspects of standard analysis
 Typical analyses of these data use the last 60 trials in a series of
-200 trials. A silent assumption that is made in these analyses is that
-behavior has stabilized after 140 trials of learning; this could very
-well be wrong and it is highly likely that there are individual
+200 trials.  A silent assumption that is made in these analyses is
+that behavior has stabilized after 140 trials of learning; this could
+very well be wrong and it is highly likely that there are individual
 differences in this learning process.
@@
 % proposed analysis here
-Single participant choices analyzed.
+Single participant IGT learning data are analyzed with the aim to
+establish 1) at which trial no more learning occurs, i.e., at which
+trial behavior has stabilized, and 2) what strategies participants use
+during and at the end of learning.  Models with an increasing number
+(from 2 through 5) of latent states were fitted to the time series of
+various participants.  The responses were fitted with multinomial
+logistic models for the 4 possible choices that participants make in
+this task.  There were no covariates on any of the parameters.  The
+only constaint that was imposed on the models' parameters was that
+there were designated begin and end states.  This means that the
+initial state probabilities were fixed to one for the first state and
+to zero for the remaning states.  For the transition matrix this means
+that the final state was an absorbing state: the probability of
+transitioning out of that state is zero.  This was done to ensure that
+there would be a final state in which participants end which provides
+the possibility of immediately seeing participants' behavior at the
+end of the task.  Models were selected using the Akaike Information
+Criterion (AIC; Akaike, 1973).
 
-\includegraphics[width=7cm]{graphs/igtdata4.pdf}
+\nocite{Akaike1973} % check the year!!
 
-\begin{itemize}
-	\item N=200
-	\item 4 choice data, displayed in blocks of 20 trials
-	\item optimal strategy is choosing C or D
-\end{itemize}
+We chose to analyze two participants that used different strategies at
+the end of the task as analyzed in the manner proposed by Huizenga et
+al\.  (2007).  The first participants' data were best described by a
+4-state model.  This model's transition matrix is: 
+$$
+\mat{A} = \begin{pmatrix} 
+				0.64 & 0.33 & 0.00 & 0.03 \\
+				0.10 & 0.89 & 0.01 & 0.00 \\
+				0.05 & 0.00 & 0.95 & 0.00 \\
+				0.00 & 0.00 & 0.00 & 1.00
+		  \end{pmatrix}
+$$
+As can be seen from the transition matrix the final state is
+absorbind.  States 2 and 3 are also fairly stable with high diagonal
+probabilities of 0.89 and 0.95 respectively.  The response parameters
+are presented in Table~ref{tab:igt4} below.
 
-Results are the 4-state model (best by AIC/BIC), model predicted
-probabilities are in the figure ???. States are characterized by
-different types of behavior, shifting from B/D strategy to C/D
-(optimal) strategy.
+\begin{table}
+	\caption{Estimates for the Iowa Gambling Task model}
+	\label{tab:igt4}
+	\begin{tabular}{lcccccccc} \hline
+		state & p(A) & p(B) & p(C) & p(D) \\
+		$S_1$ & 0    & 0.06 & 0.31 & 0.63 \\
+		$S_2$ & 0.33 & 0.67 & 0    & 0 \\ 
+		$S_3$ & 0    & 0    & 1    & 0 \\
+		$S_4$ & 0    & 0.06 & 0.94 & 0 
+	\end{tabular}
+\end{table}
 
-\includegraphics[width=7cm]{graphs/igtmodels4.pdf}
+The states have clear interpretations. States 3 and 4 are both
+dominated by C responses, and state 4 has a low probability of B
+responses. States 1 and 2 are dominated C and D and A and B responses 
+respectively. To get a clearer intpretation of these states it is
+necessary to consider when they are visited during the learning
+process. The viterbi algorithm (REFERENCE) provides us with the
+posterior state sequence, i.e., the sequence of states that the
+participant is in at each trial. This sequence is depicted in
+Figure~\ref{fig:post4}.
+
+\begin{figure}
+\begin{center}
+	\scalebox{0.9}{\includegraphics*{graphs/post4.pdf}}
+	\label{fig:post4}
+	\caption{Posterior state sequence for the 4 state model.}
+\end{center}
+\end{figure}
+
+At the start of training there are many switches between states 1 and 
+2, indicating that behavior is mostly random choices between all four 
+categories, albeit with a slight preference for B and D over A and C
+choices. This preference indicates a focus on the frequency of loss
+associated with each of the doors as B and D have low frequency of
+loss. After this, there is a period of stable state 2 behavior
+associated with responses A and B. After that there is stable state 3 
+behavior, consisting only of C responses, a short transitional period 
+and then state 4 behavior consting of 94 \% C choices and some B
+choices, i.e., almost optimal behavior. The choice proportions and the
+model predicted choice proportions are depicted in
+Figure~\ref{fig:igt4}.
+
+\begin{figure}
+\begin{center}
+	\scalebox{1}{\includegraphics*{graphs/igt4.pdf}}
+	\label{fig:igt4}
+	\caption{Data and model predicted proportions of door choices in
+	the IGT.}
+\end{center}
+\end{figure}
+
+The initial preference for B and D choices confirms the theory
+expressed in Huizenga et al. (2007) that frequency of loss is the
+dominant dimension in the IGT. The final states with mostly C 
+choices reprensents one of the optimal strategies, which consists of
+both C and D responses. Note that both C and D reponses generate equal
+profits in the long run.
 
@@
 
-The Weather Prediction Task (WPT, Knowlton, Squire \& Gluck, 1994)
-is a probabilistic categorization task in which participants learn to
-predict (or categorize) the state of the weather (sun or rain) on the basis
-of four ``tarot'' cards (cards with abstract geometrical patterns). On a
-given trial, one, two, or three cues are present. There are a total of
-14 possible cue patterns, and each one is associated with a particular
-probability distribution over the states of the weather. In order to
-perform in the task, participants must predict the weather in accordance
-with these conditional probabilities.
+The Weather Prediction Task (WPT, Knowlton, Squire \& Gluck, 1994) is
+a probabilistic categorization task in which participants learn to
+predict (or categorize) the state of the weather (sun or rain) on the
+basis of four ``tarot'' cards (cards with abstract geometrical
+patterns).  On a given trial, one, two, or three cues are present.
+There are a total of 14 possible cue patterns, and each one is
+associated with a particular probability distribution over the states
+of the weather.  In order to perform in the task, participants must
+predict the weather in accordance with these conditional
+probabilities.
 
@@
 rule-learning, participants are taken to extract rules by which to
-categorize the different cue patterns. Gluck, Shohamy and Myers (2002)
-proposed a number of such rules (or strategies). A main
+categorize the different cue patterns.  Gluck, Shohamy and Myers
+(2002) proposed a number of such rules (or strategies).  A main
 difference between these is whether responses are based on the
 presence/absence of a single cue, or whether responses are based on
-cue patterns. Gluck et al. formulated all strategies in a deterministic and
-optimal manner (e.g., the multi-cue strategy corresponded to giving the optimal
-response to each cue pattern). Meeter et al. allowed for probabilistic
-responding (a small probability of giving the non-optimal response).
+cue patterns.  Gluck et al.  formulated all strategies in a
+deterministic and optimal manner (e.g., the multi-cue strategy
+corresponded to giving the optimal response to each cue pattern).
+Meeter et al.  allowed for probabilistic responding (a small
+probability of giving the non-optimal response).
 
-Alternative non-strategy based analyses of the WPT (Lagnado et al, Speekenbrink
-et al) have estimated response strategies by logistic regression, allowing the
-regression coefficients to change over time.
+Alternative non-strategy based analyses of the WPT (Lagnado et al,
+Speekenbrink et al) have estimated response strategies by logistic
+regression, allowing the regression coefficients to change over time.
 
@@
 Here, we analyze the behavior of a single individual performing the
-WPT for 200 trials. We chose to analyse the ``average'' participant (the
-participant with performance closest to the group average) in a large
-unpublished dataset. We let each state be characterized by a GLM with a
-Binomial distributed response and logistic link function (i.e., a logistic
-regression model). We are particularly interested in evidence for
-strategy switching and whether a DMM can recover a strategy model
-in line with Gluck et al. (2002).
+WPT for 200 trials.  We chose to analyse the ``average'' participant
+(the participant with performance closest to the group average) in a
+large unpublished dataset.  We let each state be characterized by a
+GLM with a Binomial distributed response and logistic link function
+(i.e., a logistic regression model).  We are particularly interested
+in evidence for strategy switching and whether a DMM can recover a
+strategy model in line with Gluck et al.  (2002).
 
 As we fitted a DMM to the data of a single subject, it was necessary
-to place some constraints on the model. Specifically, we constrain the state
-transitions to be in a ``left-right'' format (states can only proceed
-to the immediately adjacent state and never back, and must start in
-the initial state). We fitted a single, two and three state model to
-the data. This showed that a two state model was better than a single
-or three state model.
+to place some constraints on the model.  Specifically, we constrain
+the state transitions to be in a ``left-right'' format (states can
+only proceed to the immediately adjacent state and never back, and
+must start in the initial state).  We fitted a single, two and three
+state model to the data.  This showed that a two state model was
+better than a single or three state model.
 
@@
 
-Investigation of the parameter estimates (see Table~\ref{tab:WPT}) showed that
-the regression coefficient for the first cue was of much
+Investigation of the parameter estimates (see Table~\ref{tab:WPT})
+showed that the regression coefficient for the first cue was of much
 larger magnitude than that of the other cues.  As such, the first
-state seems representative of single cue strategy. Alternatively,
-as all regression coefficients are positive, it could indicate a ``counting''
-heuristic, where the propensity of ``sun'' responses increases when
-more cues are present (regardless of which cues they are). However, in that
-case, we would expect the regression coefficients to be of roughly identical
-magnitude. The second state seemed to represent a multi-cue strategy, as all
-cues had regression coefficients of reasonable magnitude, with differing
-directions in line with the objective cue validities.
+state seems representative of single cue strategy.  Alternatively, as
+all regression coefficients are positive, it could indicate a
+``counting'' heuristic, where the propensity of ``sun'' responses
+increases when more cues are present (regardless of which cues they
+are).  However, in that case, we would expect the regression
+coefficients to be of roughly identical magnitude.  The second state
+seemed to represent a multi-cue strategy, as all cues had regression
+coefficients of reasonable magnitude, with differing directions in
+line with the objective cue validities.
 
 To reduce the degrees of freedom, and improve parameter estimates, we
-implemented constraints to force state 1 into a single cue strategy (fixing the
-coefficients of the remaining three cues to 0) and state 2 in a multi-cue
-strategy (forcing the intercept to 0). These restrictions resulted in a better
-AIC value of AIC=185.24 (df=7). Interestingly, the single cue strategy was
-somewhat different than described by Gluck et al. Parameter estimates indicated
-relatively more consistent predictions of ``rain'' in the absence of cue 1
-($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun'' in the
-presence of cue 1 ($Pr(\text{sun}) = 0.60$). The cue weights of the multi-cue
-strategy were in the direction of the optimal weights. The Viterbi state
-sequence indicated that the participant used the single cue strategy for the
-first 60 trials, and then switched to the multi-cue strategy.
+implemented constraints to force state 1 into a single cue strategy
+(fixing the coefficients of the remaining three cues to 0) and state 2
+in a multi-cue strategy (forcing the intercept to 0).  These
+restrictions resulted in a better AIC value of AIC=185.24 (df=7).
+Interestingly, the single cue strategy was somewhat different than
+described by Gluck et al.  Parameter estimates indicated relatively
+more consistent predictions of ``rain'' in the absence of cue 1
+($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun''
+in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
+of the multi-cue strategy were in the direction of the optimal
+weights.  The Viterbi state sequence indicated that the participant
+used the single cue strategy for the first 60 trials, and then
+switched to the multi-cue strategy.
 

### 2008-11-10 papers/individual/individual.tex
added new bib file and updated refs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@238 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 changes in the strategies that children apply in all kinds of tasks
-such as the conservation learning and the balance scale task.  Van der
-Maas et al.  (1992) developed a catastrophe model to describe phase
+such as the conservation learning and the balance scale task.
+\citet{Maas1992} developed a catastrophe model to describe phase
 transitions in learning and developmental processes.  They applied the
@@
 tend to ignore the width dimension and hence always choose the glass
-with the highest level of liquid (REFERENCE??).  Van der Maas et al.
-(1992) showed that there is a sudden transition to a new strategy in
-which the children also take the width of the glasses into account
-when judging the volume of liquids.
+with the highest level of liquid (REFERENCE??).  \citet{Maas1992}
+showed that there is a sudden transition to a new strategy in which
+the children also take the width of the glasses into account when
+judging the volume of liquids.
 
 % balance scale rules
-Jansen and Van der Maas (2001) applied the catastrophe model to
+\citet{Jansen2001} applied the catastrophe model to
 development of strategies on the balance scale task (Siegler, 1981).
@@
 number of weights on each side of the fulcrum.  This strategy for
-solving balance scale items is called Rule 1 (Siegler, 1981).  Older
+solving balance scale items is called Rule 1 \cite{Siegler1981}.  Older
 children include the distance dimension in determining their response
@@
 the number of weights is equal on both sides of the balance scale.
-This strategy is called Rule 2 (Siegler, 1981).
+This strategy is called Rule 2 \cite{Siegler1981}.
 
 % hysteresis on the balance scale
-Jansen and Van der Maas (2001) found clear evidence for stage-wise
+\citet{Jansen2001} found clear evidence for stage-wise
 transitions between Rule 1 and Rule 2 by testing criteria that were
@@
 in-between strategies: children apply either Rule 1 or Rule 2 and
-there is no in-between option.  Jansen and Van der Maas also found
+there is no in-between option.  \citet{Jansen2001} also found
 evidence for hysteresis: the phenomenon that switching between
@@
 Also in animal learning and conditioning, evidence is found for sudden
-changes in response behavior (Gallistel et al 2004).  In particular,
+changes in response behavior \citet{Gallistel2004}.  In particular,
 in their study, evidence was found for sudden onset of learning: at
@@
 form or color.  This kind of learning is referred to as all-or-none
-learning or concept identification learning.  Raijmakers et al (2001)
+learning or concept identification learning.  \citet{Raijmakers2001}
 found evidence for different strategies applied by children when faced
-with such a learning task.  Schmittmann et al (2006) reanalyzed their
+with such a learning task.  \citet{Schmittmann2006} reanalyzed their
 data using hidden Markov models to show that both strategies are
@@
 many repeated measurements, or time series, observed in only a few
-participants or even just a single participant.  For example, Visser,
-Raijmakers, \& Van der Maas (2008) analyzed data from a single
-participant in an experimental task that manipulates the trade-off
-between speed and accuracy.  The data consisted of three time series
-with each around 150 repeated measurements of both reaction time and
-accuracy.  Below we provide examples of analyzing time series from
-single participants from two experiments; one from the Iowa Gambling
-Task and one from the weather prediction task.  The interest in these
-tasks is to show that participants develop different strategies over
-time in responding to the stimuli and that the transition from one
-strategy to the next is a discrete event.  Before providing these
-illustrations, below we give a formalization of dependent mixture
-models and a brief overview of the DepmixS4 package that was developed
-to specify and fit such models.
+participants or even just a single participant.  For example,
+\citet{Visser2009} analyzed data from a single participant in an
+experimental task that manipulates the trade-off between speed and
+accuracy.  The data consisted of three time series with each around
+150 repeated measurements of both reaction time and accuracy.  Below
+we provide examples of analyzing time series from single participants
+from two experiments; one from the Iowa Gambling Task and one from the
+weather prediction task.  The interest in these tasks is to show that
+participants develop different strategies over time in responding to
+the stimuli and that the transition from one strategy to the next is a
+discrete event.  Before providing these illustrations, below we give a
+formalization of dependent mixture models and a brief overview of the
+DepmixS4 package that was developed to specify and fit such models.
 
@@
 The Iowa gambling task (IGT) is an experimental paradigm designed to
-mimic real-life decision-making situations (Bechara, Damasio, Damasio
-\& Anderson, 1994), in the way that it factors uncertainty, reward and
-punishment (Dunn, Dalgleish, \& Lawrence, 2006).  The task requires
-the selection of cards from four decks.  Each deck is characterized by
-a certain amount of gain (delivered on each draw), frequency of loss,
-and amount of loss.  Two decks (A and B) yield consistently high
-rewards, but also high, probabilistic penalties and are both (equally)
-disadvantageous in the long run.  The other two decks (C and D) yield
-consistently smaller rewards, but also low, probabilistic penalties
-and are both (equally) advantageous in the long run.  It is assumed
-that the ventromedial prefrontal cortex (VMPFC) is active in the IGT
-as VMPC patients show impaired task performance.  Their preference for
-the decks with immediate high rewards indicates ``myopia for the
-future''.
+mimic real-life decision-making situations \cite{Bechara1994}, in the
+way that it factors uncertainty, reward and punishment
+\cite{Dunn2006}.  The task requires the selection of cards from four
+decks.  Each deck is characterized by a certain amount of gain
+(delivered on each draw), frequency of loss, and amount of loss.  Two
+decks (A and B) yield consistently high rewards, but also high,
+probabilistic penalties and are both (equally) disadvantageous in the
+long run.  The other two decks (C and D) yield consistently smaller
+rewards, but also low, probabilistic penalties and are both (equally)
+advantageous in the long run.  It is assumed that the ventromedial
+prefrontal cortex (VMPFC) is active in the IGT as VMPC patients show
+impaired task performance.  Their preference for the decks with
+immediate high rewards indicates ``myopia for the future''.
 
 % what is the HDT? developmental trends and relevance?
-Crone \& van der Molen (2004) designed a developmentally appropriate
-analogue of the IGT, the Hungry Donkey Task (HDT), with a similar win
-and loss schedule although the abolute amounts were redcuced by a
-factor of 25.  The HDT is a pro-social game inviting the player to
-assist a hungry donkey to collect as many apples as possible, by
-opening one of four doors.  Again, doors A and B are characterized by
-a high constant gain (10 apples), whereas doors C and D deliver a low
-constant gain (2 apples).  At doors A and C, a loss of 50 apples (A)
-or 10 apples (C) is delivered in 50\% of the trials.  For doors B and
-D, frequency of loss is only 10\%.  The median loss of doors B and D
-is 10 and 2, respectively.  Crone and van der Molen administered the
-HDT to children from four age groups (6-9, 10-12, 13-15, and 18-25
-year-olds) and concluded that children also fail to consider future
-consequences.
+\cite{Crone2004} designed a developmentally appropriate analogue of
+the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
+schedule although the abolute amounts were redcuced by a factor of 25.
+The HDT is a pro-social game inviting the player to assist a hungry
+donkey to collect as many apples as possible, by opening one of four
+doors.  Again, doors A and B are characterized by a high constant gain
+(10 apples), whereas doors C and D deliver a low constant gain (2
+apples).  At doors A and C, a loss of 50 apples (A) or 10 apples (C)
+is delivered in 50\% of the trials.  For doors B and D, frequency of
+loss is only 10\%.  The median loss of doors B and D is 10 and 2,
+respectively.  \cite{Crone2004} administered the HDT to children from
+four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
+that children also fail to consider future consequences.
 
@@
 % finite mixture analysis
-A reanalysis of this dataset (Huizenga, Crone, \& Jansen, 2007)
-indicated that participants might solve the task by sequentially
-considering the three dimensions (constant gain, frequency of loss,
-and amount of loss) in order to choose a door.  Most youngest children
-in the dataset seem to focus on the dominant dimension in the task,
-frequency of loss, resulting in equal preference for doors B and D.
-Older participants seem to use a two-dimensional rule where
-participants first focus on the frequency of loss and then consider
-amount of loss, resulting in a preference for door D. A third very
-small subgroup seems to use an integrative rule where participants
-combine all three dimensions in the appropriate way.  Participants
-using the integrative rule pick cards from doors C and D, which are
-advantageous in the long run.
+A reanalysis of this dataset \cite{Huizenga2007} indicated that
+participants might solve the task by sequentially considering the
+three dimensions (constant gain, frequency of loss, and amount of
+loss) in order to choose a door.  Most youngest children in the
+dataset seem to focus on the dominant dimension in the task, frequency
+of loss, resulting in equal preference for doors B and D. Older
+participants seem to use a two-dimensional rule where participants
+first focus on the frequency of loss and then consider amount of loss,
+resulting in a preference for door D. A third very small subgroup
+seems to use an integrative rule where participants combine all three
+dimensions in the appropriate way.  Participants using the integrative
+rule pick cards from doors C and D, which are advantageous in the long
+run.
 
@@
 We chose to analyze two participants that used different strategies at
-the end of the task as analyzed in the manner proposed by Huizenga et
-al\.  (2007).  The first participants' data were best described by a
-4-state model.  This model's transition matrix is: 
+the end of the task as analyzed in the manner proposed by
+\citet{Huizenga2007}.  The first participants' data were best
+described by a 4-state model.  This model's transition matrix is:
 $$
@@
 \begin{center}
-	\scalebox{0.9}{\includegraphics*{graphs/post4.pdf}}
+	\includegraphics[width=\textwidth]{graphs/post4.pdf}
 	\label{fig:post4}
@@
 \begin{center}
-	\scalebox{1}{\includegraphics*{graphs/igt4.pdf}}
+	\includegraphics[width=\textwidth]{graphs/igt4.pdf}
 	\label{fig:igt4}
@@
 The initial preference for B and D choices confirms the theory
-expressed in Huizenga et al. (2007) that frequency of loss is the
+expressed in \citet{Huizenga2007} that frequency of loss is the
 dominant dimension in the IGT. The final states with mostly C 
@@
 
+\nocite{Knowlton1994}
+
 %The WPT has been popular in neuropsychological research,
@@
 rule-learning, participants are taken to extract rules by which to
-categorize the different cue patterns.  Gluck, Shohamy and Myers
-(2002) proposed a number of such rules (or strategies).  A main
-difference between these is whether responses are based on the
-presence/absence of a single cue, or whether responses are based on
-cue patterns.  Gluck et al.  formulated all strategies in a
-deterministic and optimal manner (e.g., the multi-cue strategy
-corresponded to giving the optimal response to each cue pattern).
-Meeter et al.  allowed for probabilistic responding (a small
-probability of giving the non-optimal response).
+categorize the different cue patterns.  \cite{Gluck2002} proposed a
+number of such rules (or strategies).  A main difference between these
+is whether responses are based on the presence/absence of a single
+cue, or whether responses are based on cue patterns.  Gluck et al.
+formulated all strategies in a deterministic and optimal manner (e.g.,
+the multi-cue strategy corresponded to giving the optimal response to
+each cue pattern).  Meeter et al.  allowed for probabilistic
+responding (a small probability of giving the non-optimal response).
 
-Alternative non-strategy based analyses of the WPT (Lagnado et al,
-Speekenbrink et al) have estimated response strategies by logistic
-regression, allowing the regression coefficients to change over time.
+Alternative non-strategy based analyses of the WPT
+\cite{Lagnado??,Speekenbrink??} have estimated response strategies by
+logistic regression, allowing the regression coefficients to change
+over time.
 
@@
 in evidence for strategy switching and whether a DMM can recover a
-strategy model in line with Gluck et al.  (2002).
+strategy model in line with \cite{Gluck2002}.
 
@@
 Interestingly, the single cue strategy was somewhat different than
-described by Gluck et al.  Parameter estimates indicated relatively
+described by \citet{Gluck2002}  Parameter estimates indicated relatively
 more consistent predictions of ``rain'' in the absence of cue 1
@@
 
-References
-
-Bechara, A., Damasio, A. R., Damasio, H., \& Anderson, S. W.(1994).
-Insensitivity to future consequences following damage to human
-prefrontal cortex.  Cognition, 50(1Ð3), 7Ð15.
-
-Chambers, J. M. (1998). Programming with Data: A Guide to the S Lan-
-guage. New York: Springer-Verlag.
-
-Crone, E. A.,  \& van der Molen, M. W. (2004).  Developmental changes in
-real life decision making: Performance on a gambling task previously
-shown to depend on the ventromedial prefrontal cortex.  Developmental
-Neuropsychology, 25(3), 251-279.
-
-Dunn, B. D., Dalgleish, T.,  \& Lawrence, A. D. (2006).  The somatic
-marker hypothesis: A critical evaluation.  Neuroscience and
-Biobehavioral Reviews, 30(2), 239-271.
-
-Gluck, M. A., Shohamy, D., \& Myers, C. (2002). How do people solve the
-weather prediction task?: Individual variability in strategies for
-probabilistic category learning. Learning \& Memory, 9, 408-418.
-
-Huizenga, H. M., Crone, E. A., \& Jansen, B. R. J. (2007).
-Decision-making in healthy children, adolescents and adults explained
-by the use of increasingly complex proportional reasoning rules.
-Developmental Science, 10(6), 814-825.
-
-Knowlton, B. J., Squire, L. R., \& Gluck, M. A. (1994).
-Probabilistic classification learning in amnesia.
-Learning \& Memory, 1 , 106-120.
-
-Siegler, R. S. (1981).  Developmental sequences within and between
-concepts.  Monographs of the Society for Research in Child
-Development, 46(2, Serial No.  189).
-
-Visser, Raijmakers, \& Van der Maas (2008). Dynamics book chapter.
-
 \section*{Author note}
@@
 
-\bibliography{all,ingmar}
+\bibliography{all,ingmar,individual}
 

### 2008-11-11 papers/individual/individual.tex
Updated the individual paper and removed unneccesary graphics files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@239 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 the experimental situation.  The model is illustrated with an example
-of participants' learning in the weather prediction task.}
+of participants' learning in the Weather Prediction Task and in the 
+Iowa Gambling Task.}
 
@@
 % phenomena of discrete change in development/learning
-Piagetian developmental theory (REFERENCE??)  assumes step-wise
-changes in the strategies that children apply in all kinds of tasks
-such as the conservation learning and the balance scale task.
-\citet{Maas1992} developed a catastrophe model to describe phase
-transitions in learning and developmental processes.  They applied the
-catastrophe model to learning in the conservation of liquid task
-(REFERENCE??)  in which children have to judge relative volumes of
-liquid in glasses of different heights and widths.  Young children
-tend to ignore the width dimension and hence always choose the glass
-with the highest level of liquid (REFERENCE??).  \citet{Maas1992}
-showed that there is a sudden transition to a new strategy in which
-the children also take the width of the glasses into account when
-judging the volume of liquids.
+Piagetian developmental theory assumes step-wise changes in the
+strategies that children apply in all kinds of tasks such as the
+conservation learning and the balance scale task (REFERENCE??).
+\citet{Maas1992} developed a catastrophe model to describe step-wise
+changes or phase transitions in learning and developmental processes.
+They applied the catastrophe model to learning in the conservation of
+liquid task (REFERENCE??)  in which children have to judge relative
+volumes of liquid in glasses of different heights and widths.  Young
+children tend to ignore the width dimension and hence always choose
+the glass with the highest level of liquid (REFERENCE??).
+\citet{Maas1992} showed that there is a sudden transition to a new
+strategy in which the children also take the width of the glasses into
+account when judging the volume of liquids.
 
@@
 Also in animal learning and conditioning, evidence is found for sudden
-changes in response behavior \citet{Gallistel2004}.  In particular,
-in their study, evidence was found for sudden onset of learning: at
-the start of the learning experiment, the pigeons did not learn
-anything and performance was stable; after a number of trials,
-learning kicks in and there are large increases in performance.  The
-interest here is in modeling the distribution of onset times: that is,
-the trials at which learning suddenly takes off.  A similar interest
-in process onset times is found in addiction research.  For example,
-(REFERENCE) study the age at which children start using alcohol and
-how this related to eventual outcomes in terms of addiction.
+changes in response behavior \citet{Gallistel2004}.  In particular, in
+their study, evidence was found for sudden onset of learning: at the
+start of the learning experiment, the pigeons did not learn anything
+and performance was stable; after a number of trials, learning kicks
+in and there are large increases in performance.  The interest here is
+in modeling the distribution of onset times: that is, the trials at
+which learning suddenly takes off.  A similar interest in process
+onset times is found in addiction research.  For example,
+\cite{Sher2004} study the age at which children start using alcohol
+and how this related to eventual outcomes in terms of addiction.
 
@@
 from two experiments; one from the Iowa Gambling Task and one from the
-weather prediction task.  The interest in these tasks is to show that
+Weather Prediction Task.  The interest in these tasks is to show that
 participants develop different strategies over time in responding to
@@
 time series data.  The dependent mixture model is similar to, but
-slightly different from two other types of models that are in use for
+slightly different from, two other types of models that are in use for
 modelling discrete change: the hidden and the latent Markov models.
@@
 these models, the focus is on survey type data: a few repeated
-measurements taken in a large sample.  \citeA{Langeheine1990}
-discusses latent Markov models and their use in sociology ands
-political science (see also McCutcheon, 1987).  Latent transition
-models, for example, been used in studying development of math skills
-\citep{Collins1992} and in medical applications \citep{Reboussin1998};
-\citet{Kaplan2008} provides an overview of such models, that are
-called stage-sequential models in the developmental psychology
-literature.
+measurements taken in a large sample.  \citeA{Langeheine1990} discuss
+latent Markov models and their use in sociology and political science
+(see also McCutcheon, 1987).  Latent transition models have been used
+in studying development of math skills \citep{Collins1992} and in
+medical applications \citep{Reboussin1998}; \citet{Kaplan2008}
+provides an overview of such models, that are called stage-sequential
+models in the developmental psychology literature.
 
@@
 Hidden Markov models (HMM) tend to be used in the analysis of long
-univariate and individual timeseries.  For example, HMMs are the model
+univariate (individual) timeseries.  For example, HMMs are the model
 of choice in speech recognition applications \cite{Rabiner1989}.  In
-biology, HMMs are used to analyze DNA sequences and in econometric
-science, to analyze changes in stock market prices and commodities
-\cite{Kim1994}.
+biology, HMMs are used to analyze DNA sequences \cite{Krogh1998} and
+in econometric science, to analyze changes in stock market prices and
+commodities \cite{Kim1994}.
 
@@
 \end{enumerate}
+Here $\xi_{t}$ and $\zeta_{t}$ are independent error processes 
+\cite{Elliott1995}.
 
@@
 no knowledge of which features are important in categorization.
+Hence, the states in the state space represent knowledge states of the
+participants, such that different knowledge states lead to different
+observed behavior or responses.
 
@@
 so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
-rather by the forward part of this algorithm.  \cite{Lystig2002}
+rather by the forward part of this algorithm.  \citet{Lystig2002}
 changed the forward algorithm in such a way as to allow computing the
@@
 
-To compute the log-likelihood, \cite{Lystig2002} define the following
+To compute the log-likelihood, \citet{Lystig2002} define the following
 (forward) recursion:
@@
 be fitted on mixed multivariate data with multinomial and/or gaussian
-distributions.  Parameters can be estimated subject to general linear
-constraints.  Parameter estimation is done through an EM algorithm or
-by a direct optimization approach with gradients using the
-\code{Rdonlp2} optimization routine when contraints are imposed on the
-parameters.
+distributions.
 
@@
 models are used commonly in the social sciences, but no comprehensive
-package was available for fitting such models.  Common programs for
-Markovian models include Panmark \citep{Pol1996}, and for latent class
-models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
-number of important features.  There are currently some packages in R
-that handle hidden Markov models but they lack a number of features
-that we needed in our research.  In particular, \pkg{depmixS4} was
-designed to meet the following goals:
+package was available for fitting such models in R. Common programs
+for Markovian models include Panmark \citep{Pol1996}, and for latent
+class models Latent Gold \citep{Vermunt2003}.  Those programs are
+lacking a number of features that were needed in our own research.  In
+particular, \pkg{depmixS4} was designed to meet the following goals:
 \begin{enumerate}
@@
 	\item to be able to fit transition models with covariates, i.e.,
-	to have time-dependent transition matrices
+	to have time-dependent transition matrices.
 
 	\item to be able to include covariates in the prior or initial
-	state probabilities of models
+	state probabilities of models.
 
-	\item to allow for easy extensibility, in particular, to be able
+	\item to allow for easy extensibility; in particular, to be able
 	to add new response distributions, both univariate and
 	multivariate, and similarly to be able to allow for the addition
-	of other transition models, e.g., continuous time observation
-	models
+	of other transition models, e.g., models for observations in 
+	continuous time.
 
@@
 time dependencies between observed data, and the model reduces to a
-finite mixture model, or a latent class model.  Although there are
-other specialized (R) packages to deal with mixture data, one specific
-feature that we needed which is not available in other packages is the
-possibility to include covariates on the prior probabilities of class
-membership.
+finite mixture model \cite{McLachlan2000}, or a latent class model
+\cite{McCutcheon1987}.  Although there are other specialized (R-)
+packages to deal with mixture data, one specific feature that we
+needed which is not available in other packages is the possibility to
+include covariates on the prior probabilities of class membership.
 
-\subsection{Response distributions and parameters}
+The package is built using S4 classes (object oriented classes in R)
+to allow easy extensibility (Chambers, 1998). A depmix model consists 
+of the following: 
+\begin{enumerate} 
+	\item A model for the initial probabilities or prior 
+	\item A list of transition models; one model for each row of the 
+	transition matrix
+	\item A list of response models for each of the states of the model; 
+	this is a list of lists in the case of multivariate responses. 
+\end{enumerate} 
+Each of these are adressed briefly below.
 
-% this needs some more on the design, interface to glm and similar
-% other models etc.
-The package is built using S4 classes (object oriented classes in R)
-to allow easy extensibility (Chambers, 1998).
+\subsection{Transition and prior probabilities models}
 
-Each row of the transition matrix and the initial state probabilities:
-\begin{itemize}
-	\item is modeled as a multinomial distribution
-	\item uses the logistic link function to include covariates
-	\item can have time-dependent covariates
-\end{itemize}
+% transition and initial state probs
+Each row of the transition matrix and the vector of initial state 
+probabilities is modelled as a baseline-category multinomial logistic 
+model \cite<see>[chapter 7]{Agresti2002}. This means that covariates 
+can be used as predictors in these models. In particular, the model 
+for each multinomial is:
+\begin{equation} 
+	\pi_{i} = \frac{\exp(\alpha_{i}+\mathbf{\beta}^{`}_{i}\vc{x})}
+				{1+\sum_{j=1}^{J-1}\exp(\alpha_{j}+\mathbf{\beta}^{`}_{j}\vc{x})},
+\end{equation}
+where $\pi_{i}$ is the probability (e.g.\ the probability of the
+$i$-th initial state); the $\alpha_{i}$ are the category intercepts;
+the $\mathbf{\beta}_{i}$ are the category regression coefficients;
+$\vc{x}$ is a vector of covariates or predictors; $J$ is the number
+of states; in this example, $J$ serves as the baseline-category,
+meaning that $\alpha_{J}$ and $\mathbf{\beta}_{J}$ are zero.
 
-% response distributions include the possibility to have direct
-% effects of covariates or predictors as well as indirect effects
-% through the use of covariates on the transition matrix and the
-% initial state probabilities
-Current options for the response models are models from the
-generalized linear modeling framework:
+\subsection{Response models}
 
-\begin{itemize}
-	\item normal distribution; continuous, gaussian data
-	\item binomial logistic; binary data
-	\item Poisson distribution; count data
-	\item multinomial logistic; multiple choice data
-\end{itemize}
+% response models
+Response models in \pkg{depmixS4} interface with the \code{glm} 
+functions available in R and with the \pkg{nnet}-package for the 
+multinomial logistic models. Normally distributed data can hence be 
+modelled with direct effects included as well. For example, normal 
+data are modelled as:
+\begin{equation}
+	O_{t}|S_{t} = \mu + \mathbf{\beta}^{`}\vc{x} + \epsilon_{t},
+\end{equation}
+where $\mu$ the (state-dependent) mean; $\vc{x}$ is a vector of
+covariates or predictors; $\mathbf{\beta}$ is the vector of
+(state-dependent) regression coefficients; $\epsilon_{t}$ is a
+normally distributed error with (state-dependent) standard deviation
+$\sigma$.
 
-All response models have the option of including covariates
-Other link functions may be used; eg the probit for binary data.
+Multinomial responses are modeled by the same multinomial 
+baseline-category logistic model as is used for the transition 
+probabilities and the initial state probabilities. Other response 
+functions from the generalized linear model family that can be used 
+include: binomial logistic regression, poisson regression, gamma and 
+inverse gaussian. There is also support for the multivariate normal 
+distribution.
 
-% possibility of multivariate responses: local independence or true
-% multivariate distributions, eg multivariate normal
+% MAARTEN: de multinomial logistic models maken weer gebruik van andere
+% packages toch, kun je daar de referentie van geven?
 
@@
 Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
-through the use of a general Newton-Raphson optimizer.  The EM
-algorithm however has some drawbacks.  First, it can be slow to
-converge towards the end of optimization (although it is usually
-faster than direct optimization at the start, so possibly a
-combination of EM and direct optimization is fastest).  Second,
-applying constraints to parameters can be problematic; in particular,
-EM can lead to wrong parameter estimates when applying constraints.
-Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
-models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
-\cite{Tamura2007,Spellucci2002}, because it handles general linear
-(in)equality constraints, and optionally also non-linear constraints.
+through the use of a general Newton-Raphson optimizer.  EM is used by
+default in unconstrained models, but otherwise, direct optimization is
+done using \pkg{Rdonlp2} \cite{Tamura2007,Spellucci2002}, because it
+handles general linear (in)equality constraints, and optionally also
+non-linear constraints.
 
@@
 
+General issues
 \begin{itemize}
-	\item depmixS4 can be downloaded from: http://r-forge.r-project.org/depmix/
-	\item It is feasible to fit hidden Markov models in moderate length time series
-	\item Many applications in experimental psychology
-	\item Future developments:
+	\item It is feasible to fit hidden Markov models in moderate length individual time series
+	\item illustrations show results that are consistent with usual group based analyses AND 
+	provide extra information on switching between strategies over the course of learning
+	\item Many applications in experimental psychology: examples????
+\end{itemize}
+
+Specific insights from the illustrations: suggestions welcome
+
+Possible future extensions in depmixS4
 \begin{enumerate}
@@
 \end{enumerate}
-	\item standard errors of parameters
-\end{itemize}
 

### 2008-11-12 papers/individual/individual.tex
Updated authors in individual paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@240 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@241 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\documentclass[a4paper,12pt,man]{apa} % nobf, doc, apa
+\documentclass[a4paper,12pt,man,english]{apa} % nobf, doc, apa
+
+\usepackage[english]{babel} % ik krijg anders foute citaties...in Zweeds formaat???
 
@@
 
-\author{Ingmar Visser \& Maarten Speekenbrink}
+\author{Ingmar Visser, Brenda R. J. Jansen, \& Maarten Speekenbrink}
 
@@
 
-The Weather Prediction Task (WPT, Knowlton, Squire \& Gluck, 1994) is
+The Weather Prediction Task \cite<WPT>{Knowlton1994} %(WPT, Knowlton, Squire \& Gluck, 1994) 
+is
 a probabilistic categorization task in which participants learn to
@@
 Alternative non-strategy based analyses of the WPT
-\cite{Lagnado??,Speekenbrink??} have estimated response strategies by
+\cite{Lagnado2006,Speekenbrink2008} have estimated response strategies by
 logistic regression, allowing the regression coefficients to change
@@
 in evidence for strategy switching and whether a DMM can recover a
-strategy model in line with \cite{Gluck2002}.
+strategy model in line with \citeA{Gluck2002}.
 

### 2008-11-17 papers/individual/individual.tex
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@242 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %&LaTeX
-
 \documentclass[a4paper,12pt,man,english]{apa} % nobf, doc, apa
@@
 
-\author{Ingmar Visser, Brenda R. J. Jansen, \& Maarten Speekenbrink}
+\twoauthors{Ingmar Visser \& Brenda R. J. Jansen}{Maarten Speekenbrink}
 
@@
 
-\affiliation{Department of Psychology, University of Amsterdam\\
- Correspondence concerning this article should be addressed to:
+\twoaffiliations{Department of Psychology, University of Amsterdam}{Department of Cognitive, Perceptual and Brain Sciences, University College London}
+
+\note{Correspondence concerning this article should be addressed to:
  \\
@@
 include time-varying predictors.  In effect, this model consists of
-mixtures of general linear models with Markovian depencies over time
+mixtures of generalized linear models with Markovian dependencies over time
 to model the change process.  In addition, transition parameters can
@@
 the experimental situation.  The model is illustrated with an example
-of participants' learning in the Weather Prediction Task and in the 
-Iowa Gambling Task.}
+of participants' learning in the Iowa Gambling Task and in the Weather 
+Prediction Task.}
 
@@
 discrimination learning and in conditioning.  This chapter is
-concerned with detecting the time points of change in (individual)
+concerned with detecting the nature and time points of change in (individual)
 time series.  We present a framework of dependent mixture models that can
@@
 transitions between Rule 1 and Rule 2 by testing criteria that were
-derived from the catastrophe model.  In particular, they found bimodal
+derived from the catastrophe model.  In particular, she found bimodal
 test scores and inaccessibility.  The latter means that there are no
@@
 evidence for hysteresis: the phenomenon that switching between
-strategies is assymetric.  Children can switch from Rule 1 to Rule 2
+strategies is asymmetric.  Children can switch from Rule 1 to Rule 2
 and back, but this occurs at different trials.  In particular, if the
@@
 Also in animal learning and conditioning, evidence is found for sudden
-changes in response behavior \citet{Gallistel2004}.  In particular, in
+changes in response behavior \cite{Gallistel2004}.  In particular, in
 their study, evidence was found for sudden onset of learning: at the
-start of the learning experiment, the pigeons did not learn anything
-and performance was stable; after a number of trials, learning kicks
-in and there are large increases in performance.  The interest here is
-in modeling the distribution of onset times: that is, the trials at
+start of the learning experiment, pigeons did not learn anything
+and performance was stable; after a number of trials, learning kicked
+in and there were large increases in performance.  \citeauthor{Gallistel2004} 
+focused on modeling the distribution of onset times: that is, the trials at
 which learning suddenly takes off.  A similar interest in process
 onset times is found in addiction research.  For example,
-\cite{Sher2004} study the age at which children start using alcohol
+\citet{Sher2004} study the age at which children start using alcohol
 and how this related to eventual outcomes in terms of addiction.
@@
 formalization of dependent mixture models and a brief overview of the
-DepmixS4 package that was developed to specify and fit such models.
+\pkg{depmixS4} package that was developed to specify and fit such models.
 
@@
 Markov models have been used extensively in the social sciences; for
-example, in analyzing language learning \cite{Miller1952,Miller1963},
+example, in analyzing language learning \cite{Miller1952,Miller1963} and
 in the analysis of paired associate learning \cite{Wickens1982}.  In
@@
 class models Latent Gold \citep{Vermunt2003}.  Those programs are
-lacking a number of features that were needed in our own research.  In
+lacking a number of features that were needed in our research.  In
 particular, \pkg{depmixS4} was designed to meet the following goals:
@@
 finite mixture model \cite{McLachlan2000}, or a latent class model
-\cite{McCutcheon1987}.  Although there are other specialized (R-)
-packages to deal with mixture data, one specific feature that we
-needed which is not available in other packages is the possibility to
-include covariates on the prior probabilities of class membership.
+\cite{McCutcheon1987}.  Although there are other specialized (R)
+packages to deal with mixture data, these don't allow the inclusion of 
+covariates on the prior probabilities of class membership.
 
-The package is built using S4 classes (object oriented classes in R)
+\pkg{depmixS4} is built using S4 classes (object oriented classes in R)
 to allow easy extensibility (Chambers, 1998). A depmix model consists 
@@
 \begin{enumerate} 
-	\item A model for the initial probabilities or prior 
+	\item A model for the initial or prior probabilities 
 	\item A list of transition models; one model for each row of the 
@@
 % transition and initial state probs
-Each row of the transition matrix and the vector of initial state 
+By default, each row of the transition matrix and the vector of initial state 
 probabilities is modelled as a baseline-category multinomial logistic 
@@
 through the use of a general Newton-Raphson optimizer.  EM is used by
-default in unconstrained models, but otherwise, direct optimization is
+default in unconstrained models, but otherwise direct optimization is
 done using \pkg{Rdonlp2} \cite{Tamura2007,Spellucci2002}, because it
@@
 % what is the HDT? developmental trends and relevance?
-\cite{Crone2004} designed a developmentally appropriate analogue of
+\citet{Crone2004} designed a developmentally appropriate analogue of
 the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
-schedule although the abolute amounts were redcuced by a factor of 25.
+schedule although the absolute amounts were reduced by a factor of 25.
 The HDT is a pro-social game inviting the player to assist a hungry
@@
 loss is only 10\%.  The median loss of doors B and D is 10 and 2,
-respectively.  \cite{Crone2004} administered the HDT to children from
+respectively.  \citet{Crone2004} administered the HDT to children from
 four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
@@
 three dimensions (constant gain, frequency of loss, and amount of
-loss) in order to choose a door.  Most youngest children in the
-dataset seem to focus on the dominant dimension in the task, frequency
+loss) in order to choose a door.  The youngest children in the
+dataset mostly seem to focus on the dominant dimension in the task, frequency
 of loss, resulting in equal preference for doors B and D. Older
@@
 Typical analyses of these data use the last 60 trials in a series of
-200 trials.  A silent assumption that is made in these analyses is
-that behavior has stabilized after 140 trials of learning; this could
+200 trials.  A silent assumption made in these analyses is
+that behavior has stabilized after 140 learning trials; this could
 very well be wrong and it is highly likely that there are individual
@@
 % proposed analysis here
-Single participant IGT learning data are analyzed with the aim to
-establish 1) at which trial no more learning occurs, i.e., at which
+Here, we analyze IGT learning data from a single participant with the aim to
+establish 1) at which trial learning stops, i.e., at which
 trial behavior has stabilized, and 2) what strategies participants use
@@
 this task.  There were no covariates on any of the parameters.  The
-only constaint that was imposed on the models' parameters was that
+only constraint imposed on the models' parameters was that
 there were designated begin and end states.  This means that the
 initial state probabilities were fixed to one for the first state and
-to zero for the remaning states.  For the transition matrix this means
+to zero for the remaining states.  For the transition matrix this means
 that the final state was an absorbing state: the probability of
@@
 
-We chose to analyze two participants that used different strategies at
-the end of the task as analyzed in the manner proposed by
-\citet{Huizenga2007}.  The first participants' data were best
+%We chose to analyze two participants that used different strategies at
+%the end of the task as analyzed in the manner proposed by
+%\citet{Huizenga2007}.  The first participants' data 
+The data were best 
 described by a 4-state model.  This model's transition matrix is:
@@
 As can be seen from the transition matrix the final state is
-absorbind.  States 2 and 3 are also fairly stable with high diagonal
+absorbing.  States 2 and 3 are also fairly stable with high diagonal
 probabilities of 0.89 and 0.95 respectively.  The response parameters
-are presented in Table~ref{tab:igt4} below.
+are presented in Table~\ref{tab:igt4} below.
 
@@
 responses. States 1 and 2 are dominated C and D and A and B responses 
-respectively. To get a clearer intpretation of these states it is
+respectively. To get a clearer interpretation of these states it is
 necessary to consider when they are visited during the learning
-process. The viterbi algorithm (REFERENCE) provides us with the
+process. The Viterbi algorithm (REFERENCE) provides us with the
 posterior state sequence, i.e., the sequence of states that the
@@
 behavior, consisting only of C responses, a short transitional period 
-and then state 4 behavior consting of 94 \% C choices and some B
+and then state 4 behavior consisting of 94 \% C choices and some B
 choices, i.e., almost optimal behavior. The choice proportions and the
@@
 dominant dimension in the IGT. The final states with mostly C 
-choices reprensents one of the optimal strategies, which consists of
-both C and D responses. Note that both C and D reponses generate equal
+choices represents one of the optimal strategies, which consists of
+both C and D responses. Note that both C and D responses generate equal
 profits in the long run.
@@
 
-There are different accounts of probabilistic category learning.
+There are different accounts of probabilistic category learning 
+\cite<see e.g.>[for an overview]{Ashby2005}.
 According to instance or exemplar learning theories, participants
-learn by storing each encountered cue-outcome pairing.  When presented
-with a cue pattern, these exemplars are retrieved from memory, and
-weighted according to their similarity to the probe cue pattern, to
-form a classification.  According to associative theories,
-participants gradually learn by gradually associating the individual
-cues (or cue patterns in configural learning) to the outcomes.  In
+learn by storing each encountered cue-outcome pair.  When presented
+with a probe cue pattern, a response is made by retrieving these 
+exemplars are retrieved from memory and weighting them according 
+to their similarity to the probe cue pattern. According to associative theories,
+participants learn by gradually associating the individual
+cues (or cue patterns in configural learning) to the outcomes. Finally, in
 rule-learning, participants are taken to extract rules by which to
 categorize the different cue patterns.  \cite{Gluck2002} proposed a
-number of such rules (or strategies).  A main difference between these
-is whether responses are based on the presence/absence of a single
-cue, or whether responses are based on cue patterns.  Gluck et al.
+number of such rules (or strategies). The rules differ in the way
+the cues are combined into a response. A main difference is whether responses are
+based on the presence/absence of a single cue, or whether information 
+from multiple cues is integrated. \citeauthor{Gluck2002}
 formulated all strategies in a deterministic and optimal manner (e.g., 
 the multi-cue strategy corresponded to giving the optimal response to
-each cue pattern).  Meeter et al.  allowed for probabilistic
-responding (a small probability of giving the non-optimal response).
-
-Alternative non-strategy based analyses of the WPT
-\cite{Lagnado2006,Speekenbrink2008} have estimated response strategies by
-logistic regression, allowing the regression coefficients to change
-over time.
+each cue pattern). \citet{Meeter2006}  allowed for probabilistic
+responding (a small probability of responding non-optimally) as well
+as switches between strategies during learning. However, both strategies 
+and response probabilities remained predefined. Alternative non-strategy based 
+analyses of the WPT \cite{Lagnado2006,Speekenbrink2008} have estimated 
+response strategies through logistic regression, allowing the regression 
+coefficients to change smoothly over time.
 
 %associative, rule-based.
-Here, we analyze the behavior of a single individual performing the
-WPT for 200 trials.  We chose to analyse the ``average'' participant
-(the participant with performance closest to the group average) in a
-large unpublished dataset.  We let each state be characterized by a
-GLM with a Binomial distributed response and logistic link function
-(i.e., a logistic regression model).  We are particularly interested
-in evidence for strategy switching and whether a DMM can recover a
-strategy model in line with \citeA{Gluck2002}.
+Here, we combine the regression and strategy-based approaches, and analyse 
+the behavior of a single individual performing the
+WPT for 200 trials.  We are particularly interested
+in evidence for strategy switching and whether a DMM can recover
+strategies in accordance with \citeA{Gluck2002}. We chose to analyse the 
+``average'' participant (the participant with performance closest to 
+the group average) in a large unpublished dataset. We let each state 
+be characterized by a GLM with a Binomial distributed response and 
+logistic link function (i.e., a logistic regression model). The regression
+coefficients of a model relating the cues to the state of the weather are 
+given in column ```validity'' of Table~\ref{tab:WPT}. Note that indentical
+coefficients for the model relating the cues to responses would indicate 
+probability matching. A maximizing strategy is indicated by more
+extreme regression coefficients.
 
 As we fitted a DMM to the data of a single subject, it was necessary
-to place some constraints on the model.  Specifically, we constrain
+to place some constraints on the model.  Specifically, we constrained
 the state transitions to be in a ``left-right'' format (states can
@@
 \label{tab:WPT}
-\begin{tabular}{lcccccccc} \hline
- & & \multicolumn{1}{c}{1 state} & & \multicolumn{2}{c}{2 state} &&
- \multicolumn{2}{c}{2 state (constr.)} \\ \cline{3-3} \cline{5-6} \cline{8-9}
-parameter & & $S_1$ & & $S_1$ & $S_2$ & & $S_1$ & $S_2$ \\ \hline
-(intercept) & & -0.69 & & -2.73 & 0.88 & & -1.24 & 0 \\
-cue 1 && 1.69 && 2.12 & 1.60 && 1.65 & 1.97 \\
-cue 2 && 1.12 && 0.97 & 1.63 && 0 & 1.92 \\
-cue 3 && -0.49 && 0.91 & -2.03 && 0 & -1.58 \\
-cue 4 && -1.32 && 0.69 & -3.16 && 0 & -2.67 \\ \hline
- & & \multicolumn{1}{c}{AIC=204.47} & & \multicolumn{2}{c}{AIC=187.50} &&
+\begin{tabular}{lcccccccccc} \hline
+ & & \multicolumn{1}{c}{} && \multicolumn{1}{c}{1 state} & & \multicolumn{2}{c}{2 state} &&
+ \multicolumn{2}{c}{2 state (constr.)} \\ \cline{5-5} \cline{7-8} \cline{10-11}
+parameter && validity && $S_1$ & & $S_1$ & $S_2$ & & $S_1$ & $S_2$ \\ \hline
+(intercept) && 0 && -0.69 & & -2.73 & 0.88 & & -1.24 & 0 \\
+cue 1 && 2.10 && 1.69 && 2.12 & 1.60 && 1.65 & 1.97 \\
+cue 2 && 0.58 && 1.12 && 0.97 & 1.63 && 0 & 1.92 \\
+cue 3 && -0.58 && -0.49 && 0.91 & -2.03 && 0 & -1.58 \\
+cue 4 && -2.10 && -1.32 && 0.69 & -3.16 && 0 & -2.67 \\ \hline
+ & & & & \multicolumn{1}{c}{AIC=204.47} & & \multicolumn{2}{c}{AIC=187.50} &&
  \multicolumn{2}{c}{AIC=185.24}
@@
 Investigation of the parameter estimates (see Table~\ref{tab:WPT})
-showed that the regression coefficient for the first cue was of much
+showed that the regresion coefficient for the first cue was of much
 larger magnitude than that of the other cues. As such, the first
@@
 seemed to represent a multi-cue strategy, as all cues had regression
-coefficients of reasonable magnitude, with differing directions in
-line with the objective cue validities.
+coefficients of reasonable magnitude in the direction of the objective
+cue validities.
 
-To reduce the degrees of freedom, and improve parameter estimates, we
+To reduce the degrees of freedom and improve parameter estimates, we
 implemented constraints to force state 1 into a single cue strategy
@@
 Interestingly, the single cue strategy was somewhat different than
-described by \citet{Gluck2002}  Parameter estimates indicated relatively
+described by \citet{Gluck2002}.  Parameter estimates indicated relatively
 more consistent predictions of ``rain'' in the absence of cue 1

### 2008-11-17 papers/individual/individual.tex
-added paragraph in discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@243 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+Analyzing the WPT data on an individual level allowed us to precisely estimate idiographic strategies and their progression during learning.
+Moreover, using DMMs, we can combine an individual and group level analysis, increasing reliability of the individual analyses whilst allowing for substantial individual differences.
+Previous analyses \cite<e.g.,>{Meeter2006} pre-defined individual strategies, based on participants' self-reports in \citet{Gluck2002}.
+Interestingly, \citet{Gluck2002} noted that these self-reports often did not correspond to the strategies evident in participants' responses.
+It is well known that people may find it difficult to verbalize the way in which they integrate multiple sources of information.
+As such, estimating strategies directly from participants' responses will result in more valid assessment of their response strategies, and provides a direct test of the validity of the strategy set identified by \citeauthor{Gluck2002}.
+
 Possible future extensions in depmixS4

### 2008-11-26 papers/individual/individual.tex
updated individual paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@244 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 measurement models.  In this paper, the basic model is extended to
-include arbitrary distributions for the observed variables, including
-multi-variate distributions.  Moreover, there is optional support to
-include time-varying predictors.  In effect, this model consists of
-mixtures of generalized linear models with Markovian dependencies over time
-to model the change process.  In addition, transition parameters can
-be made to depend on covariates as well, such that the switching
-regime between states depends on characteristics of the individual or
-the experimental situation.  The model is illustrated with an example
-of participants' learning in the Iowa Gambling Task and in the Weather 
-Prediction Task.}
+include various distributions for the observed variables, including
+mixed multi-variate distributions.  Moreover, there is an option to
+include time-varying predictors on the observed distributions.  In
+effect, this model consists of mixtures of generalized linear models
+with Markovian dependencies over time to model the change process.  In
+addition, the transition parameters can be estimated with covariates,
+such that the switching regime between states depends on
+characteristics of the individual or the experimental situation.  The
+model is illustrated with examples of participants' learning in the
+Iowa Gambling Task and in the Weather Prediction Task.}
 
@@
 Discrete change frequently occurs in learning and development: in
-learning concepts, in performance on Piagetian tasks, in
-discrimination learning and in conditioning.  This chapter is
-concerned with detecting the nature and time points of change in (individual)
-time series.  We present a framework of dependent mixture models that can
-be used to differentiate between gradual and discrete learning events
-in individual time series data.  Before presenting the model in formal
-terms and providing some illustrations, we first review some examples
-in which discrete change is found.
+learning concepts and discrimination learning, in performance on
+Piagetian tasks such as the conservation of liquid task and the
+balance scale task, and in conditioning.  This chapter is concerned
+with detecting and characterizing such discrete changes in
+(individual) time series data.  We present a framework of dependent
+mixture models that can be used to test whether discrete learning
+events are present and, if so, to test at which point in time those
+changes are taking place.  Before presenting the model in formal terms
+and providing some illustrations, we first review some examples from
+the developmental psychology literature in which discrete change is
+typically found.
 
 % phenomena of discrete change in development/learning
-Piagetian developmental theory assumes step-wise changes in the
-strategies that children apply in all kinds of tasks such as the
-conservation learning and the balance scale task (REFERENCE??).
+Piagetian developmental theory assumes step-wise changes in the rules
+that children apply in all kinds of tasks such as in conservation
+learning and in the balance scale task (SieglerAlibali19??).
 \citet{Maas1992} developed a catastrophe model to describe step-wise
-changes or phase transitions in learning and developmental processes.
-They applied the catastrophe model to learning in the conservation of
-liquid task (REFERENCE??)  in which children have to judge relative
-volumes of liquid in glasses of different heights and widths.  Young
-children tend to ignore the width dimension and hence always choose
-the glass with the highest level of liquid (REFERENCE??).
-\citet{Maas1992} showed that there is a sudden transition to a new
-strategy in which the children also take the width of the glasses into
-account when judging the volume of liquids.
+changes predicted to occur by Piagetian theory; importantly, they
+derived precise statistical criteria for testing whether two
+discretely different stages exist.  They applied the catastrophe model
+to development on the conservation of liquid task (InhelderPiaget19??)
+in which children have to judge relative volumes of liquid in glasses
+of different heights and widths.  Young children tend to ignore the
+width dimension and hence always choose the glass with the highest
+level of liquid (InhelderPiaget19??).  It is assumed that there is a
+sudden transition to a new rule, for example, a rule in which the
+children also take the width of the glasses into account when judging
+the volume of liquids.
+
+% \citet{Maas1992} 
+% opzoeken: Siegler & Alibali 19??, intro textbook
+% opzoeken: Inhelder & Piaget, 19??
 
@@
 \citet{Jansen2001} applied the catastrophe model to
-development of strategies on the balance scale task (Siegler, 1981).
+development of reasoning on the balance scale task (Siegler, 1981).
 In the balance scale task participants have to judge which side of a
@@
 
+% Jansen 2001: aanpassen in de paper ref ipv het proefschrift
+
 % hysteresis on the balance scale
-\citet{Jansen2001} found clear evidence for stage-wise
-transitions between Rule 1 and Rule 2 by testing criteria that were
-derived from the catastrophe model.  In particular, she found bimodal
-test scores and inaccessibility.  The latter means that there are no
-in-between strategies: children apply either Rule 1 or Rule 2 and
-there is no in-between option.  \citet{Jansen2001} also found
-evidence for hysteresis: the phenomenon that switching between
+\citet{Jansen2001} found evidence for stage-wise transitions, or
+discrete change in our model, between Rule 1 and Rule 2 by testing
+criteria that were derived from the catastrophe model.  In particular,
+she found bimodal test scores and inaccessibility.  The latter means
+that there are no in-between strategies: children apply either Rule 1
+or Rule 2 and there is no in-between option.  \citet{Jansen2001} also
+found evidence for hysteresis: the phenomenon that switching between
 strategies is asymmetric.  Children can switch from Rule 1 to Rule 2
-and back, but this occurs at different trials.  In particular, if the
-distance dimension in the balance scale problems is made more salient
-by increasing the distance difference between weights on either side
-of the balance scale, children may switch from Rule 1 to Rule 2.  If
-subsequently the distance difference is decreased again, children may
-switch back to using Rule 1.  Hysteresis is the phenomenon that this
-switch back occurs at a different value of the control variable, in
-this case the distance difference.
+and back, but this occurs at different values of a continuously
+changing independent variable.  In particular, if the distance
+dimension in the balance scale problems is made increasingly more
+salient by increasing the distance difference between weights on
+either side of the balance scale, children may switch from Rule 1 to
+Rule 2.  If subsequently the distance difference is decreased again,
+children may switch back to using Rule 1.  Hysteresis is the
+phenomenon that this switch back occurs at a different value of the
+control variable, in this case the distance difference.
 
@@
 their study, evidence was found for sudden onset of learning: at the
-start of the learning experiment, pigeons did not learn anything
-and performance was stable; after a number of trials, learning kicked
-in and there were large increases in performance.  \citeauthor{Gallistel2004} 
-focused on modeling the distribution of onset times: that is, the trials at
-which learning suddenly takes off.  A similar interest in process
-onset times is found in addiction research.  For example,
-\citet{Sher2004} study the age at which children start using alcohol
-and how this related to eventual outcomes in terms of addiction.
+start of the learning experiment, pigeons did not learn anything and
+performance was stable; after a number of trials, learning kicked in
+and there were large increases in performance.
+\citeauthor{Gallistel2004} focused on modeling the distribution of
+onset times: that is, the trials at which learning suddenly takes off.
+A similar interest in process onset times is found in addiction
+research.  For example, \citet{Sher2004} study the age at which
+children start using alcohol and how this related to eventual outcomes
+in terms of addiction.
 

### 2008-11-28 papers/individual/individual.tex
Updated individual paper with Brenda's comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@245 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %&LaTeX
-\documentclass[a4paper,12pt,man,english]{apa} % nobf, doc, apa
+\documentclass[a4paper,12pt,man]{apa} % nobf, doc, apa
 
-\usepackage[english]{babel} % ik krijg anders foute citaties...in Zweeds formaat???
+%\usepackage[english]{babel} % ik krijg anders foute citaties...in Zweeds formaat???
 
@@
 
-\twoaffiliations{Department of Psychology, University of Amsterdam}{Department of Cognitive, Perceptual and Brain Sciences, University College London}
+\twoaffiliations{Department of Psychology, University of Amsterdam}
+{Department of Cognitive, Perceptual and Brain Sciences, University College London}
 
@@
 % balance scale rules
-\citet{Jansen2001} applied the catastrophe model to
+Jansen \& Maas (2001) applied the catastrophe model to
 development of reasoning on the balance scale task (Siegler, 1981).
@@
 number of weights on each side of the fulcrum.  This strategy for
-solving balance scale items is called Rule 1 \cite{Siegler1981}.  Older
+solving balance scale items is called Rule 1 \citet{Siegler1981}.  Older
 children include the distance dimension in determining their response
@@
 Also in animal learning and conditioning, evidence is found for sudden
-changes in response behavior \cite{Gallistel2004}.  In particular, in
-their study, evidence was found for sudden onset of learning: at the
-start of the learning experiment, pigeons did not learn anything and
-performance was stable; after a number of trials, learning kicked in
-and there were large increases in performance.
-\citeauthor{Gallistel2004} focused on modeling the distribution of
-onset times: that is, the trials at which learning suddenly takes off.
-A similar interest in process onset times is found in addiction
-research.  For example, \citet{Sher2004} study the age at which
-children start using alcohol and how this related to eventual outcomes
-in terms of addiction.
+changes in response behavior.  For example, \citet{Gallistel2004}
+found evidence for sudden onset of learning: at the start of the
+learning experiment, pigeons did not learn anything and performance
+was stable; after a number of trials, learning kicked in and there
+were large increases in performance.  \citet{Gallistel2004}
+focused on modeling the distribution of onset times: that is, the
+trials at which learning suddenly takes off.  A similar interest in
+process onset times is found in addiction research.  For example,
+\citet{Sher2004} study the age at which children start using alcohol
+and how this related to eventual outcomes in terms of addiction.
 
@@
 discriminate a number of stimuli based on a single dimension such as
-form or color.  This kind of learning is referred to as all-or-none
-learning or concept identification learning.  \citet{Raijmakers2001}
-found evidence for different strategies applied by children when faced
-with such a learning task.  \citet{Schmittmann2006} reanalyzed their
-data using hidden Markov models to show that both strategies are
-characterized by sudden transitions in the learning process.
+form or color.  \citet{Raijmakers2001} found evidence for two
+different strategies applied by children when faced with such a
+learning task.  \citet{Schmittmann2006} reanalyzed their data using
+hidden Markov models to show that both strategies are characterized by
+sudden transitions in the learning process.
 
@@
 the stimuli and that the transition from one strategy to the next is a
-discrete event.  Before providing these illustrations, below we give a
+discrete event.  Before providing these illustrations, we give a
 formalization of dependent mixture models and a brief overview of the
-\pkg{depmixS4} package that was developed to specify and fit such models.
+\pkg{depmixS4} package that was developed to specify and fit such
+models.
 
@@
 % history of Markov models/LMM applications
-Markov models have been used extensively in the social sciences; for
-example, in analyzing language learning \cite{Miller1952,Miller1963} and
-in the analysis of paired associate learning \cite{Wickens1982}.  In
-these models, the focus is on survey type data: a few repeated
-measurements taken in a large sample.  \citeA{Langeheine1990} discuss
-latent Markov models and their use in sociology and political science
-(see also McCutcheon, 1987).  Latent transition models have been used
-in studying development of math skills \citep{Collins1992} and in
-medical applications \citep{Reboussin1998}; \citet{Kaplan2008}
-provides an overview of such models, that are called stage-sequential
-models in the developmental psychology literature.
+Markov models have been used in the social sciences; for example, in
+analyzing language learning \cite{Miller1952,Miller1963} and in the
+analysis of paired associate learning \cite{Wickens1982}.  In these
+models, the focus is on survey type data: a few repeated measurements
+taken in a large sample.  \citeA{Langeheine1990} discuss latent Markov
+models and their use in sociology and political science (see also
+McCutcheon, 1987).  Latent Markov models have been used in studying
+development of math skills \citep{Collins1992}, and in medical
+applications \citep{Reboussin1998}; \citet{Kaplan2008} provides an
+overview of such models, also called stage-sequential models in the
+developmental psychology literature.
 
@@
 % hidden Markov models
-Hidden Markov models (HMM) tend to be used in the analysis of long
+Hidden Markov models (HMMs) tend to be used in the analysis of long
 univariate (individual) timeseries.  For example, HMMs are the model
@@
 % depmix model
-The dependent mixture model that we propose here spans the range from
-latent Markov models for few repeated measurements with many
+The dependent mixture model (DMM) that we propose here spans the range
+from latent Markov models for few repeated measurements with many
 participants to hidden Markov models for individual time series.  In
-addition, the dependent mixture model includes multivariate responses.
-The dependent mixture model consists of the following elements:
+addition, the dependent mixture model includes multivariate responses
+and covariates.  The dependent mixture model for a time series $O_{t}$
+consists of the following elements:
 \begin{enumerate}
@@
 	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix
-	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density
+	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density.
 \end{enumerate}
@@
 consideration can be in.  In the balance scale example mentioned
-above, children are applying one of two possible strategies in
-responding to the items.  The states are characterized by their
-corresponding observation densities.  Using for example Rule 1 in the
-balance scale task leads to correct answers on some items and
-incorrect answers on others.  A different strategy may lead to correct
-answers on some items and to guessing behavior on other items.  In
+above, children are applying one of two possible rules in responding
+to the items.  The states are characterized by their corresponding
+observation densities.  Using a particular rule in the balance scale
+task leads to correct answers on some items and incorrect answers on
+others.  A different rule may lead to correct answers on some items
+and to guessing behavior on other items.  As another example, in
 analyzing categorization learning data, in which participants learn to
@@
 participants, such that different knowledge states lead to different
-observed behavior or responses.
+observed behaviors or responses.
 
 % transitions/matrix
-The transition matrix $\mat{A}$ describes the transitions between
-states over repeated measures or trials.  This matrix summarizes the
-probabilities of transitioning from one state to another which
-represents learning or development.  The transition model contains the
-Markov assumption:
+Transition matrix $\mat{A}$ describes the transitions between states
+over repeated measurements.  This matrix summarizes the probabilities
+of transitioning from one state to another, which represents learning
+or development.  The transition model contains the Markov assumption:
 $$Pr(S_{t}|S_{t-1}, \ldots, S_{1}) = Pr(S_{t}|S_{t-1}),$$
-which means that the current state (at time $t$) only depends on the
-previous state $S_{t-1}$, and not on earlier states.
+where $Pr$ is the probability.  The Markov assumption means that the
+current state (at time $t$) only depends on the previous state
+$S_{t-1}$, and not on earlier states.
 
@@
 characterize the state, and in our examples, these characterize the
-strategy that a participant is using at a given measurement occasion.
+rule that a participant is using at a given measurement occasion.
 
-% log-likelihood
+% log-likelihood: kan dit helemaal weg???
 The log-likelihood of DMMs is usually computed by the
@@
 	\item to be able to fit transition models with covariates, i.e.,
-	to have time-dependent transition matrices.
+	such that the transition probabilities depend on a time-dependent 
+	covariate.
 
@@
 % transition and initial state probs
-By default, each row of the transition matrix and the vector of initial state 
-probabilities is modelled as a baseline-category multinomial logistic 
-model \cite<see>[chapter 7]{Agresti2002}. This means that covariates 
-can be used as predictors in these models. In particular, the model 
-for each multinomial is:
+By default, each row of the transition matrix and the vector of
+initial state probabilities is modelled as a baseline-category
+multinomial logistic model \cite<see>[chapter 7]{Agresti2002}.  This
+means that covariates can be used as predictors in these models.  In
+particular, the model for each multinomial is:
 \begin{equation} 
@@
 the $\mathbf{\beta}_{i}$ are the category regression coefficients;
-$\vc{x}$ is a vector of covariates or predictors; $J$ is the number
-of states; in this example, $J$ serves as the baseline-category,
-meaning that $\alpha_{J}$ and $\mathbf{\beta}_{J}$ are zero.
+$\vc{x}$ is a vector of covariates or predictors; $J$ is the number of
+states; in this example, $J$ serves as the baseline-category, meaning
+that $\alpha_{J}$ and $\mathbf{\beta}_{J}$ are zero.
 
@@
 \end{equation}
-where $\mu$ the (state-dependent) mean; $\vc{x}$ is a vector of
+where $\mu$ is the (state-dependent) mean; $\vc{x}$ is a vector of
 covariates or predictors; $\mathbf{\beta}$ is the vector of
 (state-dependent) regression coefficients; $\epsilon_{t}$ is a
-normally distributed error with (state-dependent) standard deviation
-$\sigma$.
+normally distributed error term with (state-dependent) standard
+deviation $\sigma$.
 
@@
 
+% referentie: nnet
+
 \subsection{Parameter Estimation}
@@
 \section{Illustrations}
-
-% \subsection{Toy data: parameter retrieval} ????
 
@@
 impaired task performance.  Their preference for the decks with
-immediate high rewards indicates ``myopia for the future''.
+immediate high rewards indicates ``myopia for the future''
+\cite{Bechara1994}.
 
@@
 (10 apples), whereas doors C and D deliver a low constant gain (2
-apples).  At doors A and C, a loss of 50 apples (A) or 10 apples (C)
+apples).  At doors A and C, a loss of 10 apples (A) or 2 apples (C)
 is delivered in 50\% of the trials.  For doors B and D, frequency of
-loss is only 10\%.  The median loss of doors B and D is 10 and 2,
+loss is only 10\%.  The median loss of doors B and D is 50 and 10,
 respectively.  \citet{Crone2004} administered the HDT to children from
 four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
-that children also fail to consider future consequences.
+that children have difficulty to take future consequences into account.
 
@@
 % finite mixture analysis
-A reanalysis of this dataset \cite{Huizenga2007} indicated that
+A reanalysis of this dataset \cite{Huizenga2007b} indicated that
 participants might solve the task by sequentially considering the
 three dimensions (constant gain, frequency of loss, and amount of
-loss) in order to choose a door.  The youngest children in the
-dataset mostly seem to focus on the dominant dimension in the task, frequency
-of loss, resulting in equal preference for doors B and D. Older
-participants seem to use a two-dimensional rule where participants
-first focus on the frequency of loss and then consider amount of loss,
-resulting in a preference for door D. A third very small subgroup
-seems to use an integrative rule where participants combine all three
-dimensions in the appropriate way.  Participants using the integrative
-rule pick cards from doors C and D, which are advantageous in the long
-run.
+loss) in order to choose a door.  Most of the youngest children seem
+to focus frequency of loss, resulting in equal preference for doors B
+and D. It's assumed that frequency of loss is the dominant dimension
+in the task.  Older participants seem to use a two-dimensional rule
+where participants first focus on the frequency of loss and then
+consider amount of loss, resulting in a preference for door D. A third
+very small subgroup seems to use an integrative rule where
+participants combine all three dimensions in the appropriate way.
+Participants using the integrative rule pick cards from doors C and D,
+which are advantageous in the long run.
+
+% Check ref Huizenga 2007b
 
 % problematic aspects of standard analysis
-Typical analyses of these data use the last 60 trials in a series of
-200 trials.  A silent assumption made in these analyses is
-that behavior has stabilized after 140 learning trials; this could
-very well be wrong and it is highly likely that there are individual
-differences in this learning process.
-
-% proposed analysis here
-Here, we analyze IGT learning data from a single participant with the aim to
-establish 1) at which trial learning stops, i.e., at which
+\citet{Huizenga2007b} used the last 60 trials in a series of 200
+trials, assuming that behavior has stabilized after 140 trials of
+learning; this could very well be wrong and it is highly likely that
+there are individual differences in this learning process.  Hence,
+here we analyze IGT learning data from a single participant with the
+aim to establish 1) at which trial learning stops, i.e., at which
 trial behavior has stabilized, and 2) what strategies participants use
@@
 various participants.  The responses were fitted with multinomial
-logistic models for the 4 possible choices that participants make in
+logistic models for the four possible choices that participants make in
 this task.  There were no covariates on any of the parameters.  The
-only constraint imposed on the models' parameters was that
-there were designated begin and end states.  This means that the
-initial state probabilities were fixed to one for the first state and
-to zero for the remaining states.  For the transition matrix this means
-that the final state was an absorbing state: the probability of
-transitioning out of that state is zero.  This was done to ensure that
-there would be a final state in which participants end which provides
-the possibility of immediately seeing participants' behavior at the
-end of the task.  Models were selected using the Akaike Information
-Criterion (AIC; Akaike, 1973).
+only constraint imposed on the models' parameters was that there were
+designated begin and end states.  This means that the initial state
+probabilities were fixed to one for the first state and to zero for
+the remaining states.  For the transition matrix this means that the
+final state was an absorbing state: the probability of transitioning
+out of that state is zero.  This was done to ensure that there would
+be a final state in which participants end which provides the
+possibility of immediately seeing participants' behavior at the end of
+the task.  Models were selected using the Akaike Information Criterion
+(AIC; Akaike, 1973).
 
@@
 
-%We chose to analyze two participants that used different strategies at
-%the end of the task as analyzed in the manner proposed by
-%\citet{Huizenga2007}.  The first participants' data 
-The data were best 
-described by a 4-state model.  This model's transition matrix is:
+The data were best described by a 4-state model.  This model's
+transition matrix is:
 $$
@@
 are presented in Table~\ref{tab:igt4} below.
+
+% Table refs checken
 
@@
 dominated by C responses, and state 4 has a low probability of B
-responses. States 1 and 2 are dominated C and D and A and B responses 
-respectively. To get a clearer interpretation of these states it is
-necessary to consider when they are visited during the learning
-process. The Viterbi algorithm (REFERENCE) provides us with the
-posterior state sequence, i.e., the sequence of states that the
-participant is in at each trial. This sequence is depicted in
-Figure~\ref{fig:post4}.
+responses.  States 1 and 2 are dominated by C and D, and by A and B
+responses, respectively.  To get a clearer interpretation of these
+states it is necessary to consider when they are visited during the
+learning process.  The Viterbi algorithm \citep{Rabiner1989} provides
+us with the posterior state sequence, i.e., the sequence of states
+that the participant is in at each trial.  This sequence is depicted
+in Figure~\ref{fig:post4}.
+
+% figure refs checken
 
@@
 choices. This preference indicates a focus on the frequency of loss
-associated with each of the doors as B and D have low frequency of
+associated with each of the doors as B and D both have low frequency of
 loss. After this, there is a period of stable state 2 behavior
@@
 The Weather Prediction Task \cite<WPT>{Knowlton1994} %(WPT, Knowlton, Squire \& Gluck, 1994) 
-is
-a probabilistic categorization task in which participants learn to
+is a probabilistic categorization task in which participants learn to
 predict (or categorize) the state of the weather (sun or rain) on the
@@
 patterns).  On a given trial, one, two, or three cues are present.
-There are a total of 14 possible cue patterns, and each one is
+There are a total of 14 possible cue patterns, and each cue pattern is
 associated with a particular probability distribution over the states
-of the weather.  In order to perform in the task, participants must
-predict the weather in accordance with these conditional
+of the weather.  In order to perform well in the task, participants
+must predict the weather in accordance with these conditional
 probabilities.
-
-\nocite{Knowlton1994}
 
@@
 There are different accounts of probabilistic category learning
-\cite<see e.g.>[for an overview]{Ashby2005}.
-According to instance or exemplar learning theories, participants
-learn by storing each encountered cue-outcome pair.  When presented
-with a probe cue pattern, a response is made by retrieving these 
-exemplars are retrieved from memory and weighting them according 
-to their similarity to the probe cue pattern. According to associative theories,
-participants learn by gradually associating the individual
-cues (or cue patterns in configural learning) to the outcomes. Finally, in
+\citep<see e.g.>[for an overview]{Ashby2005}.  According to instance or
+exemplar learning theories, participants learn by storing each
+encountered cue-outcome pair.  When presented with a probe cue
+pattern, a response is made by retrieving these exemplars are
+retrieved from memory and weighting them according to their similarity
+to the probe cue pattern.  According to associative theories,
+participants learn by gradually associating the individual cues (or
+cue patterns in configural learning) to the outcomes.  Finally, in
 rule-learning, participants are taken to extract rules by which to
-categorize the different cue patterns.  \cite{Gluck2002} proposed a
-number of such rules (or strategies). The rules differ in the way
-the cues are combined into a response. A main difference is whether responses are
-based on the presence/absence of a single cue, or whether information 
-from multiple cues is integrated. \citeauthor{Gluck2002}
-formulated all strategies in a deterministic and optimal manner (e.g., 
-the multi-cue strategy corresponded to giving the optimal response to
-each cue pattern). \citet{Meeter2006}  allowed for probabilistic
-responding (a small probability of responding non-optimally) as well
-as switches between strategies during learning. However, both strategies 
-and response probabilities remained predefined. Alternative non-strategy based 
-analyses of the WPT \cite{Lagnado2006,Speekenbrink2008} have estimated 
-response strategies through logistic regression, allowing the regression 
+categorize the different cue patterns.  \citet{Gluck2002} proposed a
+number of such rules (or strategies).  The rules differ in the way the
+cues are combined into a response.  A main difference is whether
+responses are based on the presence/absence of a single cue, or
+whether information from multiple cues is integrated.
+\citet{Gluck2002} formulated all strategies in a deterministic
+and optimal manner (e.g., the multi-cue strategy corresponded to
+giving the optimal response to each cue pattern).  \citet{Meeter2006}
+allowed for probabilistic responding (a small probability of
+responding non-optimally) as well as switches between strategies
+during learning.  However, both strategies and response probabilities
+remained predefined.  Alternative non-strategy based analyses of the
+WPT \cite{Lagnado2006,Speekenbrink2008} have estimated response
+distributions through logistic regression, allowing the regression
 coefficients to change smoothly over time.
@@
 %associative, rule-based.
-Here, we combine the regression and strategy-based approaches, and analyse 
-the behavior of a single individual performing the
-WPT for 200 trials.  We are particularly interested
-in evidence for strategy switching and whether a DMM can recover
-strategies in accordance with \citeA{Gluck2002}. We chose to analyse the 
-``average'' participant (the participant with performance closest to 
-the group average) in a large unpublished dataset. We let each state 
-be characterized by a GLM with a Binomial distributed response and 
-logistic link function (i.e., a logistic regression model). The regression
-coefficients of a model relating the cues to the state of the weather are 
-given in column ```validity'' of Table~\ref{tab:WPT}. Note that indentical
-coefficients for the model relating the cues to responses would indicate 
-probability matching. A maximizing strategy is indicated by more
-extreme regression coefficients.
+Here, we combine the regression and strategy-based approaches, and
+analyse the behavior of a single individual performing the WPT for 200
+trials.  We are particularly interested in evidence for strategy
+switching and whether a DMM can recover strategies in accordance with
+\citet{Gluck2002}.  We chose to analyse the ``average'' participant
+(the participant with performance closest to the group average) in a
+large unpublished dataset.  We let each state be characterized by a
+GLM with a Binomial distributed response and logistic link function
+(i.e., a logistic regression model).  The regression coefficients of a
+model relating the cues to the state of the weather are given in
+column ```validity'' of Table~\ref{tab:WPT}.  Note that indentical
+coefficients for the model relating the cues to responses would
+indicate probability matching.  A maximizing strategy is indicated by
+more extreme regression coefficients.
 
@@
 
-Analyzing the WPT data on an individual level allowed us to precisely estimate idiographic strategies and their progression during learning.
-Moreover, using DMMs, we can combine an individual and group level analysis, increasing reliability of the individual analyses whilst allowing for substantial individual differences.
-Previous analyses \cite<e.g.,>{Meeter2006} pre-defined individual strategies, based on participants' self-reports in \citet{Gluck2002}.
-Interestingly, \citet{Gluck2002} noted that these self-reports often did not correspond to the strategies evident in participants' responses.
-It is well known that people may find it difficult to verbalize the way in which they integrate multiple sources of information.
-As such, estimating strategies directly from participants' responses will result in more valid assessment of their response strategies, and provides a direct test of the validity of the strategy set identified by \citeauthor{Gluck2002}.
+% IGT
+confirmation of existence of different strategies
+
+dynamics of learning now known
+
+different strategies/rules observed during learning (eg AB and 
+guessing strategies)
+
+% WPT
+Analyzing the WPT data on an individual level allowed us to precisely
+estimate idiographic strategies and their progression during learning.
+Moreover, using DMMs, we can combine an individual and group level
+analysis, increasing reliability of the individual analyses whilst
+allowing for substantial individual differences.  Previous analyses
+\cite<e.g.,>{Meeter2006} pre-defined individual strategies, based on
+participants' self-reports in \citet{Gluck2002}.  Interestingly,
+\citet{Gluck2002} noted that these self-reports often did not
+correspond to the strategies evident in participants' responses.  It
+is well known that people may find it difficult to verbalize the way
+in which they integrate multiple sources of information.  As such,
+estimating strategies directly from participants' responses will
+result in more valid assessment of their response strategies, and
+provides a direct test of the validity of the strategy set identified
+by \citet{Gluck2002}.
 

### 2008-12-05 papers/individual/individual.tex
Updated individual paper, almost complete discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@246 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 that children apply in all kinds of tasks such as in conservation
-learning and in the balance scale task (SieglerAlibali19??).
+learning and in the balance scale task \citep{Siegler2005}.
 \citet{Maas1992} developed a catastrophe model to describe step-wise
@@
 discretely different stages exist.  They applied the catastrophe model
-to development on the conservation of liquid task (InhelderPiaget19??)
+to development on the conservation of liquid task \citep{Inhelder1958}
 in which children have to judge relative volumes of liquid in glasses
@@
 width dimension and hence always choose the glass with the highest
-level of liquid (InhelderPiaget19??).  It is assumed that there is a
-sudden transition to a new rule, for example, a rule in which the
-children also take the width of the glasses into account when judging
-the volume of liquids.
-
-% \citet{Maas1992} 
-% opzoeken: Siegler & Alibali 19??, intro textbook
-% opzoeken: Inhelder & Piaget, 19??
+level of liquid \citep{Inhelder1958}.  It is assumed that there is a
+sudden transition, or a discrete change, to a new rule, for example, a
+rule in which the children also take the width of the glasses into
+account when judging the volume of liquids.
 
 % balance scale rules
-Jansen \& Maas (2001) applied the catastrophe model to
-development of reasoning on the balance scale task (Siegler, 1981).
-In the balance scale task participants have to judge which side of a
-balance goes down when the number of weights and their distances to
-the fulcrum are varied over trials.  Younger children tend to ignore
-the distance dimension in this task, and instead focus solely on the
+\citet{Jansen2001b} applied the catastrophe model to development of
+reasoning on the balance scale task \citep{Siegler1981}.  In the
+balance scale task participants have to judge which side of a balance
+goes down when the number of weights and their distances to the
+fulcrum are varied over trials.  Younger children tend to ignore the
+distance dimension in this task, and instead focus solely on the
 number of weights on each side of the fulcrum.  This strategy for
-solving balance scale items is called Rule 1 \citet{Siegler1981}.  Older
-children include the distance dimension in determining their response
-to balance scale problems; however, they only do so when the weight
-dimension does not differ between the sides of the balance, i.e., when
-the number of weights is equal on both sides of the balance scale.
-This strategy is called Rule 2 \cite{Siegler1981}.
-
-% Jansen 2001: aanpassen in de paper ref ipv het proefschrift
+solving balance scale items is called Rule 1 \citep{Siegler1981}.
+Older children include the distance dimension in determining their
+response to balance scale problems; however, they only do so when the
+weight dimension does not differ between the sides of the balance,
+i.e., when the number of weights is equal on both sides of the balance
+scale.  This strategy is called Rule 2 \cite{Siegler1981}.
 
 % hysteresis on the balance scale
-\citet{Jansen2001} found evidence for stage-wise transitions, or
+\citet{Jansen2001b} found evidence for stage-wise transitions, or
 discrete change in our model, between Rule 1 and Rule 2 by testing
 criteria that were derived from the catastrophe model.  In particular,
-she found bimodal test scores and inaccessibility.  The latter means
-that there are no in-between strategies: children apply either Rule 1
-or Rule 2 and there is no in-between option.  \citet{Jansen2001} also
-found evidence for hysteresis: the phenomenon that switching between
-strategies is asymmetric.  Children can switch from Rule 1 to Rule 2
-and back, but this occurs at different values of a continuously
-changing independent variable.  In particular, if the distance
-dimension in the balance scale problems is made increasingly more
-salient by increasing the distance difference between weights on
+they found bimodal test scores and inaccessibility.  The latter means
+that there are no in-between strategies: children apply {\em either}
+Rule 1 {\em or} Rule 2 and there is no in-between option.
+\citet{Jansen2001b} also found evidence for hysteresis: the phenomenon
+that switching between strategies is asymmetric.  Children can switch
+from Rule 1 to Rule 2 and back, but this occurs at different values of
+a continuously changing independent variable.  In particular, if the
+distance dimension in the balance scale problems is made increasingly
+more salient by increasing the distance difference between weights on
 either side of the balance scale, children may switch from Rule 1 to
@@
 % conditioning and addiction research
-Also in animal learning and conditioning, evidence is found for sudden
-changes in response behavior.  For example, \citet{Gallistel2004}
-found evidence for sudden onset of learning: at the start of the
-learning experiment, pigeons did not learn anything and performance
-was stable; after a number of trials, learning kicked in and there
-were large increases in performance.  \citet{Gallistel2004}
-focused on modeling the distribution of onset times: that is, the
-trials at which learning suddenly takes off.  A similar interest in
-process onset times is found in addiction research.  For example,
-\citet{Sher2004} study the age at which children start using alcohol
-and how this related to eventual outcomes in terms of addiction.
+Also in animal learning and conditioning, evidence is found for
+discrete changes in response behavior.  For example,
+\citet{Gallistel2004} found evidence for sudden onset of learning: at
+the start of the learning experiment, pigeons did not learn anything
+and performance was stable; after a number of trials, learning kicked
+in and there were large increases in performance.
+\citet{Gallistel2004} focused on modeling the distribution of onset
+times: that is, the trials at which learning suddenly takes off.  A
+similar interest in process onset times is found in addiction
+research.  For example, \citet{Sher2004} study the age at which
+children start using alcohol and how this relates to eventual outcomes
+in terms of (severity of) addiction.
 
@@
 hidden Markov models to show that both strategies are characterized by
-sudden transitions in the learning process.
+discrete changes in the learning process.
 
@@
 participants or even just a single participant.  For example,
-\citet{Visser2009} analyzed data from a single participant in an
+\citet{Visser2009c} analyzed data from a single participant in an
 experimental task that manipulates the trade-off between speed and
@@
 In this section we describe a class of models which are especially
-suitable for describing and testing discrete change in (individual)
+suitable for describing and testing discrete changes in (individual)
 time series data.  The dependent mixture model is similar to, but
@@
 % history of Markov models/LMM applications
-Markov models have been used in the social sciences; for example, in
-analyzing language learning \cite{Miller1952,Miller1963} and in the
-analysis of paired associate learning \cite{Wickens1982}.  In these
-models, the focus is on survey type data: a few repeated measurements
-taken in a large sample.  \citeA{Langeheine1990} discuss latent Markov
-models and their use in sociology and political science (see also
-McCutcheon, 1987).  Latent Markov models have been used in studying
-development of math skills \citep{Collins1992}, and in medical
-applications \citep{Reboussin1998}; \citet{Kaplan2008} provides an
-overview of such models, also called stage-sequential models in the
-developmental psychology literature.
-
-\nocite{McCutcheon1987}
+Markov models have been used for a long time in the social sciences;
+for example, in analyzing language learning
+\cite{Miller1952,Miller1963} and in the analysis of paired associate
+learning \cite{Wickens1982}.  In these models, the focus is on survey
+type data: a few repeated measurements taken in a large sample.
+\citeA{Langeheine1990} discuss latent Markov models and their use in
+sociology and political science \citep<see also>{McCutcheon1987}.  Latent
+Markov models have been used in studying development of math skills
+\citep{Collins1992}, and in medical applications
+\citep{Reboussin1998}.  \citet{Kaplan2008} provides an overview of
+such models, also called stage-sequential models, and their
+application in developmental psychology.
 
@@
 \begin{enumerate}
-	\item $S$ is a collection of discrete states
-	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix
+	\item $S$ is a collection of discrete states, forming the state space.
+	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix.
 	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density.
@@
 % state space
-The state space, which is a set of discrete states, captures the
-different states that the learning or developmental process under
-consideration can be in.  In the balance scale example mentioned
-above, children are applying one of two possible rules in responding
-to the items.  The states are characterized by their corresponding
-observation densities.  Using a particular rule in the balance scale
-task leads to correct answers on some items and incorrect answers on
-others.  A different rule may lead to correct answers on some items
-and to guessing behavior on other items.  As another example, in
-analyzing categorization learning data, in which participants learn to
-categorize a set of objects, a typical initial state is that
-participants are guessing because at the start of the task they have
-no knowledge of which features are important in categorization.
-Hence, the states in the state space represent knowledge states of the
-participants, such that different knowledge states lead to different
-observed behaviors or responses.
+The state space $S$ captures the different states that the learning or
+developmental process under consideration can be in.  In the balance
+scale example mentioned above, children are applying one of two
+possible rules in responding to the items.  The states are
+characterized by their corresponding observation densities.  Using a
+particular rule in the balance scale task leads to correct answers on
+some items and incorrect answers on others.  A different rule may lead
+to correct answers on some items and to guessing behavior on others.
+As another example, in analyzing categorization learning data, in
+which participants learn to categorize a set of objects, a typical
+initial state is that participants are guessing because at the start
+of the task they have no knowledge of which features are important in
+categorization.  Hence, the states in the state space represent
+knowledge states of the participants, such that different knowledge
+states lead to different observed behaviors or responses.
 
@@
 
-% log-likelihood: kan dit helemaal weg???
-The log-likelihood of DMMs is usually computed by the
-so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
-rather by the forward part of this algorithm.  \citet{Lystig2002}
-changed the forward algorithm in such a way as to allow computing the
-gradients of the log-likelihood at the same time.  They start by
-rewriting the likelihood as follows (for ease of exposition the
-dependence on the model parameters is dropped here):
-\begin{equation}
-	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
-Pr(\vc{O}_{t}|\vc{O}_{1},
-	\ldots, \vc{O}_{t-1}),
-	\label{condLike}
-\end{equation}
-where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a
-simple, i.e.\ observed, Markov chain these probabilities reduce to
-$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots,
-\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
-The log-likelihood can now be expressed as:
-\begin{equation}
-	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
-\vc{O}_{t-1})].
-	\label{eq:condLogl}
-\end{equation}
-
-To compute the log-likelihood, \citet{Lystig2002} define the following
-(forward) recursion:
-\begin{align}
-	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1})
-	\label{eq:fwd1} \\
-\begin{split}
-	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots,
-\vc{O}_{t-1}) \\
-	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times
-(\Phi_{t-1})^{-1},
-	\label{eq:fwdt}
-\end{split}
-\end{align}
-where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining
-$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and
-equation~(\ref{eq:condLogl}) gives the following expression for the
-log-likelihood:
-\begin{equation}
-	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
-	\label{eq:logl}
-\end{equation}
-
-Note that so far no assumptions have been made about the response
-distributions $b_{j}$, hence these can be arbitrary univariate or
-multivariate distributions.
+% % log-likelihood: kan dit helemaal weg???
+% The log-likelihood of DMMs is usually computed by the
+% so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
+% rather by the forward part of this algorithm.  \citet{Lystig2002}
+% changed the forward algorithm in such a way as to allow computing the
+% gradients of the log-likelihood at the same time. 
+% 
+% 
+% They start by rewriting the likelihood as follows (for ease of
+% exposition the dependence on the model parameters is dropped here):
+% \begin{equation}
+% 	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T}
+% Pr(\vc{O}_{t}|\vc{O}_{1},
+% 	\ldots, \vc{O}_{t-1}),
+% 	\label{condLike}
+% \end{equation}
+% where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a
+% simple, i.e.\ observed, Markov chain these probabilities reduce to
+% $Pr(\vc{O}_{t}|\vc{O}_{1},\ldots,
+% \vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
+% The log-likelihood can now be expressed as:
+% \begin{equation}
+% 	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots,
+% \vc{O}_{t-1})].
+% 	\label{eq:condLogl}
+% \end{equation}
+% 
+% To compute the log-likelihood, \citet{Lystig2002} define the following
+% (forward) recursion:
+% \begin{align}
+% 	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1})
+% 	\label{eq:fwd1} \\
+% \begin{split}
+% 	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots,
+% \vc{O}_{t-1}) \\
+% 	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times
+% (\Phi_{t-1})^{-1},
+% 	\label{eq:fwdt}
+% \end{split}
+% \end{align}
+% where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining
+% $\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and
+% equation~(\ref{eq:condLogl}) gives the following expression for the
+% log-likelihood:
+% \begin{equation}
+% 	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
+% 	\label{eq:logl}
+% \end{equation}
+% 
+% Note that so far no assumptions have been made about the response
+% distributions $b_{j}$, hence these can be arbitrary univariate or
+% multivariate distributions.
 
@@
 % in detail below
-\pkg{depmixS4} implements a general framework for defining and fitting
-dependent mixture models in the R programming language \citep{R2008}.
-This includes standard Markov models, latent/hidden Markov models, and
-latent class and finite mixture distribution models.  The models can
-be fitted on mixed multivariate data with multinomial and/or gaussian
-distributions.
+\pkg{depmixS4} \citep{Visser2008a} implements a general framework for
+defining and fitting dependent mixture models in the R programming
+language \citep{R2008}.  This includes standard Markov models,
+latent/hidden Markov models, and latent class and finite mixture
+distribution models.  The models can be fitted on mixed multivariate
+data with error distriutions from the generalized linear model
+framework, such as the gaussian, binomial, multinomial logistic et
+cetera.
 
@@
 \begin{enumerate} 
-	\item A model for the initial or prior probabilities 
+	\item A model for the initial or prior probabilities.
 	\item A list of transition models; one model for each row of the 
-	transition matrix
+	transition matrix.
 	\item A list of response models for each of the states of the model; 
@@
 Response models in \pkg{depmixS4} interface with the \code{glm}
-functions available in R and with the \pkg{nnet}-package for the 
-multinomial logistic models. Normally distributed data can hence be 
-modelled with direct effects included as well. For example, normal 
-data are modelled as:
+functions \citep{R2008} available in R and with the \pkg{nnet}-package
+\citep{Venables2002} for the multinomial logistic models.  Normally
+distributed data can hence be modelled with direct effects included as
+well.  For example, normal data are modelled as:
 \begin{equation}
@@
 
-% MAARTEN: de multinomial logistic models maken weer gebruik van andere
-% packages toch, kun je daar de referentie van geven?
-
 % referentie: nnet
@@
 Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
-through the use of a general Newton-Raphson optimizer.  EM is used by
-default in unconstrained models, but otherwise direct optimization is
-done using \pkg{Rdonlp2} \cite{Tamura2007,Spellucci2002}, because it
-handles general linear (in)equality constraints, and optionally also
-non-linear constraints.
+alternatively through the use of a general Newton-Raphson optimizer.
+EM is used by default in unconstrained models, but otherwise direct
+optimization is done using \pkg{Rdonlp2}
+\cite{Tamura2007,Spellucci2002}, because it handles general linear
+(in)equality constraints, and optionally also non-linear constraints.
+Both the forward algorithm \citep{Baum1966,Rabiner1989} and the
+version by \citet{Lystig2002} are implemented in \pkg{depmixS4}; the
+latter is implemented to allow easy computation of the gradients of
+the parameters.
 
@@
 In both of these, participants learn different strategies through
-trial and error.
+trial and error.  The main goal of these illustrations is to establish
+the possibility of analyzing single participant data using the
+\pkg{depmixS4}-framework.
 
@@
 (10 apples), whereas doors C and D deliver a low constant gain (2
-apples).  At doors A and C, a loss of 10 apples (A) or 2 apples (C)
-is delivered in 50\% of the trials.  For doors B and D, frequency of
-loss is only 10\%.  The median loss of doors B and D is 50 and 10,
-respectively.  \citet{Crone2004} administered the HDT to children from
-four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
-that children have difficulty to take future consequences into account.
+apples) at each trial.  At doors A and C, a loss of 10 apples (A) or 2
+apples (C) is delivered in 50\% of the trials.  For doors B and D,
+frequency of loss is only 10\%.  The median loss of doors B and D is
+50 and 10, respectively.  \citet{Crone2004} administered the HDT to
+children from four age groups (6-9, 10-12, 13-15, and 18-25 year-olds)
+and concluded that children have difficulty to take future
+consequences into account.
 
@@
 loss) in order to choose a door.  Most of the youngest children seem
-to focus frequency of loss, resulting in equal preference for doors B
-and D. It's assumed that frequency of loss is the dominant dimension
-in the task.  Older participants seem to use a two-dimensional rule
-where participants first focus on the frequency of loss and then
-consider amount of loss, resulting in a preference for door D. A third
-very small subgroup seems to use an integrative rule where
-participants combine all three dimensions in the appropriate way.
-Participants using the integrative rule pick cards from doors C and D,
-which are advantageous in the long run.
+to focus on frequency of loss, resulting in equal preference for doors
+B and D. It is assumed that frequency of loss is the dominant
+dimension in this task.  Older participants seem to use a
+two-dimensional rule where participants first focus on the frequency
+of loss and then consider amount of loss, resulting in a preference
+for door D. A third very small subgroup seems to use an integrative
+rule where participants combine all three dimensions in the
+appropriate way.  Participants using the integrative rule pick cards
+from doors C and D, which are both equally advantageous in the long
+run, although door D shows a higher volatility.
 
@@
 \citet{Huizenga2007b} used the last 60 trials in a series of 200
-trials, assuming that behavior has stabilized after 140 trials of
-learning; this could very well be wrong and it is highly likely that
-there are individual differences in this learning process.  Hence,
-here we analyze IGT learning data from a single participant with the
-aim to establish 1) at which trial learning stops, i.e., at which
-trial behavior has stabilized, and 2) what strategies participants use
+trials, as is typically done in analyzing this task.  Doing so
+silently assumes that behavior has stabilized after 140 trials of
+learning; this may not be the case, and it is highly likely that there
+are individual differences in this learning process.  Hence, here we
+analyze IGT learning data from a single participant with the aim to
+establish 1) at which trial learning stops, i.e., at which trial
+behavior has stabilized, and 2) what strategies participants use
 during and at the end of learning.  Models with an increasing number
 (from 2 through 5) of latent states were fitted to the time series of
-various participants.  The responses were fitted with multinomial
-logistic models for the four possible choices that participants make in
-this task.  There were no covariates on any of the parameters.  The
+a single participant.  The responses were fitted with multinomial
+logistic models for the four possible choices that participants make
+in this task.  There were no covariates on any of the parameters.  The
 only constraint imposed on the models' parameters was that there were
@@
 final state was an absorbing state: the probability of transitioning
-out of that state is zero.  This was done to ensure that there would
-be a final state in which participants end which provides the
-possibility of immediately seeing participants' behavior at the end of
-the task.  Models were selected using the Akaike Information Criterion
-(AIC; Akaike, 1973).
-
-\nocite{Akaike1973} % check the year!!
+out of the final state was set to zero.  This was done to ensure that
+there would be a final state in which participants end which provides
+the possibility of immediately seeing participants' behavior at the
+end of the task.  Models were selected using the Akaike Information
+Criterion \citep<AIC;>{Akaike1973}.
 
@@
 $$
-As can be seen from the transition matrix the final state is
+As can be seen from the transition matrix, the final state is
 absorbing.  States 2 and 3 are also fairly stable with high diagonal
@@
 \begin{table}
-	\caption{Estimates for the Iowa Gambling Task model}
+	\caption{Estimates for response probabilities in the IGT model.}
 	\label{tab:igt4}
@@
 responses, respectively.  To get a clearer interpretation of these
-states it is necessary to consider when they are visited during the
+states, it is necessary to consider when they are visited during the
 learning process.  The Viterbi algorithm \citep{Rabiner1989} provides
@@
 that the participant is in at each trial.  This sequence is depicted
-in Figure~\ref{fig:post4}.
+in Figure~1.
 
@@
 	\includegraphics[width=\textwidth]{graphs/post4.pdf}
+	\caption{Posterior state sequence for the 4 state model of the IGT 
+	data.}
 	\label{fig:post4}
-	\caption{Posterior state sequence for the 4 state model.}
 \end{center}
@@
 choices.  This preference indicates a focus on the frequency of loss
-associated with each of the doors as B and D both have low frequency of
-loss. After this, there is a period of stable state 2 behavior
+associated with each of the doors as B and D both have low frequency
+of loss.  After this, there is a period of stable state 2 behavior
 associated with responses A and B. After that there is stable state 3
@@
 and then state 4 behavior consisting of 94 \% C choices and some B
-choices, i.e., almost optimal behavior. The choice proportions and the
-model predicted choice proportions are depicted in
-Figure~\ref{fig:igt4}.
+choices, i.e., almost optimal behavior.  The choice proportions and
+the model predicted choice proportions are depicted in Figure~2.
 
@@
 	\includegraphics[width=\textwidth]{graphs/igt4.pdf}
-	\label{fig:igt4}
 	\caption{Data and model predicted proportions of door choices in
 	the IGT.}
+	\label{fig:igt4}
 \end{center}
@@
 The initial preference for B and D choices confirms the theory
-expressed in \citet{Huizenga2007} that frequency of loss is the
+expressed in \citet{Huizenga2007b} that frequency of loss is the
 dominant dimension in the IGT. The final states with mostly C 
 choices represents one of the optimal strategies, which consists of
-both C and D responses. Note that both C and D responses generate equal
+both C and D responses. Note that C and D responses generate equal
 profits in the long run.
@@
 
-The Weather Prediction Task \cite<WPT>{Knowlton1994} %(WPT, Knowlton, Squire \& Gluck, 1994) 
-is a probabilistic categorization task in which participants learn to
+The Weather Prediction Task \cite<WPT;>{Knowlton1994} is a
+probabilistic categorization task in which participants learn to
 predict (or categorize) the state of the weather (sun or rain) on the
@@
 There are different accounts of probabilistic category learning
-\citep<see e.g.>[for an overview]{Ashby2005}.  According to instance or
-exemplar learning theories, participants learn by storing each
+\citep<see e.g.>[for an overview]{Ashby2005}.  According to instance
+or exemplar learning theories, participants learn by storing each
 encountered cue-outcome pair.  When presented with a probe cue
-pattern, a response is made by retrieving these exemplars are
-retrieved from memory and weighting them according to their similarity
-to the probe cue pattern.  According to associative theories,
+pattern, a response is made by retrieving these exemplars from memory
+and weighting them according to their similarity to the probe cue
+pattern.  In contrast, according to associative learning theories,
 participants learn by gradually associating the individual cues (or
@@
 number of such rules (or strategies).  The rules differ in the way the
-cues are combined into a response.  A main difference is whether
-responses are based on the presence/absence of a single cue, or
-whether information from multiple cues is integrated.
-\citet{Gluck2002} formulated all strategies in a deterministic
-and optimal manner (e.g., the multi-cue strategy corresponded to
-giving the optimal response to each cue pattern).  \citet{Meeter2006}
-allowed for probabilistic responding (a small probability of
-responding non-optimally) as well as switches between strategies
-during learning.  However, both strategies and response probabilities
-remained predefined.  Alternative non-strategy based analyses of the
-WPT \cite{Lagnado2006,Speekenbrink2008} have estimated response
+cues are combined into a response.  The main difference concerns
+whether responses are based on the presence/absence of a single cue,
+or whether information from multiple cues is integrated.
+\citet{Gluck2002} formulated all strategies in a deterministic and
+optimal manner (e.g., the multi-cue strategy corresponded to giving
+the optimal response to each cue pattern).  \citet{Meeter2006} allowed
+for probabilistic responding (a small probability of responding
+non-optimally) as well as switches between strategies during learning.
+However, both strategies and response probabilities remained
+predefined.  Alternative non-strategy based analyses of the WPT
+\cite{Lagnado2006,Speekenbrink2008} have estimated response
 distributions through logistic regression, allowing the regression
@@
 GLM with a Binomial distributed response and logistic link function
-(i.e., a logistic regression model).  The regression coefficients of a
-model relating the cues to the state of the weather are given in
-column ```validity'' of Table~\ref{tab:WPT}.  Note that indentical
-coefficients for the model relating the cues to responses would
-indicate probability matching.  A maximizing strategy is indicated by
-more extreme regression coefficients.
+(i.e., a logistic regression model, see equation 1).  The regression
+coefficients of a model relating the cues to the state of the weather
+are given in column ```validity'' of Table~\ref{tab:WPT}.  Note that
+identical coefficients for the model relating the cues to responses
+would indicate probability matching.  A maximizing strategy is
+indicated by more extreme regression coefficients.
 
-As we fitted a DMM to the data of a single subject, it was necessary
-to place some constraints on the model.  Specifically, we constrained
-the state transitions to be in a ``left-right'' format (states can
-only proceed to the immediately adjacent state and never back, and
-must start in the initial state).  We fitted a single, two and three
-state model to the data.  This showed that a two state model was
-better than a single or three state model.
+As we fitted a DMM to the data of a single participant, it was
+necessary to place some constraints on the model.  Specifically, we
+constrained the state transitions to be in a ``left-right'' format
+(states can only proceed to the immediately adjacent state and never
+back, and must start in the initial state).  We fitted models with
+one, two or three states to the data.  The AIC values of these models
+showed that the two state model outperformed the one and three state
+models.
 
 \begin{table}
-\caption{Estimates for the weather prediction task}
+\caption{Parameter estimates of models for the WPT data.}
 \label{tab:WPT}
@@
 showed that the regresion coefficient for the first cue was of much
-larger magnitude than that of the other cues. As such, the first
-state seems representative of single cue strategy.  Alternatively, as
-all regression coefficients are positive, it could indicate a
-``counting'' heuristic, where the propensity of ``sun'' responses
-increases when more cues are present (regardless of which cues they
-are).  However, in that case, we would expect the regression
+larger magnitude than that of the other cues.  Because of this, the
+first state seems representative of single cue strategy.
+Alternatively, as all regression coefficients are positive, it could
+indicate a ``counting'' heuristic, where the propensity of ``sun''
+responses increases when more cues are present (regardless of which
+cues they are).  However, in that case, we would expect the regression
 coefficients to be of roughly identical magnitude.  The second state
-seemed to represent a multi-cue strategy, as all cues had regression
+represents a multi-cue strategy, as all cues had regression
 coefficients of reasonable magnitude in the direction of the objective
@@
 
-General issues
-\begin{itemize}
-	\item It is feasible to fit hidden Markov models in moderate length individual time series
-	\item illustrations show results that are consistent with usual group based analyses AND 
-	provide extra information on switching between strategies over the course of learning
-	\item Many applications in experimental psychology: examples????
-\end{itemize}
+Discrete change is hypothesized to occur in many domains in
+experimental psychology and especially so in learning and
+developmental tasks.  Piagetian theory predicts discrete changes to
+occur in a number of developmental tasks \citep{Inhelder1958}.  There
+is a growing body of evidence showing that indeed discrete change
+occurs during development and learning
+\citep{Jansen2001b,Maas1992,Raijmakers2001,Schmittmann2006}.  Discrete
+strategies or rules have recently been hypothesized to play an
+important role in learning tasks such as the Iowa Gambling Task and
+the Weather Prediction Task.  Until now, such data were always
+analyzed on a group basis, i.e., pooling data of many participants
+together.  Given that it is extremely likely that there are individual
+differences between participants, such pooling of data may not be
+warranted and may lead to sub optimal results \citep{Molenaar2005}.
 
-Specific insights from the illustrations: suggestions welcome
+In the current study we have shown the feasibility of analyzing single
+participant data in two common experimental tasks, thereby avoiding
+the pitfalls of pooling data from individuals that may differ in
+important respects.  The illustrations show that single participant
+data only comprising 200 trials can be analyzed and lead to
+non-trivial models with multiple states.  This opens up possiblities
+for analyzing data from other experimental tasks on an individual
+basis.
+
+In addition to analyzing single participant data, the \pkg{depmixS4}
+package \citep{Visser2008b} that we have presented here offers the
+possibility of further combining models of single participant data
+into larger models and testing whether for example some strategies or
+rules are identical between (groups of) participants.  Performing such
+analyses and aggregating models afterwards based on commonalities and
+differences between participants avoids making unwarranted assumptions
+about homogeneity between participants.  A similar approach using
+continuous latent variable models, i.e. factor models, was used by
+\citet{Hamaker2007} to integrate within and between person variability
+in the domain of personality research.
 
 % IGT
-confirmation of existence of different strategies
-
-dynamics of learning now known
-
-different strategies/rules observed during learning (eg AB and 
-guessing strategies)
+The illustration with the IGT data lead to a four state model.  The
+states of the model corresponded very well with earlier analyses by
+\citet{Huizenga2007b}, showing a number of discrete strategies.  Even
+within a single participant, a number of different strategies were
+observed during the acquisition process; learning started with
+behavior close to guessing, which could be interpreted as exploring
+behavior, and was followed by two states with close to optimal
+performance.  In addition to confirming the existence of discrete
+strategies, the use of DMM analyses provides us with information of
+the dynamics of knowledge acquisition in this task.
 
@@
 
+General issues
+\begin{itemize}
+	\item illustrations show results that are consistent with usual group based analyses AND 
+	provide extra information on switching between strategies over the course of learning
+	\item Many applications in experimental psychology: examples????
+\end{itemize}
+
+\newpage
+
 \section*{Author note}
@@
 Economic Learning and Social Evolution (ELSE).  Thanks to Hilde
-Huizenga, Brenda Jansen and Anna van Duijvenvoorde for the Iowa
+Huizenga, Brenda Jansen, and Anna van Duijvenvoorde for the Iowa
 Gambling task data and David Lagnado for the Weather Prediction Task

### 2008-12-08 papers/individual/individual.tex
Finished individual paper, submitted version
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@247 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\shorttitle{Modelling Discrete Change}
+\shorttitle{Modeling Discrete Change}
 
@@
 events are present and, if so, to test at which point in time those
-changes are taking place.  Before presenting the model in formal terms
-and providing some illustrations, we first review some examples from
-the developmental psychology literature in which discrete change is
-typically found.
+changes are taking place.  We first review some examples from the
+developmental psychology literature in which discrete change is
+typically found, and then turn to describing the model more formally.
 
@@
 they found bimodal test scores and inaccessibility.  The latter means
-that there are no in-between strategies: children apply {\em either}
-Rule 1 {\em or} Rule 2 and there is no in-between option.
+that there are no in-between rules: children apply {\em either} Rule 1
+{\em or} Rule 2 and there is no in-between option.
 \citet{Jansen2001b} also found evidence for hysteresis: the phenomenon
-that switching between strategies is asymmetric.  Children can switch
-from Rule 1 to Rule 2 and back, but this occurs at different values of
-a continuously changing independent variable.  In particular, if the
+that switching between rules is asymmetric.  Children can switch from
+Rule 1 to Rule 2 and back, but this occurs at different values of a
+continuously changing independent variable.  In particular, if the
 distance dimension in the balance scale problems is made increasingly
@@
 form or color.  \citet{Raijmakers2001} found evidence for two
-different strategies applied by children when faced with such a
+different rules applied by children when faced with such a
 learning task.  \citet{Schmittmann2006} reanalyzed their data using
-hidden Markov models to show that both strategies are characterized by
-discrete changes in the learning process.
+hidden Markov models to show that both rules are characterized by
+discrete changes during the learning process.
 
@@
 Weather Prediction Task.  The interest in these tasks is to show that
-participants develop different strategies over time in responding to
-the stimuli and that the transition from one strategy to the next is a
-discrete event.  Before providing these illustrations, we give a
-formalization of dependent mixture models and a brief overview of the
-\pkg{depmixS4} package that was developed to specify and fit such
+participants develop different rules over time in responding to the
+stimuli and that the transition from one rule to the next is a
+discrete event.  Before presenting these illustrations, we provide a
+brief formal definition of dependent mixture models and an overview of
+the \pkg{depmixS4} package that was developed to specify and fit such
 models.
@@
 In this section we describe a class of models which are especially
-suitable for describing and testing discrete changes in (individual)
+suitable for describing and testing discrete change in (individual)
 time series data.  The dependent mixture model is similar to, but
 slightly different from, two other types of models that are in use for
-modelling discrete change: the hidden and the latent Markov models.
+modeling discrete change: the hidden and the latent Markov models.
 
@@
 \cite{Miller1952,Miller1963} and in the analysis of paired associate
-learning \cite{Wickens1982}.  In these models, the focus is on survey
-type data: a few repeated measurements taken in a large sample.
-\citeA{Langeheine1990} discuss latent Markov models and their use in
-sociology and political science \citep<see also>{McCutcheon1987}.  Latent
-Markov models have been used in studying development of math skills
-\citep{Collins1992}, and in medical applications
-\citep{Reboussin1998}.  \citet{Kaplan2008} provides an overview of
-such models, also called stage-sequential models, and their
-application in developmental psychology.
+learning \cite{Wickens1982}.  \citeA{Langeheine1990} discuss latent
+Markov models and their use in sociology and political science
+\citep<see also>{McCutcheon1987}.  Latent Markov models have been used
+in studying development of math skills \citep{Collins1992}, and in
+medical applications \citep{Reboussin1998}.  \citet{Kaplan2008}
+provides an overview of such models, also called stage-sequential
+models, and their application in developmental psychology.
 
@@
 Hidden Markov models (HMMs) tend to be used in the analysis of long
-univariate (individual) timeseries.  For example, HMMs are the model
+(univariate, individual) time series.  For example, HMMs are the model
 of choice in speech recognition applications \cite{Rabiner1989}.  In
@@
 	\item $S$ is a collection of discrete states, forming the state space.
-	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, \mat{A}$, a transition matrix.
-	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, \mat{B}$,  an observation density.
+	\item $S_{t} = \mat{A}S_{t-1}+\xi_{t}, with \mat{A}$ a transition matrix.
+	\item $O_{t} = \mat{B}(S_{t}) + \zeta_{t}, with \mat{B}$  an observation density.
 \end{enumerate}
@@
 model; these describe the distributions of the observations
-conditional on the current state. Hence, these distributions
+conditional on the current state $S_{t}$. Hence, these distributions
 characterize the state, and in our examples, these characterize the
@@
 % in detail below
-\pkg{depmixS4} \citep{Visser2008a} implements a general framework for
-defining and fitting dependent mixture models in the R programming
-language \citep{R2008}.  This includes standard Markov models,
-latent/hidden Markov models, and latent class and finite mixture
-distribution models.  The models can be fitted on mixed multivariate
-data with error distriutions from the generalized linear model
-framework, such as the gaussian, binomial, multinomial logistic et
-cetera.
+The \pkg{depmixS4} package \citep{Visser2008a}, which is used to
+analyze data in the illustrations below, implements a general
+framework for defining and fitting dependent mixture models in the R
+programming language \citep{R2008}.  This includes standard Markov
+models, latent/hidden Markov models, and latent class and finite
+mixture distribution models.  The models can be fitted on mixed
+multivariate data with error distributions from the generalized linear
+model framework, such as the gaussian, binomial, multinomial logistic
+et cetera.
 
@@
 
+	\item to be able to fit models on arbitrary length individual time 
+	series.
+
 \end{enumerate}
@@
 
-\pkg{depmixS4} is built using S4 classes (object oriented classes in R)
-to allow easy extensibility (Chambers, 1998). A depmix model consists 
-of the following: 
+\pkg{depmixS4} is built using S4 classes \citep<object oriented
+classes in R>{Chambers1998} to allow easy extensibility.  A depmix
+model consists of the following:
 \begin{enumerate} 
-	\item A model for the initial or prior probabilities.
+	\item A model for the initial or prior probabilities, possibly 
+	depending on covariates.
 	\item A list of transition models; one model for each row of the 
-	transition matrix.
+	transition matrix, possibly depending on covariates. 
 	\item A list of response models for each of the states of the model; 
-	this is a list of lists in the case of multivariate responses. 
+	this is a list of lists in the case of multivariate responses (see 
+	details below).
 \end{enumerate} 
-Each of these are adressed briefly below.
+Each of these are addressed briefly below.
 
@@
 By default, each row of the transition matrix and the vector of
-initial state probabilities is modelled as a baseline-category
+initial state probabilities is modeled as a baseline-category
 multinomial logistic model \cite<see>[chapter 7]{Agresti2002}.  This
@@
 \begin{equation} 
-	\pi_{i} = \frac{\exp(\alpha_{i}+\mathbf{\beta}^{`}_{i}\vc{x})}
+	p_{i} = \frac{\exp(\alpha_{i}+\mathbf{\beta}^{`}_{i}\vc{x})}
 				{1+\sum_{j=1}^{J-1}\exp(\alpha_{j}+\mathbf{\beta}^{`}_{j}\vc{x})},
 \end{equation}
-where $\pi_{i}$ is the probability (e.g.\ the probability of the
+where $p_{i}$ is the probability (e.g.\ the probability of the
 $i$-th initial state); the $\alpha_{i}$ are the category intercepts;
@@
 \citep{Venables2002} for the multinomial logistic models.  Normally
-distributed data can hence be modelled with direct effects included as
-well.  For example, normal data are modelled as:
+distributed data can hence be modeled with direct effects included as
+well.  For example, normal data are modeled as:
 \begin{equation}
@@
 
-Two illustrations are provided below of models that analyze single
+Two illustrations are provided below of models used to analyze single
 participant time series data from two common experimental paradigms.
-In both of these, participants learn different strategies through
-trial and error.  The main goal of these illustrations is to establish
-the possibility of analyzing single participant data using the
+In both of these, participants learn different rules through trial and
+error.  The main goal of these illustrations is to establish the
+possibility of analyzing single participant data using the
 \pkg{depmixS4}-framework.
@@
 % finite mixture analysis
-A reanalysis of this dataset \cite{Huizenga2007b} indicated that
+A reanalysis of this data set \cite{Huizenga2007b} indicated that
 participants might solve the task by sequentially considering the
@@
 establish 1) at which trial learning stops, i.e., at which trial
-behavior has stabilized, and 2) what strategies participants use
+behavior has stabilized, and 2) what rules participants use
 during and at the end of learning.  Models with an increasing number
@@
 probabilities of 0.89 and 0.95 respectively.  The response parameters
-are presented in Table~\ref{tab:igt4} below.
+are presented in Table~\ref{tab:igt4}.
 
@@
 dominant dimension in the IGT. The final states with mostly C 
-choices represents one of the optimal strategies, which consists of
+choices represents one of the optimal rules, which consists of
 both C and D responses. Note that C and D responses generate equal
@@
 categorize the different cue patterns.  \citet{Gluck2002} proposed a
-number of such rules (or strategies).  The rules differ in the way the
+number of such rules.  The rules differ in the way the
 cues are combined into a response.  The main difference concerns
@@
 or whether information from multiple cues is integrated.
-\citet{Gluck2002} formulated all strategies in a deterministic and
-optimal manner (e.g., the multi-cue strategy corresponded to giving
+\citet{Gluck2002} formulated all rules in a deterministic and
+optimal manner (e.g., the multi-cue rule corresponded to giving
 the optimal response to each cue pattern).  \citet{Meeter2006} allowed
 for probabilistic responding (a small probability of responding
-non-optimally) as well as switches between strategies during learning.
-However, both strategies and response probabilities remained
-predefined.  Alternative non-strategy based analyses of the WPT
+non-optimally) as well as switches between rules during learning.
+However, both rules and response probabilities remained
+predefined.  Alternative non-rule based analyses of the WPT
 \cite{Lagnado2006,Speekenbrink2008} have estimated response
@@
 %associative, rule-based.
-Here, we combine the regression and strategy-based approaches, and
-analyse the behavior of a single individual performing the WPT for 200
-trials.  We are particularly interested in evidence for strategy
-switching and whether a DMM can recover strategies in accordance with
-\citet{Gluck2002}.  We chose to analyse the ``average'' participant
+Here, we combine the regression and rule-based approaches, and
+analyze the behavior of a single individual performing the WPT for 200
+trials.  We are particularly interested in evidence for rule
+switching and whether a DMM can recover rules in accordance with
+\citet{Gluck2002}.  We chose to analyze the ``average'' participant
 (the participant with performance closest to the group average) in a
@@
 Investigation of the parameter estimates (see Table~\ref{tab:WPT})
-showed that the regresion coefficient for the first cue was of much
+showed that the regression coefficient for the first cue was of much
 larger magnitude than that of the other cues.  Because of this, the
-first state seems representative of single cue strategy.
+first state seems representative of single cue rule.
 Alternatively, as all regression coefficients are positive, it could
@@
 coefficients to be of roughly identical magnitude.  The second state
-represents a multi-cue strategy, as all cues had regression
+represents a multi-cue rule, as all cues had regression
 coefficients of reasonable magnitude in the direction of the objective
@@
 To reduce the degrees of freedom and improve parameter estimates, we
-implemented constraints to force state 1 into a single cue strategy
+implemented constraints to force state 1 into a single cue rule
 (fixing the coefficients of the remaining three cues to 0) and state 2
-in a multi-cue strategy (forcing the intercept to 0).  These
+in a multi-cue rule (forcing the intercept to 0).  These
 restrictions resulted in a better AIC value of AIC=185.24 (df=7).
-Interestingly, the single cue strategy was somewhat different than
+Interestingly, the single cue rule was somewhat different than
 described by \citet{Gluck2002}.  Parameter estimates indicated relatively
@@
 in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
-of the multi-cue strategy were in the direction of the optimal
+of the multi-cue rule were in the direction of the optimal
 weights.  The Viterbi state sequence indicated that the participant
-used the single cue strategy for the first 60 trials, and then
-switched to the multi-cue strategy.
+used the single cue rule for the first 60 trials, and then
+switched to the multi-cue rule.
 
@@
 \citep{Jansen2001b,Maas1992,Raijmakers2001,Schmittmann2006}.  Discrete
-strategies or rules have recently been hypothesized to play an
-important role in learning tasks such as the Iowa Gambling Task and
-the Weather Prediction Task.  Until now, such data were always
-analyzed on a group basis, i.e., pooling data of many participants
-together.  Given that it is extremely likely that there are individual
-differences between participants, such pooling of data may not be
-warranted and may lead to sub optimal results \citep{Molenaar2005}.
+rules have recently been hypothesized to play an important role in
+learning tasks such as the Iowa Gambling Task \citep{Huizenga2007b}
+and the Weather Prediction Task \citep{Gluck2002}.  Until now, such
+data were always analyzed on a group basis, i.e., pooling data of many
+participants together.  Given that it is extremely likely that there
+are individual differences between participants, such pooling of data
+may not be warranted and may lead to sub optimal results
+\citep{Gallistel2004,Molenaar2005}.
 
@@
 data only comprising 200 trials can be analyzed and lead to
-non-trivial models with multiple states.  This opens up possiblities
+non-trivial models with multiple states.  This opens up possibilities
 for analyzing data from other experimental tasks on an individual
@@
 possibility of further combining models of single participant data
-into larger models and testing whether for example some strategies or
-rules are identical between (groups of) participants.  Performing such
-analyses and aggregating models afterwards based on commonalities and
+into larger models and testing whether, for example, some rules are
+identical between (groups of) participants.  Performing such analyses
+and aggregating models afterwards based on commonalities and
 differences between participants avoids making unwarranted assumptions
@@
 states of the model corresponded very well with earlier analyses by
-\citet{Huizenga2007b}, showing a number of discrete strategies.  Even
-within a single participant, a number of different strategies were
+\citet{Huizenga2007b}, showing a number of discrete rules.  Even
+within a single participant, a number of different rules were
 observed during the acquisition process; learning started with
-behavior close to guessing, which could be interpreted as exploring
+behavior close to guessing, which is best interpreted as exploring
 behavior, and was followed by two states with close to optimal
 performance.  In addition to confirming the existence of discrete
-strategies, the use of DMM analyses provides us with information of
+rules, the use of DMM analyses provides us with information of
 the dynamics of knowledge acquisition in this task.
@@
 Analyzing the WPT data on an individual level allowed us to precisely
-estimate idiographic strategies and their progression during learning.
+estimate idiographic rules and their progression during learning.
 Moreover, using DMMs, we can combine an individual and group level
@@
 allowing for substantial individual differences.  Previous analyses
-\cite<e.g.,>{Meeter2006} pre-defined individual strategies, based on
+\cite<e.g.,>{Meeter2006} used pre-defined individual rules, based on
 participants' self-reports in \citet{Gluck2002}.  Interestingly,
 \citet{Gluck2002} noted that these self-reports often did not
-correspond to the strategies evident in participants' responses.  It
-is well known that people may find it difficult to verbalize the way
-in which they integrate multiple sources of information.  As such,
-estimating strategies directly from participants' responses will
-result in more valid assessment of their response strategies, and
-provides a direct test of the validity of the strategy set identified
-by \citet{Gluck2002}.
+correspond to the rules evident in participants' responses.  It is
+well known that people may find it difficult to verbalize the way in
+which they integrate multiple sources of information.  This problem is
+even more severe in studying young children.  Because of this,
+estimating rules directly from participants' responses results in a
+more valid assessment of their response rules, and provides a direct
+test of the validity of the rule set identified by \citet{Gluck2002}.
 
-Possible future extensions in depmixS4
-\begin{enumerate}
-	\item richer measurement models, eg factor models, AR models etc
-	\item richer transition models, eg continuous time measurement occasions
-	\item explicit state durations
-	\item identifiability of models
-	\item model selection
-\end{enumerate}
-
-General issues
-\begin{itemize}
-	\item illustrations show results that are consistent with usual group based analyses AND 
-	provide extra information on switching between strategies over the course of learning
-	\item Many applications in experimental psychology: examples????
-\end{itemize}
+% DMM: their great future, conclusion
+In sum, DMMs have been shown to be feasible models of individual time
+series.  Moreover, \pkg{depmixS4} can be used to build models of
+groups of participants combined and allows for testing whether pooling
+individual participants' data is warranted.  Due to the use of S4
+classes, the \pkg{depmixS4} package is highly flexible and easily
+extensible.  Future options are to include other measurement models
+such as the factor model or models with autoregressive coefficients.
+Also, extensions of the transition model are easily implemented; for
+example, the addition of support for continuous time measurement in
+contrast to equally space measurement occasions as were used in the
+presented illustrations.  Dependent mixture models hence have great
+potential in enlarging our insights into important issues in learning
+and development, in particular in modeling and testing the occurrence
+of discrete change.
 

### 2009-07-08 papers/jss/depmix-manual.Rnw
Added .Rnw file of the jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@280 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
%&LaTeX

\documentclass[]{report}

\usepackage[]{amsmath, amsfonts, amstext, amsthm}

\batchmode

\usepackage{Sweave}

\usepackage{Rd}

\usepackage[round]{natbib}

\title{depmix: An R-package for fitting mixture models on mixed
multivariate data with Markov dependencies}

\author{Ingmar Visser\thanks{
Correspondence concerning this manual should be adressed to:
Ingmar Visser, Department of Psychology, University of Amsterdam,
Roetersstraat 15, 1018 WB, Amsterdam, The Netherlands}\\
Department of Psychology, University of Amsterdam\\
i.visser@uva.nl}

\date{\today}

%%\VignetteIndexEntry{depmix: An R-package for fitting mixture models on mixed multivariate data with Markov dependencies}
%%\VignetteDepends{depmix}
%%\VignetteKeywords{R, latent/hidden Markov model, latent class model, finite mixture model, direct optimization, gradients}
%%\VignettePackage{depmix}

\SweaveOpts{prefix.string=graphs/depmix}

\newcommand{\vc}{\mathbf}
\newcommand{\mat}{\mathbf}

% \newcommand{\pkg}{\texttt}
% \newcommand{\code}{\texttt}

\begin{document}

\bibliographystyle{plainnat}

\maketitle

\begin{abstract}

	\pkg{depmix} implements a general class of mixture models with
	Markovian dependencies between them in the R programming language
	\citep{R2007}.  This includes standard Markov models, latent/hidden
	Markov models, and latent class and finite mixture distribution models.
	The models can be fitted on mixed multivariate data with multinomial
	and/or gaussian distributions.  Parameters can be estimated subject to
	general linear constraints, and with optional inclusion of regression
	on (time-dependent) covariates.  Parameters estimation is done through
	a direct optimization approach with gradients using the \code{Rdonlp2}
	optimization routine.  A number of illustrative examples are included.

\end{abstract}

\tableofcontents

\chapter{Introduction}

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982}, for an
overview, and \citet{Schmittmann2006} for a recent application.  In
economics, latent Markov models are commonly used as regime switching
models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.  Further
applications include speech recognition \citep{Rabiner1989}, EEG analysis
\citep{Rainer2000}, and genetics \citep{Krogh1998}.  In those latter areas
of application, latent Markov models are usually referred to as hidden
Markov models.

The \pkg{depmix} package was motivated by the fact that Markov models
are used commonly in the social sciences, but no comprehensive package
was available for fitting such models.  Common programs for Markovian
models include Panmark \citep{Panmark3-96}, and for latent class models
Latent Gold 
\citep{LatentGold3-2003}.  Those programs are lacking a number of
important features, besides not being freely available.  In
particular, \pkg{depmix}: 1) handles multiple case, or multiple group,
analysis; 2) handles arbitrarily long time series; 3) estimates models
with general linear constraints between parameters; 4) analyzes mixed
distributions, i.e., combinations of categorical and continuous
observed variables; 5) fits mixtures of latent Markov models to deal
with population heterogeneity; 6) can fit models with covariates.
Although \pkg{depmix} is specifically meant for dealing with
longitudinal or time series data, for say $T>100$, it can also handle
the limit case with $T=1$.  In those cases, there are no time
dependencies between observed data, and the model reduces to a finite
mixture model, or a latent class model.  In the next chapter, an outline
is provided of the model and the likelihood equations.  In the
chapters after that a number of examples are presented.

\section*{Acknowledgements}

I am indebted to many people for providing help in writing this package.
First and foremost Maartje Raijmakers and Verena Schmittmann tested
countless earlier versions, spotted bugs and suggested many features.
Moreover, Maartje Raijmakers provided the discrimination data set.  Han van
der Maas provided the speed-accuracy data and thereby neccessitated
implementing models with time-dependent covariates.  Conor Dolan and Raoul
Grasman both provided valuable advice on statistics in general and
optimization in particular.

\chapter{Dependent mixture models}

The data considered here, has the general form $O_{1}^{1}, \ldots,
O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1}, \ldots,
O_{T}^{m}$ for an $m$-variate time series of length $T$.  As an
example, consider a time series of responses generated by a single
subject in a reaction time experiment.  The data consists of three
variables, reaction time, accuracy and a covariate which is a pay-off
factor which determines the reward for speed and accuracy.  These
variables are measured at 168, 134 and 137 occasions respectively.
Below, a summary is provided for these data, as well as a plot of the
first timeseries, which is selected by \code{nind=1}.

<<echo=FALSE, results=hide>>=
.libPaths(new="/Users/ivisser/Library/R/library/")
library(depmix)
load("models.Rda")
@

<<>>=
data(speed)
summary(speed)
@
\begin{figure}[htbp]
  \begin{center}
<<fig=TRUE>>=
plot(speed,nind=1) 
@
	\caption{Reaction times, accuracy and pay-off values for
	the first series of responses in dataset \texttt{speed}.}
  \end{center}
\end{figure}

The latent Markov model is commonly associated with data of this type,
albeit usually only multinomial variables are considered.  However,
common estimation procedures, such as those implemented in
\citet{Panmark3-96} are not suitable for long time series due to
underflow problems.  In contrast, the hidden Markov model is typically
only used for `long' univariate time series.  In the next section, the
likelihood and estimation procedure for the hidden Markov model is
described, given data of the above form.

The dependent mixture model is defined by the following elements:
\begin{enumerate}

	\item a set $\vc{S}$ of latent classes or states $S_{i},\, i=1,
	\ldots , n$,

	\item a matrix $\mat{A}$ of transition probabilities $a_{ij}$ for
	the transition from state $S_{i}$ to state $S_{j}$,

	\item a set $\vc{B}$ of observation functions $b_{j}(\cdot)$ that
	provide the conditional probabilities associated with latent state
	$S_{j}$,

	\item a vector $\pmb{\pi}$ of latent state initial probabilities
	$\pi_{i}$
\end{enumerate}
When transitions are added to the latent class model, it is more 
appropriate to refer to the classes as states. The word class is 
rather more associated with a stable trait-like attribute whereas a 
state can change over time.

\section{Likelihood}

The loglikelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Bau66,Rab89}, or rather by
the forward part of this algorithm.  \cite{Lys2002} changed the forward
algorithm in such a way as to allow computing the gradients of the
loglikelihood at the same time.  They start by rewriting the
likelihood as follows (for ease of exposition the dependence on the
model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
Pr(\vc{O}_{t}|\vc{O}_{1}, 
	\ldots, \vc{O}_{t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lys2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

The above forward recursion can readily be generalized to mixture 
models, in which it is assumed that the data are realizations of a 
number of different LMMs and the goal is to assign posterior 
probabilities to sequences of observations. This situation occurs, 
for example, in learning data where different learning strategies may 
lead to different answer patterns. From an observed sequence of 
responses, it may not be immediately clear from which learning 
process they stem. Hence, it is interesting to consider a mixture of 
latent Markov models which incorporate restrictions that are 
consistent with each of the learning strategies.

To compute the likelihood of a mixture of $K$ models, define the 
forward recursion variables as follows (these variables now have an 
extra index $k$ indicating that observation and transition 
probabilities are from latent model $k$):
\begin{align}
\begin{split}
\phi_{1}(j_{k}) &=  Pr(\vc{O}_{1}, 
S_{1}=j_{k})=p_{k}\pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}).
\end{split}\label{eq:fwd1mix} \\
\begin{split}
\phi_{t}(j_{k})   &=   Pr(\vc{O}_{t}, S_{t}=j_{k}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
			&= \left[ \sum_{k=1}^{K} \sum_{i=1}^{n_{k}} \phi_{t-1}(i_{k}) 
			a_{ij_{k}}b_{j_{k}}(\vc{O}_{t}) \right] \times (\Phi_{t-1})^{-1},
\end{split}\label{eq:fwdtmix} 
\end{align}
where $\Phi_{t} = \sum_{k=1}^{K}\sum_{i=1}^{n_{k}} \phi_{t}(j_{k})$.
Note that the double sum over $k$ and $n_{k}$ is simply an enumeration
of all the states of the model.  Now, because $a_{ij_{k}}=0$ whenever
$S_{i}$ is not part of component $k$, the sum over $k$ can be dropped
and hence equation~\ref{eq:fwdtmix} reduces to:
\begin{equation}
	\phi_{t}(j_{k}) = \left[ \sum_{i=1}^{n_{k}} \phi_{t-1}(i_{k}) 
			a_{ij_{k}}b_{j_{k}}(\vc{O}_{t}) \right] \times (\Phi_{t-1})^{-1}
\end{equation}
The loglikelihood is computed by applying equation~\ref{eq:logl} on
these terms.  For multiple cases, the log-likelihood is simply the sum
over the individual log-likelihoods.

\paragraph{Computational considerations} From
equations~(\ref{eq:fwd1}--\ref{eq:fwdt}), it can be seen that
computing the forward variables, and hence the log-likelihood, takes
$O(Tn^{2})$ computations, for an $n$-state model and a time series of
length $T$.  Consider a mixture of two components, one with two states
and the other with three states.  Using
equations~(\ref{eq:fwd1}--\ref{eq:fwdt}) to compute the log-likelihood
of this model one needs $O(Tn^{2})=O(T\times 25)$ computations whereas
with the mixture equations~(\ref{eq:fwd1mix}--\ref{eq:fwdtmix}),
$\sum_{n_{i}} O(n_{i}^{2}T)$ computations are needed, in this case
$O(T \times 13)$.  So, it can be seen that in this easy example the
computational cost is almost halved.

\section{Gradients}

\newcommand{\fpp}{\frac{\partial} {\partial \lambda_{1}}}

See equations 10--12  in \cite{Lys2002} for the score recursion 
functions of the hidden Markov model for a univariate time series. 
Here the corresponding score recursion for the multivariate mixture 
case are provided. The $t=1$ components of this score recursion are 
defined as (for an arbitrary parameter $\lambda_{1}$):
\begin{align}
\psi_{1}(j_{k};\lambda_{1}) &:=  \fpp Pr(\vc{O}_{1}|S_{1}=j_{k}) \\
\begin{split} 
	&= \left[  \fpp p_{k} \right] \pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}) + 
	p_{k}\left[ \fpp \pi_{j_{k}} \right] b_{j_{k}}(\vc{O}_{1}) \\
	& \qquad  + p_{k}\pi_{j_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{1})\right],
\end{split} \label{eq:psi1}
\end{align}
and for $t>1$ the definition is:
\begin{align}
\psi_{t}(j_{k};\lambda_{1})  & =  \frac{\fpp Pr(\vc{O}_{1}, \ldots, 
\vc{O}_{t}, S_{t}=j_{k})}
			{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}  \\
\begin{split} 
	& =  
			 \sum_{i=1}^{n_{k}} \Bigg\{ \psi_{t-1}(i;\lambda_{1})a_{ij_{k}} 
			 b_{j_{k}}(\vc{O}_{t}) \\ 
			 &\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] b_{j_{k}} 
(\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
(\vc{O}_{t}) \right] \Bigg\} 
			\times (\Phi_{t-1})^{-1}.
\end{split} \label{eq:psit}
\end{align}

Using above equations, \cite{Lys2002} derive the following equation 
for the partial derivative of the likelihood:
\begin{equation}
	\fpp l_{T}= 	
		\frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}},
\end{equation}
where $\Psi_{t}=\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} 
\psi_{t}(j_{k};\lambda_{1})$. 
Starting from the equation from the logarithm of the likelihood, this 
is easily seen to be correct: 
\begin{eqnarray*}
	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})^{-1} 
	\fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) \\
	&=&  \frac{Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1})}{Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})}  \Psi_{T} (\lambda_{1}) \\
	&=&  \frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}}.
\end{eqnarray*}

Further, to actually compute the gradients, the partial derivatives of
the parameters and observation distribution functions are neccessary,
i.e., $\fpp p_{k}$, $\fpp \pi_{i}$, $\fpp a_{ij}$, and $\fpp
\vc{b}_{i}(\vc{O}_{t})$.  Only the latter case requires some
attention.  We need the following derivatives $\fpp
\vc{b}_{j}(\vc{O}_{t})=\fpp \vc{b}_{j}(O_{t}^{1}, \ldots, O_{t}^{m})$, for
arbitrary parameters $\lambda_{1}$. To stress that $\vc{b}_{j}$ is a
vector of functions, we here used boldface. First note that because of local
independence we can write:
\begin{equation*}
	\fpp \left[ b_{j}(O_{t}^{1}, \ldots, O_{t}^{m}) \right] = \frac{\partial} 
{\partial \lambda_{1} } \left[ b_{j}(O_{t}^{1}) \right] \times  
\left[ b_{j}(O_{t}^{2}) \right], \ldots,  \left[ b_{j}(O_{t}^{m}) \right].  
\end{equation*}
Applying the chain rule for products we get:
\begin{equation}
	\fpp [b_{j}(O_{t}^{1}, \ldots, O_{t}^{m})] =
	\sum_{l=1}^{m} \left[ \prod_{i=1, \ldots, \hat{l}, \ldots, m} 
b_{j}(O_{t}^{i}) \right] \times
	\fpp  [b_{j}(O_{t}^{l})],
	\label{partialProd}
\end{equation}
where $\hat{l}$ means that that term is left out of the product. 
These latter terms, $\frac{\partial} {\partial \lambda_{1} }  
[b_{j}(O_{t}^{k})]$, are easy to compute given either multinomial or 
gaussian observation densities $b_{j}(\dot)$

\section{Parameter estimation}

Parameters are estimated in \pkg{depmix} using a direct optimization
approach instead of the EM algorithm which is frequently used for this type
of model.  The EM algorithm however has some drawbacks.  First, it can be
slow to converge.  Second, applying constraints to parameters can be
problmatic.  The EM algorithm can sometimes lead to incorrect estimates
when constraints are applied to parameters in the M-step of the algorithm.
The package was designed to be used with the \code{npsol}-library, the
main reason being that it handles general linear (in-)equality
constraints very well. Unfortunately, \code{npsol} is not freeware and
hence is not distributed with \pkg{depmix}. Two other options are
available for optimization using  \code{nlm} and \code{optim}
respectively. Linear equality constraints are fitted through
reparametrization. Inequality constraints are fitted through adding a 
penalty to the likelihood depending on the amount by which a
constraint is not satisfied. The argument \code{vfactor} to the
fitting function can be used to control this bahavior. See details of 
this in the chapter on fitting models.

\chapter{Using depmix}

Three steps are involved in using \pkg{depmix} which are illustrated
below with examples:
\begin{enumerate}
	\item  data specification with function \code{markovdata}

	\item  model specification with function \code{dmm}

	\item  model fitting with function \code{fitdmm}
\end{enumerate}

To be able to fit models, data need to in a specific format created
for this package.  Basically, data should be in the form of a matrix
with each row corresponding to measures taken at a single measurement
occasion for a single subject.  The function \code{markovdata} further
only requires one argument providing the itemtypes, being one of
categorical, continuous or covariate.  A markovdata object is a matrix
with a number of attributes.

\section{Creating data sets}

As an example we make a dataset with two variables measured at two
times 50 occasions.

<<>>=
x=rnorm(100,10,2)
y=ifelse(runif(100)<0.5,0,1)
z=matrix(c(x,y),100,2)
md=markovdata(z,itemtypes=c("cont","cat"),ntimes=c(50,50))
md[1:10,]
@

In the example below, we split the dataset \code{speed} into three
separate datasets, which we later use as an example to do multi-group
analysis.

<<>>=
data(speed)
r1=markovdata(dat=speed[1:168,],itemt=itemtypes(speed))
r2=markovdata(dat=speed[169:302,],itemt=itemtypes(speed))
r3=markovdata(dat=speed[303:439,],itemt=itemtypes(speed))
summary(r2)
@

Here is the full specification of the \code{markovdata} function.

\input{markovdata}

\section{Data set speed}

Throughout this manual we will use a data set called speed, and hence
we provide some background information on how these data were
gathered.

\input{speed}

Interesting hypotheses to test are: is the switching regime symmetric?
Is there evidence for two states or does one state suffice?  Is the
guessing state actually a guessing state, i.e., is the probability
correct at chance level of 0.5?

\section{Defining models}

A dependent mixture model is defined by the number of states, and by
the item distribution functions, and can be created with the
\code{dmm}-function as follows:

<<>>= 
mod <-dmm(nstates=2,itemtypes=c("gaus",2))
summary(mod)
@

Here \code{itemtypes} is a vector of length the number of items
measured at each occasion specifying the desired distributions, in
this case the first item is to follow a normal distribution, and the
second item follows a bernouilli distribution.  Allowable
distributions are listed in Table~\ref{tbl:dist}, along with their
internal code, and the parameter names.  The R-internal code is used
for estimating these parameters.  Specifics of these distributions and
their estimation can be found in their respective help files.
Itemtypes can be specified by their name or by their internal code,
except in the case of multinomial items, which have to be specified by
a number.

\begin{table}[tbp]
	\centering
	\begin{tabular}{lcc}
		\hline
		distribution & code & parameters \\
		\hline
		\hline
		multinomial & $2,3,4, \ldots $ & $p_1, p_2, p_3, \ldots $  \\
		\hline
		gaussian, normal & 1  & $ \mu, \sigma $ \\
		\hline
		lognormal & -21 & $l\mu$ , $l\sigma$  \\
		\hline
		weibull & -22 & shape (a), scale (b) \\
		\hline
		gamma & -23 & shape (a), scale (s)  \\
		\hline
		3lognormal & -31 &  $l\mu$ , $l\sigma$, shift \\
		\hline
		3weibull & -32 & shape (a), scale (b), shift \\
		\hline
		3gamma & -33 & shape (a), scale (s), shift \\
		\hline
	\end{tabular}
	\caption{Allowable distribution names, internal codes, and number 
	of parameters.}
	\label{tbl:dist}
\end{table}

The function \code{dmm} returns an object of class \code{dmm} which
has its own summary function providing the parameter values of the
model. See the help files for further details. Except in simple 
cases, starting values can be a problem in latent Markov models, and
so in general it's best to provide them if you have a fairly good idea
of what to expect. Providing starting values is done through the stval
argument:

<<>>=
st <- c(1,0.9,0.1,0.2,0.8,2,1,0.7,0.3,5,2,0.2,0.8,0.5,0.5)
mod <- dmm(nsta=2,itemt=c(1,2), stval=st)
summary(mod)
@

\input{dmm}

\subsection{Generating data}

The \code{dmm}-class has a \code{generate} method that can be used to 
generate data according to a specified model.

<<>>=
gen<-generate(c(100,50),mod)
summary(gen)
@
\begin{figure}[htbp]
  \begin{center}
<<fig=TRUE>>=
plot(gen)
@
	\caption{Two timeseries generated by 2-state model with one 
	gaussian item and one binary item.}
  \end{center}
\end{figure}

\input{generate}

\section{Fitting models\label{fitmod}}

Fitting models is done using the function \code{fitdmm}. The standard
call only requires a dataset and a model as in:

<<>>= 
data(speed)
mod <- dmm(nstates=2,itemtypes=c(1,2))
fitex <- fitdmm(speed,mod)
@

Calling \code{fitdmm} produces some online ouput about the progress of
the optimization which can be controlled with the \code{printlevel}
argument.  Its default value of 1 just gives the first and the last
iteration of the optimization; 0 gives no output, and setting it to 5
or higher values will produce output at each iteration.  These values
correspond with the 0,1, and 2 printlevel of \code{nlm}.  When using
\code{optim}, the \code{printlevel} argument is used to set the
\code{REPORT} argument of \code{optim} (see its help page for
details).  Printlevel 0 gives report 0, printlevel 1 gives report 10,
printlevels 2--4 give report 5 and printlevel>4 gives report 1,
producing output at every iteration.  Printlevels starting from 15 and
higher produce increasingly annoying output from the C-routines that
compute the loglikelihood.

\code{Fitdmm} returns an object of class \code{fit} which has a
summary method showing the estimated parameter values, along with
standard errors, and t-ratios whenever those are available.  Along
with the log-likelihood, the AIC and BIC values are provided.  Apart
from the printed values (see summary below), a \code{fit}-object has a
number of other fields.  Most importantly, it contains a copy of the
fitted model in \code{mod} and it has a field \code{post} containing
posterior state estimates.  That is, for each group $g$,
\code{post\$states[[g]]} is a matrix with dimensions the number of
states of the model + 2, and \code{sum(ntimes(dat))}.  The first
column contains the a posteriori component model, the second column
has the state number within the component, and the other columns are
used for the a posteriori probabilities of each of the states.

<<>>=
summary(fitex)
@

\input{fitdmm}

\chapter{Extending and constraining models}

\section{Fixing and constraining parameters}

Continuing the example from above, it can be seen that in one of the
states, the probability of a correct answer is about .5, as is the
probability of an incorrect answer, i.e., these are parameters Item2,p1 and
Item2,p2. This latent state, is supposed to be a guessing state, and hence 
it makes sense to constrain these parameters to their theoretical values of 
.5. Similarly, the initial state probability for the slow state is one, and
zero for the other state, and hence it makes sense to fix these parameters.
The third constraint that we consider here is an equality constraint
between the transition parameters. Using this constraint, we can test the
hypothesis whether the switching between states is a symmetric process or
not. Hence, we constrain the transition parameters $a_{11}$ and $a_{22}$.

Constraining and fixing parameters is done in a similar fashion as the
\code{pa} command that is used in LISREL \citep{Lis99}. The
\code{conpat} argument to the \code{fitdmm}-function specifies for
each parameter in the model whether it's fixed (0) or free (1 or
higher). Equality constraints can be imposed by having two parameters have 
the same number in the \code{conpat} vector. When only fixed values are
required the \code{fixed} argument can be used instead of \code{conpat},
with zeroes for fixed parameters and other values (ones e.g.) for non-fixed
parameters.

Fitting the models subject to these constraints is mostly done through
reparametrization.  Inequality constraints are enforced by adding a
penalty to the loglikelihood when the constraint is not satisfied.
The penalty is linear in the amount by which the constraint is not
satisfied, and not logarithmic or something similar which is often
used (see e.g.\ the documentation for \code{constrOptim} which uses a
logarithmic boundary for inequality constraints).  This has advantages
and disadvantages.  There are two marked disadvantages.  First, the
loglikelihood is not smooth at the boundary of the paramter space.
Second, it can happen that the constraint is not satisfied.  Whenever
cosntraints are not satisfied \code{fitdmm} exits with a warning
stating the amount by which it is not satisfied.  This can be remedied
by upping the \code{vfactor} argument which simply increases the
penalty by this factor (its default value is 5).  An advantage is that
using a linear penalty, it is possible that the parameter is estimated
at the boundary, which is prohibited with logarithmic boundaries.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete vectors should be supplied, i.e., these
vectors should have length of the number of parameters of the model.
Parameters are numbered in the following order:
\begin{enumerate}
	\item  the mixing proportions of a mixture of latent Markov
	models, i.e., just one parameter for a single component model
	which has value 1 and is fixed

	\item  the component parameters for each component consisting of
	the following:
	\begin{enumerate}
		\item  transition parameters in row major order, $a_{11},
		a_{12}, a_{13}, \ldots, a_{21}, a_{22}, a_{23}, \ldots $

		\item  the observation parameters per state and per item, in
		the order listed in Table~\ref{tbl:dist}

		\item  the initial state probabilities per state
	\end{enumerate}

\end{enumerate}

<<>>=
conpat=rep(1,15)
conpat[1]=0 
conpat[14:15]=0 
conpat[8:9]=0
conpat[2]=conpat[5]=2
stv=c(1,.896,.104,.084,.916,5.52,.20,.5,.5,6.39,.24,.098,.902,0,1)
mod=dmm(nstates=2,itemt=c("n",2),stval=stv,conpat=conpat)
@

In the example above \code{conpat} is used to specify a number of
constraints.  First, \code{conpat[1]=0} specifies that the mixing
proportion of the model should be fixed (at its starting value of 1),
which is always the case for single component models.  Second,
\code{conpat[14:15]=0} fixes the initial state probabilities to zero
and one respectively.  Similarly, for \code{conpat[8:9]=0}, which are
the guessing state parameters for the accuracy scores.  They are both
fixed at 0.5 so as to make the guessing state an actual guessing
state.  Finally, by invoking \code{conpat[2]=conpat[5]=2}, transition
parameters $a_{11}$ and $a_{22}$ are set to be equal.  Whenever
equality constraints are not sufficient, general linear constraints
can be specified using the \code{conrows} argument.

The constrained model has the following estimated
parameters\footnote{Note that in running this example with the
starting values from the unconstrained model, the initial
loglikelihood is worse than the final loglikelihood because the
initial likelihood is based on parameters that do not satisfy the
constraints.}:

% <<results=hide>>=
% fitfix <- fitdmm(dat=speed,dmm=mod)
% @
% results in models.Rda

<<>>=
summary(fitfix)
@

\section{Multi group/case analysis}

<<echo=FALSE>>=
conpat=rep(1,15)
conpat[1]=0
conpat[8:9]=0
conpat[14:15]=0
stv=c(1,0.9,0.1,0.1,0.9,5.5,0.2,0.5,0.5,6.4,0.25,0.9,0.1,0.5,0.5)
mod<-dmm(nstates=2,itemt=c("n",2),stval=stv,conpat=conpat)
@

\code{depmix} can handle multiple cases or multiple groups. A
multigroup model is specified using the function \code{mgdmm} as
follows:

<<>>=
mgr <- mgdmm(dmm=mod,ng=3,trans=TRUE,obser=FALSE)
mgrfree <- mgdmm(dmm=mod,ng=3,trans=FALSE)
@

The \code{ng} argument specifies the number of groups, and the
\code{dmm} argument specifies the model for each group.  \code{dmm}
can be either a single model or list of models of \code{length(ng)}.
If it is a single model, each group has an identical structural model
(same fixed and constrained parameters), and else each group has its
model. Three further arguments can be used to constrain parameters
between groups, \code{trans}, \code{obser}, and \code{init}
respectively. By setting either of these to \code{TRUE}, the
corresponding transition, observation, and initial state parameters
are estimated equal between groups\footnote{There is at this moment no way of 
fine-tuning this to restrict equalities to individual parameters.
However, this can be accomplished by manually changing the linear
constraint matrix, and the corresponding upper and lower boundaries.}.

In this example, the model from above was used and fitted on the three
observed series, and the \code{trans=TRUE} ensures that the
transition matrix parameters are constrained to be equal between the
models for these series, whereas the observation parameters are freely
estimated, i.e.\ to capture learning effects. The resulting
parameters are:

% <<>>=
% fitmg <- fitdmm(dat=list(r1,r2,r3),dmm=mgr,ses=FALSE)
% fitmgfree <- fitdmm(dat=list(r1,r2,r3),dmm=mgrfree,ses=FALSE)
% @

<<>>=
summary(fitmg)
@

The loglikelihood ratio statistic can be used to test whether
constraining these transition parameters significantly reduces the
goodness-of-fit of the model.  The statistic has value
$LR=\Sexpr{round(2*(fitmgfree$loglike-fitmg$loglike),3)}$, and it has
an approximate $\chi^{2}$ distribution with $df=4$ because in each but
the first model, two transition matrix parameters were estimated equal
to the parameters in the first model (note that the other two
transition parameters were already had to be constrained to ensure
that the rows of the transition matrices sum to 1).  The associated
$p$-value for the statistic is
$p=\Sexpr{round(pchisq(2*(fitmgfree$loglike-fitmg$loglike),4,lower.tail=FALSE),3)}$,
indicating that constraining the transition matrix parameters does not
significantly worsen the goodness-of-fit of the model.

\input{mgdmm}

\section{Models with time-dependent covariates}

Specifying a model with covariates is done by including two arguments 
in a call to \code{dmm} called \code{tdfix} and \code{tdst}, where
\code{td} means time dependent. \code{tdfix} is a logical vector of
length the number of parameters of the model, specifying which
parameters are to be estimated time-dependent. For an arbitrary
parameter $\lambda$, the model that is estimated has the form:
\begin{equation}
	\lambda_{t}=\lambda_{0}+\beta x_{t},
\end{equation}
where $\lambda_{0}$ is the intercept of the parameter, $\beta$ is the 
regression coefficient, and $x_{t}$ is the time-dependent covariate.
The covariate has to be scaled to lie between 0 and 1; this is
neccessary to be able to impose the right constraints on $\beta$ in
order to ensure that $\lambda_{t}$ is always appropriate, ie within
its lower and upper bounds (mostly 0 and 1 for multinomial item
parameters and transition parameters etc). The current version of
\code{depmix} does not have non-time-dependent covariates, which can
simply be faked by having $x_{t}$ be constant, and there is only
support for a single covariate.

In the example below, the transition parameters (numbers 2--5) are
defined to depend on the covariate which is the pay-off for accuracy. 
Providing starting values for the covariates is optional. If not
provided they are chosen at random around 0 which usually works just
fine.

<<>>=
conpat=rep(1,15)
conpat[1]=0
conpat[8:9]=0
conpat[14:15]=0
conpat[2]=2
conpat[5]=2
stv=c(1,0.9,0.1,0.1,0.9,5.5,0.2,0.5,0.5,6.4,0.25,0.9,0.1,0,1)
tdfix=rep(0,15)
tdfix[2:5]=1
tdst=rep(0,15)
tdst[2:5]=c(-0.4,0.4,0.15,-0.15)

mod<-dmm(nstates=2,itemt=c("n",2),stval=stv,conpat=conpat,tdfix=tdfix,tdst=tdst,modname="twoboth+cov")
@

% <<results=hide>>=
% fittd <-
% fitdmm(dat=speed,dmm=mod,tdcov=1,print=5,der=0,ses=0,vf=130)
% @
% results in models.Rda

<<>>=
summary(fittd)
@

\chapter{Special topics}

\section{Starting values}

Although providing your own starting values is preferable,
\pkg{depmix} has a routine for generating starting values using the
\code{kmeans}-function from the \pkg{stats}-package.  This will
usually provide reasonable starting values, but can be way off in a
number of cases.  First, for univariate categorical time series,
\code{kmeans} does not work at all, and \pkg{depmix} will provide a
warning.  Second, for multivariate series with unordered categorical
items with more than 2 categories, \code{kmeans} may provide good
starting values, but they may similarly be completely off, due to the
implicit assumption in \code{kmeans} that the categories are
indicating an underlying continuum.  Starting values using
\code{kmeans} are automatically provided when a model is specified
without starting values.  The argument \code{kmst} to the
\code{fitdmm}-function can be used to control this behavior.

Starting values of the paramaters, either user provided or generated,
are further boosted by using posterior estimates.  That is, first the
a posteriori latent states are generated from the current parameter
values for the data at hand.  Next, from the a posteriori latent
states, new parameter estimates are derived.  This is done by default
and can be controlled by the \code{postst} argument.  Provided that
the starting values were close to their true values, using this
procedure further pushes those parameters in the right direction.  If
however the original values were bad, this procedure may result in bad
estimates, i.e., optimization will lead to some non-optimal local
maximum of the loglikelihood.

\section{Finite mixtures and latent class models}

The function \code{lca} can be used to specify latent class models
and/or finite mixture models. It is simply a wrapper for the \code{dmm}
function, and all it does is adding appropriate numbers of zeroes and 
ones to the parameter specification vectors for starting values, fixed
values and linear constraints. When a model has class \code{lca} the
summary function does not print the transition matrix (because it is
fixed and/or not estimated).

\section{Mixtures of latent Markov models}

\code{depmix} provides support for fitting mixtures of latent Markov
models using the \code{mixdmm} function; it takes a list of
\code{dmm}'s as argument, possibly together with the starting values
for the mixing proportions for each component model. There's an
example in the helpfiles.

\input{mixdmm}

An example of fitting a mixture of \code{dmm}'s is in the \code{fitdmm}
helpfile. It fits the model in the example to data from a
discrimination learning experiment which is provided as data set
\code{discrimination}.

\input{discrimination}

\nocite{Raijmakers2001}
\nocite{Maas2005b}
\nocite{Schmittmann2005b}
\nocite{McCutcheon1987}

\bibliography{all}

\end{document}

### 2009-07-08 papers/jss/depmix-intro.Rnw
Moved around files in the papers directory
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@285 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2009-07-08 papers/jss/depmix-manual.Rnw
@@
 
-\batchmode
+%\batchmode
 
@@
 
-\author{Ingmar Visser\thanks{
-Correspondence concerning this manual should be adressed to:
-Ingmar Visser, Department of Psychology, University of Amsterdam,
-Roetersstraat 15, 1018 WB, Amsterdam, The Netherlands}\\
+\author{Version: 0.9.1 \\ Ingmar Visser\thanks{Correspondence concerning this manual
+should be adressed to: Ingmar Visser, Department of Psychology,
+University of Amsterdam, Roetersstraat 15, 1018 WB, Amsterdam, The
+Netherlands}\\
+Developmental Processes Research Group\\
 Department of Psychology, University of Amsterdam\\
@@
 \begin{abstract}
-
 	\pkg{depmix} implements a general class of mixture models with
 	Markovian dependencies between them in the R programming language
-	\citep{R2007}.  This includes standard Markov models, latent/hidden
-	Markov models, and latent class and finite mixture distribution models.
-	The models can be fitted on mixed multivariate data with multinomial
-	and/or gaussian distributions.  Parameters can be estimated subject to
-	general linear constraints, and with optional inclusion of regression
-	on (time-dependent) covariates.  Parameters estimation is done through
-	a direct optimization approach with gradients using the \code{Rdonlp2}
-	optimization routine.  A number of illustrative examples are included.
-
+	\citep{R2006}.  This includes standard Markov models,
+	latent/hidden Markov models, and latent class and finite mixture
+	distribution models.  The models can be fitted on mixed
+	multivariate data from a number of distributions including the
+	binomial, multinomial, gaussian, lognormal and Weibull
+	distributions.  Parameters can be estimated subject to general
+	linear constraints, and with optional inclusion of regression on
+	(time-dependent) covariates.  Parameter estimation is done
+	through a direct optimization approach with gradients using the
+	\code{nlm} optimization routine.  Optional support for using the NPSOL
+	optimization routines is provided as well.  A number of illustrative
+	examples are included.
 \end{abstract}
@@
 are used for modelling learning processes, see \citet{Wickens1982}, for an
-overview, and \citet{Schmittmann2006} for a recent application.  In
+overview, and \citet{Schmittmann2005} for a recent application.  In
 economics, latent Markov models are commonly used as regime switching
@@
 was available for fitting such models.  Common programs for Markovian
-models include Panmark \citep{Panmark3-96}, and for latent class models
+models include Panmark \citep{Pol1996}, and for latent class models
 Latent Gold 
-\citep{LatentGold3-2003}.  Those programs are lacking a number of
+\citep{Vermunt2003}.  Those programs are lacking a number of
 important features, besides not being freely available.  In
@@
 \section*{Acknowledgements}
+
+Development of this pacakge was supported by European Commission grant
+51652 (NEST) and by a VENI grant from the Dutch Organization for Scientific
+Research (NWO) to Ingmar Visser.
 
@@
 common estimation procedures, such as those implemented in
-\citet{Panmark3-96} are not suitable for long time series due to
+\citet{Pol1996} are not suitable for long time series due to
 underflow problems.  In contrast, the hidden Markov model is typically
@@
 The loglikelihood of hidden Markov models is usually computed by the
-so-called forward-backward algorithm \citep{Bau66,Rab89}, or rather by
-the forward part of this algorithm.  \cite{Lys2002} changed the forward
+so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or rather by
+the forward part of this algorithm.  \cite{Lystig2002} changed the forward
 algorithm in such a way as to allow computing the gradients of the
@@
 
-To compute the log-likelihood, \cite{Lys2002} define the following 
+To compute the log-likelihood, \cite{Lystig2002} define the following 
 (forward) recursion:
@@
 
-See equations 10--12  in \cite{Lys2002} for the score recursion 
+See equations 10--12  in \cite{Lystig2002} for the score recursion 
 functions of the hidden Markov model for a univariate time series. 
@@
 
-Using above equations, \cite{Lys2002} derive the following equation 
+Using above equations, \cite{Lystig2002} derive the following equation 
 for the partial derivative of the likelihood:
@@
 Constraining and fixing parameters is done in a similar fashion as the
-\code{pa} command that is used in LISREL \citep{Lis99}. The
+\code{pa} command that is used in LISREL \citep{Joreskog1999}. The
 \code{conpat} argument to the \code{fitdmm}-function specifies for
@@
 
+\section{Known issues and future plans}
+
+\paragraph{Constraint violations} Parameter optimization is done by adding
+a penalty to the log likelihood whenever constraints are not satisfied, ie
+linear inequality and box constraints.  Equality constraints are fitted by
+reparametrization and do not suffer from this problem.  Refer to the
+section on parameter estimation and the \code{fitdmm} help page on how to
+deal with constraint violations. Future plans include using Lagrange
+multipliers to overcome this problem.
+
 \nocite{Raijmakers2001}
 \nocite{Maas2005b}
-\nocite{Schmittmann2005b}
+\nocite{Schmittmann2006}
 \nocite{McCutcheon1987}
+\nocite{Visser2006b}
 
-\bibliography{all}
+\bibliography{all,ingmar}
 

### 2009-07-08 papers/jss/article.tex
Written section/subsections into jss style paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@290 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Ingmar Visser\\University of Amsterdam \And 
        Maarten Speekenbrink\\University College London }
\title{\pkg{depmixS4}: An \proglang{R}-package for hidden Markov models}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
\Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting
\Shorttitle{depmixS4 Hidden Markov Models} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
	\pkg{depmixS4} implements a general framework for definining and
	fitting hidden Markov mixture model in the R programming language
	\citep{R2009}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the
	\pkg{Rdonlp2} optimization routine when contraints are imposed on
	the parameters.  A number of illustrative examples are included.

}
\Keywords{hidden Markov model, dependent mixture model, mixture model, \proglang{R}}
\Plainkeywords{hidden Markov model, dependent mixture model, mixture model,R} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Ingmar Visser\\
Department of Psychology\\
University of Amsterdam,
Roetersstraat 15\\
1018 WB, Amsterdam\\
The Netherlands\\
E-mail: \email{i.visser@uva.nl} \\
URL: \url{http://www.ingmar.org/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

\section[]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

Why did we develop depmixS4?

\section{The dependent mixture model}

Describe the model formulae

\subsection{Likelihood}

Give some refs to computing the likelihood

\subsection{Parameter estimation}

Give some refs to EM/forward/backbward and possibly others.

\section{Using \pkg{depmixS4}}

Explain the three steps involved in fitting a model.

\subsection{Start-up example}

Just the RT data.

\subsection{Adding covariates on transition parameters}

Give an example on how to do this.

\subsection{Add covariates to prior model}

Give an example with the balance data.

\section{Extending \pkg{depmixS4}}

Introduce the exgaus example.

Explain the use of makeDepmix for having full control of every aspect of the model.

\section{Conclusion \& future work}

What are our next plans?

\bibliography{all,ingmar}

\end{document}

### 2009-07-09 papers/jss/article.tex
Filled in most sections of the paper, added first 2 examples.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@291 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 }
+
 %% It is also possible to add a telephone and fax number
@@
 
+\batchmode
+
 \begin{document}
@@
 
-Why did we develop depmixS4?
+Markov and latent Markov models are frequently used in the social
+sciences, in different areas and applications.  In psychology, they
+are used for modelling learning processes, see \citet{Wickens1982},
+for an overview, and \citet{Schmittmann2006} for a recent application.
+In economics, latent Markov models are commonly used as regime
+switching models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.
+Further applications include speech recognition \citep{Rabiner1989},
+EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
+those latter areas of application, latent Markov models are usually
+referred to as hidden Markov models.
+
+The \pkg{depmixS4} package was motivated by the fact that Markov models
+are used commonly in the social sciences, but no comprehensive package
+was available for fitting such models.  Common programs for Markovian
+models include Panmark \citep{Pol1996}, and for latent class models
+Latent Gold \citep{Vermunt2003}.  Those programs are lacking a number
+of important features, besides not being freely available.  There are
+currently some packages in R that handle hidden Markov models but they
+lack a number of features that we needed in our research.  In
+particular, \pkg{depmixS4} was designed to meet the following goals: 
+\begin{enumerate}
+
+	\item to be able to handle parameter estimates subject to general
+	linear (in)equality constraints
+
+	\item to be able to fit transition models with covariates, i.e.,
+	to have time-dependent transition matrices
+
+	\item to be able to include covariates in the prior or initial
+	state probabilities of models
+
+	\item to allow for easy extensibility, in particular, to be able
+	to add new response distributions, both univariate and
+	multivariate, and similarly to be able to allow for the addition
+	of other transition models, e.g., continuous time observation
+	models
+
+\end{enumerate}
+
+Although \pkg{depmixS4} is designed to deal with
+longitudinal or time series data, for say $T>100$, it can also handle
+the limit case with $T=1$.  In those cases, there are no time
+dependencies between observed data, and the model reduces to a finite
+mixture model, or a latent class model.  Although there are other
+specialized packages to deal with mixture data, one specific feature
+that we needed ourselves which is to the best of our knowledge not
+available in other packages is the possibility to include covariates
+on the prior probabilities of class membership.  In the next section,
+an outline is provided of the model and the likelihood equations.
+
+\subsection*{Acknowledgements}
+
+Ingmar Visser was supported by an EC Framework 6 grant, project 516542
+(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
+Economic Learning and Social Evolution (ELSE).  Han van der Maas
+provided the speed-accuracy data \cite{Maas2005b} and thereby
+neccessitated implementing models with time-dependent covariates.
+Brenda Jansen provided the balance scale data set \citep{Jansen2001}
+which was the perfect opportunity to test the covariates on the prior
+model parameters.  The examples in the help files use both of these
+data sets.
 
@@
 
-Describe the model formulae
+The data considered here, has the general form $O_{1}^{1}, \ldots,
+O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1}, \ldots,
+O_{T}^{m}$ for an $m$-variate time series of length $T$.  As an
+example, consider a time series of responses generated by a single
+subject in a reaction time experiment.  The data consists of three
+variables, reaction time, accuracy and a covariate which is a pay-off
+factor which determines the reward for speed and accuracy.  These
+variables are measured on 168, 134 and 137 occasions respectively 
+(in Figure~\ref{fig:speed} the first part of this series is plotted).
+
+\begin{figure}[htbp]
+  \begin{center}
+	  \scalebox{1}{\includegraphics*{graphs/speed1.pdf}}
+	  \caption{Reaction times, accuracy and pay-off values for
+	  the first series of responses in dataset \texttt{speed}.}
+	  \label{fig:speed}
+  \end{center}
+\end{figure}
+
+The latent Markov model is commonly associated with data of this type,
+albeit usually only multinomial variables are considered.  However,
+common estimation procedures, such as those implemented in
+\citet{Pol1996} are not suitable for long time series due to underflow
+problems.  In contrast, the hidden Markov model is typically only used
+for `long' univariate time series.  In the next sections, the
+likelihood and estimation procedure for the hidden Markov model is
+described, given data of the above form.  These models are called
+dependent mixture models because one of the authors (Ingmar Visser)
+thought it was time for a new name for these models\footnote{Only later 
+did I find out that \citet{Leroux1992} already coined the term dependent 
+mixture models in an application with hidden Markov mixtures of Poisson 
+count data.}
+
+The dependent mixture model is defined by the following elements:
+\begin{enumerate}
+
+	\item a set $\vc{S}$ of latent classes or states $S_{i},\, i=1,
+	\ldots , n$,
+
+	\item matrices $\mat{A}_t$ of transition probabilities $a_{ij,t}$ for
+	the transition from state $S_{i}$ to state $S_{j}$ at time $t$,
+
+	\item a set $\vc{B}_t$ of observation functions $b_j^k(\cdot)$ that
+	provide the conditional probabilities of observations $O_{t}^k$ 
+	associated with latent state $S_{j}$,
+
+	\item a vector $\pmb{\pi}$ of latent state initial probabilities
+	$\pi_{i}$
+\end{enumerate}
+When transitions are added to the latent class model, it is more 
+appropriate to refer to the classes as states. The word class is 
+rather more associated with a stable trait-like attribute whereas a 
+state can change over time.
+
+Need model equations: maybe from Fruhwirth-Schnatter?
 
@@
 
-Give some refs to computing the likelihood
+The log-likelihood of hidden Markov models is usually computed by the
+so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
+rather by the forward part of this algorithm.  \cite{Lystig2002}
+changed the forward algorithm in such a way as to allow computing the
+gradients of the log-likelihood at the same time.  They start by
+rewriting the likelihood as follows (for ease of exposition the
+dependence on the model parameters is dropped here):
+\begin{equation}
+	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
+Pr(\vc{O}_{t}|\vc{O}_{1}, 
+	\ldots, \vc{O}_{t-1}), 
+	\label{condLike}
+\end{equation}
+where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
+simple, i.e.\ observed, Markov chain these probabilities reduce to 
+$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
+\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
+The log-likelihood can now be expressed as:
+\begin{equation}
+	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
+\vc{O}_{t-1})].
+	\label{eq:condLogl}
+\end{equation}
+
+To compute the log-likelihood, \cite{Lystig2002} define the following 
+(forward) recursion:
+\begin{align}
+	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
+	\label{eq:fwd1} \\
+\begin{split}
+	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
+\vc{O}_{t-1}) \\
+	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
+(\Phi_{t-1})^{-1},
+	\label{eq:fwdt} 
+\end{split} 
+\end{align}
+where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
+$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
+equation~(\ref{eq:condLogl}) gives the following expression for the 
+log-likelihood:
+\begin{equation}
+	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
+	\label{eq:logl}
+\end{equation}
 
@@
 
-Give some refs to EM/forward/backbward and possibly others.
+Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
+through the use of a general Newton-Raphson optimizer.  The EM
+algorithm however has some drawbacks.  First, it can be slow to
+converge towards the end of optimization (although it is usually
+faster than direct optimization at the start, so possibly a
+combination of EM and direct optimization is fastest).  Second,
+applying constraints to parameters can be problematic; in particular,
+EM can lead to wrong parameter estimates when applying constraints.
+Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
+models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
+\cite{Tamura2007,Spellucci2002}, because it handles general linear
+(in)equality constraints, and optionally also non-linear constraints.
+
+Need some more on EM and how/why it is justified to do separate weighted
+fits of the response models and transition and prior models.
+
+Also mention use of glm, nnet and possibly other packages that we use.
 
@@
 
-Explain the three steps involved in fitting a model.
+Two steps are involved in using \pkg{depmixS4} which are illustrated
+below with examples:
+\begin{enumerate}
+	\item model specification with function \code{depmix}
 
-\subsection{Start-up example}
+	\item  model fitting with function \code{fit}
+\end{enumerate}
 
-Just the RT data.
+\subsection{Example data: speed}
+
+Throughout this manual a data set called \code{speed} is used.  It
+consists of three time series with three variables: reaction time,
+accuracy, and a covariate Pacc which defines the relative pay-off for
+speeded and accurate responding. The participant in this experiment 
+switches between fast responding at chance level and relatively 
+slower responding at a high level of accuracy.
+
+Interesting hypotheses to test are: is the switching regime symmetric?
+Is there evidence for two states or does one state suffice?  Is the
+guessing state actually a guessing state, i.e., is the probability
+of a correct response at chance level (0.5)?
+
+\subsection{A simple example}
+
+A dependent mixture model is defined by the number of states, and by
+the response distribution functions, and can be created with the
+\code{depmix}-function as follows (see the help files for other
+options):
+
+\begin{verbatim} 
+set.seed(1)
+mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
+fm <- fit(mod)
+\end{verbatim}
+
+The function returns an object of class \code{depmix.fitted} which
+extends the \code{depmix} class, adding convergence information (and
+information about constraints if these were applied, see section on Constraints below).  The
+\code{print} method provides summary information on convergence, the log likelihood
+and the AIC and BIC values. These statistics may be extracted using  \code{logLik},
+\code{AIC} and \code{BIC}, respectively.
+
+\begin{verbatim}
+> fm 
+Convergence info: Log likelihood converged to within tol. 
+'log Lik.' -84.34175 (df=7)
+AIC:  182.6835 
+BIC:  211.275 
+\end{verbatim}
+
+The \code{summary} method of fitted models provides the parameter estimates, first for
+the prior probabilities model, second for the transition model, and third for the response models.
+
+\begin{verbatim}
+> summary(fm)
+Initial state probabilties model 
+Model of type multinomial, formula: ~1
+Coefficients: 
+     [,1]      [,2]
+[1,]    0 -11.25688
+Probalities at zero values of the covariates.
+0.999987 1.291797e-05
+
+Transition model for state (component) 1 
+Model of type multinomial, formula: ~1
+Coefficients: 
+[1]  0.000000 -2.392455
+Probalities at zero values of the covariates.
+0.9162501 0.08374986
+
+Transition model for state (component) 2 
+Model of type multinomial, formula: ~1
+Coefficients: 
+[1] 0.000000 2.139255
+Probalities at zero values of the covariates.
+0.1053396 0.8946604
+
+Response model(s) for state 1
+
+Response model for response 1 
+Model of type gaussian, formula: rt ~ 1
+Coefficients: 
+[1] 6.385492
+sd  0.2439376
+
+Response model(s) for state 2
+
+Response model for response 1 
+Model of type gaussian, formula: rt ~ 1
+Coefficients: 
+[1] 5.511151
+sd  0.1926063 
+\end{verbatim}
 
@@
 
-Give an example on how to do this.
+The transition matrix is parametrized as a list of multinomial
+logistic models.  The initial state probabilities are similarly
+parametrized as a multinomial logistic model.  Both models use a baseline
+category parametrization, meaning that the parameter for the base
+category is fixed at zero.  The default base category is the first
+state.  Hence, for example, for a 3-state model, the initial state
+probability model would have three parameters of which the first is
+fixed at zero and the other two are freely estimated.
+
+See \citet[see][p.\ 267 ff.]{Agresti2002} for multinomial logistic models and various
+parameterizations.
+
+Covariates can be specified using a one-sided formula as in the
+following example:
+
+\begin{verbatim}
+set.seed(1)
+mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, family=list(gaussian(), multinomial()),
+transition=~Pacc, instart=runif(2))
+fm <- fit(mod)
+\end{verbatim}
 
@@
 
-Give an example with the balance data.
+Add example of balance data.
+
+Note that this can be done for the initial state probabilities by
+specifying \code{prior=~X1}, where X1 is the desired covariate.  The result
+of this is that the transition probabilities are now dependent on the
+covariate Pacc (which is an experimenter controlled variable to induce
+switching between guessing and accurate responding).
+
+\begin{verbatim}
+data(balance)
+set.seed(1)
+mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
+        family=list(multinomial(),multinomial(),multinomial(),multinomial()),
+        respstart=runif(16))
+fm <- fit(mod)
+\end{verbatim}
+
+\subsection{Fixing and constraining parameters}
+
+Using package \pkg{Rdonlp2}, parameters may be fitted subject to 
+general linear (in-)equality constraints. In particular, the following
+constraints may be applied:
+
+Constraining and fixing parameters is done using the \code{conpat}
+argument to the \code{depmix.fit}-function, which specifies for each
+parameter in the model whether it's fixed (0) or free (1 or higher).
+Equality constraints can be imposed by having two parameters have the
+same number in the \code{conpat} vector.  When only fixed values are
+required the \code{fixed} argument can be used instead of
+\code{conpat}, with zeroes for fixed parameters and other values (ones
+e.g.) for non-fixed parameters.  Fitting the models subject to these
+constraints is handled by the optimization routine \code{donlp2}.
+
+\paragraph{Parameter numbering} When using the \code{conpat} and
+\code{fixed} arguments, complete parameter vectors should be supplied,
+i.e., these vectors should have length of the number of parameters of
+the model, which can be obtained by calling \code{npar(object)}.
+Parameters are numbered in the following order:
+\begin{enumerate}
+	\item  the prior model parameters
+
+	\item  the parameters for the transition models
+
+	\item  the response model parameters per state (and subsequently
+	per response in the case of multivariate time series)
+
+\end{enumerate}
+
+To see the ordering of parameters use the following:
+\begin{verbatim}
+mod <- setpars(mod, value=1:npar(mod))mod
+\end{verbatim}
+
+To see which parameters are fixed (by default only baseline parameters
+in the multinomial logistic models for the transition models and the
+initial state probabilities model:
+\begin{verbatim}
+mod <- setpars(mod,
+getpars(mod,which="fixed"))mod
+\end{verbatim}
 
@@
 
+Adding gradients for speed and computation of the Hessian.
+
 \bibliography{all,ingmar}

### 2009-07-13 papers/jss/article.tex
Added all the code examples, albeit with minimal text.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@293 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	\Abstract{
+
 	\pkg{depmixS4} implements a general framework for definining and
@@
 	Parameter estimation is done through an EM algorithm or by a
-	direct optimization approach using the
-	\pkg{Rdonlp2} optimization routine when contraints are imposed on
-	the parameters.  A number of illustrative examples are included.
+	direct optimization approach using the \pkg{Rdonlp2} optimization
+	routine when contraints are imposed on the parameters.  A number
+	of illustrative examples are included.
 
 }
+
 \Keywords{hidden Markov model, dependent mixture model, mixture model, \proglang{R}}
+
 \Plainkeywords{hidden Markov model, dependent mixture model, mixture model,R} %% without formatting
+
 %% at least one keyword must be supplied
@@
 
-The \pkg{depmixS4} package was motivated by the fact that Markov models
-are used commonly in the social sciences, but no comprehensive package
-was available for fitting such models.  Common programs for Markovian
-models include Panmark \citep{Pol1996}, and for latent class models
-Latent Gold \citep{Vermunt2003}.  Those programs are lacking a number
-of important features, besides not being freely available.  There are
-currently some packages in R that handle hidden Markov models but they
-lack a number of features that we needed in our research.  In
-particular, \pkg{depmixS4} was designed to meet the following goals: 
+The \pkg{depmixS4} package was motivated by the fact that Markov
+models are used commonly in the social sciences, but no comprehensive
+package was available for fitting such models.  Common programs for
+Markovian models include Panmark \citep{Pol1996}, and for latent class
+models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
+number of important features, besides not being freely available.
+There are currently some packages in R that handle hidden Markov
+models but they lack a number of features that we needed in our
+research.  In particular, \pkg{depmixS4} was designed to meet the
+following goals:
 \begin{enumerate}
@@
 
-Although \pkg{depmixS4} is designed to deal with
-longitudinal or time series data, for say $T>100$, it can also handle
-the limit case with $T=1$.  In those cases, there are no time
-dependencies between observed data, and the model reduces to a finite
-mixture model, or a latent class model.  Although there are other
-specialized packages to deal with mixture data, one specific feature
-that we needed ourselves which is to the best of our knowledge not
-available in other packages is the possibility to include covariates
-on the prior probabilities of class membership.  In the next section,
-an outline is provided of the model and the likelihood equations.
+Although \pkg{depmixS4} is designed to deal with longitudinal or time
+series data, for say $T>100$, it can also handle the limit case with
+$T=1$.  In those cases, there are no time dependencies between
+observed data, and the model reduces to a finite mixture model, or a
+latent class model.  Although there are other specialized packages to
+deal with mixture data, one specific feature that we needed ourselves
+which is to the best of our knowledge not available in other packages
+is the possibility to include covariates on the prior probabilities of
+class membership.  In the next section, an outline is provided of the
+model and the likelihood equations.
 
@@
 The data considered here, has the general form $O_{1}^{1}, \ldots,
-O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1}, \ldots,
-O_{T}^{m}$ for an $m$-variate time series of length $T$.  As an
-example, consider a time series of responses generated by a single
+O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
+\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
+an example, consider a time series of responses generated by a single
 subject in a reaction time experiment.  The data consists of three
@@
 factor which determines the reward for speed and accuracy.  These
-variables are measured on 168, 134 and 137 occasions respectively 
-(in Figure~\ref{fig:speed} the first part of this series is plotted).
+variables are measured on 168, 134 and 137 occasions respectively (in
+Figure~\ref{fig:speed} the first part of this series is plotted).
 
@@
   \begin{center}
-	  \scalebox{1}{\includegraphics*{graphs/speed1.pdf}}
-	  \caption{Reaction times, accuracy and pay-off values for
+	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
+	  \caption{Response times, accuracy and pay-off values for
 	  the first series of responses in dataset \texttt{speed}.}
@@
 dependent mixture models because one of the authors (Ingmar Visser)
-thought it was time for a new name for these models\footnote{Only later 
-did I find out that \citet{Leroux1992} already coined the term dependent 
-mixture models in an application with hidden Markov mixtures of Poisson 
-count data.}
+thought it was time for a new name for these models\footnote{Only
+later did I find out that \citet{Leroux1992} already coined the term
+dependent mixture models in an application with hidden Markov mixtures
+of Poisson count data.}
 
@@
 
-Need model equations: maybe from Fruhwirth-Schnatter?
+The dependent mixture model is then given by the following equations: 
+\begin{align}
+	S_{t} = A S_{t-1}, t=2, \ldots, T \\
+	O_{t} = f(O_{t}|S_{t}),
+\end{align}
+where $S_{t}$ is a sequence of hidden states, $A$ is a transition
+matrix, $O_{t}$ is an (possibly multivariate) observation and $f$ is a
+density function for $O_{t}$ conditional on the hidden state $S_{t}$.
+In the example data above, $f$ could be a gaussian distribution
+function for the response time variable, and a Bernouilli for the
+accuracy data.  In the models we are considering here, both the
+transition probabilities $A$ and the initial state probabilities $\pi$
+may depend on covariates as well as the response distributions $f$. 
+See for example \citet{Fruhwirth-Schnatter2006} for an overview of 
+hidden Markov models with extensions.
 
@@
 Throughout this manual a data set called \code{speed} is used.  It
-consists of three time series with three variables: reaction time,
+consists of three time series with three variables: response time,
 accuracy, and a covariate Pacc which defines the relative pay-off for
 speeded and accurate responding.  The participant in this experiment
-switches between fast responding at chance level and relatively 
-slower responding at a high level of accuracy.
-
-Interesting hypotheses to test are: is the switching regime symmetric?
-Is there evidence for two states or does one state suffice?  Is the
-guessing state actually a guessing state, i.e., is the probability
-of a correct response at chance level (0.5)?
+switches between fast responding at chance level and relatively slower
+responding at a high level of accuracy.  Interesting hypotheses to
+test are: is the switching regime symmetric?  Is there evidence for
+two states or does one state suffice?  Is the guessing state actually
+a guessing state, i.e., is the probability of a correct response at
+chance level (0.5)?
 
@@
 the response distribution functions, and can be created with the
-\code{depmix}-function as follows (see the help files for other
-options):
+\code{depmix}-function as follows:
 
@@
 mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
+\end{verbatim}
+The \code{depmix} function returns an object of class \code{depmix}
+which contains the model specification (and not a fitted model!).
+Note also that start values for the transition parameters are provided
+in this call using the \code{trstart} argument. The package does not 
+provide automatic starting values.
+
+The so-defined models needs to be \code{fit}ted with the following: 
+\begin{verbatim}
 fm <- fit(mod)
@@
 
-The function returns an object of class \code{depmix.fitted} which
-extends the \code{depmix} class, adding convergence information (and
-information about constraints if these were applied, see section on Constraints below).  The
-\code{print} method provides summary information on convergence, the log likelihood
-and the AIC and BIC values. These statistics may be extracted using  \code{logLik},
+The \code{fit}-function returns an object of class
+\code{depmix.fitted} which extends the \code{depmix} class, adding
+convergence information (and information about constraints if these
+were applied, see below).  The \code{print} method provides summary
+information on convergence, the log likelihood and the AIC and BIC
+values.  These statistics may be extracted using \code{logLik},
 \code{AIC} and \code{BIC}, respectively.
@@
 
-The \code{summary} method of fitted models provides the parameter estimates, first for
-the prior probabilities model, second for the transition model, and third for the response models.
+The \code{summary} method of \code{fit}ted models provides the parameter
+estimates, first for the prior probabilities model, second for the
+transition model, and third for the response models.
 
@@
 
-\subsection{Adding covariates on transition parameters}
+\subsection{Transition parameters}
 
@@
 logistic models.  The initial state probabilities are similarly
-parametrized as a multinomial logistic model.  Both models use a baseline
-category parametrization, meaning that the parameter for the base
-category is fixed at zero.  The default base category is the first
-state.  Hence, for example, for a 3-state model, the initial state
-probability model would have three parameters of which the first is
-fixed at zero and the other two are freely estimated.
+parametrized as a multinomial logistic model.  Both models use a
+baseline category parametrization, meaning that the parameter for the
+base category is fixed at zero (see \citet[see][p.\ 267
+ff.]{Agresti2002} for multinomial logistic models and various
+parameterizations).  The default base category is the first state.
+Hence, for example, for a 3-state model, the initial state probability
+model would have three parameters of which the first is fixed at zero
+and the other two are freely estimated.
 
-See \citet[see][p.\ 267 ff.]{Agresti2002} for multinomial logistic models and various
-parameterizations.
-
-Covariates can be specified using a one-sided formula as in the
-following example:
-
+Covariates on the transition probabilities can be specified using a
+one-sided formula as in the following example:
 \begin{verbatim}
 set.seed(1)
-mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, family=list(gaussian(), multinomial()),
-transition=~Pacc, instart=runif(2))
+mod <- depmix(rt~1, data=speed, nstates=2, 
+family=gaussian(), transition=~scale(Pacc), 
+instart=runif(2))
 fm <- fit(mod)
@@
 
-\subsection{Add covariates to prior model}
+\begin{verbatim}
+	Initial state probabilties model 
+	Model of type multinomial, formula: ~1
+	Coefficients: 
+		 [,1]     [,2]
+	[1,]    0 10.71779
+	Probalities at zero values of the covariates.
+	2.214681e-05 0.9999779
 
-Add example of balance data.
+	Transition model for state (component) 1 
+	Model of type multinomial, formula: ~scale(Pacc)
+	Coefficients: 
+		 [,1]       [,2]
+	[1,]    0 -0.9215182
+	[2,]    0  1.8649734
+	Probalities at zero values of the covariates.
+	0.7153513 0.2846487
 
-Note that this can be done for the initial state probabilities by
-specifying \code{prior=~X1}, where X1 is the desired covariate.  The result
-of this is that the transition probabilities are now dependent on the
-covariate Pacc (which is an experimenter controlled variable to induce
-switching between guessing and accurate responding).
+	Transition model for state (component) 2 
+	Model of type multinomial, formula: ~scale(Pacc)
+	Coefficients: 
+		 [,1]     [,2]
+	[1,]    0 2.471442
+	[2,]    0 3.570856
+	Probalities at zero values of the covariates.
+	0.07788458 0.9221154
 
+	Response model(s) for state 1
+
+	Response model for response 1 
+	Model of type gaussian, formula: rt ~ 1
+	Coefficients: 
+	[1] 5.512179
+	sd  0.1921098
+
+	Response model(s) for state 2
+
+	Response model for response 1 
+	Model of type gaussian, formula: rt ~ 1
+	Coefficients: 
+	[1] 6.3885
+	sd  0.2402693
+
+\end{verbatim}
+
+The summary provides all parameters of the model, also the 
+(redundant) zeroes for the base-line category in the multinomial model. 
+The summary also prints the transition probabilities at the zero value 
+of the covariate. Note that scaling of the covariate is useful in this 
+regard as it makes interpretation of these intercept probabilities 
+easier.
+
+\subsection{Multivariate data}
+
+Multivariate data can be modelled by providing a list of formulae as 
+well as a list of family objects for the distributions of the various 
+responses. In above examples we have only used the response times 
+which were modelled with the gaussian distribution. The accuracy data 
+are in the \code{speed} data are modelled with a multinomial by 
+specifying the following: 
 \begin{verbatim}
-data(balance)
+set.seed(1)
+mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
+	family=list(gaussian(),multinomial()), transition=~scale(Pacc), 
+	instart=runif(2))
+fm <- fit(mod)
+\end{verbatim}
+which provides the following fitted model parameters (only the 
+response parameters are given here): 
+\begin{verbatim}
+	> summary(fm)
+	\ldots 
+	Response model(s) for state 1
+
+	Response model for response 1 
+	Model of type gaussian, formula: rt ~ 1
+	Coefficients: 
+	[1] 5.52169
+	sd  0.2028857
+
+	Response model for response 2 
+	Model of type multinomial, formula: corr ~ 1
+	Coefficients: 
+		 [,1]      [,2]
+	[1,]    0 0.1030554
+	Probalities at zero values of the covariates.
+	0.4742589 0.5257411
+
+	Response model(s) for state 2
+
+	Response model for response 1 
+	Model of type gaussian, formula: rt ~ 1
+	Coefficients: 
+	[1] 6.39369
+	sd  0.2373650
+
+	Response model for response 2 
+	Model of type multinomial, formula: corr ~ 1
+	Coefficients: 
+		 [,1]     [,2]
+	[1,]    0 2.245514
+	Probalities at zero values of the covariates.
+	0.09573715 0.9042629 	
+\end{verbatim}
+As can be seen, state 1 has fast response times around chance level, 
+whereas state 2 corresponds with slower responding at higher accuracy 
+levels.
+
+\subsection{Adding covariates on the prior probabilities}
+
+The \pkg{depmixS4} contains another data set which is used to
+illustrate the use of covariates on the prior probabilities of models,
+in this case a latent class model. The \code{balance} data consists 
+of 4 binary items (correct-incorrect) on a balance scale task 
+\citet{Siegler198?}. The data form a subset of the data published in 
+\citet{Jansen199?}.
+
+Similarly to the transition matrix, covariates on the prior 
+probabilities of the latent states (or classes in this case), are 
+defined by using a one-sided formula: 
+\begin{verbatim}
+balance$age <- balance$age-5
 set.seed(1)
 mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
-        family=list(multinomial(),multinomial(),multinomial(),multinomial()),
-        respstart=runif(16))
+        family=list(multinomial(), multinomial(), multinomial(),
+		multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
+		prior=~age, initdata=balance)
 fm <- fit(mod)
 \end{verbatim}
+Note here that we define a \code{mix} model instead of a \code{depmix}
+models as these data form independent observations.
+
+The summary of the \code{fit}ted model gives (only the prior model is 
+shown here): 
+\begin{verbatim}
+> summary(fm)
+Mixture probabilities model 
+Model of type multinomial, formula: ~age
+		[,1]   [,2]
+[1,]    0 -2.5182573
+[2,]    0  0.5512996
+Probalities at zero values of the covariates.
+0.9254119 0.07458815 
+\end{verbatim}	
+Hence at young ages, children have a high probability of incorrect 
+responding in class~1, whereas the prior probability for class~2 
+increases with age.
 
@@
 Using package \pkg{Rdonlp2}, parameters may be fitted subject to
-general linear (in-)equality constraints. In particular, the following
-constraints may be applied:
-
-Constraining and fixing parameters is done using the \code{conpat}
-argument to the \code{depmix.fit}-function, which specifies for each
-parameter in the model whether it's fixed (0) or free (1 or higher).
-Equality constraints can be imposed by having two parameters have the
-same number in the \code{conpat} vector.  When only fixed values are
+general linear (in-)equality constraints.  Constraining and fixing
+parameters is done using the \code{conpat} argument to the
+\code{depmix.fit}-function, which specifies for each parameter in the
+model whether it's fixed (0) or free (1 or higher).  Equality
+constraints can be imposed by having two parameters have the same
+number in the \code{conpat} vector.  When only fixed values are
 required the \code{fixed} argument can be used instead of
@@
 	\item  the prior model parameters
-
 	\item  the parameters for the transition models
-
 	\item  the response model parameters per state (and subsequently
 	per response in the case of multivariate time series)
-
 \end{enumerate}
@@
 \begin{verbatim}
-mod <- setpars(mod, value=1:npar(mod))mod
+mod <- setpars(mod, value=1:npar(mod))
+mod
 \end{verbatim}
@@
 in the multinomial logistic models for the transition models and the
-initial state probabilities model:
+initial state probabilities model):
 \begin{verbatim}
-mod <- setpars(mod,
-getpars(mod,which="fixed"))mod
+mod <- setpars(mod, getpars(mod,which="fixed"))
+mod
 \end{verbatim}
+
+When fitting constraints it is useful to have good starting values 
+for the parameters and hence we first fit the following model: 
+\begin{verbatim}
+trst=c(0.9,0.1,0,0,0.1,0.9,0,0)
+mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
+	nstates=2,family=list(gaussian(),multinomial()),
+	trstart=trst,inst=c(.999,0.001))
+fm <- fit(mod)
+\end{verbatim}
+After this, we use the fitted values from this model to constrain the 
+regression coefficients on the transition matrix:
+\begin{verbatim}
+# start with fixed and free parameters
+conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
+# constrain the beta's on the transition parameters to be equal
+conpat[6] <- conpat[10] <- 2
+fm <- fit(fm,equal=conpat)
+\end{verbatim}
+Using \code{summary} on the fitted model shows that the regression 
+coefficients are now estimated at the same value of 12.66. 
+Note that these arguments provide the possibility for arbitrary 
+constraints, also between, e.g., a multinomial regression coefficient 
+for the transition matrix and the mean of a gaussian response model. 
+Whether such constraints make sense is hence the responsibility of 
+the user.
 
@@
 
-Introduce the exgaus example.
+The \pkg{depmixS4} package was built with the aim of having the 
+possibility of adding new response distributions (and possibly also 
+other models for the prior and transition probabilities models). 
+Therefore, the EM algorithm simply calls \code{fit} functions from the 
+separate response models without knowing what they are. As a 
+consequence, adding user-specified response models is straightforward. 
+User-defined distributions should extend the \code{response}-class 
+and have the following slots:
+\begin{enumerate}
+	\item y: the response variable
+	\item x: the design matrix, possibly only an intercept
+	\item paramaters: a named list with the coefficients and possibly 
+	other parameters, e.g., the standard deviation in the gaussian 
+	response model
+	\item fixed: a vector of logicals indicating whether parameters are 
+	fixed
+	\item npar: numerical indicating the number of parameters of the model
+\end{enumerate}
 
-Explain the use of makeDepmix for having full control of every aspect of the model.
+In the \code{speed} data example, it may be more appropriate to model
+the response times as an exgaus distribution rather than using the
+gaussian.  We can do so as follows, first we define our own
+exgaus-class extending the response-class and set it to be a generic:
+\begin{verbatim}
+	setClass("exgaus", contains="response")
+	setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
+	standardGeneric("exgaus"))
+\end{verbatim}
+
+The so-defined class now needs a number of methods: 
+\begin{enumerate}
+	\item constructor: function to create instances of the class 
+	with starting values
+	\item show method: to print the model to the terminal
+	\item dens: the function that computes the density of the responses
+	\item getpars and setpars: to get and set parameters 
+	\item predict: to generate predict'ed values 
+	\item fit: function to fit the model using posterior weights (used 
+	by the EM algorithm)
+\end{enumerate}
+
+Only the constructor function and the fit function are provided here
+and the complete code is provided in the helpfile for \code{makeDepmix}. 
+The \code{exgaus} example uses the \pkg{gamlss.distr} package for 
+computing the \code{dens}ity and for fitting the parameters.
+
+The constructor function is defined as: 
+\begin{verbatim}
+setMethod("exgaus", 
+	signature(y="ANY"), 
+	function(y,pstart=NULL,fixed=NULL, ...) {
+		y <- matrix(y,length(y))
+		x <- matrix(1)
+		parameters <- list()
+		npar <- 3
+		if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
+		if(!is.null(pstart)) {
+			if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
+			parameters$mu <- pstart[1]
+			parameters$sigma <- log(pstart[2])
+			parameters$nu <- log(pstart[3])
+		}
+		mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
+		mod
+	}
+)
+\end{verbatim}
+
+The \code{fit} function is defined as follows: 
+\begin{verbatim}
+setMethod("fit","exgaus",
+	function(object,w) {
+		if(missing(w)) w <- NULL
+		y <- object@y
+		fit <- gamlss(y~1,weights=w,family=exGAUS(),
+			control=gamlss.control(n.cyc=100,trace=FALSE),
+			mu.start=object@parameters$mu,
+			sigma.start=exp(object@parameters$sigma),
+			nu.start=exp(object@parameters$nu))
+		pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
+		object <- setpars(object,pars)
+		object
+	}
+)
+The \code{fit} function defines a trivial \code{gamlss} model with 
+only an intercept to be estimated and then sets the fitted parameters 
+back into their respective slots in the `exgaus' object. See the help 
+for \pkg{gamlss.distr} for interpretation of these parameters.
+
+After defining all the necessary methods for the new response model, 
+we can  now define the dependent mixture model using this reponse. 
+The function \code{makeDepmix} is added to \pkg{depmixS4} to have 
+full control over model specification, and we need it here.
+
+We first create all the response models that we want as a double list: 
+\begin{verbatim}
+rModels <- list(
+	list(
+		exgaus(rt,pstart=c(5,.1,.1)),
+		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
+	),
+	list(
+		exgaus(rt,pstart=c(6,.1,.1)),
+		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
+	)
+)
+\end{verbatim}
+Next, we define the transition and prior probability models using the 
+transInit function (which produces a transInit model, which also extends 
+the response class): 
+\begin{verbatim}
+trstart=c(0.9,0.1,0.1,0.9)
+transition <- list()
+transition[[1]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.9,0.1,0,0))
+transition[[2]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.1,0.9,0,0))
+inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
+\end{verbatim}
+Finally, we put everything together using \code{makeDepmix} and fit 
+the model: 
+\begin{verbatim}
+mod <- makeDepmix(response=rModels,transition=transition,
+prior=inMod,stat=FALSE)
+fm <- fit(mod)
+\end{verbatim}
+
+Using \code{summary} will print the fitted parameters. Note that the 
+use of \code{makeDepmix} allows the possibility of, say, fitting a 
+gaussian in one state and an exgaus distribution in another state.
 

### 2009-07-13 papers/jss/article.tex
Updated refs in the jss paper (still have compile problems ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@295 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
+%&LaTeX
 \documentclass[article]{jss}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\usepackage[]{amsmath, amsfonts, amstext, amsthm} 
+%\usepackage{amssymb}
+%\usepackage[]{graphics} 
+%\usepackage{epsfig}
+%\usepackage{epstopdf}
 
@@
 \author{Ingmar Visser\\University of Amsterdam \And 
-        Maarten Speekenbrink\\University College London }
-\title{\pkg{depmixS4}: An \proglang{R}-package for hidden Markov models}
+        Maarten Speekenbrink\\University College London}
+
+\title{\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models}
 
@@
 \Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
+
 \Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting
-\Shorttitle{depmixS4 Hidden Markov Models} %% a short title (if necessary)
+
+\Shorttitle{depmixS4: Hidden Markov Models} %% a short title (if necessary)
 
@@
 \Abstract{ 	
-
-	\pkg{depmixS4} implements a general framework for definining and
+	depmixS4 implements a general framework for definining and
 	fitting hidden Markov mixture model in the R programming language
-	\citep{R2009}.  This includes standard Markov models,
+	R2009.  This includes standard Markov models,
 	latent/hidden Markov models, and latent class and finite mixture
@@
 	Parameter estimation is done through an EM algorithm or by a
-	direct optimization approach using the \pkg{Rdonlp2} optimization
+	direct optimization approach using the Rdonlp2 optimization
 	routine when contraints are imposed on the parameters.  A number
 	of illustrative examples are included.
-
 }
 
-\Keywords{hidden Markov model, dependent mixture model, mixture model, \proglang{R}}
+\Keywords{hidden markov model, dependent mixture model, mixture model}
 
-\Plainkeywords{hidden Markov model, dependent mixture model, mixture model,R} %% without formatting
-
+\Plainkeywords{hidden markov model, dependent mixture model, mixture model} %% without formatting
 %% at least one keyword must be supplied
@@
 Department of Psychology\\
-University of Amsterdam,
+University of Amsterdam\\
 Roetersstraat 15\\
@@
 
-%% for those who use Sweave please include the following line (with % symbols):
-%% need no \usepackage{Sweave.sty}
+\newcommand{\vc}{\mathbf}
+\newcommand{\mat}{\mathbf}
 
@@
 
-\section[]{Introduction}
+% Refs to check: Jansen 2002, Dutilh 2009
+
+\section{Introduction}
 %% Note: If there is markup in \(sub)section, then it has to be escape as above.
@@
 model and the likelihood equations.
-
-\subsection*{Acknowledgements}
-
-Ingmar Visser was supported by an EC Framework 6 grant, project 516542
-(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
-Economic Learning and Social Evolution (ELSE).  Han van der Maas
-provided the speed-accuracy data \cite{Maas2005b} and thereby
-neccessitated implementing models with time-dependent covariates.
-Brenda Jansen provided the balance scale data set \citep{Jansen2001}
-which was the perfect opportunity to test the covariates on the prior
-model parameters.  The examples in the help files use both of these
-data sets.
 
@@
 may depend on covariates as well as the response distributions $f$. 
-See for example \citet{Fruhwirth-Schnatter2006} for an overview of 
+See for example \citet{Fruhwirth2006} for an overview of 
 hidden Markov models with extensions.
@@
 of 4 binary items (correct-incorrect) on a balance scale task 
-\citet{Siegler198?}. The data form a subset of the data published in 
-\citet{Jansen199?}.
+\citet{Siegler1981}. The data form a subset of the data published in 
+\citet{Jansen2002}.
 
@@
 gaussian.  We can do so as follows, first we define our own
-exgaus-class extending the response-class and set it to be a generic:
+exgaus-class extending the response-class:
 \begin{verbatim}
 	setClass("exgaus", contains="response")
-	setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
-	standardGeneric("exgaus"))
 \end{verbatim}
@@
 \begin{verbatim}
+setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
+	standardGeneric("exgaus"))
 setMethod("exgaus", 
@@
 
+\section*{Acknowledgements}
+
+Ingmar Visser was supported by an EC Framework 6 grant, project 516542
+(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
+Economic Learning and Social Evolution (ELSE).  Han van der Maas
+provided the speed-accuracy data \cite{Dutilh2009} and thereby
+neccessitated implementing models with time-dependent covariates.
+Brenda Jansen provided the balance scale data set \citep{Jansen2002}
+which was the perfect opportunity to test the covariates on the prior
+model parameters.  The examples in the help files use both of these
+data sets.
+
 \bibliography{all,ingmar}

### 2009-07-13 papers/jss/article.tex
added EM bit
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@296 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass[article]{jss}
+\usepackage{amsmath}
 
@@
 %\usepackage{epstopdf}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
-
 \Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting
@@
 	depmixS4 implements a general framework for definining and
-	fitting hidden Markov mixture model in the R programming language
+	estimating hidden Markov mixture model in the R programming language
 	R2009.  This includes standard Markov models,
@@
 The latent Markov model is commonly associated with data of this type,
-albeit usually only multinomial variables are considered.  However,
+although usually only multinomial variables are considered.  However,
 common estimation procedures, such as those implemented in
@@
 Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
-through the use of a general Newton-Raphson optimizer.  The EM
-algorithm however has some drawbacks.  First, it can be slow to
+through the use of a general Newton-Raphson optimizer.
+
+In the EM algorithm, parameters are estimated by iteratively maximising the 
+expected joint likelihood of the parameters given the observations and states. 
+Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
+be the general parameter vector consisting of three subvectors with parameters 
+for the prior model, transition model, and response model respectively. The 
+joint log likelihood can be written as
+\begin{equation}
+\log Pr(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log Pr(S_1|\greekv{\theta}_1) 
++ \sum_{t=2}^{T} \log Pr(S_t|S_{t-1},\greekv{\theta}_2) 
++ \sum_{t=1}^{T} \log Pr(O_t|S_t,\greekv{\theta}_3)
+\end{equation}
+This likelihood depends on the unobserved states $S_t$. In the Expectation step,
+we replace these with their expected values given a set of (initial) parameters 
+$\greekv{\theta}' = (\greekv{\theta}'_1, \greekv{\theta}'_2,\greekv{\theta}'_3)$
+and observations $O_{1:T}$. The expected log likelihood 
+\begin{equation}
+Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
+(\log Pr(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
+\end{equation}
+can be written as
+%\begin{equation}
+\begin{multline}
+\label{eq:Q}
+Q(\greekv{\theta},\greekv{\theta}') = 
+\sum_{j=1}^n \gamma_1(j) \log Pr(S_1=j|\greekv{\theta}_1) \\ 
++ \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log Pr(S_t = k|S_{t-1} 
+= j,\greekv{\theta}_2) \\ + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
+\ln Pr(O_t|S_t=j,\greekv{\theta}_3),
+\end{multline}
+%\end{equation}
+where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|O_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|O_{1:T},\greekv{\theta}')$ can be computed effectively by the Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}.
+The Maximisation step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.
+As the r.h.s.
+of (\ref{eq:Q}) consists of three separate parts, we can maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and $\greekv{\theta}_3$.
+In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{nnet}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.
+
+%The maximising values of $a_{jk}$ are \cite<e.g.,>{Rabiner89}
+%\begin{equation}
+%\label{eq:MLA}
+%\hat{a}_{jk} = \frac{1}{N(T-1)} \sum_{i=1}^N \sum_{t=2}^T  \frac{\xi^i_t(j,k)}{\gamma_{t-1}^i(j)}.
+%\end{equation}
+%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does not affect the estimation of the other elements.
+When elements $a_{jk}$ are assumed identical, we simply extend the summation in (\ref{eq:MLA}) to include all those elements, changing the denominator $N(T-1)$ accordingly.
+
+%To estimate $\greekv{\lambda}$, we note that the term containing this parameter has the form of a weighted likelihood, where $\gamma_t^i(j)$ can be interpreted as the number of replications of $r^i_t$.
+Hence, we can rely on standard ML estimates of the logistic regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$ as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To be more specific, we replicate each observation $r^i_t$ a total of 6 times, once for each of the states besides the random state (which offers no information regarding $\lambda_i$).
+For the $j$-th replication (corresponding to the $j$-th state), we used $v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a case weight.
+All these replications were used to obtain the maximum likelihood estimate of $\lambda$ from a single GLM, using the ``glm.fit'' function in R \cite{R}.}.
+The resulting parameter estimates are then used in a new Expectation step, followed by another Maximisation step, until convergence.
+Under mild regularity conditions, it can be shown that the EM algorithm converges to a (local) maximum of the likelihood \cite{EM}.
+
+The EM algorithm however has some drawbacks.  First, it can be slow to
 converge towards the end of optimization (although it is usually
@@
 Need some more on EM and how/why it is justified to do separate weighted
-fits of the response models and transition and prior models.
+iuuuuuuuuuuucvfits of the response models and transition and prior models.
 

### 2009-07-14 papers/jss/article.tex
added EM bit
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@297 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
some fixes for the jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@298 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \Abstract{ 	
-	depmixS4 implements a general framework for definining and
+	\pkg{depmixS4} implements a general framework for definining and
 	estimating hidden Markov mixture model in the R programming language
-	R2009.  This includes standard Markov models,
+	\citep{R2009}.  This includes standard Markov models,
 	latent/hidden Markov models, and latent class and finite mixture
@@
 	Parameter estimation is done through an EM algorithm or by a
-	direct optimization approach using the Rdonlp2 optimization
+	direct optimization approach using the \pkg{Rdonlp2} optimization
 	routine when contraints are imposed on the parameters.  A number
@@
 
-\Keywords{hidden markov model, dependent mixture model, mixture model}
+\Keywords{hidden Markov model, dependent mixture model, mixture model}
 
@@
 \newcommand{\mat}{\mathbf}
+\newcommand{\greekv}[1]{\mbox{\boldmath$\mathrm{#1}$}}
+\newcommand{\greekm}[1]{\mbox{\boldmath$#1$}}
 
@@
 
-Need some more on EM and how/why it is justified to do separate weighted
-iuuuuuuuuuuucvfits of the response models and transition and prior models.
+%Need some more on EM and how/why it is justified to do separate weighted
+%iuuuuuuuuuuucvfits of the response models and transition and prior models.
 
-Also mention use of glm, nnet and possibly other packages that we use.
+%Also mention use of glm, nnet and possibly other packages that we use.
 
-\section{Using \pkg{depmixS4}}
+\section[Using depmixS4]{Using \pkg{depmixS4}}
 
@@
 
-\begin{verbatim} 
-set.seed(1)
-mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+> set.seed(1)
+> mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
+\end{CodeInput}
+\end{CodeChunk}
+
 The \code{depmix} function returns an object of class \code{depmix}
@@
 The so-defined models needs to be \code{fit}ted with the following: 
-\begin{verbatim}
-fm <- fit(mod)
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+> fm <- fit(mod)
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 > fm
+\end{CodeInput}
+\begin{CodeOutput}
 Convergence info: Log likelihood converged to within tol. 
@@
 BIC:  211.275 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 > summary(fm)
+\end{CodeInput}
+\begin{CodeOutput}
 Initial state probabilties model 
@@
 Probalities at zero values of the covariates.
-0.999987 1.291797e-05
+0.999987 1.291798e-05
 
@@
 sd  0.1926063 
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 one-sided formula as in the following example:
-\begin{verbatim}
-set.seed(1)
-mod <- depmix(rt~1, data=speed, nstates=2, 
-family=gaussian(), transition=~scale(Pacc), 
-instart=runif(2))
-fm <- fit(mod)
-\end{verbatim}
-
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+> set.seed(1)
+> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
+  transition=~scale(Pacc), instart=runif(2))
+> fm <- fit(mod)
+\end{CodeInput}
+\begin{CodeOutput}
 Initial state probabilties model 
@@
 sd  0.2402693 
-
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 
@@
 
-\section{Extending \pkg{depmixS4}}
+\section[Extending depmixS4]{Extending \pkg{depmixS4}}
 
@@
 )
+\end{verbatim}
+
 The \code{fit} function defines a trivial \code{gamlss} model with 
@@
 back into their respective slots in the `exgaus' object. See the help 
-for \pkg{gamlss.distr} for interpretation of these parameters.
+for \code{gamlss.distr} for interpretation of these parameters.
 
@@
 
-\section{Conclusion \& future work}
+\section[Conclusion and future work]{Conclusion \& future work}
 

### 2009-07-14 papers/jss/article.tex
Remove pdf (which leads to svn conflicts and cannot be merged).
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@300 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Yet another version of the jss paper, now with conclusions as well.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@301 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \Abstract{
-	\pkg{depmixS4} implements a general framework for definining and
-	estimating hidden Markov mixture model in the R programming language
-	\citep{R2009}.  This includes standard Markov models,
-	latent/hidden Markov models, and latent class and finite mixture
-	distribution models.  The models can be fitted on mixed
-	multivariate data with multinomial and/or gaussian distributions.
+
+	\pkg{depmixS4} implements a general framework for defining and estimating
+	 dependent mixture models in the \proglang{R} programming language
+	\citep{R2009}. This includes standard Markov models, latent/hidden Markov
+	models, and latent class and finite mixture distribution models. The models can
+	be fitted on mixed multivariate data with distributions from the \code{glm} 
+	family, the logistic multinomial, or the multivariate normal distribution. Other 
+	distributions can be added easily, and an example is provided. Parameter 
+	estimation is done through an EM algorithm or by a direct optimization approach using
+	the\pkg{Rdonlp2} optimization routine when contraints are imposed on the parameters. 
 	Parameters can be estimated subject to general linear constraints.  
-	Parameter estimation is done through an EM algorithm or by a
-	direct optimization approach using the \pkg{Rdonlp2} optimization
-	routine when contraints are imposed on the parameters.  A number
-	of illustrative examples are included.
 }
@@
 those latter areas of application, latent Markov models are usually
-referred to as hidden Markov models.
+referred to as hidden Markov models. See e.g.\ \citet[][chapter~1]{Cappe2005}
+for an overview of different applications.
 
@@
 following goals:
+
 \begin{enumerate}
@@
 is the possibility to include covariates on the prior probabilities of
-class membership.  In the next section, an outline is provided of the
-model and the likelihood equations.
+class membership and the transition probabilities. In the next section, 
+an outline is provided of the model and the likelihood equations.
 
@@
 an example, consider a time series of responses generated by a single
-subject in a reaction time experiment.  The data consists of three
-variables, reaction time, accuracy and a covariate which is a pay-off
-factor which determines the reward for speed and accuracy.  These
+participant in a response time experiment.  The data consists of three
+variables, response time, accuracy and a covariate which is a pay-off
+determining the relative reward for speeded and/or accuracy responding. These
 variables are measured on 168, 134 and 137 occasions respectively (in
-Figure~\ref{fig:speed} the first part of this series is plotted).
+Figure~\ref{fig:speed} the first part of this series is plotted). These data
+are more fully described in \citet{Dutilh2009}.
 
@@
 	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
-	  \caption{Response times, accuracy and pay-off values for
-	  the first series of responses in dataset \texttt{speed}.}
+	  \caption{Response times (rt), accuracy (corr) and pay-off values (Pacc) for
+	  the first series of responses in dataset \code{speed}.}
 	  \label{fig:speed}
@@
 The latent Markov model is commonly associated with data of this type,
-although usually only multinomial variables are considered.  However,
+although usually only multinomial response variables are considered.  However,
 common estimation procedures, such as those implemented in
@@
 problems.  In contrast, the hidden Markov model is typically only used
-for `long' univariate time series.  In the next sections, the
-likelihood and estimation procedure for the hidden Markov model is
-described, given data of the above form.  These models are called
+for `long' univariate time series 
+\cite[see e.g.][, chapter~1 for an overview of examples]{Cappe2005}.  
+In the next sections, the
+likelihood and estimation procedures for the dependent mixture model is
+described for data of the above form.  These models are called
 dependent mixture models because one of the authors (Ingmar Visser)
@@
 dependent mixture models in an application with hidden Markov mixtures
-of Poisson count data.}
+of Poisson count data.}.
 
@@
 
-	\item matrices $\mat{A}_t$ of transition probabilities $a_{ij,t}$ for
-	the transition from state $S_{i}$ to state $S_{j}$ at time $t$,
+	\item matrices $\mat{A}$ of transition probabilities $a_{ij}$ for
+	the transition from state $S_{i}$ to state $S_{j}$,
 
-	\item a set $\vc{B}_t$ of observation functions $b_j^k(\cdot)$ that
-	provide the conditional probabilities of observations $O_{t}^k$ 
+	\item a set $\vc{B}$ of observation density functions $b_j^k(\cdot)$ that
+	provide the conditional densities of observations $O_{t}^k$ 
 	associated with latent state $S_{j}$,
@@
 \end{enumerate}
-When transitions are added to the latent class model, it is more
-appropriate to refer to the classes as states.  The word class is
-rather more associated with a stable trait-like attribute whereas a
-state can change over time.
 
-The dependent mixture model is then given by the following equations: 
+The dependent mixture model is defined by the following equations: 
 \begin{align}
-	S_{t} = A S_{t-1}, t=2, \ldots, T \\
-	O_{t} = f(O_{t}|S_{t}),
+	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
+	O_{t} &= b(O_{t}|S_{t}), 
 \end{align}
 where $S_{t}$ is a sequence of hidden states, $A$ is a transition
-matrix, $O_{t}$ is an (possibly multivariate) observation and $f$ is a
+matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
 density function for $O_{t}$ conditional on the hidden state $S_{t}$.
-In the example data above, $f$ could be a gaussian distribution
+In the example data above, $b$ could be a gaussian distribution
 function for the response time variable, and a Bernouilli for the
 accuracy data.  In the models we are considering here, both the
-transition probabilities $A$ and the initial state probabilities $\pi$
-may depend on covariates as well as the response distributions $f$. 
+transition probabilities $\mat{A}$ and the initial state probabilities $\pi$
+may depend on covariates as well as the response distributions $\vc{B}$. 
 See for example \citet{Fruhwirth2006} for an overview of 
@@
 Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
-through the use of a general Newton-Raphson optimizer.
-
-In the EM algorithm, parameters are estimated by iteratively maximising the 
+through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
+parameters are estimated by iteratively maximising the 
 expected joint likelihood of the parameters given the observations and states. 
@@
 \begin{equation}
-\log Pr(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log Pr(S_1|\greekv{\theta}_1) 
-+ \sum_{t=2}^{T} \log Pr(S_t|S_{t-1},\greekv{\theta}_2) 
-+ \sum_{t=1}^{T} \log Pr(O_t|S_t,\greekv{\theta}_3)
+\log \Prob(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log \Prob(S_1|\greekv{\theta}_1) 
++ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\greekv{\theta}_2) 
++ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\greekv{\theta}_3)
 \end{equation}
@@
 Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
-(\log Pr(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
+(\log \Prob(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
 \end{equation}
@@
 Q(\greekv{\theta},\greekv{\theta}') = 
-\sum_{j=1}^n \gamma_1(j) \log Pr(S_1=j|\greekv{\theta}_1) \\ 
-+ \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log Pr(S_t = k|S_{t-1} 
+\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\greekv{\theta}_1) \\ 
++ \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
 = j,\greekv{\theta}_2) \\ + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
-\ln Pr(O_t|S_t=j,\greekv{\theta}_3),
+\ln \Prob(O_t|S_t=j,\greekv{\theta}_3),
 \end{multline}
@@
 The EM algorithm however has some drawbacks.  First, it can be slow to
-converge towards the end of optimization (although it is usually
-faster than direct optimization at the start, so possibly a
-combination of EM and direct optimization is fastest).  Second,
+converge towards the end of optimization.  Second,
 applying constraints to parameters can be problematic; in particular,
@@
 models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
-\cite{Tamura2007,Spellucci2002}, because it handles general linear
+\cite{Tamura2009,Spellucci2002}, because it handles general linear
 (in)equality constraints, and optionally also non-linear constraints.
@@
 \begin{enumerate}
-	\item model specification with function \code{depmix}
+	\item model specification with function \code{depmix} (or with \code{mix}
+	for latent class and finite mixture models, see example below on adding
+	covariates to prior probabilities)
 
@@
 \end{enumerate}
+We have separated the stages of model specification and model fitting because
+fitting large models can be fairly time-consuming and it is hence useful to be
+able to check the model specification before actually fitting the model.
 
@@
 
-\subsection{A simple example}
+\subsection{A simple model}
 
@@
 
-The so-defined models needs to be \code{fit}ted with the following: 
+The so-defined models needs to be \code{fit}ted with the following
+line of code:
 \begin{CodeChunk}
@@
 information on convergence, the log likelihood and the AIC and BIC
-values.  These statistics may be extracted using \code{logLik},
-\code{AIC} and \code{BIC}, respectively.
-
+values:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
+These statistics may be extracted using \code{logLik},
+\code{AIC} and \code{BIC}, respectively.
 
@@
 
-\subsection{Transition parameters}
+\subsection{Covariates on transition parameters}
 
-The transition matrix is parametrized as a list of multinomial
-logistic models.  The initial state probabilities are similarly
-parametrized as a multinomial logistic model.  Both models use a
-baseline category parametrization, meaning that the parameter for the
+The transition probabilities and the initial state probabilities
+are parameterized using the multinomial
+logistic model.   In particular, each row of the transition matrix is
+parameterized by a baseline category logistic multinomial, 
+meaning that the parameter for the
 base category is fixed at zero (see \citet[see][p.\ 267
 ff.]{Agresti2002} for multinomial logistic models and various
-parameterizations).  The default base category is the first state.
+parameterizations). See also \citet{Chung2007} for similar models, latent transition models using logistic regression on the transition parameters. They fit such models on repeated measurement
+data ($T=2$) using Bayesian methods.  The default baseline category is the first state.
 Hence, for example, for a 3-state model, the initial state probability
@@
 \begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
+set.seed(1)
+mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
   transition=~scale(Pacc), instart=runif(2))
-> fm <- fit(mod)
+fm <- fit(mod)
 \end{CodeInput}
+\end{CodeChunk}
+
+Applying \code{summary} to the \code{fit}ted object gives (only transition models
+printed here): 
+\begin{CodeChunk}
 \begin{CodeOutput}
-Initial state probabilties model 
-Model of type multinomial, formula: ~1
-Coefficients: 
-	 [,1]     [,2]
-[1,]    0 10.71779
-Probalities at zero values of the covariates.
-2.214681e-05 0.9999779
-
+...
 Transition model for state (component) 1 
@@
 0.07788458 0.9221154
-
-Response model(s) for state 1
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 5.512179
-sd  0.1921098
-
-Response model(s) for state 2
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 6.3885
-sd  0.2402693 
+...
 \end{CodeOutput}
 \end{CodeChunk}
-
 The summary provides all parameters of the model, also the 
 (redundant) zeroes for the base-line category in the multinomial model. 
-The summary also prints the transition probabilities at the zero value 
-of the covariate. Note that scaling of the covariate is useful in this 
-regard as it makes interpretation of these intercept probabilities 
+The summary also prints the transition probabilities
+at the zero value of the covariate. Note that scaling of the covariate 
+is useful in this regard as it makes interpretation of these intercept probabilities 
 easier.
@@
 mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
-	family=list(gaussian(),multinomial()), transition=~scale(Pacc), 
-	instart=runif(2))
+	family=list(gaussian(),multinomial()),
+	 transition=~scale(Pacc),instart=runif(2))
 fm <- fit(mod)
@@
 	> summary(fm)
-	\ldots 
+	...
 	Response model(s) for state 1
@@
 
-The \pkg{depmixS4} contains another data set which is used to
-illustrate the use of covariates on the prior probabilities of models,
-in this case a latent class model. The \code{balance} data consists 
+To illustrate the use of covariates on the prior probabilities we have included
+another data set with \pkg{depmixS4}. The \code{balance} data consists 
 of 4 binary items (correct-incorrect) on a balance scale task 
@@
 0.9254119 0.07458815 
+...
 \end{verbatim}	
@@
 
-Using package \pkg{Rdonlp2}, parameters may be fitted subject to
+Using package \pkg{Rdonlp2} by \citet{Tamura2009}, parameters may be fitted subject to
 general linear (in-)equality constraints.  Constraining and fixing
 parameters is done using the \code{conpat} argument to the
-\code{depmix.fit}-function, which specifies for each parameter in the
+\code{fit}-function, which specifies for each parameter in the
 model whether it's fixed (0) or free (1 or higher).  Equality
@@
 number in the \code{conpat} vector.  When only fixed values are
-required the \code{fixed} argument can be used instead of
+required, the \code{fixed} argument can be used instead of
 \code{conpat}, with zeroes for fixed parameters and other values (ones
@@
 constraints is handled by the optimization routine \code{donlp2}.
+To be able to construct the \code{conpat} and/or \code{fixed} vectors 
+one needs the correct ordering of parameters which is briefly discussed 
+next before proceeding with an example.
 
@@
 i.e., these vectors should have length of the number of parameters of
-the model, which can be obtained by calling \code{npar(object)}.
+the model, which can be obtained by calling \code{npar(object)}. Note that
+this is not the same as the degrees of freedom used e.g.\ in the \code{logLik}
+function because \code{npar} also counts the baseline category zeroes
+from the multinomial logistic models. 
 Parameters are numbered in the following order:
@@
 When fitting constraints it is useful to have good starting values 
-for the parameters and hence we first fit the following model: 
+for the parameters and hence we first fit the following model without
+constraints:
 \begin{verbatim}
@@
 	trstart=trst,inst=c(.999,0.001))
-fm <- fit(mod)
+fm1 <- fit(mod)
 \end{verbatim}
 After this, we use the fitted values from this model to constrain the 
-regression coefficients on the transition matrix:
+regression coefficients on the transition matrix (parameters numbers~6 and~10):
 \begin{verbatim}
@@
 conpat[6] <- conpat[10] <- 2
-fm <- fit(fm,equal=conpat)
+fm2 <- fit(fm,equal=conpat)
 \end{verbatim}
 Using \code{summary} on the fitted model shows that the regression 
-coefficients are now estimated at the same value of 12.66. 
-Note that these arguments provide the possibility for arbitrary 
+coefficients are now estimated at the same value of 12.66. The function
+\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
+associated $p$-value with appropriate degrees of freedom for testing the
+tenability of constraints \citep{Dannemann2007}. Note that these arguments 
+provide the possibility for arbitrary 
 constraints, also between, e.g., a multinomial regression coefficient 
@@
 and the complete code is provided in the helpfile for \code{makeDepmix}. 
-The \code{exgaus} example uses the \pkg{gamlss.distr} package for 
-computing the \code{dens}ity and for fitting the parameters.
+The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} packages \cite{Stasinopoulos2009a,Stasinopoulos2009b} for 
+computing the \code{dens}ity and for \code{fit}ting the parameters.
 
@@
 
-The \code{fit} function defines a trivial \code{gamlss} model with 
+The \code{fit} function defines a \code{gamlss} model with 
 only an intercept to be estimated and then sets the fitted parameters 
@@
 
-\section[Conclusion and future work]{Conclusion \& future work}
+\section[Conclusions and future work]{Conclusions \& future work}
 
-What are our next plans?
+\pkg{depmixS4} provides a flexible framework for fitting dependent mixture models
+for a large variety of response distributions. It can also fit latent class regression
+and finite mixture models, although it should be noted that more specialized packages 
+are available for doing so, e.g.\ \pkg{FlexMix} \cite{Leisch2004}. The package is intended
+for modeling of (individual) time series data with the aim of characterizing the transition 
+processes underlying the data. The possibility to use covariates on the transition matrix
+greatly enhances the flexibility of the model. The EM algorithm uses a very general 
+interface that allows easy addition of new response models.
 
-Adding gradients for speed and computation of the Hessian.
+We are currently working on implementing the gradients for response and transition models 
+with two goals in mind. First, to speed up (constrained) parameter optimization using 
+\pkg{Rdonlp2}. Second, analytic gradients are useful in computing the Hessian of the
+estimated parameters so as to arrive at standard errors for those. We are also planning to
+implement goodness-of-fit statistics \citep{Titman2008}.
 

### 2009-07-14 papers/jss/article.tex
minor stuff, typo's and the like
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@303 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	estimation is done through an EM algorithm or by a direct optimization approach using
-	the\pkg{Rdonlp2} optimization routine when contraints are imposed on the parameters. 
+	the \pkg{Rdonlp2} optimization routine when contraints are imposed on the parameters. 
 	Parameters can be estimated subject to general linear constraints.  
@@
 for `long' univariate time series 
-\cite[see e.g.][, chapter~1 for an overview of examples]{Cappe2005}.  
+\citep[][, chapter~1]{Cappe2005}.  
 In the next sections, the
@@
 + \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
-= j,\greekv{\theta}_2) \\ + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
+= j,\greekv{\theta}_2)  \\
+ + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
 \ln \Prob(O_t|S_t=j,\greekv{\theta}_3),

### 2009-07-15 papers/jss/article.tex
some changes to the jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@305 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \Abstract{
-
 	\pkg{depmixS4} implements a general framework for defining and estimating
@@
 	\citep{R2009}. This includes standard Markov models, latent/hidden Markov
-	models, and latent class and finite mixture distribution models. The models can
-	be fitted on mixed multivariate data with distributions from the \code{glm} 
-	family, the logistic multinomial, or the multivariate normal distribution. Other 
-	distributions can be added easily, and an example is provided. Parameter 
-	estimation is done through an EM algorithm or by a direct optimization approach using
-	the \pkg{Rdonlp2} optimization routine when contraints are imposed on the parameters. 
-	Parameters can be estimated subject to general linear constraints.  
+	models, and latent class and finite mixture distribution models. The models 
+	can be fitted on mixed multivariate data with distributions from the 
+	\code{glm} family, the logistic multinomial, or the multivariate normal 
+	distribution. Other distributions can be added easily, and an example is 
+	provided. Parameter are estimated by the EM algorithm or, when (linear) 
+	constraints are imposed on the parameters, by direct numerical optimization 
+	with the \pkg{Rdonlp2} routine.
 }
@@
 
-\Plainkeywords{hidden markov model, dependent mixture model, mixture model} %% without formatting
+\Plainkeywords{hidden Markov model, dependent mixture model, mixture model} %% without formatting
 %% at least one keyword must be supplied
@@
 for an overview, and \citet{Schmittmann2006} for a recent application.
-In economics, latent Markov models are commonly used as regime
-switching models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.
+In economics, common latent Markov models are so-called regime
+switching models (see e.g., \citealp{Kim1994} and \citealp{Ghysels1994}).
 Further applications include speech recognition \citep{Rabiner1989},
@@
 those latter areas of application, latent Markov models are usually
-referred to as hidden Markov models. See e.g.\ \citet[][chapter~1]{Cappe2005}
-for an overview of different applications.
+referred to as hidden Markov models. For more examples of applications, 
+see e.g.\ \citet[][chapter~1]{Cappe2005}.
 
-The \pkg{depmixS4} package was motivated by the fact that Markov
-models are used commonly in the social sciences, but no comprehensive
-package was available for fitting such models.  Common programs for
-Markovian models include Panmark \citep{Pol1996}, and for latent class
+The \pkg{depmixS4} package was motivated by the fact that while Markov
+models are used commonly in the social sciences, no comprehensive
+package was available for fitting such models.  Common software for 
+estimating Markovian models include Panmark \citep{Pol1996}, and for latent class
 models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
 number of important features, besides not being freely available.
-There are currently some packages in R that handle hidden Markov
+There are currently some packages in \proglang{R} that handle hidden Markov
 models but they lack a number of features that we needed in our
@@
 
-	\item to be able to handle parameter estimates subject to general
+	\item to be able to estimate parameters subject to general
 	linear (in)equality constraints
@@
 
-	\item to allow for easy extensibility, in particular, to be able
-	to add new response distributions, both univariate and
-	multivariate, and similarly to be able to allow for the addition
-	of other transition models, e.g., continuous time observation
-	models
+	\item to be easily extensible, in particular, to allow users to
+  easily add new uni- or multivariate response distributions, and similarly 
+  add of other transition models, e.g., continuous time observation
+	models.
 
@@
 
-Although \pkg{depmixS4} is designed to deal with longitudinal or time
-series data, for say $T>100$, it can also handle the limit case with
-$T=1$.  In those cases, there are no time dependencies between
-observed data, and the model reduces to a finite mixture model, or a
+Although \pkg{depmixS4} was designed to deal with longitudinal or time
+series data, for say $T>100$, it can also handle the limit case when
+$T=1$.  In this case, there are no time dependencies between
+observed data and the model reduces to a finite mixture or
 latent class model. Although there are other specialized packages to
-deal with mixture data, one specific feature that we needed ourselves
-which is to the best of our knowledge not available in other packages
-is the possibility to include covariates on the prior probabilities of
-class membership and the transition probabilities. In the next section, 
-an outline is provided of the model and the likelihood equations.
+deal with mixture data, as far as we know these 
+don't allow the inclusion of covariates on the prior probabilities of
+class membership. The possibility to estimating the effects of covariates on 
+prior and transition probabilities is a distinguishing feature of 
+\pkg{depmixS4}. In the next section, an outline is provided of the model and 
+the likelihood equations.
 
@@
 
-The data considered here, has the general form $O_{1}^{1}, \ldots,
+The data considered here have the general form $O_{1}^{1}, \ldots,
 O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
@@
 participant in a response time experiment. The data consists of three
-variables, response time, accuracy and a covariate which is a pay-off
-determining the relative reward for speeded and/or accuracy responding. These
-variables are measured on 168, 134 and 137 occasions respectively (in
+variables, response time, response accuracy, and a covariate which is a pay-off
+variable reflecting the relative reward for speeded and/or accuracy responding. 
+These variables are measured on 168, 134 and 137 occasions respectively (in
 Figure~\ref{fig:speed} the first part of this series is plotted). These data
@@
 common estimation procedures, such as those implemented in
-\citet{Pol1996} are not suitable for long time series due to underflow
+\citet{Pol1996}, are not suitable for long time series due to underflow
 problems.  In contrast, the hidden Markov model is typically only used
@@
 In the next sections, the
-likelihood and estimation procedures for the dependent mixture model is
-described for data of the above form.  These models are called
-dependent mixture models because one of the authors (Ingmar Visser)
+likelihood equation and estimation procedures for the dependent mixture model 
+are described for data of the above form. We use the term
+``dependent mixture model'' because one of the authors (Ingmar Visser)
 thought it was time for a new name for these models\footnote{Only
@@
 
-	\item a set $\vc{S}$ of latent classes or states $S_{i},\, i=1,
+	\item a set $\mathcal{S}$ of latent classes or states $S_{i},\, i=1,
 	\ldots , n$,
 
-	\item matrices $\mat{A}$ of transition probabilities $a_{ij}$ for
-	the transition from state $S_{i}$ to state $S_{j}$,
+	%\item a set of matrices $\mat{A}_t$ of transition probabilities $a_{ijt} = P_t(S_j|S_i)$ 
+	%for the transition from state $S_{i}$ to state $S_{j}$ at time $t$,
 
-	\item a set $\vc{B}$ of observation density functions $b_j^k(\cdot)$ that
+	\item a set $\mathcal{A}$ of transition probability functions $a_{ij}(\vc{x}_t) = 
+	Pr(S_t=j|S_{t-1}=i,\vc{x}_t)$, giving the probability of a transition from state 
+	$i$ to state $j$ with covariate $\vc{x}_t$,
+
+	\item a set $\mathcal{B}$ of observation density functions $b_j^k(\vc{x}_t) = p(O_{t}^k|S_t = j, \vc{x}_t)$ that
 	provide the conditional densities of observations $O_{t}^k$ 
-	associated with latent state $S_{j}$,
+	associated with latent class/state $j$ and covariate $\vc{x}_t$,
 
-	\item a vector $\pmb{\pi}$ of latent state initial probabilities
-	$\pi_{i}$
+	\item a set $\mathcal{P}$ of latent class/state initial probability functions
+	$\pi_{i}(\vc{x}_1) = Pr(S_1 = i|\vc{x}_1)$, giving the probability of class/state
+	$i$ at time $t=1$ with covariate $\vc{x}_1$.
+
 \end{enumerate}
@@
 The dependent mixture model is defined by the following equations: 
+%\begin{align}
+%	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
+%	O_{t} &= b(O_{t}|S_{t}), 
+%\end{align}
 \begin{align}
-	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
-	O_{t} &= b(O_{t}|S_{t}), 
+	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
+	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
 \end{align}
-where $S_{t}$ is a sequence of hidden states, $A$ is a transition
-matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
-density function for $O_{t}$ conditional on the hidden state $S_{t}$.
-In the example data above, $b$ could be a gaussian distribution
-function for the response time variable, and a Bernouilli for the
+%where $S_{t}$ is a sequence of hidden states, $A$ is a transition
+%matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
+%density function for $O_{t}$ conditional on the hidden state $S_{t}$.
+In the example data above, $b_j^k$ could be a Gaussian distribution
+function for the response time variable, and a Bernoulli for the
 accuracy data.  In the models we are considering here, both the
@@
 \sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\greekv{\theta}_1) \\ 
-+ \sum_{t=2}^T \sum_{j=1}^M \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
++ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
 = j,\greekv{\theta}_2)  \\
- + \sum_{t=1}^T \sum_{j=1}^n \gamma_t^i(j) 
-\ln \Prob(O_t|S_t=j,\greekv{\theta}_3),
+ + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
+\ln \Prob(O^k_t|S_t=j,\greekv{\theta}_3),
 \end{multline}
@@
 of (\ref{eq:Q}) consists of three separate parts, we can maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and $\greekv{\theta}_3$.
-In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{nnet}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.
+In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{MASS}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.
 
@@
 \begin{CodeInput}
-set.seed(1)
-mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
+> set.seed(1)
+> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
   transition=~scale(Pacc), instart=runif(2))
-fm <- fit(mod)
+> fm <- fit(mod)
 \end{CodeInput}
@@
 responses. In above examples we have only used the response times 
-which were modelled with the gaussian distribution. The accuracy data 
+which were modelled with the Gaussian distribution. The accuracy data 
 are in the \code{speed} data are modelled with a multinomial by 
 specifying the following: 
-\begin{verbatim}
-set.seed(1)
-mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
+\begin{CodeChunk}
+\begin{CodeInput}
+> set.seed(1)
+> mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
 	  family=list(gaussian(),multinomial()),
 	  transition=~scale(Pacc),instart=runif(2))
-fm <- fit(mod)
-\end{verbatim}
+> fm <- fit(mod)
+\end{CodeInput}
+\end{CodeChunk}
 which provides the following fitted model parameters (only the 
 response parameters are given here): 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 > summary(fm)
+\end{CodeInput}
+\begin{CodeOutput}
 ...
@@
 0.09573715 0.9042629 	
-\end{verbatim}
+\end{CodeOutput}
+\end{CodeChunk}
 As can be seen, state 1 has fast response times around chance level, 
@@
 levels.
+
+Note that by specifying multivariate observations in terms of a list, the 
+variables are considered conditionally independent (given the states). 
+Conditionally \emph{dependent} variables must be handled as a single element in
+the list. Effectively, this means specifying a multivariate response model. The 
+only multivariate response model currently implemented in \pkg{depmix} is for
+multivariate normal variables.
 
@@
 defined by using a one-sided formula: 
-\begin{verbatim}
-balance$age <- balance$age-5
-set.seed(1)
-mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
+\begin{CodeChunk}
+\begin{CodeInput}
+> balance$age <- balance$age-5
+> set.seed(1)
+> mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
     family=list(multinomial(), multinomial(), multinomial(),
@@
     prior=~age, initdata=balance)
-fm <- fit(mod)
-\end{verbatim}
+> fm <- fit(mod)
+\end{CodeInput}
+\end{CodeChunk}
 Note here that we define a \code{mix} model instead of a \code{depmix}
@@
 shown here): 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 > summary(fm)
+\end{CodeInput}
+\begin{CodeOutput}
 Mixture probabilities model 
@@
 ...
-\end{verbatim}	
+\end{CodeOutput}
+\end{CodeChunk}	
 Hence at young ages, children have a high probability of incorrect 
@@
 To see the ordering of parameters use the following:
-\begin{verbatim}
-mod <- setpars(mod, value=1:npar(mod))
-mod
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+> mod <- setpars(mod, value=1:npar(mod))
+> mod
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 initial state probabilities model):
-\begin{verbatim}
-mod <- setpars(mod, getpars(mod,which="fixed"))
-mod
-\end{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
+> mod <- setpars(mod, getpars(mod,which="fixed"))
+> mod
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 constraints:
-\begin{verbatim}
-trst=c(0.9,0.1,0,0,0.1,0.9,0,0)
-mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
+\begin{CodeChunk}
+\begin{CodeInput}
+> trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
+> mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
 	  nstates=2,family=list(gaussian(),multinomial()),
 	  trstart=trst,inst=c(.999,0.001))
-fm1 <- fit(mod)
-\end{verbatim}
+> fm1 <- fit(mod)
+\end{CodeInput}
+\end{CodeChunk}
 After this, we use the fitted values from this model to constrain the 
 regression coefficients on the transition matrix (parameters numbers~6 and~10):
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 # start with fixed and free parameters
-conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
+> conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
 # constrain the beta's on the transition parameters to be equal
-conpat[6] <- conpat[10] <- 2
-fm2 <- fit(fm,equal=conpat)
-\end{verbatim}
+> conpat[6] <- conpat[10] <- 2
+> fm2 <- fit(fm,equal=conpat)
+\end{CodeInput}
+\end{CodeChunk}
 Using \code{summary} on the fitted model shows that the regression 
@@
 the response times as an exgaus distribution rather than using the
-gaussian.  We can do so as follows, first we define our own
+Gaussian.  We can do so as follows, first we define our own
 exgaus-class extending the response-class:
-\begin{verbatim}
+\begin{CodeChunk}
 	setClass("exgaus", contains="response")
-\end{verbatim}
+\end{CodeChunk}
 
@@
 The constructor function is defined as: 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
@@
 )
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 
 The \code{fit} function is defined as follows: 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 setMethod("fit","exgaus",
@@
 )
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 We first create all the response models that we want as a double list: 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 rModels <- list(
@@
 )
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 Next, we define the transition and prior probability models using the 
@@
 the response class): 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 trstart=c(0.9,0.1,0.1,0.9)
@@
 inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 Finally, we put everything together using \code{makeDepmix} and fit 
 the model: 
-\begin{verbatim}
+\begin{CodeChunk}
+\begin{CodeInput}
 mod <- makeDepmix(response=rModels,transition=transition,
@@
 fm <- fit(mod)
-\end{verbatim}
+\end{CodeInput}
+\end{CodeChunk}
 
@@
 use of \code{makeDepmix} allows the possibility of, say, fitting a 
-gaussian in one state and an exgaus distribution in another state.
+Gaussian in one state and an exgaus distribution in another state.
 
@@
 and finite mixture models, although it should be noted that more specialized packages 
-are available for doing so, e.g.\ \pkg{FlexMix} \cite{Leisch2004}. The package is intended
+are available for doing so, e.g.\ \pkg{FlexMix} \citep{Leisch2004}. The package is intended
 for modeling of (individual) time series data with the aim of characterizing the transition 

### 2009-08-05 papers/jss/article.tex
Minor changes to jss paper, bib update.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@306 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Notation issues ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@307 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \Abstract{
-	\pkg{depmixS4} implements a general framework for defining and estimating
-	 dependent mixture models in the \proglang{R} programming language
-	\citep{R2009}. This includes standard Markov models, latent/hidden Markov
-	models, and latent class and finite mixture distribution models. The models 
-	can be fitted on mixed multivariate data with distributions from the 
-	\code{glm} family, the logistic multinomial, or the multivariate normal 
-	distribution. Other distributions can be added easily, and an example is 
-	provided. Parameter are estimated by the EM algorithm or, when (linear) 
-	constraints are imposed on the parameters, by direct numerical optimization 
-	with the \pkg{Rdonlp2} routine.
-}
+
+	\pkg{depmixS4} implements a general framework for defining and
+	estimating dependent mixture models in the \proglang{R}
+	programming language \citep{R2009}.  This includes standard Markov
+	models, latent/hidden Markov models, and latent class and finite
+	mixture distribution models.  The models can be fitted on mixed
+	multivariate data with distributions from the \code{glm} family,
+	the logistic multinomial, or the multivariate normal distribution.
+	Other distributions can be added easily, and an example is
+	provided with the exgaus distribution.  Parameter are estimated by
+	the EM algorithm or, when (linear) constraints are imposed on the
+	parameters, by direct numerical optimization with the
+	\pkg{Rdonlp2} routine.  }
 
@@
 
+\begin{center}
+\bf{DRAFT: DO NOT QUOTE WITHOUT CONTACTING AUTHOR}
+\end{center}
+
 \section{Introduction}
@@
 are used for modelling learning processes, see \citet{Wickens1982},
-for an overview, and \citet{Schmittmann2006} for a recent application.
-In economics, common latent Markov models are so-called regime
-switching models (see e.g., \citealp{Kim1994} and \citealp{Ghysels1994}).
-Further applications include speech recognition \citep{Rabiner1989},
-EEG analysis \citep{Rainer2000}, and genetics \citep{Krogh1998}.  In
-those latter areas of application, latent Markov models are usually
-referred to as hidden Markov models. For more examples of applications, 
-see e.g.\ \citet[][chapter~1]{Cappe2005}.
+for an overview, and e.g.\ \citet{Schmittmann2006} for a recent
+application.  In economics, latent Markov models are so-called regime
+switching models (see e.g., \citealp{Kim1994} and
+\citealp{Ghysels1994}).  Further applications include speech
+recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
+genetics \citep{Krogh1998}.  In those latter areas of application,
+latent Markov models are usually referred to as hidden Markov models.
+For more examples of applications, see e.g.\
+\citet[][chapter~1]{Cappe2005}.
 
@@
 package was available for fitting such models.  Common software for
-estimating Markovian models include Panmark \citep{Pol1996}, and for latent class
-models Latent Gold \citep{Vermunt2003}.  Those programs are lacking a
-number of important features, besides not being freely available.
-There are currently some packages in \proglang{R} that handle hidden Markov
-models but they lack a number of features that we needed in our
-research. In particular, \pkg{depmixS4} was designed to meet the
-following goals:
+estimating Markovian models include Panmark \citep{Pol1996}, and for
+latent class models Latent Gold \citep{Vermunt2003}.  Those programs
+are lacking a number of important features, besides not being freely
+available.  There are currently some packages in \proglang{R} that
+handle hidden Markov models but they lack a number of features that we
+needed in our research.  In particular, \pkg{depmixS4} was designed to
+meet the following goals:
 
@@
 	\item to be easily extensible, in particular, to allow users to
-  easily add new uni- or multivariate response distributions, and similarly 
-  add of other transition models, e.g., continuous time observation
-	models.
+	easily add new uni- or multivariate response distributions, and
+	similarly for the addition of other transition models, e.g.,
+	continuous time observation models
 
@@
 
-The data considered here have the general form $O_{1}^{1}, \ldots,
+The data considered here have the general form $\vc{O}_{1:T}=O_{1}^{1}, \ldots,
 O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
@@
 The latent Markov model is commonly associated with data of this type,
-although usually only multinomial response variables are considered.  However, 
-common estimation procedures, such as those implemented in
-\citet{Pol1996}, are not suitable for long time series due to underflow
-problems.  In contrast, the hidden Markov model is typically only used
-for `long' univariate time series 
-\citep[][, chapter~1]{Cappe2005}.  
-In the next sections, the
-likelihood equation and estimation procedures for the dependent mixture model 
-are described for data of the above form. We use the term
-``dependent mixture model'' because one of the authors (Ingmar Visser)
-thought it was time for a new name for these models\footnote{Only
-later did I find out that \citet{Leroux1992} already coined the term
-dependent mixture models in an application with hidden Markov mixtures
-of Poisson count data.}.
+although usually only multinomial response variables are considered.
+However, common estimation procedures, such as those implemented in
+\citet{Pol1996}, are not suitable for long time series due to
+underflow problems.  In contrast, the hidden Markov model is typically
+only used for `long' univariate time series \citep[][,
+chapter~1]{Cappe2005}.  In the next paragraphs, the likelihood
+equation and estimation procedures for the dependent mixture model are
+described for data of the above form.  We use the term ``dependent
+mixture model'' because one of the authors (Ingmar Visser) thought it
+was time for a new name for these models\footnote{Only later did I
+find out that \citet{Leroux1992} already coined the term dependent
+mixture models in an application with hidden Markov mixtures of
+Poisson count data.}.
 
-The dependent mixture model is defined by the following elements:
+The likelihood of the dependent mixture model conditional on the
+unknown (or hidden) state sequence $S_{1:T}$ and the model is given
+by:
+\begin{equation}
+	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
+	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
+\end{equation}
+where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
+at the likelihood of the data given the parameter vector
+$\greekv{\theta}$, i.e.\ without the state sequence, we need to sum
+above likelihood over all possible state sequences. This likelihood 
+is written as: 
+\begin{equation}
+	\Prob(\vc{O}_{1:T}|\vc{S}_{1:T},\greekv{\theta}) 
+	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
+	\prod_{t=1}^{T-1} a_{ij}(z_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
+\end{equation}
+where we have the following elements:
 \begin{enumerate}
 
-	\item a set $\mathcal{S}$ of latent classes or states $S_{i},\, i=1,
-	\ldots , n$,
+	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
+	of $n$ latent classes or states; we write for short $S_{i}$ to
+	denote $S_{t}=i$.
 
-	%\item a set of matrices $\mat{A}_t$ of transition probabilities $a_{ijt} = P_t(S_j|S_i)$ 
-	%for the transition from state $S_{i}$ to state $S_{j}$ at time $t$,
+	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
+	giving the probability of class/state $i$ at time $t=1$ with
+	covariate $\vc{z}_1$.
 
-	\item a set $\mathcal{A}$ of transition probability functions $a_{ij}(\vc{x}_t) = 
-	Pr(S_t=j|S_{t-1}=i,\vc{x}_t)$, giving the probability of a transition from state 
-	$i$ to state $j$ with covariate $\vc{x}_t$,
+	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
+	provides the probability of a transition from state $i$ to state
+	$j$ with covariate $\vc{z}_t$,
 
-	\item a set $\mathcal{B}$ of observation density functions $b_j^k(\vc{x}_t) = p(O_{t}^k|S_t = j, \vc{x}_t)$ that
-	provide the conditional densities of observations $O_{t}^k$ 
-	associated with latent class/state $j$ and covariate $\vc{x}_t$,
-
-	\item a set $\mathcal{P}$ of latent class/state initial probability functions
-	$\pi_{i}(\vc{x}_1) = Pr(S_1 = i|\vc{x}_1)$, giving the probability of class/state
-	$i$ at time $t=1$ with covariate $\vc{x}_1$.
-
+	\item $\vc{b}_{j}$ is a vector of observation densities
+	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
+	conditional densities of observations $O_{t}^k$ associated with
+	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
+	$k=1, \ldots, m$. 
 \end{enumerate}
 
-The dependent mixture model is defined by the following equations: 
-%\begin{align}
-%	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
-%	O_{t} &= b(O_{t}|S_{t}), 
-%\end{align}
-\begin{align}
-	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
-	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
-\end{align}
+% The dependent mixture model is defined by the following equations: 
+% %\begin{align}
+% %	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
+% %	O_{t} &= b(O_{t}|S_{t}), 
+% %\end{align}
+% \begin{align}
+% 	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
+% 	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
+% \end{align}
 %where $S_{t}$ is a sequence of hidden states, $A$ is a transition
@@
 %density function for $O_{t}$ conditional on the hidden state $S_{t}$.
+
 In the example data above, $b_j^k$ could be a Gaussian distribution
@@
 accuracy data.  In the models we are considering here, both the
-transition probabilities $\mat{A}$ and the initial state probabilities $\pi$
-may depend on covariates as well as the response distributions $\vc{B}$. 
-See for example \citet{Fruhwirth2006} for an overview of 
-hidden Markov models with extensions.
+transition probability functions $a_{ij}$and the initial state
+probability functions $\greekv{\pi}$ may depend on covariates as well
+as the response distributions $\vc{b_{j}^{k}}$.  See for example
+\citet{Fruhwirth2006} for an overview of hidden Markov models with
+extensions.
 
@@
 \begin{equation}
-	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
-Pr(\vc{O}_{t}|\vc{O}_{1}, 
-	\ldots, \vc{O}_{t-1}), 
+	L_{T} = Pr(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
+	Pr(\vc{O}_{t}|\vc{O}_{1:t-1}), 
 	\label{condLike}
@@
 \begin{equation}
-	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
+	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
 \vc{O}_{t-1})].
@@
 \begin{align}
-	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
+	\phi_{1}(j) &:= \Prob(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
 	\label{eq:fwd1} \\
 \begin{split}
-	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
+	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
 \vc{O}_{t-1}) \\
@@
 where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
-$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
+$\Phi_{t}=\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
 equation~(\ref{eq:condLogl}) gives the following expression for the 

### 2009-08-05 papers/jss/article.tex
- changes to EM description
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@308 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{equation}
-\log \Prob(O_{1:T}, S_{1:T}|\greekv{\theta}) = \log \Prob(S_1|\greekv{\theta}_1) 
-+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\greekv{\theta}_2) 
-+ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\greekv{\theta}_3)
+\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
+\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
++ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
++ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\vc{z}_t,\greekv{\theta}_3)
 \end{equation}
-This likelihood depends on the unobserved states $S_t$. In the Expectation step,
-we replace these with their expected values given a set of (initial) parameters 
-$\greekv{\theta}' = (\greekv{\theta}'_1, \greekv{\theta}'_2,\greekv{\theta}'_3)$
-and observations $O_{1:T}$. The expected log likelihood 
+This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
+Expectation step, we replace these with their expected values given a set of 
+(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
+\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $O_{1:T}$. The expected 
+log likelihood 
 \begin{equation}
 Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
-(\log \Prob(O_{1:T},S_{1:T}|O_{1:T},\greekv{\theta}))
+(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
 \end{equation}
@@
 Q(\greekv{\theta},\greekv{\theta}') = 
-\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\greekv{\theta}_1) \\ 
+\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
 + \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
-= j,\greekv{\theta}_2)  \\
+= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
  + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
-\ln \Prob(O^k_t|S_t=j,\greekv{\theta}_3),
+\ln \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
 \end{multline}
 %\end{equation}
-where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|O_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|O_{1:T},\greekv{\theta}')$ can be computed effectively by the Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}.
-The Maximisation step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.
-As the r.h.s.
-of (\ref{eq:Q}) consists of three separate parts, we can maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and $\greekv{\theta}_3$.
-In common models, maximisation for $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in \pkg{MASS}, and maximisation for $\greekv{\theta}_3$ by the \code{glm} routine.
+where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|\vc{O}_{1:T},
+\vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|\vc{O}_{1:T},
+\vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
+Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
+step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
+right hand side of (\ref{eq:Q}) consists of three separate parts, we can 
+maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and 
+$\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and 
+$\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in the 
+\pkg{nnet} package \citep{Venables2002}, and maximisation for 
+$\greekv{\theta}_3$ by the standard \code{glm} routine. Note that for the latter 
+maximisation, the expected values $\gamma_t(j)$ are used as prior weights of the 
+observations $O^k_t$.
 
@@
 models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
-\cite{Tamura2009,Spellucci2002}, because it handles general linear
+\citep{Tamura2009,Spellucci2002}, because it handles general linear
 (in)equality constraints, and optionally also non-linear constraints.
@@
 
-Throughout this manual a data set called \code{speed} is used.  It
+Throughout this article a data set called \code{speed} is used.  It
 consists of three time series with three variables: response time,
@@
 Note also that start values for the transition parameters are provided
-in this call using the \code{trstart} argument. The package does not 
-provide automatic starting values.
+in this call using the \code{trstart} argument. At this time, the package does 
+not provide automatic starting values.
 
@@
 \end{CodeChunk}
-These statistics may be extracted using \code{logLik},
+These statistics can also be extracted using \code{logLik},
 \code{AIC} and \code{BIC}, respectively.
@@
 meaning that the parameter for the
-base category is fixed at zero (see \citet[see][p.\ 267
-ff.]{Agresti2002} for multinomial logistic models and various
-parameterizations). See also \citet{Chung2007} for similar models, latent transition models using logistic regression on the transition parameters. They fit such models on repeated measurement
-data ($T=2$) using Bayesian methods.  The default baseline category is the first state.
+base category is fixed at zero \citep[see][p.\ 267
+ff., for multinomial logistic models and various
+parameterizations]{Agresti2002}. See also \citet{Chung2007} for similar models, 
+latent transition models using logistic regression on the transition parameters. 
+They fit such models on repeated measurement
+data ($T=2$) using Bayesian methods.  The default baseline category is the 
+first state.
 Hence, for example, for a 3-state model, the initial state probability

### 2009-08-13 papers/jss/article.tex
- more changes to jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@309 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 sciences, in different areas and applications.  In psychology, they
-are used for modelling learning processes, see \citet{Wickens1982},
-for an overview, and e.g.\ \citet{Schmittmann2006} for a recent
-application.  In economics, latent Markov models are so-called regime
+are used for modelling learning processes (see \citealp{Wickens1982},
+for an overview, and e.g. \citealp{Schmittmann2006}, for a recent
+application).  In economics, latent Markov models are so-called regime
 switching models (see e.g., \citealp{Kim1994} and
@@
 recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
-genetics \citep{Krogh1998}.  In those latter areas of application,
+genetics \citep{Krogh1998}.  In these latter areas of application,
 latent Markov models are usually referred to as hidden Markov models. 
-For more examples of applications, see e.g.\
+See for example \citet{Fruhwirth2006} for an overview of hidden Markov models 
+with extensions. Further examples of applications can be found in e.g.\
 \citet[][chapter~1]{Cappe2005}.
@@
 models are used commonly in the social sciences, no comprehensive
-package was available for fitting such models.  Common software for
+package was available for fitting such models.  Existing software for
 estimating Markovian models include Panmark \citep{Pol1996}, and for
-latent class models Latent Gold \citep{Vermunt2003}.  Those programs
-are lacking a number of important features, besides not being freely
+latent class models Latent Gold \citep{Vermunt2003}. These programs
+lack a number of important features, besides not being freely
 available. There are currently some packages in \proglang{R} that
@@
 	\item to be able to include covariates in the prior or initial
-	state probabilities of models
+	state probabilities
 
 	\item to be easily extensible, in particular, to allow users to
-	easily add new uni- or multivariate response distributions, and
-	similarly for the addition of other transition models, e.g.,
-	continuous time observation models
+	easily add new uni- or multivariate response distributions and
+	new transition models, e.g., continuous time observation models
 
@@
 observed data and the model reduces to a finite mixture or
-latent class model. Although there are other specialized packages to
+latent class model. While there are specialized packages to
 deal with mixture data, as far as we know these 
 don't allow the inclusion of covariates on the prior probabilities of
-class membership. The possibility to estimating the effects of covariates on 
+class membership. The possibility to estimate the effects of covariates on 
 prior and transition probabilities is a distinguishing feature of 
-\pkg{depmixS4}. In the next section, an outline is provided of the model and 
-the likelihood equations.
+\pkg{depmixS4}. In the next section, we provide an outline of the model and 
+likelihood equations.
 
@@
 
-The data considered here have the general form $\vc{O}_{1:T}=O_{1}^{1}, \ldots,
+The data considered here have the general form $\vc{O}_{1:T}= (O_{1}^{1}, \ldots,
 O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
-\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
+\ldots, O_{T}^{m})$ for an $m$-variate time series of length $T$.  As
 an example, consider a time series of responses generated by a single
-participant in a response time experiment. The data consists of three
+participant in a psychological response time experiment. The data consists of three
 variables, response time, response accuracy, and a covariate which is a pay-off
-variable reflecting the relative reward for speeded and/or accuracy responding. 
-These variables are measured on 168, 134 and 137 occasions respectively (in
-Figure~\ref{fig:speed} the first part of this series is plotted). These data
+variable reflecting the relative reward for speeded and/or accurate responding. 
+These variables are measured on 168, 134 and 137 occasions respectively (the 
+first part of this series is plotted in
+Figure~\ref{fig:speed}). These data
 are more fully described in \citet{Dutilh2009}.
@@
 
-The latent Markov model is commonly associated with data of this type,
-although usually only multinomial response variables are considered.
-However, common estimation procedures, such as those implemented in
-\citet{Pol1996}, are not suitable for long time series due to
+The latent Markov model is usually associated with data of this type,
+in particular for multinomially distributed responses. % response variables are considered.
+However, commonly employed estimation procedures
+\citep[e.g.,][]{Pol1996}, are not suitable for long time series due to
 underflow problems. In contrast, the hidden Markov model is typically
-only used for `long' univariate time series \citep[][,
-chapter~1]{Cappe2005}.  In the next paragraphs, the likelihood
-equation and estimation procedures for the dependent mixture model are
-described for data of the above form.  We use the term ``dependent
-mixture model'' because one of the authors (Ingmar Visser) thought it
-was time for a new name for these models\footnote{Only later did I
+only used for `long' univariate time series \citep[][chapter~1]{Cappe2005}. 
+We use the term ``dependent mixture model'' because one of the authors 
+(Ingmar Visser) thought it
+was time for a new name to relate these models\footnote{Only later did I
 find out that \citet{Leroux1992} already coined the term dependent
@@
 
-The likelihood of the dependent mixture model conditional on the
-unknown (or hidden) state sequence $S_{1:T}$ and the model is given
-by:
+The fundamental assumption of a dependent mixture model is that at any time 
+point, the observations are distributed as a mixture with $n$ components 
+(or states), and that time-dependencies between the observations are due to 
+time-dependencies between the mixture components (i.e., transition probabilities
+between the components). These latter dependencies are assumed to
+follow a first-order Markov process. In the models we are considering here, the 
+mixture distributions, the initial mixture probabilities and transition 
+probabilities can all dependent on covariates $\vc{z}_t$.
+
+%transition probability functions $a_{ij}$ and the initial state
+%probability functions $\greekv{\pi}$ may depend on covariates as well
+%as the response distributions $\vc{b_{j}^{k}}$.  See for example
+%\citet{Fruhwirth2006} for an overview of hidden Markov models with
+%extensions.
+
+In a dependent mixture model, the joint likelihood of observations $\vc{O}_{1:T}$ and 
+latent states $\vc{S}_{1:T} = (S_1,\ldots,S_T)$, given model parameters 
+$\greekvec{\theta}$ and covariates $\vc{z}_{1:T} = (\vc{z}_1,\ldots,\vc{z}_T)$, 
+can be written as
 \begin{equation}
-	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
-	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
-\end{equation}
-where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
-at the likelihood of the data given the parameter vector
-$\greekv{\theta}$, i.e.\ without the state sequence, we need to sum
-above likelihood over all possible state sequences. This likelihood 
-is written as: 
-\begin{equation}
-	\Prob(\vc{O}_{1:T}|\vc{S}_{1:T},\greekv{\theta}) 
-	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
-	\prod_{t=1}^{T-1} a_{ij}(z_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
+	\Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\greekv{\theta},\vc{z}_{1:T}) =  
+	\pi_{i}(\vc{z}_1) \vc{b}_{S_t}(\vc{O}_1|\vc{z}_{1})
+	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{S_t}(\vc{O}_{t+1}|\vc{z}_{t+1}),
 \end{equation}
@@
 	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
-	of $n$ latent classes or states; we write for short $S_{i}$ to
-	denote $S_{t}=i$.
+	of $n$ latent classes or states. %; we write for short $S_{i}$ to
+	%denote $S_{t}=i$.
 
@@
 
-	\item $\vc{b}_{j}$ is a vector of observation densities
-	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
+	\item $\vc{b}_{S_t}$ is a vector of observation densities
+	$b_{j}^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
 	conditional densities of observations $O_{t}^k$ associated with
@@
 \end{enumerate}
+
+%In the next paragraphs, the likelihood
+%equation and estimation procedures for the dependent mixture model are
+%described for data of the above form.
+
+%The likelihood of the dependent mixture model conditional on the
+%unknown (or hidden) state sequence $\vc{S}_{1:T} = (S_1,\ldots,S_T)$ 
+%and the model is given
+%by:
+%\begin{equation}
+%	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
+%	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
+%\end{equation}
+%where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
+%at the marginal likelihood of the data for parameter vector
+%$\greekv{\theta}$, i.e.\ unconditional on the state sequence, we need to sum
+%above likelihood over all possible state sequences. This likelihood 
+%is written as: 
+%\begin{equation}
+%	\Prob(\vc{O}_{1:T}|\greekv{\theta}) =  
+%	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
+%	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
+%\end{equation}
+%where we have the following elements:
+%\begin{enumerate}
+%	
+%	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
+%	of $n$ latent classes or states; we write for short $S_{i}$ to
+%	denote $S_{t}=i$.
+%	
+%	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
+%	giving the probability of class/state $i$ at time $t=1$ with
+%	covariate $\vc{z}_1$.
+%	
+%	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
+%	provides the probability of a transition from state $i$ to state
+%	$j$ with covariate $\vc{z}_t$,
+%	
+%	\item $\vc{b}_{j}$ is a vector of observation densities
+%	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
+%	conditional densities of observations $O_{t}^k$ associated with
+%	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
+%	$k=1, \ldots, m$. 
+%\end{enumerate}
 
@@
 
-In the example data above, $b_j^k$ could be a Gaussian distribution
-function for the response time variable, and a Bernoulli for the
-accuracy data.  In the models we are considering here, both the
-transition probability functions $a_{ij}$and the initial state
+For the example data above, $b_j^k$ could be a Gaussian distribution
+function for the response time variable, and a Bernoulli distribution for the
+accuracy variable.  In the models we are considering here, both the
+transition probability functions $a_{ij}$ and the initial state
 probability functions $\greekv{\pi}$ may depend on covariates as well
-as the response distributions $\vc{b_{j}^{k}}$.  See for example
+as the response distributions $b_{j}^{k}$.  See for example
 \citet{Fruhwirth2006} for an overview of hidden Markov models with
@@
 
-The log-likelihood of hidden Markov models is usually computed by the
+To obtain maximum likelihood estimates of the model parameters, we need the
+marginal likelihood of the observations. For hidden Markov models, this 
+marginal (log-)likelihood is usually computed by the
 so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
@@
 rewriting the likelihood as follows (for ease of exposition the
-dependence on the model parameters is dropped here):
+dependence on the model parameters and covariates is dropped here):
 \begin{equation}
-	L_{T} = Pr(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
-	Pr(\vc{O}_{t}|\vc{O}_{1:t-1}), 
+	L_{T} = \Prob(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
+	\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}), 
 	\label{condLike}
 \end{equation}
-where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
+where $\Prob(\vc{O}_{1}|\vc{O}_{0}):=\Prob(\vc{O}_{1})$. Note that for a 
 simple, i.e.\ observed, Markov chain these probabilities reduce to 
-$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
-\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
+$\Prob(\vc{O}_{t}|\vc{O}_{1},\ldots, 
+\vc{O}_{t-1})=\Prob(\vc{O}_{t}|\vc{O}_{t-1})$.
 The log-likelihood can now be expressed as:
 \begin{equation}
-	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
-\vc{O}_{t-1})].
+	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)}].
 	\label{eq:condLogl}
@@
 	\label{eq:fwd1} \\
-\begin{split}
-	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
-\vc{O}_{t-1}) \\
-	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
+%\begin{split}
+	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1:(t-1)}) %\\
+	= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
 (\Phi_{t-1})^{-1},
 	\label{eq:fwdt} 
-\end{split} 
+%\end{split} 
 \end{align}
 where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
-$\Phi_{t}=\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
+$\Phi_{t}=\Prob(\vc{O}_{t}|\vc{O}_{1:(t-1)})$, and 
 equation~(\ref{eq:condLogl}) gives the following expression for the 
@@
 for the prior model, transition model, and response model respectively. The 
-joint log likelihood can be written as
+joint log-likelihood can be written as
 \begin{equation}
@@
 + \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
-+ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\vc{z}_t,\greekv{\theta}_3)
++ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
 \end{equation}
@@
 \sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
-+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
++ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
 = j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
@@
 \vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
-Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
+forward-backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
 step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
@@
 
-Throughout this article a data set called \code{speed} is used.  It
+Throughout this article a data set called \code{speed} is used.  As already 
+indicated in the Introduction, it
 consists of three time series with three variables: response time,
@@
 
-A dependent mixture model is defined by the number of states, and by
-the response distribution functions, and can be created with the
-\code{depmix}-function as follows:
+A dependent mixture model is defined by the number of states and the initial 
+state, state transition, and response distribution functions. A default 
+dependent mixture model can be created with the \code{depmix}-function as 
+follows:
 
@@
 were applied, see below).  The \code{print} method provides summary
-information on convergence, the log likelihood and the AIC and BIC
+information on convergence, the log-likelihood and the AIC and BIC
 values:
@@
 
+Note that, since no further arguments were specified, the initial state, 
+state transition and response distributions were set to their defaults 
+(multinomial distributions for the first two, and Gaussian distributions for the
+response distributions).
+
 \subsection{Covariates on transition parameters}
 
-The transition probabilities and the initial state probabilities
+By default, the transition probabilities and the initial state probabilities
 are parameterized using the multinomial
-logistic model.   In particular, each row of the transition matrix is
+logistic model. More precisely, each row of the transition matrix is
 parameterized by a baseline category logistic multinomial, 
@@
 ff., for multinomial logistic models and various
-parameterizations]{Agresti2002}. See also \citet{Chung2007} for similar models, 
-latent transition models using logistic regression on the transition parameters. 
-They fit such models on repeated measurement
-data ($T=2$) using Bayesian methods.  The default baseline category is the 
-first state.
-Hence, for example, for a 3-state model, the initial state probability
+parameterizations]{Agresti2002}. The default baseline category is the 
+first state. Hence, for example, for a 3-state model, the initial state probability
 model would have three parameters of which the first is fixed at zero
@@
 
+The multinomial logistic model allows us to include covariates 
+on the initial state and transition probabilities. \citet{Chung2007} discuss a 
+related latent transition model for repeated measurement
+data ($T=2$) using logistic regression on the transition 
+parameters; they rely on Bayesian methods of estimation. 
 Covariates on the transition probabilities can be specified using a
@@
 \begin{CodeInput}
-> mod <- setpars(mod, value=1:npar(mod))
-> mod
+> setpars(mod, value=1:npar(mod))
 \end{CodeInput}
@@
 \begin{CodeInput}
-> mod <- setpars(mod, getpars(mod,which="fixed"))
-> mod
+> setpars(mod, getpars(mod,which="fixed"))
 \end{CodeInput}
@@
 constraints, also between, e.g., a multinomial regression coefficient 
-for the transition matrix and the mean of a gaussian response model. 
+for the transition matrix and the mean of a Gaussian response model. 
 Whether such constraints make sense is hence the responsibility of 
@@
 
-The \pkg{depmixS4} package was built with the aim of having the 
-possibility of adding new response distributions (and possibly also 
-other models for the prior and transition probabilities models). 
-Therefore, the EM algorithm simply calls \code{fit} functions from the 
-separate response models without knowing what they are. As a 
+The \pkg{depmixS4} package was designed with the aim of making it relatively
+easy to add new response distributions (as well as possibly new prior and 
+transition models). 
+To make this possible, the EM routine simply calls the \code{fit} methods 
+of the separate response models without needing to know how they work. As a 
 consequence, adding user-specified response models is straightforward. 
@@
 	\item paramaters: a named list with the coefficients and possibly 
-	other parameters, e.g., the standard deviation in the gaussian 
+	other parameters, e.g., the standard deviation in the Gaussian 
 	response model
@@
 In the \code{speed} data example, it may be more appropriate to model
-the response times as an exgaus distribution rather than using the
-Gaussian.  We can do so as follows, first we define our own
-exgaus-class extending the response-class:
+the response times with an exgaus rather than Gaussian distribution.  
+To do so, we first define an exgaus-class extending the response-class:
 \begin{CodeChunk}
+\begin{CodeInput}
 setClass("exgaus", contains="response")
+\end{CodeInput}
 \end{CodeChunk}
@@
 
-Only the constructor function and the fit function are provided here
-and the complete code is provided in the helpfile for \code{makeDepmix}. 
-The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} packages \cite{Stasinopoulos2009a,Stasinopoulos2009b} for 
+Only the constructor and the fit method are provided here; the complete code 
+can be found in the helpfile for \code{makeDepmix}. 
+The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} 
+packages \citep{Stasinopoulos2009a,Stasinopoulos2009b} for 
 computing the \code{dens}ity and for \code{fit}ting the parameters.
 
-The constructor function is defined as: 
+The constructor method is defined as: 
 \begin{CodeChunk}
@@
   standardGeneric("exgaus"))
+
 setMethod("exgaus", 
@@
 
-The \code{fit} function is defined as follows: 
+The \code{fit} method is defined as follows: 
 \begin{CodeChunk}
@@
 
-The \code{fit} function defines a \code{gamlss} model with 
+The \code{fit} method defines a \code{gamlss} model with 
 only an intercept to be estimated and then sets the fitted parameters 
@@
 After defining all the necessary methods for the new response model, 
-we can  now define the dependent mixture model using this reponse. 
-The function \code{makeDepmix} is added to \pkg{depmixS4} to have 
+we can  now define the dependent mixture model using this reponse model. 
+The function \code{makeDepmix} is included in \pkg{depmixS4} to have 
 full control over model specification, and we need it here.
@@
 Next, we define the transition and prior probability models using the 
-transInit function (which produces a transInit model, which also extends 
+\code{transInit} function (which produces a transInit model, which also extends 
 the response class): 
@@
 \begin{CodeInput}
-mod <- makeDepmix(response=rModels,transition=transition,
+> mod <- makeDepmix(response=rModels,transition=transition,
     prior=inMod,stat=FALSE)
-fm <- fit(mod)
+> fm <- fit(mod)
 \end{CodeInput}
@@
 and finite mixture models, although it should be noted that more specialized packages 
-are available for doing so, e.g.\ \pkg{FlexMix} \citep{Leisch2004}. The package is intended
+are available for this such as \pkg{FlexMix} \citep{Leisch2004}. The package is intended
 for modeling of (individual) time series data with the aim of characterizing the transition 

### 2009-08-19 papers/jss/article.tex
... jss final version, to be submitted.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@310 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	Other distributions can be added easily, and an example is
-	provided with the exgaus distribution.  Parameter are estimated by
+	provided with the exgaus distribution.  Parameters are estimated by
 	the EM algorithm or, when (linear) constraints are imposed on the
 	parameters, by direct numerical optimization with the
-	\pkg{Rdonlp2} routine.  }
+	\pkg{Rdonlp2} routine.}
 
@@
 latent Markov models are usually referred to as hidden Markov models.
-See for example \citet{Fruhwirth2006} for an overview of hidden Markov models 
-with extensions. Further examples of applications can be found in e.g.\
-\citet[][chapter~1]{Cappe2005}.
+See for example \citet{Fruhwirth2006} for an overview of hidden Markov
+models with extensions.  Further examples of applications can be found
+in e.g.\ \citet[][chapter~1]{Cappe2005}.
 
@@
 series data, for say $T>100$, it can also handle the limit case when
-$T=1$.  In this case, there are no time dependencies between
-observed data and the model reduces to a finite mixture or
-latent class model. While there are specialized packages to
-deal with mixture data, as far as we know these 
-don't allow the inclusion of covariates on the prior probabilities of
-class membership. The possibility to estimate the effects of covariates on 
-prior and transition probabilities is a distinguishing feature of 
-\pkg{depmixS4}. In the next section, we provide an outline of the model and 
-likelihood equations.
+$T=1$.  In this case, there are no time dependencies between observed
+data and the model reduces to a finite mixture or latent class model.
+While there are specialized packages to deal with mixture data, as far
+as we know these don't allow the inclusion of covariates on the prior
+probabilities of class membership.  The possibility to estimate the
+effects of covariates on prior and transition probabilities is a
+distinguishing feature of \pkg{depmixS4}.  In the next section, we
+provide an outline of the model and likelihood equations.
 
@@
 
-The data considered here have the general form $\vc{O}_{1:T}= (O_{1}^{1}, \ldots,
-O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
-\ldots, O_{T}^{m})$ for an $m$-variate time series of length $T$.  As
-an example, consider a time series of responses generated by a single
-participant in a psychological response time experiment. The data consists of three
-variables, response time, response accuracy, and a covariate which is a pay-off
-variable reflecting the relative reward for speeded and/or accurate responding. 
-These variables are measured on 168, 134 and 137 occasions respectively (the 
-first part of this series is plotted in
-Figure~\ref{fig:speed}). These data
-are more fully described in \citet{Dutilh2009}.
+The data considered here have the general form $\vc{O}_{1:T}=
+(O_{1}^{1}, \ldots, O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$,
+\ldots, $O_{T}^{1}, \ldots, O_{T}^{m})$ for an $m$-variate time series
+of length $T$.  As an example, consider a time series of responses
+generated by a single participant in a psychological response time
+experiment.  The data consists of three variables, response time,
+response accuracy, and a covariate which is a pay-off variable
+reflecting the relative reward for speeded and/or accurate responding.
+These variables are measured on 168, 134 and 137 occasions
+respectively (the first part of this series is plotted in
+Figure~\ref{fig:speed}).  These data are more fully described in
+\citet{Dutilh2009}.
 
@@
 The latent Markov model is usually associated with data of this type,
-in particular for multinomially distributed responses. % response variables are considered.
-However, commonly employed estimation procedures
-\citep[e.g.,][]{Pol1996}, are not suitable for long time series due to
-underflow problems. In contrast, the hidden Markov model is typically
-only used for `long' univariate time series \citep[][chapter~1]{Cappe2005}. 
-We use the term ``dependent mixture model'' because one of the authors 
-(Ingmar Visser) thought it
-was time for a new name to relate these models\footnote{Only later did I
-find out that \citet{Leroux1992} already coined the term dependent
-mixture models in an application with hidden Markov mixtures of
-Poisson count data.}.
-
-The fundamental assumption of a dependent mixture model is that at any time 
-point, the observations are distributed as a mixture with $n$ components 
-(or states), and that time-dependencies between the observations are due to 
-time-dependencies between the mixture components (i.e., transition probabilities
-between the components). These latter dependencies are assumed to
-follow a first-order Markov process. In the models we are considering here, the 
-mixture distributions, the initial mixture probabilities and transition 
-probabilities can all dependent on covariates $\vc{z}_t$.
+in particular for multinomially distributed responses.  However,
+commonly employed estimation procedures \citep[e.g.,][]{Pol1996}, are
+not suitable for long time series due to underflow problems.  In
+contrast, the hidden Markov model is typically only used for `long'
+univariate time series \citep[][chapter~1]{Cappe2005}.  We use the
+term ``dependent mixture model'' because one of the authors (Ingmar
+Visser) thought it was time for a new name to relate these
+models\footnote{Only later did I find out that \citet{Leroux1992}
+already coined the term dependent mixture models in an application
+with hidden Markov mixtures of Poisson count data.}.
+
+The fundamental assumption of a dependent mixture model is that at any
+time point, the observations are distributed as a mixture with $n$
+components (or states), and that time-dependencies between the
+observations are due to time-dependencies between the mixture
+components (i.e., transition probabilities between the components).
+These latter dependencies are assumed to follow a first-order Markov
+process.  In the models we are considering here, the mixture
+distributions, the initial mixture probabilities and transition
+probabilities can all depend on covariates $\vc{z}_t$.
 
@@
 
-In a dependent mixture model, the joint likelihood of observations $\vc{O}_{1:T}$ and 
-latent states $\vc{S}_{1:T} = (S_1,\ldots,S_T)$, given model parameters 
-$\greekvec{\theta}$ and covariates $\vc{z}_{1:T} = (\vc{z}_1,\ldots,\vc{z}_T)$, 
-can be written as
+In a dependent mixture model, the joint likelihood of observations
+$\vc{O}_{1:T}$ and latent states $\vc{S}_{1:T} = (S_1,\ldots,S_T)$,
+given model parameters $\greekv{\theta}$ and covariates $\vc{z}_{1:T}
+= (\vc{z}_1,\ldots,\vc{z}_T)$, can be written as:
 \begin{equation}
@@
 	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
-	of $n$ latent classes or states. %; we write for short $S_{i}$ to
-	%denote $S_{t}=i$.
-
-	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
-	giving the probability of class/state $i$ at time $t=1$ with
-	covariate $\vc{z}_1$.
+	of $n$ latent classes or states.
+
+	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$, giving the
+	probability of class/state $i$ at time $t=1$ with covariate
+	$\vc{z}_1$.
 
@@
 	\item $\vc{b}_{S_t}$ is a vector of observation densities
-	$b_{j}^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
-	conditional densities of observations $O_{t}^k$ associated with
-	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
-	$k=1, \ldots, m$. 
+	$b_{j}^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that
+	provide the conditional densities of observations $O_{t}^k$
+	associated with latent class/state $j$ and covariate $\vc{z}_t$,
+	$j=1, \ldots, n$, $k=1, \ldots, m$.
+
 \end{enumerate}
@@
 For the example data above, $b_j^k$ could be a Gaussian distribution
-function for the response time variable, and a Bernoulli distribution for the
-accuracy variable.  In the models we are considering here, both the
-transition probability functions $a_{ij}$ and the initial state
-probability functions $\greekv{\pi}$ may depend on covariates as well
-as the response distributions $b_{j}^{k}$.  See for example
-\citet{Fruhwirth2006} for an overview of hidden Markov models with
-extensions.
+function for the response time variable, and a Bernoulli distribution
+for the accuracy variable.  In the models we are considering here,
+both the transition probability functions $a_{ij}$ and the initial
+state probability functions $\greekv{\pi}$ may depend on covariates as
+well as the response distributions $b_{j}^{k}$.
 
@@
 
-To obtain maximum likelihood estimates of the model parameters, we need the
-marginal likelihood of the observations. For hidden Markov models, this 
-marginal (log-)likelihood is usually computed by the
+To obtain maximum likelihood estimates of the model parameters, we
+need the marginal likelihood of the observations.  For hidden Markov
+models, this marginal (log-)likelihood is usually computed by the
 so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
@@
 Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
-through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
-parameters are estimated by iteratively maximising the 
-expected joint likelihood of the parameters given the observations and states. 
-Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
-be the general parameter vector consisting of three subvectors with parameters 
-for the prior model, transition model, and response model respectively. The 
-joint log-likelihood can be written as
+through the use of a general Newton-Raphson optimizer.  In the EM
+algorithm, parameters are estimated by iteratively maximising the
+expected joint likelihood of the parameters given the observations and
+states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
+\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
+consisting of three subvectors with parameters for the prior model,
+transition model, and response model respectively.  The joint
+log-likelihood can be written as
 \begin{equation}
@@
 (initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
-\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $O_{1:T}$. The expected 
+\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. The expected 
 log likelihood 
@@
 The EM algorithm however has some drawbacks.  First, it can be slow to
-converge towards the end of optimization.  Second,
-applying constraints to parameters can be problematic; in particular,
-EM can lead to wrong parameter estimates when applying constraints.
-Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
-models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
+converge towards the end of optimization.  Second, applying
+constraints to parameters can be problematic; in particular, EM can
+lead to wrong parameter estimates when applying constraints.  Hence,
+in \pkg{depmixS4}, EM is used by default in unconstrained models, but
+otherwise, direct optimization is done using \pkg{Rdonlp2}
 \citep{Tamura2009,Spellucci2002}, because it handles general linear
@@
 \end{enumerate}
-We have separated the stages of model specification and model fitting because
-fitting large models can be fairly time-consuming and it is hence useful to be
-able to check the model specification before actually fitting the model.
-
-\subsection{Example data: speed}
-
-Throughout this article a data set called \code{speed} is used.  As already 
-indicated in the Introduction, it
-consists of three time series with three variables: response time,
-accuracy, and a covariate Pacc which defines the relative pay-off for
-speeded and accurate responding.  The participant in this experiment
-switches between fast responding at chance level and relatively slower
-responding at a high level of accuracy.  Interesting hypotheses to
-test are: is the switching regime symmetric?  Is there evidence for
-two states or does one state suffice?  Is the guessing state actually
-a guessing state, i.e., is the probability of a correct response at
-chance level (0.5)?
+We have separated the stages of model specification and model fitting
+because fitting large models can be fairly time-consuming and it is
+hence useful to be able to check the model specification before
+actually fitting the model.
+
+\subsection[Example data: speed]{Example data: \code{speed}}
+
+Throughout this article a data set called \code{speed} is used.  As
+already indicated in the Introduction, it consists of three time
+series with three variables: response time, accuracy, and a covariate
+Pacc which defines the relative pay-off for speeded versus accurate
+responding.  The participant in this experiment switches between fast
+responding at chance level and relatively slower responding at a high
+level of accuracy.  Interesting hypotheses to test are: is the
+switching regime symmetric?  Is there evidence for two states or does
+one state suffice?  Is the guessing state actually a guessing state,
+i.e., is the probability of a correct response at chance level (0.5)?
 
@@
 
-A dependent mixture model is defined by the number of states and the initial 
-state, state transition, and response distribution functions. A default 
-dependent mixture model can be created with the \code{depmix}-function as 
-follows:
+A dependent mixture model is defined by the number of states and the
+initial state, state transition, and response distribution functions.
+A dependent mixture model can be created with the
+\code{depmix}-function as follows:
 
@@
 Note also that start values for the transition parameters are provided
-in this call using the \code{trstart} argument. At this time, the package does 
-not provide automatic starting values.
-
-The so-defined models needs to be \code{fit}ted with the following
+in this call using the \code{trstart} argument.  At this time, the
+package does not provide automatic starting values.
+
+The so-defined model needs to be \code{fit}ted with the following
 line of code:
@@
 Convergence info: Log likelihood converged to within tol. 
-'log Lik.' -84.34175 (df=7)
-AIC:  182.6835 
-BIC:  211.275 
+'log Lik.' -84.34 (df=7)
+AIC:  182.68 
+BIC:  211.28 
 \end{CodeOutput}
 \end{CodeChunk}
-These statistics can also be extracted using \code{logLik},
-\code{AIC} and \code{BIC}, respectively.
-
-The \code{summary} method of \code{fit}ted models provides the parameter
-estimates, first for the prior probabilities model, second for the
-transition model, and third for the response models.
+These statistics can also be extracted using \code{logLik}, \code{AIC}
+and \code{BIC}, respectively.  By comparison, a 1-state model for
+these data, i.e. assuming there is no mixture, has a log-likelihood of
+-305.33, and 614.66, and 622.83 for the AIC and BIC respectively.
+Hence, the 2-state model fits the data much better than the 1-state
+model.  Note that the 1-state model can be specified using \code{mod <-
+depmix(rt~1, data=speed, nstates=1)}, although this model is trivial
+as it will simply return the mean and the sd of the rt variable.
+
+The \code{summary} method of \code{fit}ted models provides the
+parameter estimates, first for the prior probabilities model, second
+for the transition model, and third for the response models.
 
@@
      [,1]      [,2]
-[1,]    0 -11.25688
+[1,]    0 -11.25
 Probalities at zero values of the covariates.
-0.999987 1.291798e-05
+0.999 1.292e-05
 
@@
 Coefficients: 
-[1]  0.000000 -2.392455
+[1]  0.000 -2.392
 Probalities at zero values of the covariates.
-0.9162501 0.08374986
+0.9163 0.0837
 
@@
 Coefficients: 
-[1] 0.000000 2.139255
+[1] 0.000 2.139
 Probalities at zero values of the covariates.
-0.1053396 0.8946604
+0.105 0.895
 
@@
 Coefficients: 
-[1] 6.385492
-sd  0.2439376
+[1] 6.385
+sd  0.244
 
@@
 Coefficients: 
-[1] 5.511151
-sd  0.1926063 
+[1] 5.511
+sd  0.193 
 \end{CodeOutput}
@@
 
-Note that, since no further arguments were specified, the initial state, 
-state transition and response distributions were set to their defaults 
-(multinomial distributions for the first two, and Gaussian distributions for the
-response distributions).
+Note that, since no further arguments were specified, the initial
+state, state transition and response distributions were set to their
+defaults (multinomial distributions for the first two, and Gaussian
+distributions for the response distributions).
 
@@
 
-By default, the transition probabilities and the initial state probabilities
-are parameterized using the multinomial
-logistic model. More precisely, each row of the transition matrix is
-parameterized by a baseline category logistic multinomial, 
-meaning that the parameter for the
-base category is fixed at zero \citep[see][p.\ 267
-ff., for multinomial logistic models and various
+By default, the transition probabilities and the initial state
+probabilities are parameterized using the multinomial logistic model.
+More precisely, each row of the transition matrix is parameterized by
+a baseline category logistic multinomial, meaning that the parameter
+for the base category is fixed at zero \citep[see][p.\ 267 ff., for
+multinomial logistic models and various
 parameterizations]{Agresti2002}.  The default baseline category is the
-first state. Hence, for example, for a 3-state model, the initial state probability
-model would have three parameters of which the first is fixed at zero
-and the other two are freely estimated.
-
-The multinomial logistic model allows us to include covariates 
-on the initial state and transition probabilities. \citet{Chung2007} discuss a 
-related latent transition model for repeated measurement
-data ($T=2$) using logistic regression on the transition 
-parameters; they rely on Bayesian methods of estimation. 
-Covariates on the transition probabilities can be specified using a
-one-sided formula as in the following example:
+first state.  Hence, for example, for a 3-state model, the initial
+state probability model would have three parameters of which the first
+is fixed at zero and the other two are freely estimated.
+
+The multinomial logistic model allows us to include covariates on the
+initial state and transition probabilities.  \citet{Chung2007} discuss
+a related latent transition model for repeated measurement data
+($T=2$) using logistic regression on the transition parameters; they
+rely on Bayesian methods of estimation.  Covariates on the transition
+probabilities can be specified using a one-sided formula as in the
+following example:
 \begin{CodeChunk}
@@
      [,1]       [,2]
-[1,]    0 -0.9215182
-[2,]    0  1.8649734
+[1,]    0 -0.9215
+[2,]    0  1.865
 Probalities at zero values of the covariates.
-0.7153513 0.2846487
+0.7154 0.2846
 
@@
      [,1]     [,2]
-[1,]    0 2.471442
-[2,]    0 3.570856
+[1,]    0 2.471
+[2,]    0 3.571
 Probalities at zero values of the covariates.
-0.07788458 0.9221154
+0.0779 0.9221
 ...
@@
 \end{CodeChunk}
-The summary provides all parameters of the model, also the 
-(redundant) zeroes for the base-line category in the multinomial model. 
-The summary also prints the transition probabilities
-at the zero value of the covariate. Note that scaling of the covariate 
-is useful in this regard as it makes interpretation of these intercept probabilities 
+The summary provides all parameters of the model, also the (redundant)
+zeroes for the base-line category in the multinomial model.  The
+summary also prints the transition probabilities at the zero value of
+the covariate.  Note that scaling of the covariate is useful in this
+regard as it makes interpretation of these intercept probabilities
 easier.
@@
 responses.  In above examples we have only used the response times
-which were modelled with the Gaussian distribution. The accuracy data 
-are in the \code{speed} data are modelled with a multinomial by 
-specifying the following: 
+which were modelled with the Gaussian distribution.  The accuracy
+variable in the \code{speed} data can be modelled with a multinomial
+by specifying the following:
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-which provides the following fitted model parameters (only the 
+This provides the following fitted model parameters (only the 
 response parameters are given here): 
@@
 Coefficients: 
-[1] 5.52169
-sd  0.2028857
+[1] 5.522
+sd  0.2029
 
@@
      [,1]      [,2]
-[1,]    0 0.1030554
+[1,]    0 0.1031
 Probalities at zero values of the covariates.
-0.4742589 0.5257411
+0.4743 0.5257
 
@@
 Coefficients: 
-[1] 6.39369
-sd  0.2373650
+[1] 6.394
+sd  0.2374
 
@@
 	   [,1]     [,2]
-[1,]    0 2.245514
+[1,]    0 2.2455
 Probalities at zero values of the covariates.
-0.09573715 0.9042629 	
+0.0957 0.9043 	
 \end{CodeOutput}
 \end{CodeChunk}
-As can be seen, state 1 has fast response times around chance level, 
-whereas state 2 corresponds with slower responding at higher accuracy 
-levels.
-
-Note that by specifying multivariate observations in terms of a list, the 
-variables are considered conditionally independent (given the states). 
-Conditionally \emph{dependent} variables must be handled as a single element in
-the list. Effectively, this means specifying a multivariate response model. The 
-only multivariate response model currently implemented in \pkg{depmix} is for
-multivariate normal variables.
+As can be seen, state 1 has fast response times and accuracy is
+approximately at chance level (.474), whereas state 2 corresponds with
+slower responding at higher accuracy levels (.904).
+
+Note that by specifying multivariate observations in terms of a list,
+the variables are considered conditionally independent (given the
+states).  Conditionally \emph{dependent} variables must be handled as
+a single element in the list.  Effectively, this means specifying a
+multivariate response model.  The only multivariate response model
+currently implemented in \pkg{depmixS4} is for multivariate normal
+variables.
 
@@
 
-To illustrate the use of covariates on the prior probabilities we have included
-another data set with \pkg{depmixS4}. The \code{balance} data consists 
-of 4 binary items (correct-incorrect) on a balance scale task 
-\citet{Siegler1981}. The data form a subset of the data published in 
-\citet{Jansen2002}.
+To illustrate the use of covariates on the prior probabilities we have
+included another data set with \pkg{depmixS4}.  The \code{balance}
+data consists of 4 binary items (correct-incorrect) on a balance scale
+task \citet{Siegler1981}.  The data form a subset of the data
+published in \citet{Jansen2002}.
 
@@
 Note here that we define a \code{mix} model instead of a \code{depmix}
-models as these data form independent observations.
-
-The summary of the \code{fit}ted model gives (only the prior model is 
-shown here): 
+models as these data form independent observations.  More formally,
+\code{depmix} models extend the class of \code{mix} models by adding
+the transition models.  As for fitting \code{mix} models: as can be
+seen in equation~\ref{eq:Q}, the EM algorithm can be applied by simply
+dropping the second summand containing the transition parameters, and 
+this is implemented as such in the EM algorithms in \pkg{depmixS4}.
+
+The summary of the \code{fit}ted model gives the following (only the
+prior model is shown here):
 \begin{CodeChunk}
@@
      [,1]       [,2]
-[1,]    0 -2.5182573
-[2,]    0  0.5512996
+[1,]    0 -2.518
+[2,]    0  0.551
 Probalities at zero values of the covariates.
-0.9254119 0.07458815 
+0.9254 0.0746
 ...
@@
 
-Using package \pkg{Rdonlp2} by \citet{Tamura2009}, parameters may be fitted subject to
-general linear (in-)equality constraints.  Constraining and fixing
-parameters is done using the \code{conpat} argument to the
-\code{fit}-function, which specifies for each parameter in the
-model whether it's fixed (0) or free (1 or higher).  Equality
-constraints can be imposed by having two parameters have the same
-number in the \code{conpat} vector.  When only fixed values are
+Using package \pkg{Rdonlp2} by \citet{Tamura2009}, parameters may be
+fitted subject to general linear (in-)equality constraints.
+Constraining and fixing parameters is done using the \code{conpat}
+argument to the \code{fit}-function, which specifies for each
+parameter in the model whether it's fixed (0) or free (1 or higher).
+Equality constraints can be imposed by having two parameters have the
+same number in the \code{conpat} vector.  When only fixed values are
 required, the \code{fixed} argument can be used instead of
@@
 e.g.) for non-fixed parameters.  Fitting the models subject to these
-constraints is handled by the optimization routine \code{donlp2}.
-To be able to construct the \code{conpat} and/or \code{fixed} vectors 
-one needs the correct ordering of parameters which is briefly discussed 
+constraints is handled by the optimization routine \code{donlp2}.  To
+be able to construct the \code{conpat} and/or \code{fixed} vectors one
+needs the correct ordering of parameters which is briefly discussed
 next before proceeding with an example.
@@
 i.e., these vectors should have length of the number of parameters of
-the model, which can be obtained by calling \code{npar(object)}. Note that
-this is not the same as the degrees of freedom used e.g.\ in the \code{logLik}
-function because \code{npar} also counts the baseline category zeroes
-from the multinomial logistic models. 
-Parameters are numbered in the following order:
+the model, which can be obtained by calling \code{npar(object)}.  Note
+that this is not the same as the degrees of freedom used e.g.\ in the
+\code{logLik} function because \code{npar} also counts the baseline
+category zeroes from the multinomial logistic models.  Parameters are
+numbered in the following order:
 \begin{enumerate}
@@
 After this, we use the fitted values from this model to constrain the
-regression coefficients on the transition matrix (parameters numbers~6 and~10):
+regression coefficients on the transition matrix (parameters numbers~6
+and~10):
 \begin{CodeChunk}
@@
 
-The \pkg{depmixS4} package was designed with the aim of making it relatively
-easy to add new response distributions (as well as possibly new prior and 
-transition models). 
-To make this possible, the EM routine simply calls the \code{fit} methods 
-of the separate response models without needing to know how they work. As a 
-consequence, adding user-specified response models is straightforward. 
-User-defined distributions should extend the \code{response}-class 
-and have the following slots:
+The \pkg{depmixS4} package was designed with the aim of making it
+relatively easy to add new response distributions (as well as possibly
+new prior and transition models).  To make this possible, the EM
+routine simply calls the \code{fit} methods of the separate response
+models without needing access to the internal workings of these
+routines.  Referring to equation~\ref{eq:Q}, the EM algorithm calls
+separate fit functions for each part of the model, the prior
+probability model, the transition models, and the response models.  As
+a consequence, adding user-specified response models is
+straightforward.
+
+User-defined distributions should extend the \code{response}-class and
+have the following slots:
 \begin{enumerate}
@@
 In the \code{speed} data example, it may be more appropriate to model
-the response times with an exgaus rather than Gaussian distribution.  
-To do so, we first define an exgaus-class extending the response-class:
+the response times with an exgaus rather than a Gaussian distribution.
+To do so, we first define an exgaus-class extending the
+response-class:
 \begin{CodeChunk}
@@
 
-Only the constructor and the fit method are provided here; the complete code 
-can be found in the helpfile for \code{makeDepmix}. 
-The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} 
-packages \citep{Stasinopoulos2009a,Stasinopoulos2009b} for 
-computing the \code{dens}ity and for \code{fit}ting the parameters.
-
-The constructor method is defined as: 
+Only the constructor and the fit methods are provided here; the
+complete code can be found in the help file of the \code{makeDepmix}
+function.  The \code{exgaus} example uses the \pkg{gamlss} and
+\pkg{gamlss.distr} packages
+\citep{Stasinopoulos2009a,Stasinopoulos2009b} for computing the
+\code{dens}ity and for \code{fit}ting the parameters.
+
+The constructor method return an object of class \code{exgaus}, and is
+defined as follows:
 \begin{CodeChunk}
@@
 
-We first create all the response models that we want as a double list: 
+We first create all the response models that we need as a double list: 
 \begin{CodeChunk}
@@
     exgaus(rt,pstart=c(5,.1,.1)),
-    GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
+    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
   ),
@@
     exgaus(rt,pstart=c(6,.1,.1)),
-    GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
+    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
   )
@@
 transition <- list()
-transition[[1]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.9,0.1,0,0))
-transition[[2]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.1,0.9,0,0))
+transition[[1]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.9,0.1,0,0))
+transition[[2]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.1,0.9,0,0))
 inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
@@
 Gaussian in one state and an exgaus distribution in another state.
+Note also that according to the AIC and BIC, the model with the exgaus
+describes the data much better than the same model in which the
+response times are modeled as gaussian.
 
@@
 
-\pkg{depmixS4} provides a flexible framework for fitting dependent mixture models
-for a large variety of response distributions. It can also fit latent class regression
-and finite mixture models, although it should be noted that more specialized packages 
-are available for this such as \pkg{FlexMix} \citep{Leisch2004}. The package is intended
-for modeling of (individual) time series data with the aim of characterizing the transition 
-processes underlying the data. The possibility to use covariates on the transition matrix
-greatly enhances the flexibility of the model. The EM algorithm uses a very general 
-interface that allows easy addition of new response models.
-
-We are currently working on implementing the gradients for response and transition models 
-with two goals in mind. First, to speed up (constrained) parameter optimization using 
-\pkg{Rdonlp2}. Second, analytic gradients are useful in computing the Hessian of the
-estimated parameters so as to arrive at standard errors for those. We are also planning to
-implement goodness-of-fit statistics \citep{Titman2008}.
+\pkg{depmixS4} provides a flexible framework for fitting dependent
+mixture models for a large variety of response distributions.  It can
+also fit latent class regression and finite mixture models, although
+it should be noted that more specialized packages are available for
+this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
+intended for modeling of (individual) time series data with the aim of
+characterizing the transition processes underlying the data.  The
+possibility to use covariates on the transition matrix greatly
+enhances the flexibility of the model.  The EM algorithm uses a very
+general interface that allows easy addition of new response models.
+
+We are currently working on implementing the gradients for response
+and transition models with two goals in mind.  First, to speed up
+(constrained) parameter optimization using \pkg{Rdonlp2}.  Second,
+analytic gradients are useful in computing the Hessian of the
+estimated parameters so as to arrive at standard errors for those.  We
+are also planning to implement goodness-of-fit statistics
+\citep{Titman2008}, and automatic generation of starting values.
 

### 2010-02-22 papers/jss/dpx4Rev.Rnw
Sweave related files.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@357 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2009-08-05 papers/jss/article.tex
@@
-%&LaTeX
+% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
+
 \documentclass[article]{jss}
@@
 
+\SweaveOpts{echo=FALSE}
+\usepackage{a4wide}
+
+% \batchmode
+
 \begin{document}
+
+\maketitle
 
@@
 
-\section{Introduction}
-%% Note: If there is markup in \(sub)section, then it has to be escape as above.
+A simple example that will run in any S engine: The integers from 1 to
+10 are
+<<print=TRUE>>=
+1:10
+<<results=hide>>=
+print(1:20)
+@ % the above is just to ensure that 2 code chunks can follow each other
 
-Markov and latent Markov models are frequently used in the social
-sciences, in different areas and applications.  In psychology, they
-are used for modelling learning processes, see \citet{Wickens1982},
-for an overview, and e.g.\ \citet{Schmittmann2006} for a recent
-application.  In economics, latent Markov models are so-called regime
-switching models (see e.g., \citealp{Kim1994} and
-\citealp{Ghysels1994}).  Further applications include speech
-recognition \citep{Rabiner1989}, EEG analysis \citep{Rainer2000}, and
-genetics \citep{Krogh1998}.  In those latter areas of application,
-latent Markov models are usually referred to as hidden Markov models.
-For more examples of applications, see e.g.\
-\citet[][chapter~1]{Cappe2005}.
+We can also emulate a simple calculator:
+<<echo=TRUE,print=TRUE>>=
+1 + 1
+1 + pi
+sin(pi/2)
+@
 
-The \pkg{depmixS4} package was motivated by the fact that while Markov
-models are used commonly in the social sciences, no comprehensive
-package was available for fitting such models.  Common software for
-estimating Markovian models include Panmark \citep{Pol1996}, and for
-latent class models Latent Gold \citep{Vermunt2003}.  Those programs
-are lacking a number of important features, besides not being freely
-available.  There are currently some packages in \proglang{R} that
-handle hidden Markov models but they lack a number of features that we
-needed in our research.  In particular, \pkg{depmixS4} was designed to
-meet the following goals:
+Now we look at Gaussian data:
 
-\begin{enumerate}
+<<>>=
+library(stats)
+x <- rnorm(20)
+print(x)
+print(t1 <- t.test(x))
+@
+Note that we can easily integrate some numbers into standard text: The
+third element of vector \texttt{x} is \Sexpr{x[3]}, the
+$p$-value of the test is \Sexpr{format.pval(t1$p.value)}. % $
 
-	\item to be able to estimate parameters subject to general
-	linear (in)equality constraints
+Now we look at a summary of the famous iris data set, and we want to
+see the commands in the code chunks:
 
-	\item to be able to fit transition models with covariates, i.e.,
-	to have time-dependent transition matrices
+\SweaveOpts{echo=true}
 
-	\item to be able to include covariates in the prior or initial
-	state probabilities of models
-
-	\item to be easily extensible, in particular, to allow users to
-	easily add new uni- or multivariate response distributions, and
-	similarly for the addition of other transition models, e.g.,
-	continuous time observation models
-
-\end{enumerate}
-
-Although \pkg{depmixS4} was designed to deal with longitudinal or time
-series data, for say $T>100$, it can also handle the limit case when
-$T=1$.  In this case, there are no time dependencies between
-observed data and the model reduces to a finite mixture or
-latent class model. Although there are other specialized packages to
-deal with mixture data, as far as we know these 
-don't allow the inclusion of covariates on the prior probabilities of
-class membership. The possibility to estimating the effects of covariates on 
-prior and transition probabilities is a distinguishing feature of 
-\pkg{depmixS4}. In the next section, an outline is provided of the model and 
-the likelihood equations.
-
-\section{The dependent mixture model}
-
-The data considered here have the general form $\vc{O}_{1:T}=O_{1}^{1}, \ldots,
-O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$, \ldots, $O_{T}^{1},
-\ldots, O_{T}^{m}$ for an $m$-variate time series of length $T$.  As
-an example, consider a time series of responses generated by a single
-participant in a response time experiment. The data consists of three
-variables, response time, response accuracy, and a covariate which is a pay-off
-variable reflecting the relative reward for speeded and/or accuracy responding. 
-These variables are measured on 168, 134 and 137 occasions respectively (in
-Figure~\ref{fig:speed} the first part of this series is plotted). These data
-are more fully described in \citet{Dutilh2009}.
+% the following code is R-specific, as data(iris) will not run in Splus. 
+% Hence, we mark it as R code. 
+<<engine=R>>=
+data(iris)
+summary(iris)
+@ %def
 
@@
   \begin{center}
-	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
-	  \caption{Response times (rt), accuracy (corr) and pay-off values (Pacc) for
-	  the first series of responses in dataset \code{speed}.}
-	  \label{fig:speed}
+<<fig=TRUE>>=
+library(graphics)
+pairs(iris)
+@
+    \caption{Pairs plot of the iris data.}
   \end{center}
@@
 
-The latent Markov model is commonly associated with data of this type,
-although usually only multinomial response variables are considered.
-However, common estimation procedures, such as those implemented in
-\citet{Pol1996}, are not suitable for long time series due to
-underflow problems.  In contrast, the hidden Markov model is typically
-only used for `long' univariate time series \citep[][,
-chapter~1]{Cappe2005}.  In the next paragraphs, the likelihood
-equation and estimation procedures for the dependent mixture model are
-described for data of the above form.  We use the term ``dependent
-mixture model'' because one of the authors (Ingmar Visser) thought it
-was time for a new name for these models\footnote{Only later did I
-find out that \citet{Leroux1992} already coined the term dependent
-mixture models in an application with hidden Markov mixtures of
-Poisson count data.}.
+\begin{figure}[htbp]
+  \begin{center}
+<<fig=true>>=
+boxplot(Sepal.Length~Species, data=iris)
+@
+    \caption{Boxplot of sepal length grouped by species.}
+  \end{center}
+\end{figure}
 
-The likelihood of the dependent mixture model conditional on the
-unknown (or hidden) state sequence $S_{1:T}$ and the model is given
-by:
-\begin{equation}
-	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
-	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
-\end{equation}
-where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
-at the likelihood of the data given the parameter vector
-$\greekv{\theta}$, i.e.\ without the state sequence, we need to sum
-above likelihood over all possible state sequences. This likelihood 
-is written as: 
-\begin{equation}
-	\Prob(\vc{O}_{1:T}|\vc{S}_{1:T},\greekv{\theta}) 
-	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
-	\prod_{t=1}^{T-1} a_{ij}(z_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
-\end{equation}
-where we have the following elements:
-\begin{enumerate}
-
-	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
-	of $n$ latent classes or states; we write for short $S_{i}$ to
-	denote $S_{t}=i$.
-
-	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
-	giving the probability of class/state $i$ at time $t=1$ with
-	covariate $\vc{z}_1$.
-
-	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
-	provides the probability of a transition from state $i$ to state
-	$j$ with covariate $\vc{z}_t$,
-
-	\item $\vc{b}_{j}$ is a vector of observation densities
-	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
-	conditional densities of observations $O_{t}^k$ associated with
-	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
-	$k=1, \ldots, m$. 
-\end{enumerate}
-
-% The dependent mixture model is defined by the following equations: 
-% %\begin{align}
-% %	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
-% %	O_{t} &= b(O_{t}|S_{t}), 
-% %\end{align}
-% \begin{align}
-% 	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
-% 	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
-% \end{align}
-%where $S_{t}$ is a sequence of hidden states, $A$ is a transition
-%matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
-%density function for $O_{t}$ conditional on the hidden state $S_{t}$.
-
-In the example data above, $b_j^k$ could be a Gaussian distribution
-function for the response time variable, and a Bernoulli for the
-accuracy data.  In the models we are considering here, both the
-transition probability functions $a_{ij}$and the initial state
-probability functions $\greekv{\pi}$ may depend on covariates as well
-as the response distributions $\vc{b_{j}^{k}}$.  See for example
-\citet{Fruhwirth2006} for an overview of hidden Markov models with
-extensions.
-
-\subsection{Likelihood}
-
-The log-likelihood of hidden Markov models is usually computed by the
-so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or
-rather by the forward part of this algorithm.  \cite{Lystig2002}
-changed the forward algorithm in such a way as to allow computing the
-gradients of the log-likelihood at the same time.  They start by
-rewriting the likelihood as follows (for ease of exposition the
-dependence on the model parameters is dropped here):
-\begin{equation}
-	L_{T} = Pr(\vc{O}_{1:T}) = \prod_{t=1}^{T} 
-	Pr(\vc{O}_{t}|\vc{O}_{1:t-1}), 
-	\label{condLike}
-\end{equation}
-where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
-simple, i.e.\ observed, Markov chain these probabilities reduce to 
-$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
-\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
-The log-likelihood can now be expressed as:
-\begin{equation}
-	l_{T} = \sum_{t=1}^{T} \log[\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
-\vc{O}_{t-1})].
-	\label{eq:condLogl}
-\end{equation}
-
-To compute the log-likelihood, \cite{Lystig2002} define the following 
-(forward) recursion:
-\begin{align}
-	\phi_{1}(j) &:= \Prob(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
-	\label{eq:fwd1} \\
-\begin{split}
-	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
-\vc{O}_{t-1}) \\
-	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
-(\Phi_{t-1})^{-1},
-	\label{eq:fwdt} 
-\end{split} 
-\end{align}
-where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
-$\Phi_{t}=\Prob(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
-equation~(\ref{eq:condLogl}) gives the following expression for the 
-log-likelihood:
-\begin{equation}
-	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
-	\label{eq:logl}
-\end{equation}
-
-\subsection{Parameter estimation}
-
-Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
-through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
-parameters are estimated by iteratively maximising the 
-expected joint likelihood of the parameters given the observations and states. 
-Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
-be the general parameter vector consisting of three subvectors with parameters 
-for the prior model, transition model, and response model respectively. The 
-joint log likelihood can be written as
-\begin{equation}
-\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
-\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
-+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
-+ \sum_{t=1}^{T} \log \Prob(O_t|S_t,\vc{z}_t,\greekv{\theta}_3)
-\end{equation}
-This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
-Expectation step, we replace these with their expected values given a set of 
-(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
-\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $O_{1:T}$. The expected 
-log likelihood 
-\begin{equation}
-Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
-(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
-\end{equation}
-can be written as
-%\begin{equation}
-\begin{multline}
-\label{eq:Q}
-Q(\greekv{\theta},\greekv{\theta}') = 
-\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
-+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t^i(j,k) \log \Prob(S_t = k|S_{t-1} 
-= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
- + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
-\ln \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
-\end{multline}
-%\end{equation}
-where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|\vc{O}_{1:T},
-\vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|\vc{O}_{1:T},
-\vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
-Forward-Backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
-step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
-right hand side of (\ref{eq:Q}) consists of three separate parts, we can 
-maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and 
-$\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and 
-$\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in the 
-\pkg{nnet} package \citep{Venables2002}, and maximisation for 
-$\greekv{\theta}_3$ by the standard \code{glm} routine. Note that for the latter 
-maximisation, the expected values $\gamma_t(j)$ are used as prior weights of the 
-observations $O^k_t$.
-
-%The maximising values of $a_{jk}$ are \cite<e.g.,>{Rabiner89}
-%\begin{equation}
-%\label{eq:MLA}
-%\hat{a}_{jk} = \frac{1}{N(T-1)} \sum_{i=1}^N \sum_{t=2}^T  \frac{\xi^i_t(j,k)}{\gamma_{t-1}^i(j)}.
-%\end{equation}
-%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does not affect the estimation of the other elements.
-When elements $a_{jk}$ are assumed identical, we simply extend the summation in (\ref{eq:MLA}) to include all those elements, changing the denominator $N(T-1)$ accordingly.
-
-%To estimate $\greekv{\lambda}$, we note that the term containing this parameter has the form of a weighted likelihood, where $\gamma_t^i(j)$ can be interpreted as the number of replications of $r^i_t$.
-Hence, we can rely on standard ML estimates of the logistic regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$ as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To be more specific, we replicate each observation $r^i_t$ a total of 6 times, once for each of the states besides the random state (which offers no information regarding $\lambda_i$).
-For the $j$-th replication (corresponding to the $j$-th state), we used $v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a case weight.
-All these replications were used to obtain the maximum likelihood estimate of $\lambda$ from a single GLM, using the ``glm.fit'' function in R \cite{R}.}.
-The resulting parameter estimates are then used in a new Expectation step, followed by another Maximisation step, until convergence.
-Under mild regularity conditions, it can be shown that the EM algorithm converges to a (local) maximum of the likelihood \cite{EM}.
-
-The EM algorithm however has some drawbacks.  First, it can be slow to
-converge towards the end of optimization.  Second,
-applying constraints to parameters can be problematic; in particular,
-EM can lead to wrong parameter estimates when applying constraints.
-Hence, in \pkg{depmixS4}, EM is used by default in unconstrained
-models, but otherwise, direct optimization is done using \pkg{Rdonlp2}
-\citep{Tamura2009,Spellucci2002}, because it handles general linear
-(in)equality constraints, and optionally also non-linear constraints.
-
-%Need some more on EM and how/why it is justified to do separate weighted
-%iuuuuuuuuuuucvfits of the response models and transition and prior models.
-
-%Also mention use of glm, nnet and possibly other packages that we use.
-
-\section[Using depmixS4]{Using \pkg{depmixS4}}
-
-Two steps are involved in using \pkg{depmixS4} which are illustrated
-below with examples:
-\begin{enumerate}
-	\item model specification with function \code{depmix} (or with \code{mix}
-	for latent class and finite mixture models, see example below on adding
-	covariates to prior probabilities)
-
-	\item  model fitting with function \code{fit}
-\end{enumerate}
-We have separated the stages of model specification and model fitting because
-fitting large models can be fairly time-consuming and it is hence useful to be
-able to check the model specification before actually fitting the model.
-
-\subsection{Example data: speed}
-
-Throughout this article a data set called \code{speed} is used.  It
-consists of three time series with three variables: response time,
-accuracy, and a covariate Pacc which defines the relative pay-off for
-speeded and accurate responding.  The participant in this experiment
-switches between fast responding at chance level and relatively slower
-responding at a high level of accuracy.  Interesting hypotheses to
-test are: is the switching regime symmetric?  Is there evidence for
-two states or does one state suffice?  Is the guessing state actually
-a guessing state, i.e., is the probability of a correct response at
-chance level (0.5)?
-
-\subsection{A simple model}
-
-A dependent mixture model is defined by the number of states, and by
-the response distribution functions, and can be created with the
-\code{depmix}-function as follows:
-
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
-\end{CodeInput}
-\end{CodeChunk}
-
-The \code{depmix} function returns an object of class \code{depmix}
-which contains the model specification (and not a fitted model!).
-Note also that start values for the transition parameters are provided
-in this call using the \code{trstart} argument. At this time, the package does 
-not provide automatic starting values.
-
-The so-defined models needs to be \code{fit}ted with the following
-line of code:
-\begin{CodeChunk}
-\begin{CodeInput}
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-
-The \code{fit}-function returns an object of class
-\code{depmix.fitted} which extends the \code{depmix} class, adding
-convergence information (and information about constraints if these
-were applied, see below).  The \code{print} method provides summary
-information on convergence, the log likelihood and the AIC and BIC
-values:
-\begin{CodeChunk}
-\begin{CodeInput}
-> fm
-\end{CodeInput}
-\begin{CodeOutput}
-Convergence info: Log likelihood converged to within tol. 
-'log Lik.' -84.34175 (df=7)
-AIC:  182.6835 
-BIC:  211.275 
-\end{CodeOutput}
-\end{CodeChunk}
-These statistics can also be extracted using \code{logLik},
-\code{AIC} and \code{BIC}, respectively.
-
-The \code{summary} method of \code{fit}ted models provides the parameter
-estimates, first for the prior probabilities model, second for the
-transition model, and third for the response models.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
-Initial state probabilties model 
-Model of type multinomial, formula: ~1
-Coefficients: 
-     [,1]      [,2]
-[1,]    0 -11.25688
-Probalities at zero values of the covariates.
-0.999987 1.291798e-05
-
-Transition model for state (component) 1 
-Model of type multinomial, formula: ~1
-Coefficients: 
-[1]  0.000000 -2.392455
-Probalities at zero values of the covariates.
-0.9162501 0.08374986
-
-Transition model for state (component) 2 
-Model of type multinomial, formula: ~1
-Coefficients: 
-[1] 0.000000 2.139255
-Probalities at zero values of the covariates.
-0.1053396 0.8946604
-
-Response model(s) for state 1
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 6.385492
-sd  0.2439376
-
-Response model(s) for state 2
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 5.511151
-sd  0.1926063 
-\end{CodeOutput}
-\end{CodeChunk}
-
-\subsection{Covariates on transition parameters}
-
-The transition probabilities and the initial state probabilities
-are parameterized using the multinomial
-logistic model.   In particular, each row of the transition matrix is
-parameterized by a baseline category logistic multinomial, 
-meaning that the parameter for the
-base category is fixed at zero \citep[see][p.\ 267
-ff., for multinomial logistic models and various
-parameterizations]{Agresti2002}. See also \citet{Chung2007} for similar models, 
-latent transition models using logistic regression on the transition parameters. 
-They fit such models on repeated measurement
-data ($T=2$) using Bayesian methods.  The default baseline category is the 
-first state.
-Hence, for example, for a 3-state model, the initial state probability
-model would have three parameters of which the first is fixed at zero
-and the other two are freely estimated.
-
-Covariates on the transition probabilities can be specified using a
-one-sided formula as in the following example:
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
-  transition=~scale(Pacc), instart=runif(2))
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-
-Applying \code{summary} to the \code{fit}ted object gives (only transition models
-printed here): 
-\begin{CodeChunk}
-\begin{CodeOutput}
-...
-Transition model for state (component) 1 
-Model of type multinomial, formula: ~scale(Pacc)
-Coefficients: 
-	 [,1]       [,2]
-[1,]    0 -0.9215182
-[2,]    0  1.8649734
-Probalities at zero values of the covariates.
-0.7153513 0.2846487
-
-Transition model for state (component) 2 
-Model of type multinomial, formula: ~scale(Pacc)
-Coefficients: 
-	 [,1]     [,2]
-[1,]    0 2.471442
-[2,]    0 3.570856
-Probalities at zero values of the covariates.
-0.07788458 0.9221154
-...
-\end{CodeOutput}
-\end{CodeChunk}
-The summary provides all parameters of the model, also the 
-(redundant) zeroes for the base-line category in the multinomial model. 
-The summary also prints the transition probabilities
-at the zero value of the covariate. Note that scaling of the covariate 
-is useful in this regard as it makes interpretation of these intercept probabilities 
-easier.
-
-\subsection{Multivariate data}
-
-Multivariate data can be modelled by providing a list of formulae as 
-well as a list of family objects for the distributions of the various 
-responses. In above examples we have only used the response times 
-which were modelled with the Gaussian distribution. The accuracy data 
-are in the \code{speed} data are modelled with a multinomial by 
-specifying the following: 
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
-	  family=list(gaussian(),multinomial()),
-	  transition=~scale(Pacc),instart=runif(2))
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-which provides the following fitted model parameters (only the 
-response parameters are given here): 
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
-...
-Response model(s) for state 1
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 5.52169
-sd  0.2028857
-
-Response model for response 2 
-Model of type multinomial, formula: corr ~ 1
-Coefficients: 
-	 [,1]      [,2]
-[1,]    0 0.1030554
-Probalities at zero values of the covariates.
-0.4742589 0.5257411
-
-Response model(s) for state 2
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 6.39369
-sd  0.2373650
-
-Response model for response 2 
-Model of type multinomial, formula: corr ~ 1
-Coefficients: 
-	 [,1]     [,2]
-[1,]    0 2.245514
-Probalities at zero values of the covariates.
-0.09573715 0.9042629 	
-\end{CodeOutput}
-\end{CodeChunk}
-As can be seen, state 1 has fast response times around chance level, 
-whereas state 2 corresponds with slower responding at higher accuracy 
-levels.
-
-Note that by specifying multivariate observations in terms of a list, the 
-variables are considered conditionally independent (given the states). 
-Conditionally \emph{dependent} variables must be handled as a single element in
-the list. Effectively, this means specifying a multivariate response model. The 
-only multivariate response model currently implemented in \pkg{depmix} is for
-multivariate normal variables.
-
-\subsection{Adding covariates on the prior probabilities}
-
-To illustrate the use of covariates on the prior probabilities we have included
-another data set with \pkg{depmixS4}. The \code{balance} data consists 
-of 4 binary items (correct-incorrect) on a balance scale task 
-\citet{Siegler1981}. The data form a subset of the data published in 
-\citet{Jansen2002}.
-
-Similarly to the transition matrix, covariates on the prior 
-probabilities of the latent states (or classes in this case), are 
-defined by using a one-sided formula: 
-\begin{CodeChunk}
-\begin{CodeInput}
-> balance$age <- balance$age-5
-> set.seed(1)
-> mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
-    family=list(multinomial(), multinomial(), multinomial(),
-    multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
-    prior=~age, initdata=balance)
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-Note here that we define a \code{mix} model instead of a \code{depmix}
-models as these data form independent observations.
-
-The summary of the \code{fit}ted model gives (only the prior model is 
-shown here): 
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
-Mixture probabilities model 
-Model of type multinomial, formula: ~age
-   [,1]   [,2]
-[1,]    0 -2.5182573
-[2,]    0  0.5512996
-Probalities at zero values of the covariates.
-0.9254119 0.07458815 
-...
-\end{CodeOutput}
-\end{CodeChunk}	
-Hence at young ages, children have a high probability of incorrect 
-responding in class~1, whereas the prior probability for class~2 
-increases with age.
-
-\subsection{Fixing and constraining parameters}
-
-Using package \pkg{Rdonlp2} by \citet{Tamura2009}, parameters may be fitted subject to
-general linear (in-)equality constraints.  Constraining and fixing
-parameters is done using the \code{conpat} argument to the
-\code{fit}-function, which specifies for each parameter in the
-model whether it's fixed (0) or free (1 or higher).  Equality
-constraints can be imposed by having two parameters have the same
-number in the \code{conpat} vector.  When only fixed values are
-required, the \code{fixed} argument can be used instead of
-\code{conpat}, with zeroes for fixed parameters and other values (ones
-e.g.) for non-fixed parameters.  Fitting the models subject to these
-constraints is handled by the optimization routine \code{donlp2}.
-To be able to construct the \code{conpat} and/or \code{fixed} vectors 
-one needs the correct ordering of parameters which is briefly discussed 
-next before proceeding with an example.
-
-\paragraph{Parameter numbering} When using the \code{conpat} and
-\code{fixed} arguments, complete parameter vectors should be supplied,
-i.e., these vectors should have length of the number of parameters of
-the model, which can be obtained by calling \code{npar(object)}. Note that
-this is not the same as the degrees of freedom used e.g.\ in the \code{logLik}
-function because \code{npar} also counts the baseline category zeroes
-from the multinomial logistic models. 
-Parameters are numbered in the following order:
-\begin{enumerate}
-	\item  the prior model parameters
-	\item  the parameters for the transition models
-	\item  the response model parameters per state (and subsequently
-	per response in the case of multivariate time series)
-\end{enumerate}
-
-To see the ordering of parameters use the following:
-\begin{CodeChunk}
-\begin{CodeInput}
-> mod <- setpars(mod, value=1:npar(mod))
-> mod
-\end{CodeInput}
-\end{CodeChunk}
-
-To see which parameters are fixed (by default only baseline parameters
-in the multinomial logistic models for the transition models and the
-initial state probabilities model):
-\begin{CodeChunk}
-\begin{CodeInput}
-> mod <- setpars(mod, getpars(mod,which="fixed"))
-> mod
-\end{CodeInput}
-\end{CodeChunk}
-
-When fitting constraints it is useful to have good starting values 
-for the parameters and hence we first fit the following model without
-constraints:
-\begin{CodeChunk}
-\begin{CodeInput}
-> trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
-> mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
-	  nstates=2,family=list(gaussian(),multinomial()),
-	  trstart=trst,inst=c(.999,0.001))
-> fm1 <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-After this, we use the fitted values from this model to constrain the 
-regression coefficients on the transition matrix (parameters numbers~6 and~10):
-\begin{CodeChunk}
-\begin{CodeInput}
-# start with fixed and free parameters
-> conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
-# constrain the beta's on the transition parameters to be equal
-> conpat[6] <- conpat[10] <- 2
-> fm2 <- fit(fm,equal=conpat)
-\end{CodeInput}
-\end{CodeChunk}
-Using \code{summary} on the fitted model shows that the regression 
-coefficients are now estimated at the same value of 12.66. The function
-\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
-associated $p$-value with appropriate degrees of freedom for testing the
-tenability of constraints \citep{Dannemann2007}. Note that these arguments 
-provide the possibility for arbitrary 
-constraints, also between, e.g., a multinomial regression coefficient 
-for the transition matrix and the mean of a gaussian response model. 
-Whether such constraints make sense is hence the responsibility of 
-the user.
-
-\section[Extending depmixS4]{Extending \pkg{depmixS4}}
-
-The \pkg{depmixS4} package was built with the aim of having the 
-possibility of adding new response distributions (and possibly also 
-other models for the prior and transition probabilities models). 
-Therefore, the EM algorithm simply calls \code{fit} functions from the 
-separate response models without knowing what they are. As a 
-consequence, adding user-specified response models is straightforward. 
-User-defined distributions should extend the \code{response}-class 
-and have the following slots:
-\begin{enumerate}
-	\item y: the response variable
-	\item x: the design matrix, possibly only an intercept
-	\item paramaters: a named list with the coefficients and possibly 
-	other parameters, e.g., the standard deviation in the gaussian 
-	response model
-	\item fixed: a vector of logicals indicating whether parameters are 
-	fixed
-	\item npar: numerical indicating the number of parameters of the model
-\end{enumerate}
-
-In the \code{speed} data example, it may be more appropriate to model
-the response times as an exgaus distribution rather than using the
-Gaussian.  We can do so as follows, first we define our own
-exgaus-class extending the response-class:
-\begin{CodeChunk}
-	setClass("exgaus", contains="response")
-\end{CodeChunk}
-
-The so-defined class now needs a number of methods: 
-\begin{enumerate}
-	\item constructor: function to create instances of the class 
-	with starting values
-	\item show method: to print the model to the terminal
-	\item dens: the function that computes the density of the responses
-	\item getpars and setpars: to get and set parameters 
-	\item predict: to generate predict'ed values 
-	\item fit: function to fit the model using posterior weights (used 
-	by the EM algorithm)
-\end{enumerate}
-
-Only the constructor function and the fit function are provided here
-and the complete code is provided in the helpfile for \code{makeDepmix}. 
-The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} packages \cite{Stasinopoulos2009a,Stasinopoulos2009b} for 
-computing the \code{dens}ity and for \code{fit}ting the parameters.
-
-The constructor function is defined as: 
-\begin{CodeChunk}
-\begin{CodeInput}
-setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
-	standardGeneric("exgaus"))
-setMethod("exgaus", 
-	signature(y="ANY"), 
-	function(y,pstart=NULL,fixed=NULL, ...) {
-		y <- matrix(y,length(y))
-		x <- matrix(1)
-		parameters <- list()
-		npar <- 3
-		if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
-		if(!is.null(pstart)) {
-			if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
-			parameters$mu <- pstart[1]
-			parameters$sigma <- log(pstart[2])
-			parameters$nu <- log(pstart[3])
-		}
-		mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
-		mod
-	}
-)
-\end{CodeInput}
-\end{CodeChunk}
-
-The \code{fit} function is defined as follows: 
-\begin{CodeChunk}
-\begin{CodeInput}
-setMethod("fit","exgaus",
-	function(object,w) {
-		if(missing(w)) w <- NULL
-		y <- object@y
-		fit <- gamlss(y~1,weights=w,family=exGAUS(),
-			control=gamlss.control(n.cyc=100,trace=FALSE),
-			mu.start=object@parameters$mu,
-			sigma.start=exp(object@parameters$sigma),
-			nu.start=exp(object@parameters$nu))
-		pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
-		object <- setpars(object,pars)
-		object
-	}
-)
-\end{CodeInput}
-\end{CodeChunk}
-
-The \code{fit} function defines a \code{gamlss} model with 
-only an intercept to be estimated and then sets the fitted parameters 
-back into their respective slots in the `exgaus' object. See the help 
-for \code{gamlss.distr} for interpretation of these parameters.
-
-After defining all the necessary methods for the new response model, 
-we can  now define the dependent mixture model using this reponse. 
-The function \code{makeDepmix} is added to \pkg{depmixS4} to have 
-full control over model specification, and we need it here.
-
-We first create all the response models that we want as a double list: 
-\begin{CodeChunk}
-\begin{CodeInput}
-rModels <- list(
-	list(
-		exgaus(rt,pstart=c(5,.1,.1)),
-		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
-	),
-	list(
-		exgaus(rt,pstart=c(6,.1,.1)),
-		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
-	)
-)
-\end{CodeInput}
-\end{CodeChunk}
-Next, we define the transition and prior probability models using the 
-transInit function (which produces a transInit model, which also extends 
-the response class): 
-\begin{CodeChunk}
-\begin{CodeInput}
-trstart=c(0.9,0.1,0.1,0.9)
-transition <- list()
-transition[[1]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.9,0.1,0,0))
-transition[[2]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.1,0.9,0,0))
-inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
-\end{CodeInput}
-\end{CodeChunk}
-Finally, we put everything together using \code{makeDepmix} and fit 
-the model: 
-\begin{CodeChunk}
-\begin{CodeInput}
-mod <- makeDepmix(response=rModels,transition=transition,
-prior=inMod,stat=FALSE)
-fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-
-Using \code{summary} will print the fitted parameters. Note that the 
-use of \code{makeDepmix} allows the possibility of, say, fitting a 
-Gaussian in one state and an exgaus distribution in another state.
-
-\section[Conclusions and future work]{Conclusions \& future work}
-
-\pkg{depmixS4} provides a flexible framework for fitting dependent mixture models
-for a large variety of response distributions. It can also fit latent class regression
-and finite mixture models, although it should be noted that more specialized packages 
-are available for doing so, e.g.\ \pkg{FlexMix} \citep{Leisch2004}. The package is intended
-for modeling of (individual) time series data with the aim of characterizing the transition 
-processes underlying the data. The possibility to use covariates on the transition matrix
-greatly enhances the flexibility of the model. The EM algorithm uses a very general 
-interface that allows easy addition of new response models.
-
-We are currently working on implementing the gradients for response and transition models 
-with two goals in mind. First, to speed up (constrained) parameter optimization using 
-\pkg{Rdonlp2}. Second, analytic gradients are useful in computing the Hessian of the
-estimated parameters so as to arrive at standard errors for those. We are also planning to
-implement goodness-of-fit statistics \citep{Titman2008}.
-
-\section*{Acknowledgements}
-
-Ingmar Visser was supported by an EC Framework 6 grant, project 516542
-(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
-Economic Learning and Social Evolution (ELSE).  Han van der Maas
-provided the speed-accuracy data \cite{Dutilh2009} and thereby
-neccessitated implementing models with time-dependent covariates.
-Brenda Jansen provided the balance scale data set \citep{Jansen2002}
-which was the perfect opportunity to test the covariates on the prior
-model parameters.  The examples in the help files use both of these
-data sets.
-
-\bibliography{all,ingmar}
+% R is not S-PLUS, hence this chunk will be ignored:
+<<engine=S4>>=
+function.that.comes.only.with.Splus(x)
+@
 

### 2010-02-22 papers/jss/dpx4Rev.tex
Sweave related files.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@357 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2010-02-22 papers/jss/dpx4Rev.Rnw
@@
 
-\SweaveOpts{echo=FALSE}
 \usepackage{a4wide}
@@
 
+\usepackage{Sweave}
 \begin{document}
@@
 10 are
-<<print=TRUE>>=
-1:10
-<<results=hide>>=
-print(1:20)
-@ % the above is just to ensure that 2 code chunks can follow each other
+\begin{Schunk}
+\begin{Soutput}
+ [1]  1  2  3  4  5  6  7  8  9 10
+\end{Soutput}
+\end{Schunk}
 
 We can also emulate a simple calculator:
-<<echo=TRUE,print=TRUE>>=
-1 + 1
-1 + pi
-sin(pi/2)
-@
+\begin{Schunk}
+\begin{Sinput}
+> 1 + 1
+\end{Sinput}
+\begin{Soutput}
+[1] 2
+\end{Soutput}
+\begin{Sinput}
+> 1 + pi
+\end{Sinput}
+\begin{Soutput}
+[1] 4.141593
+\end{Soutput}
+\begin{Sinput}
+> sin(pi/2)
+\end{Sinput}
+\begin{Soutput}
+[1] 1
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<>>=
-library(stats)
-x <- rnorm(20)
-print(x)
-print(t1 <- t.test(x))
-@
+\begin{Schunk}
+\begin{Soutput}
+ [1]  0.36039435 -1.83775608  1.04278935 -1.07035396 -0.82150164 -0.32257866
+ [7]  0.97533933 -0.97446747  1.25840051 -1.01986278  1.15659229 -0.71535551
+[13]  0.76128983  0.35511419  0.30919687 -1.45434289 -0.09874472 -0.13624640
+[19] -1.57319360  1.32749121
+\end{Soutput}
+\begin{Soutput}
+	One Sample t-test
+
+data:  x 
+t = -0.5432, df = 19, p-value = 0.5933
+alternative hypothesis: true mean is not equal to 0 
+95 percent confidence interval:
+ -0.6012708  0.3534912 
+sample estimates:
+ mean of x 
+-0.1238898 
+\end{Soutput}
+\end{Schunk}
 Note that we can easily integrate some numbers into standard text: The
-third element of vector \texttt{x} is \Sexpr{x[3]}, the
-$p$-value of the test is \Sexpr{format.pval(t1$p.value)}. % $
+third element of vector \texttt{x} is 1.04278935363803, the
+$p$-value of the test is 0.59332. % $
 
@@
 
-\SweaveOpts{echo=true}
-
 % the following code is R-specific, as data(iris) will not run in Splus. 
 % Hence, we mark it as R code. 
-<<engine=R>>=
-data(iris)
-summary(iris)
-@ %def
+\begin{Schunk}
+\begin{Sinput}
+> data(iris)
+> summary(iris)
+\end{Sinput}
+\begin{Soutput}
+  Sepal.Length    Sepal.Width     Petal.Length    Petal.Width   
+ Min.   :4.300   Min.   :2.000   Min.   :1.000   Min.   :0.100  
+ 1st Qu.:5.100   1st Qu.:2.800   1st Qu.:1.600   1st Qu.:0.300  
+ Median :5.800   Median :3.000   Median :4.350   Median :1.300  
+ Mean   :5.843   Mean   :3.057   Mean   :3.758   Mean   :1.199  
+ 3rd Qu.:6.400   3rd Qu.:3.300   3rd Qu.:5.100   3rd Qu.:1.800  
+ Max.   :7.900   Max.   :4.400   Max.   :6.900   Max.   :2.500  
+       Species  
+ setosa    :50  
+ versicolor:50  
+ virginica :50  
+\end{Soutput}
+\end{Schunk}
 
@@
   \begin{center}
-<<fig=TRUE>>=
-library(graphics)
-pairs(iris)
-@
+\begin{Schunk}
+\begin{Sinput}
+> library(graphics)
+> pairs(iris)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{dpx4Rev-006}
     \caption{Pairs plot of the iris data.}
@@
   \begin{center}
-<<fig=true>>=
-boxplot(Sepal.Length~Species, data=iris)
-@
+\begin{Schunk}
+\begin{Sinput}
+> boxplot(Sepal.Length ~ Species, data = iris)
+\end{Sinput}
+\end{Schunk}
+\includegraphics{dpx4Rev-007}
     \caption{Boxplot of sepal length grouped by species.}
@@
 % R is not S-PLUS, hence this chunk will be ignored:
-<<engine=S4>>=
-function.that.comes.only.with.Splus(x)
-@
 

### 2010-02-23 papers/jss/dpx4Rev.Rnw
Completed turning paper into Sweave format; .sh file contains cmds to generate the paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@360 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2009-08-19 papers/jss/article.tex
@@
 
+\SweaveOpts{echo=FALSE}
+\usepackage{a4wide}
+
+%\usepackage{Sweave}
+
 \begin{document}
+
+%set width of figures produced by Sweave
+\setkeys{Gin}{width=0.9\textwidth}
+
+\maketitle
 
@@
 \begin{center}
-	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
+<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=
+library(depmixS4)
+data(speed)
+plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
+@
 	 \caption{Response times (rt), accuracy (corr) and pay-off values (Pacc) for
@@
 	  \label{fig:speed}
+
 \end{center}
@@
 %Need some more on EM and how/why it is justified to do separate weighted
-%iuuuuuuuuuuucvfits of the response models and transition and prior models.
+%fits of the response models and transition and prior models.
 
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
-\end{CodeInput}
-\end{CodeChunk}
+<<>>=
+library(depmixS4)
+data(speed)
+set.seed(1)
+mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
+@
 
@@
 
-The so-defined model needs to be \code{fit}ted with the following
-line of code:
-\begin{CodeChunk}
-\begin{CodeInput}
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-
+<<>>=
+fm <- fit(mod)
+@
 The \code{fit}-function returns an object of class
@@
 values:
-\begin{CodeChunk}
-\begin{CodeInput}
-> fm
-\end{CodeInput}
-\begin{CodeOutput}
-Convergence info: Log likelihood converged to within tol. 
-'log Lik.' -84.34 (df=7)
-AIC:  182.68 
-BIC:  211.28 
-\end{CodeOutput}
-\end{CodeChunk}
+<<>>=
+fm
+@
+
 These statistics can also be extracted using \code{logLik}, \code{AIC}
@@
 for the transition model, and third for the response models.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
-Initial state probabilties model 
-Model of type multinomial, formula: ~1
-Coefficients: 
-     [,1]      [,2]
-[1,]    0 -11.25
-Probalities at zero values of the covariates.
-0.999 1.292e-05
-
-Transition model for state (component) 1 
-Model of type multinomial, formula: ~1
-Coefficients: 
-[1]  0.000 -2.392
-Probalities at zero values of the covariates.
-0.9163 0.0837
-
-Transition model for state (component) 2 
-Model of type multinomial, formula: ~1
-Coefficients: 
-[1] 0.000 2.139
-Probalities at zero values of the covariates.
-0.105 0.895
-
-Response model(s) for state 1
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 6.385
-sd  0.244
-
-Response model(s) for state 2
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 5.511
-sd  0.193 
-\end{CodeOutput}
-\end{CodeChunk}
+<<>>=
+summary(fm)
+@
 
@@
 following example:
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
+<<>>=
+set.seed(1)
+mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
     transition=~scale(Pacc), instart=runif(2))
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+fm <- fit(mod)
+@
 
@@
 printed here): 
-\begin{CodeChunk}
-\begin{CodeOutput}
-...
-Transition model for state (component) 1 
-Model of type multinomial, formula: ~scale(Pacc)
-Coefficients: 
-     [,1]       [,2]
-[1,]    0 -0.9215
-[2,]    0  1.865
-Probalities at zero values of the covariates.
-0.7154 0.2846
+<<>>=
+summary(fm)
+@
 
-Transition model for state (component) 2 
-Model of type multinomial, formula: ~scale(Pacc)
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 2.471
-[2,]    0 3.571
-Probalities at zero values of the covariates.
-0.0779 0.9221
-...
-\end{CodeOutput}
-\end{CodeChunk}
 The summary provides all parameters of the model, also the (redundant)
@@
 by specifying the following:
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
+<<>>=
+set.seed(1)
+mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
     family=list(gaussian(),multinomial()),
     transition=~scale(Pacc),instart=runif(2))
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+fm <- fit(mod)
+@
+
 This provides the following fitted model parameters (only the 
 response parameters are given here): 
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
-...
-Response model(s) for state 1
+<<>>=
+summary(fm)
+@
 
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 5.522
-sd  0.2029
-
-Response model for response 2 
-Model of type multinomial, formula: corr ~ 1
-Coefficients: 
-     [,1]      [,2]
-[1,]    0 0.1031
-Probalities at zero values of the covariates.
-0.4743 0.5257
-
-Response model(s) for state 2
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 6.394
-sd  0.2374
-
-Response model for response 2 
-Model of type multinomial, formula: corr ~ 1
-Coefficients: 
-	   [,1]     [,2]
-[1,]    0 2.2455
-Probalities at zero values of the covariates.
-0.0957 0.9043 	
-\end{CodeOutput}
-\end{CodeChunk}
 As can be seen, state 1 has fast response times and accuracy is
@@
 defined by using a one-sided formula:
-\begin{CodeChunk}
-\begin{CodeInput}
-> balance$age <- balance$age-5
-> set.seed(1)
-> mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
+<<>>=
+data(balance)
+balance$age <- balance$age-5
+set.seed(1)
+mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
     family=list(multinomial(), multinomial(), multinomial(),
@@
     prior=~age, initdata=balance)
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+fm <- fit(mod)
+@
+
 Note here that we define a \code{mix} model instead of a \code{depmix}
@@
 prior model is shown here):
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
-Mixture probabilities model 
-Model of type multinomial, formula: ~age
-     [,1]       [,2]
-[1,]    0 -2.518
-[2,]    0  0.551
-Probalities at zero values of the covariates.
-0.9254 0.0746
-...
-\end{CodeOutput}
-\end{CodeChunk}	
+<<>>=
+summary(fm)
+@
+
 Hence at young ages, children have a high probability of incorrect 
@@
 To see the ordering of parameters use the following:
-\begin{CodeChunk}
-\begin{CodeInput}
-> setpars(mod, value=1:npar(mod))
-\end{CodeInput}
-\end{CodeChunk}
+<<>>=
+setpars(mod, value=1:npar(mod))
+@
 
@@
 initial state probabilities model):
-\begin{CodeChunk}
-\begin{CodeInput}
-> setpars(mod, getpars(mod,which="fixed"))
-\end{CodeInput}
-\end{CodeChunk}
+<<>>=
+setpars(mod, getpars(mod,which="fixed"))
 
+@
 When fitting constraints it is useful to have good starting values 
@@
 constraints:
-\begin{CodeChunk}
-\begin{CodeInput}
-> trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
-> mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
+<<>>=
+trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
+mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
 	  nstates=2,family=list(gaussian(),multinomial()),
 	  trstart=trst,inst=c(.999,0.001))
-> fm1 <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+fm1 <- fit(mod)
+@
+
 After this, we use the fitted values from this model to constrain the
@@
 and~10):
-\begin{CodeChunk}
-\begin{CodeInput}
+<<>>=
 # start with fixed and free parameters
-> conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
+conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
 # constrain the beta's on the transition parameters to be equal
-> conpat[6] <- conpat[10] <- 2
-> fm2 <- fit(fm,equal=conpat)
-\end{CodeInput}
-\end{CodeChunk}
+conpat[6] <- conpat[10] <- 2
+fm2 <- fit(fm1,equal=conpat)
+@
+
 Using \code{summary} on the fitted model shows that the regression 
@@
 response-class:
-\begin{CodeChunk}
-\begin{CodeInput}
+<<>>=
 setClass("exgaus", contains="response")
-\end{CodeInput}
-\end{CodeChunk}
+@
 
@@
 defined as follows:
-\begin{CodeChunk}
-\begin{CodeInput}
+<<>>=
 setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
   standardGeneric("exgaus"))
+
+require(gamlss)
+require(gamlss.dist)
 
@@
 )
-\end{CodeInput}
-\end{CodeChunk}
+@
+
+<<echo=FALSE>>=
+setMethod("dens","exgaus",
+    function(object,log=FALSE) {
+        dexGAUS(object@y, mu = predict(object), sigma = exp(object@parameters$sigma), nu = exp(object@parameters$nu), log = log)
+    }
+)
+
+setMethod("getpars","response",
+    function(object,which="pars",...) {
+        switch(which,
+            "pars" = {
+                parameters <- numeric()
+                parameters <- unlist(object@parameters)
+                pars <- parameters
+            },
+            "fixed" = {
+                pars <- object@fixed
+            }
+        )
+        return(pars)
+    }
+)
+
+setMethod("setpars","exgaus",
+    function(object, values, which="pars", ...) {
+        npar <- npar(object)
+        if(length(values)!=npar) stop("length of 'values' must be",npar)
+        # determine whether parameters or fixed constraints are being set
+		nms <- names(object@parameters)
+		switch(which,
+		  "pars"= {
+		      object@parameters$mu <- values[1]
+		      object@parameters$sigma <- values[2]
+		      object@parameters$nu <- values[3]
+		      },
+		  "fixed" = {
+		      object@fixed <- as.logical(values)
+		  }
+		  )
+        names(object@parameters) <- nms
+        return(object)
+    }
+)
+
+setMethod("predict","exgaus", 
+    function(object) {
+        ret <- object@parameters$mu
+        return(ret)
+    }
+)
+
+@
 
 The \code{fit} method is defined as follows: 
-\begin{CodeChunk}
-\begin{CodeInput}
+<<>>=
 setMethod("fit","exgaus",
@@
 )
-\end{CodeInput}
-\end{CodeChunk}
+@
 
@@
 We first create all the response models that we need as a double list: 
-\begin{CodeChunk}
-\begin{CodeInput}
-rModels <- list(
+<<>>=
+rModels <- list( list( 
+		exgaus(speed$rt,pstart=c(5,.1,.1)),	
+		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))),
 	list(
-    exgaus(rt,pstart=c(5,.1,.1)),
-    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
-  ),
-  list(
-    exgaus(rt,pstart=c(6,.1,.1)),
-    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
-  )
-)
-\end{CodeInput}
-\end{CodeChunk}
+		exgaus(speed$rt,pstart=c(6,.1,.1)),
+		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))))
+@
+
 Next, we define the transition and prior probability models using the 
@@
 the response class): 
-\begin{CodeChunk}
-\begin{CodeInput}
+<<>>=
 trstart=c(0.9,0.1,0.1,0.9)
@@
 inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
-\end{CodeInput}
-\end{CodeChunk}
+@
+
 Finally, we put everything together using \code{makeDepmix} and fit 
 the model: 
-\begin{CodeChunk}
-\begin{CodeInput}
-> mod <- makeDepmix(response=rModels,transition=transition,
+<<>>=
+mod <- makeDepmix(response=rModels,transition=transition,
     prior=inMod,stat=FALSE)
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+fm <- fit(mod)
+@
 

### 2010-02-23 papers/jss/dpx4Rev.tex
Completed turning paper into Sweave format; .sh file contains cmds to generate the paper.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@360 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2009-08-19 papers/jss/article.tex
@@
 
+\usepackage{a4wide}
+
+%\usepackage{Sweave}
+
 \begin{document}
+
+%set width of figures produced by Sweave
+\setkeys{Gin}{width=0.9\textwidth}
+
+\maketitle
 
@@
 \begin{center}
-	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
+\begin{Schunk}
+\begin{Soutput}
+##
+## Markov Chain Monte Carlo Package (MCMCpack)
+## Copyright (C) 2003-2010 Andrew D. Martin, Kevin M. Quinn, and Jong Hee Park
+##
+## Support provided by the U.S. National Science Foundation
+## (Grants SES-0350646 and SES-0350613)
+##
+\end{Soutput}
+\end{Schunk}
+\includegraphics{dpx4Rev-001}
 	 \caption{Response times (rt), accuracy (corr) and pay-off values (Pacc) for
@@
 
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
-\end{CodeInput}
-\end{CodeChunk}
-
 The \code{depmix} function returns an object of class \code{depmix}
@@
 
-The so-defined model needs to be \code{fit}ted with the following
-line of code:
-\begin{CodeChunk}
-\begin{CodeInput}
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
-
+\begin{Schunk}
+\begin{Soutput}
+iteration 0 logLik: -305.3302 
+iteration 5 logLik: -305.3251 
+iteration 10 logLik: -305.3133 
+iteration 15 logLik: -305.2824 
+iteration 20 logLik: -305.1948 
+iteration 25 logLik: -304.8879 
+iteration 30 logLik: -302.2484 
+iteration 35 logLik: -109.1782 
+iteration 40 logLik: -84.3722 
+iteration 45 logLik: -84.34183 
+iteration 50 logLik: -84.34175 
+iteration 50 logLik: -84.34175 
+\end{Soutput}
+\end{Schunk}
 The \code{fit}-function returns an object of class
@@
 values:
-\begin{CodeChunk}
-\begin{CodeInput}
-> fm
-\end{CodeInput}
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Soutput}
 Convergence info: Log likelihood converged to within tol. 
-'log Lik.' -84.34 (df=7)
-AIC:  182.68 
-BIC:  211.28 
-\end{CodeOutput}
-\end{CodeChunk}
+'log Lik.' -84.34175 (df=7)
+AIC:  182.6835 
+BIC:  211.275 
+\end{Soutput}
+\end{Schunk}
+
 These statistics can also be extracted using \code{logLik}, \code{AIC}
@@
 for the transition model, and third for the response models.
-
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Soutput}
 Initial state probabilties model 
@@
      [,1]      [,2]
-[1,]    0 -11.25
-Probalities at zero values of the covariates.
-0.999 1.292e-05
+[1,]    0 -11.25688
+Probalities at zero values of the covariates.
+0.999987 1.291797e-05
 
@@
 Coefficients: 
-[1]  0.000 -2.392
-Probalities at zero values of the covariates.
-0.9163 0.0837
+[1]  0.000000 -2.392455
+Probalities at zero values of the covariates.
+0.9162501 0.08374986
 
@@
 Coefficients: 
-[1] 0.000 2.139
-Probalities at zero values of the covariates.
-0.105 0.895
+[1] 0.000000 2.139255
+Probalities at zero values of the covariates.
+0.1053396 0.8946604
 
@@
 Coefficients: 
-[1] 6.385
-sd  0.244
+[1] 6.385492
+sd  0.2439376
 
@@
 Coefficients: 
-[1] 5.511
-sd  0.193 
-\end{CodeOutput}
-\end{CodeChunk}
+[1] 5.511151
+sd  0.1926063 
+\end{Soutput}
+\end{Schunk}
 
@@
 following example:
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(), 
-    transition=~scale(Pacc), instart=runif(2))
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+iteration 0 logLik: -305.3304 
+iteration 5 logLik: -305.2360 
+iteration 10 logLik: -288.3020 
+iteration 15 logLik: -42.88840 
+iteration 20 logLik: -38.1557 
+iteration 25 logLik: -38.12224 
+iteration 30 logLik: -38.12197 
+iteration 34 logLik: -38.12196 
+\end{Soutput}
+\end{Schunk}
 
@@
 printed here): 
-\begin{CodeChunk}
-\begin{CodeOutput}
-...
+\begin{Schunk}
+\begin{Soutput}
+Initial state probabilties model 
+Model of type multinomial, formula: ~1
+Coefficients: 
+     [,1]     [,2]
+[1,]    0 10.71779
+Probalities at zero values of the covariates.
+2.214681e-05 0.9999779
+
 Transition model for state (component) 1 
@@
      [,1]       [,2]
-[1,]    0 -0.9215
-[2,]    0  1.865
-Probalities at zero values of the covariates.
-0.7154 0.2846
+[1,]    0 -0.9215182
+[2,]    0  1.8649734
+Probalities at zero values of the covariates.
+0.7153513 0.2846487
 
@@
      [,1]     [,2]
-[1,]    0 2.471
-[2,]    0 3.571
-Probalities at zero values of the covariates.
-0.0779 0.9221
-...
-\end{CodeOutput}
-\end{CodeChunk}
+[1,]    0 2.471442
+[2,]    0 3.570856
+Probalities at zero values of the covariates.
+0.07788458 0.9221154
+
+Response model(s) for state 1
+
+Response model for response 1 
+Model of type gaussian, formula: rt ~ 1
+Coefficients: 
+[1] 5.512179
+sd  0.1921098
+
+Response model(s) for state 2
+
+Response model for response 1 
+Model of type gaussian, formula: rt ~ 1
+Coefficients: 
+[1] 6.3885
+sd  0.2402693 
+\end{Soutput}
+\end{Schunk}
+
 The summary provides all parameters of the model, also the (redundant)
@@
 by specifying the following:
-\begin{CodeChunk}
-\begin{CodeInput}
-> set.seed(1)
-> mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
-    family=list(gaussian(),multinomial()),
-    transition=~scale(Pacc),instart=runif(2))
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+iteration 0 logLik: -554.6365 
+iteration 5 logLik: -553.9195 
+iteration 10 logLik: -343.1694 
+iteration 15 logLik: -249.1893 
+iteration 20 logLik: -248.9757 
+iteration 25 logLik: -248.9724 
+iteration 30 logLik: -248.9723 
+iteration 33 logLik: -248.9723 
+\end{Soutput}
+\end{Schunk}
+
 This provides the following fitted model parameters (only the 
 response parameters are given here): 
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
-...
+\begin{Schunk}
+\begin{Soutput}
+Initial state probabilties model 
+Model of type multinomial, formula: ~1
+Coefficients: 
+     [,1]     [,2]
+[1,]    0 10.24230
+Probalities at zero values of the covariates.
+3.562961e-05 0.9999644
+
+Transition model for state (component) 1 
+Model of type multinomial, formula: ~scale(Pacc)
+Coefficients: 
+     [,1]       [,2]
+[1,]    0 -0.8924638
+[2,]    0  2.1293542
+Probalities at zero values of the covariates.
+0.7093984 0.2906016
+
+Transition model for state (component) 2 
+Model of type multinomial, formula: ~scale(Pacc)
+Coefficients: 
+     [,1]     [,2]
+[1,]    0 2.390396
+[2,]    0 3.690785
+Probalities at zero values of the covariates.
+0.08390802 0.916092
+
 Response model(s) for state 1
@@
 Coefficients: 
-[1] 5.522
-sd  0.2029
+[1] 5.52169
+sd  0.2028857
 
@@
      [,1]      [,2]
-[1,]    0 0.1031
-Probalities at zero values of the covariates.
-0.4743 0.5257
+[1,]    0 0.1030554
+Probalities at zero values of the covariates.
+0.4742589 0.5257411
 
@@
 Coefficients: 
-[1] 6.394
-sd  0.2374
+[1] 6.39369
+sd  0.2373650
 
@@
      [,1]     [,2]
-[1,]    0 2.2455
-Probalities at zero values of the covariates.
-0.0957 0.9043 	
-\end{CodeOutput}
-\end{CodeChunk}
+[1,]    0 2.245514
+Probalities at zero values of the covariates.
+0.09573715 0.9042629 
+\end{Soutput}
+\end{Schunk}
+
 As can be seen, state 1 has fast response times and accuracy is
@@
 defined by using a one-sided formula:
-\begin{CodeChunk}
-\begin{CodeInput}
-> balance$age <- balance$age-5
-> set.seed(1)
-> mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
-    family=list(multinomial(), multinomial(), multinomial(),
-    multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
-    prior=~age, initdata=balance)
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+iteration 0 logLik: -953.005 
+iteration 5 logLik: -951.2919 
+iteration 10 logLik: -951.2918 
+iteration 10 logLik: -951.2918 
+\end{Soutput}
+\end{Schunk}
+
 Note here that we define a \code{mix} model instead of a \code{depmix}
@@
 prior model is shown here):
-\begin{CodeChunk}
-\begin{CodeInput}
-> summary(fm)
-\end{CodeInput}
-\begin{CodeOutput}
+\begin{Schunk}
+\begin{Soutput}
 Mixture probabilities model 
 Model of type multinomial, formula: ~age
-     [,1]       [,2]
-[1,]    0 -2.518
-[2,]    0  0.551
-Probalities at zero values of the covariates.
-0.9254 0.0746
-...
-\end{CodeOutput}
-\end{CodeChunk}	
+Coefficients: 
+     [,1]       [,2]
+[1,]    0 -2.5182573
+[2,]    0  0.5512996
+Probalities at zero values of the covariates.
+0.9254119 0.07458815
+
+Response model(s) for state 1
+
+Response model for response 1 
+Model of type multinomial, formula: d1 ~ 1
+Coefficients: 
+     [,1]      [,2]
+[1,]    0 -3.390018
+Probalities at zero values of the covariates.
+0.9673911 0.03260889
+
+Response model for response 2 
+Model of type multinomial, formula: d2 ~ 1
+Coefficients: 
+     [,1]      [,2]
+[1,]    0 -2.311851
+Probalities at zero values of the covariates.
+0.9098538 0.0901462
+
+Response model for response 3 
+Model of type multinomial, formula: d3 ~ 1
+Coefficients: 
+     [,1]      [,2]
+[1,]    0 -3.415948
+Probalities at zero values of the covariates.
+0.9681992 0.03180076
+
+Response model for response 4 
+Model of type multinomial, formula: d4 ~ 1
+Coefficients: 
+     [,1]      [,2]
+[1,]    0 -2.247889
+Probalities at zero values of the covariates.
+0.9044683 0.0955317
+
+Response model(s) for state 2
+
+Response model for response 1 
+Model of type multinomial, formula: d1 ~ 1
+Coefficients: 
+     [,1]     [,2]
+[1,]    0 2.950105
+Probalities at zero values of the covariates.
+0.04973154 0.9502685
+
+Response model for response 2 
+Model of type multinomial, formula: d2 ~ 1
+Coefficients: 
+     [,1]     [,2]
+[1,]    0 3.065500
+Probalities at zero values of the covariates.
+0.04455298 0.955447
+
+Response model for response 3 
+Model of type multinomial, formula: d3 ~ 1
+Coefficients: 
+     [,1]     [,2]
+[1,]    0 2.808197
+Probalities at zero values of the covariates.
+0.05688286 0.9431171
+
+Response model for response 4 
+Model of type multinomial, formula: d4 ~ 1
+Coefficients: 
+     [,1]     [,2]
+[1,]    0 3.350205
+Probalities at zero values of the covariates.
+0.03388844 0.9661116 
+\end{Soutput}
+\end{Schunk}
+
 Hence at young ages, children have a high probability of incorrect 
@@
 To see the ordering of parameters use the following:
-\begin{CodeChunk}
-\begin{CodeInput}
-> setpars(mod, value=1:npar(mod))
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+Initial state probabilties model 
+Model of type multinomial, formula: ~age
+Coefficients: 
+     [,1] [,2]
+[1,]    1    2
+[2,]    3    4
+Probalities at zero values of the covariates.
+0.09893802 0.7310586
+
+Response model(s) for state 1
+
+Response model for response 1 
+Model of type multinomial, formula: d1 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]    5    6
+Probalities at zero values of the covariates.
+0.001812113 0.7310586
+
+Response model for response 2 
+Model of type multinomial, formula: d2 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]    7    8
+Probalities at zero values of the covariates.
+0.0002452428 0.7310586
+
+Response model for response 3 
+Model of type multinomial, formula: d3 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]    9   10
+Probalities at zero values of the covariates.
+3.319001e-05 0.7310586
+
+Response model for response 4 
+Model of type multinomial, formula: d4 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]   11   12
+Probalities at zero values of the covariates.
+4.491779e-06 0.7310586
+
+Response model(s) for state 2
+
+Response model for response 1 
+Model of type multinomial, formula: d1 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]   13   14
+Probalities at zero values of the covariates.
+6.078962e-07 0.7310586
+
+Response model for response 2 
+Model of type multinomial, formula: d2 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]   15   16
+Probalities at zero values of the covariates.
+8.22698e-08 0.7310586
+
+Response model for response 3 
+Model of type multinomial, formula: d3 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]   17   18
+Probalities at zero values of the covariates.
+1.113401e-08 0.7310586
+
+Response model for response 4 
+Model of type multinomial, formula: d4 ~ 1
+Coefficients: 
+     [,1] [,2]
+[1,]   19   20
+Probalities at zero values of the covariates.
+1.506824e-09 0.7310586 
+\end{Soutput}
+\end{Schunk}
 
@@
 initial state probabilities model):
-\begin{CodeChunk}
-\begin{CodeInput}
-> setpars(mod, getpars(mod,which="fixed"))
-\end{CodeInput}
-\end{CodeChunk}
-
+\begin{Schunk}
+\begin{Soutput}
+Initial state probabilties model 
+Model of type multinomial, formula: ~age
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+[2,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model(s) for state 1
+
+Response model for response 1 
+Model of type multinomial, formula: d1 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model for response 2 
+Model of type multinomial, formula: d2 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model for response 3 
+Model of type multinomial, formula: d3 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model for response 4 
+Model of type multinomial, formula: d4 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model(s) for state 2
+
+Response model for response 1 
+Model of type multinomial, formula: d1 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model for response 2 
+Model of type multinomial, formula: d2 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model for response 3 
+Model of type multinomial, formula: d3 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414
+
+Response model for response 4 
+Model of type multinomial, formula: d4 ~ 1
+Coefficients: 
+     [,1]  [,2]
+[1,] TRUE FALSE
+Probalities at zero values of the covariates.
+0.2689414 0.2689414 
+\end{Soutput}
+\end{Schunk}
 When fitting constraints it is useful to have good starting values 
@@
 constraints:
-\begin{CodeChunk}
-\begin{CodeInput}
-> trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
-> mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
-	  nstates=2,family=list(gaussian(),multinomial()),
-	  trstart=trst,inst=c(.999,0.001))
-> fm1 <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+iteration 0 logLik: -554.2052 
+iteration 5 logLik: -254.4498 
+iteration 10 logLik: -249.018 
+iteration 15 logLik: -248.973 
+iteration 20 logLik: -248.9723 
+iteration 23 logLik: -248.9723 
+\end{Soutput}
+\end{Schunk}
+
 After this, we use the fitted values from this model to constrain the
@@
 and~10):
-\begin{CodeChunk}
-\begin{CodeInput}
-# start with fixed and free parameters
-> conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
-# constrain the beta's on the transition parameters to be equal
-> conpat[6] <- conpat[10] <- 2
-> fm2 <- fit(fm,equal=conpat)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+Rdonlp2 - a wrapper library for "DONLP2 (C) Peter Spellucci"
+
+Loading required package: Rsolnp (version 0.3)
+    1 fx=   0.000000e+00 upsi=  6.7e+00 b2n= -1.0e+00 umi=  0.0e+00 nr   1 si-1
+    2 fx=   2.640247e+02 upsi=  0.0e+00 b2n=  2.1e+01 umi=  0.0e+00 nr   1 si-1
+    3 fx=   2.636495e+02 upsi=  0.0e+00 b2n=  5.4e+01 umi=  0.0e+00 nr   1 si-1
+    4 fx=   2.587358e+02 upsi=  0.0e+00 b2n=  2.7e+02 umi=  0.0e+00 nr   1 si-1
+    5 fx=   2.528309e+02 upsi=  0.0e+00 b2n=  1.2e+02 umi=  0.0e+00 nr   1 si-1
+    6 fx=   2.523880e+02 upsi=  0.0e+00 b2n=  9.5e+01 umi=  0.0e+00 nr   1 si-1
+    7 fx=   2.523802e+02 upsi=  0.0e+00 b2n=  9.4e+01 umi=  0.0e+00 nr   1 si-1
+    8 fx=   2.521969e+02 upsi=  0.0e+00 b2n=  9.2e+01 umi=  0.0e+00 nr   1 si-1
+    9 fx=   2.515470e+02 upsi=  0.0e+00 b2n=  5.7e+01 umi=  0.0e+00 nr   1 si-1
+   10 fx=   2.510761e+02 upsi=  0.0e+00 b2n=  5.1e+01 umi=  0.0e+00 nr   1 si-1
+   11 fx=   2.507338e+02 upsi=  0.0e+00 b2n=  7.1e+01 umi=  0.0e+00 nr   1 si-1
+   12 fx=   2.503019e+02 upsi=  0.0e+00 b2n=  1.9e+01 umi=  0.0e+00 nr   1 si-1
+   13 fx=   2.502725e+02 upsi=  0.0e+00 b2n=  9.5e-01 umi=  0.0e+00 nr   1 si-1
+   14 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.5e-01 umi=  0.0e+00 nr   1 si-1
+   15 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.0e-02 umi=  0.0e+00 nr   1 si-1
+   16 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.5e-02 umi=  0.0e+00 nr   1 si-1
+   17 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  6.7e-02 umi=  0.0e+00 nr   1 si-1
+   18 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.1e-02 umi=  0.0e+00 nr   1 si-1
+   19 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.3e-03 umi=  0.0e+00 nr   1 si-1
+   20 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.4e-04 umi=  0.0e+00 nr   1 si-1
+   21 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.0e-03 umi=  0.0e+00 nr   1 si-1
+   22 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-01 umi=  0.0e+00 nr   1 si-1
+   23 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  7.1e-02 umi=  0.0e+00 nr   1 si-1
+   24 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.8e-02 umi=  0.0e+00 nr   1 si-1
+   25 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.9e-02 umi=  0.0e+00 nr   1 si-1
+   26 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.3e-02 umi=  0.0e+00 nr   1 si-1
+   27 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.6e-02 umi=  0.0e+00 nr   1 si-1
+   28 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  5.3e-02 umi=  0.0e+00 nr   1 si-1
+   29 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  9.0e-02 umi=  0.0e+00 nr   1 si-1
+   30 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
+   31 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
+   32 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
+   33 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
+   34 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
+   35 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.5e-02 umi=  0.0e+00 nr   1 si-1
+   36 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.3e-02 umi=  0.0e+00 nr   1 si-1
+   37 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.1e-02 umi=  0.0e+00 nr   1 si-1
+   38 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-02 umi=  0.0e+00 nr   1 si-1
+   39 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.9e-04 umi=  0.0e+00 nr   1 si-1
+
+     n=        11    nlin=         1    nonlin=         0
+
+  epsx= 1.000e-05 sigsm= 1.490e-08
+
+startvalue
+  1.1073511e+01  -4.2221899e+00   9.1331180e+00  -3.3737739e+00   1.5804725e+01 
+  5.5217096e+00   2.0290950e-01   1.0308039e-01   6.3936995e+00   2.3736013e-01 
+  2.2456377e+00
+
+  eps=  2.22e-16  tol= 2.23e-308 del0=  1.00e+00 delm=  1.00e-06 tau0=  1.00e+00
+  tau=  1.00e-01   sd=  1.00e-01   sw=  3.67e-11  rho=  1.00e-06 rho1=  1.00e-10
+ scfm=  1.00e+04  c1d=  1.00e-02 epdi=  1.00e-08
+  nre=        11 anal=         0
+taubnd=  1.00e+00 epsfcn=  1.00e-16 difftype=2
+
+ termination reason:
+ KT-conditions satisfied, no further correction computed
+ evaluations of f                          142
+ evaluations of grad f                      39
+ evaluations of constraints                143
+ evaluations of grads of constraints         0
+ final scaling of objective           2.577477e+00
+ norm of grad(f)                      5.469761e-01
+ lagrangian violation                 1.047009e-05
+ feasibility violation                0.000000e+00
+ dual feasibility violation           0.000000e+00
+ optimizer runtime sec's              1.057179e+02
+
+ optimal value of f =   2.50272256993535e+02
+
+ optimal solution  x =
+  1.57776829062875e+01 -5.33950896009860e+00  1.26660714788775e+01
+ -2.42709038921118e+00  1.26660714788775e+01  5.52688670442125e+00
+  2.09034923400015e-01  1.03019934348728e-01  6.39615051757101e+00
+  2.36133700896234e-01  2.27764349273245e+00
+
+  multipliers are relativ to scf=1
+  nr.    constraint     multiplier norm(grad) or 1 
+    1   1.0000000e+20    0.0000000e+00 
+    2   1.0000000e+20    0.0000000e+00 
+    3   1.0000000e+20    0.0000000e+00 
+    4   1.0000000e+20    0.0000000e+00 
+    5   1.0000000e+20    0.0000000e+00 
+    6   1.0000000e+20    0.0000000e+00 
+    7   1.0000000e+20    0.0000000e+00 
+    8   1.0000000e+20    0.0000000e+00 
+    9   1.0000000e+20    0.0000000e+00 
+   10   1.0000000e+20    0.0000000e+00 
+   11   1.0000000e+20    0.0000000e+00 
+   12   1.0000000e+20    0.0000000e+00 
+   13   1.0000000e+20    0.0000000e+00 
+   14   1.0000000e+20    0.0000000e+00 
+   15   1.0000000e+20    0.0000000e+00 
+   16   1.0000000e+20    0.0000000e+00 
+   17   1.0000000e+20    0.0000000e+00 
+   18   1.0000000e+20    0.0000000e+00 
+   19   1.0000000e+20    0.0000000e+00 
+   20   1.0000000e+20    0.0000000e+00 
+   21   1.0000000e+20    0.0000000e+00 
+   22   1.0000000e+20    0.0000000e+00 
+   23   0.0000000e+00    3.8677045e-01    1.4142136e+00
+   24   0.0000000e+00    0.0000000e+00
+
+ evaluations of restrictions and their gradients
+ (   143,     0)
+ last estimate of cond.nr. of active gradients   1.000e+00
+
+ last estimate of cond.nr. of approx.  hessian   3.607e+05
+iterative steps total              39
+# of restarts                       1
+# of full regular updates          37
+# of updates                       37
+# of regularized full SQP-steps     0
+\end{Soutput}
+\end{Schunk}
+
 Using \code{summary} on the fitted model shows that the regression 
@@
 response-class:
-\begin{CodeChunk}
-\begin{CodeInput}
-setClass("exgaus", contains="response")
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+[1] "exgaus"
+\end{Soutput}
+\end{Schunk}
 
@@
 defined as follows:
-\begin{CodeChunk}
-\begin{CodeInput}
-setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
-  standardGeneric("exgaus"))
-
-setMethod("exgaus", 
-  signature(y="ANY"), 
-  function(y,pstart=NULL,fixed=NULL, ...) {
-    y <- matrix(y,length(y))
-    x <- matrix(1) 
-    parameters <- list()
-    npar <- 3
-    if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
-    if(!is.null(pstart)) {
-      if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
-      parameters$mu <- pstart[1]
-      parameters$sigma <- log(pstart[2])
-      parameters$nu <- log(pstart[3])
-    }
-    mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
-    mod
-  }
-)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+[1] "exgaus"
+\end{Soutput}
+\begin{Soutput}
+ **********   GAMLSS Version 3.0-1 ********** 
+For more on GAMLSS look at http://www.gamlss.com/ 
+Type gamlssNews() to see new features/changes/bug fixes.
+\end{Soutput}
+\begin{Soutput}
+[1] "exgaus"
+\end{Soutput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Soutput}
+[1] "dens"
+\end{Soutput}
+\begin{Soutput}
+[1] "getpars"
+\end{Soutput}
+\begin{Soutput}
+[1] "setpars"
+\end{Soutput}
+\begin{Soutput}
+[1] "predict"
+\end{Soutput}
+\end{Schunk}
 
 The \code{fit} method is defined as follows: 
-\begin{CodeChunk}
-\begin{CodeInput}
-setMethod("fit","exgaus",
-  function(object,w) {
-    if(missing(w)) w <- NULL
-    y <- object@y
-    fit <- gamlss(y~1,weights=w,family=exGAUS(),
-      control=gamlss.control(n.cyc=100,trace=FALSE),
-      mu.start=object@parameters$mu,
-      sigma.start=exp(object@parameters$sigma),
-      nu.start=exp(object@parameters$nu))
-    pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
-    object <- setpars(object,pars)
-    object
-  }
-)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+[1] "fit"
+\end{Soutput}
+\end{Schunk}
 
@@
 We first create all the response models that we need as a double list:
-\begin{CodeChunk}
-\begin{CodeInput}
-rModels <- list(
-  list(
-    exgaus(rt,pstart=c(5,.1,.1)),
-    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
-  ),
-  list(
-    exgaus(rt,pstart=c(6,.1,.1)),
-    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
-  )
-)
-\end{CodeInput}
-\end{CodeChunk}
+
 Next, we define the transition and prior probability models using the 
@@
 the response class):
-\begin{CodeChunk}
-\begin{CodeInput}
-trstart=c(0.9,0.1,0.1,0.9)
-transition <- list()
-transition[[1]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.9,0.1,0,0))
-transition[[2]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.1,0.9,0,0))
-inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
-\end{CodeInput}
-\end{CodeChunk}
+
 Finally, we put everything together using \code{makeDepmix} and fit 
 the model: 
-\begin{CodeChunk}
-\begin{CodeInput}
-> mod <- makeDepmix(response=rModels,transition=transition,
-    prior=inMod,stat=FALSE)
-> fm <- fit(mod)
-\end{CodeInput}
-\end{CodeChunk}
+\begin{Schunk}
+\begin{Soutput}
+iteration 0 logLik: -265.0144 
+iteration 5 logLik: -243.4716 
+iteration 10 logLik: -232.3604 
+iteration 15 logLik: -232.2586 
+iteration 20 logLik: -232.2537 
+iteration 25 logLik: -232.2528 
+iteration 30 logLik: -232.2527 
+iteration 35 logLik: -232.2526 
+iteration 40 logLik: -232.2526 
+iteration 45 logLik: -232.2526 
+iteration 50 logLik: -232.2526 
+iteration 52 logLik: -232.2526 
+\end{Soutput}
+\end{Schunk}
 

### 2010-02-23 papers/jss/dpx4Rev.Rnw
Various changes in response to reviewers for jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@362 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\SweaveOpts{echo=FALSE}
+\SweaveOpts{echo=TRUE}
 \usepackage{a4wide}
@@
 \ldots, $O_{T}^{1}, \ldots, O_{T}^{m})$ for an $m$-variate time series
-of length $T$.  As an example, consider a time series of responses
-generated by a single participant in a psychological response time
-experiment.  The data consists of three variables, response time,
-response accuracy, and a covariate which is a pay-off variable
-reflecting the relative reward for speeded and/or accurate responding.
-These variables are measured on 168, 134 and 137 occasions
-respectively (the first part of this series is plotted in
+of length $T$.  In the following, we use $\vc{O}_{t}$ as shorthand for
+$O_{t}^{1}, \ldots, O_{t}^{m}$.  As an example, consider a time series
+of responses generated by a single participant in a psychological
+response time experiment.  The data consists of three variables,
+response time, response accuracy, and a covariate which is a pay-off
+variable reflecting the relative reward for speeded and/or accurate
+responding.  These variables are measured on 168, 134 and 137
+occasions respectively (the first part of this series is plotted in
 Figure~\ref{fig:speed}).  These data are more fully described in
@@
 \begin{align}
-	\phi_{1}(j) &:= \Prob(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
+	\phi_{1}(j) &:= \Prob(\vc{O}_{1}, S_{1}=j) = \pi_{j} \vc{b}_{j}(\vc{O}_{1}) 
 	\label{eq:fwd1} \\
-%\begin{split}
 	\phi_{t}(j) &:= \Prob(\vc{O}_{t}, S_{t}=j|\vc{O}_{1:(t-1)}) %\\
-	= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
+	= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij} \vc{b}_{j}(\vc{O}_{t})] \times 
 (\Phi_{t-1})^{-1},
 	\label{eq:fwdt} 
-%\end{split} 
 \end{align}
@@
 algorithm, parameters are estimated by iteratively maximising the
-expected joint likelihood of the parameters given the observations and
+expected joint log-likelihood of the parameters given the observations and
 states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
@@
 (initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
-\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. The expected 
-log likelihood 
+\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. 
+The expected log-likelihood:
 \begin{equation}
@@
 can be written as
-%\begin{equation}
 \begin{multline}
@@
  + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
-\ln \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
+\log \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
 \end{multline}
-%\end{equation}
-where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|\vc{O}_{1:T},
-\vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|\vc{O}_{1:T},
-\vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
-forward-backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
-step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
-right hand side of (\ref{eq:Q}) consists of three separate parts, we can 
-maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and 
-$\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and 
-$\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in the 
-\pkg{nnet} package \citep{Venables2002}, and maximisation for 
-$\greekv{\theta}_3$ by the standard \code{glm} routine. Note that for the latter 
-maximisation, the expected values $\gamma_t(j)$ are used as prior weights of the 
+where the expected values $\xi_t(j,k) = P(S_t = k, S_{t-1} =
+j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) =
+P(S_t = j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ can be
+computed effectively by the forward-backward algorithm \citep[see
+e.g.,][]{Rabiner1989}.  The Maximisation step consists of the
+maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
+side of (\ref{eq:Q}) consists of three separate parts, we can maximise
+separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
+$\greekv{\theta}_3$.  In common models, maximisation for
+$\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
+\code{nnet.default} routine in the \pkg{nnet} package
+\citep{Venables2002}, and maximisation for $\greekv{\theta}_3$ by the
+standard \code{glm} routine.  Note that for the latter maximisation,
+the expected values $\gamma_t(j)$ are used as prior weights of the
 observations $O^k_t$.
@@
 in \pkg{depmixS4}, EM is used by default in unconstrained models, but
-otherwise, direct optimization is done using \pkg{Rdonlp2}
-\citep{Tamura2009,Spellucci2002}, because it handles general linear
-(in)equality constraints, and optionally also non-linear constraints.
+otherwise, direct optimization is used.  Two options are available for
+direct optimization using package \pkg{Rdonlp2}
+\citep{Tamura2009,Spellucci2002}, or package \pkg{Rsolnp}.  Both
+packages can handle general linear (in)equality constraints, and
+optionally also non-linear constraints.
 

### 2010-02-24 papers/jss/dpx4Rev.Rnw
Added notes on balance and speed data sets.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@367 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %\end{equation}
-%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does not affect the estimation of the other elements.
-When elements $a_{jk}$ are assumed identical, we simply extend the summation in (\ref{eq:MLA}) to include all those elements, changing the denominator $N(T-1)$ accordingly.
-
-%To estimate $\greekv{\lambda}$, we note that the term containing this parameter has the form of a weighted likelihood, where $\gamma_t^i(j)$ can be interpreted as the number of replications of $r^i_t$.
-Hence, we can rely on standard ML estimates of the logistic regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$ as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To be more specific, we replicate each observation $r^i_t$ a total of 6 times, once for each of the states besides the random state (which offers no information regarding $\lambda_i$).
-For the $j$-th replication (corresponding to the $j$-th state), we used $v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a case weight.
-All these replications were used to obtain the maximum likelihood estimate of $\lambda$ from a single GLM, using the ``glm.fit'' function in R \cite{R}.}.
-The resulting parameter estimates are then used in a new Expectation step, followed by another Maximisation step, until convergence.
-Under mild regularity conditions, it can be shown that the EM algorithm converges to a (local) maximum of the likelihood \cite{EM}.
+
+%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does
+%not affect the estimation of the other elements.  When elements
+%$a_{jk}$ are assumed identical, we simply extend the summation in
+%(\ref{eq:MLA}) to include all those elements, changing the
+%denominator $N(T-1)$ accordingly.
+
+%To estimate $\greekv{\lambda}$, we note that the term containing this
+%parameter has the form of a weighted likelihood, where
+%$\gamma_t^i(j)$ can be interpreted as the number of replications of
+%$r^i_t$.  Hence, we can rely on standard ML estimates of the logistic
+%regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$
+%as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To
+%be more specific, we replicate each observation $r^i_t$ a total of 6
+%times, once for each of the states besides the random state (which
+%offers no information regarding $\lambda_i$).  For the $j$-th
+%replication (corresponding to the $j$-th state), we used
+%$v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a
+%case weight.  All these replications were used to obtain the maximum
+%likelihood estimate of $\lambda$ from a single GLM, using the
+%``glm.fit'' function in R \cite{R}.}.  The resulting parameter
+%estimates are then used in a new Expectation step, followed by
+%another Maximisation step, until convergence.  Under mild regularity
+%conditions, it can be shown that the EM algorithm converges to a
+%(local) maximum of the likelihood \cite{EM}.
 
@@
 	for latent class and finite mixture models, see example below on adding
-	covariates to prior probabilities)
+	covariates to prior probabilities in section~\ref{sec:prior})
 
@@
 already indicated in the Introduction, it consists of three time
-series with three variables: response time, accuracy, and a covariate
-Pacc which defines the relative pay-off for speeded versus accurate
-responding.  The participant in this experiment switches between fast
-responding at chance level and relatively slower responding at a high
-level of accuracy.  Interesting hypotheses to test are: is the
-switching regime symmetric?  Is there evidence for two states or does
-one state suffice?  Is the guessing state actually a guessing state,
-i.e., is the probability of a correct response at chance level (0.5)?
+series with three variables: response time, accuracy, and a covariate,
+Pacc, which defines the relative pay-off for speeded versus accurate
+responding.  Before describing some of the models that are fitted
+to these data, we provide a brief sketch of the reasons for gathering
+these data in the first place.
+
+Response times are a very common dependent variable in psychological
+experiments and hence form the basis for infernce about many
+psychological processes.  A potential threat to such inference based
+on response times is formed by the speed-accuracy trade-off: different
+participants in an experiment may respond differently to typical
+instructions to `respond as fast and accurate as possible'.  A popular
+model which takes the speed-accuracy trade-off into account is the
+diffusion model \citep{Ratcliff1978}, which has proven to provide
+accurate descriptions of response times in many different settings.
+
+One potential problem with the diffusion model is that it predicts a
+continuous trade-off between speed and accuracy of responding, i.e.,
+when participants are pressed to respond faster and faster, the
+diffusion model predicts that this would lead to a gradual decrease in
+accuracy.  The \code{speed} data set that we analyze below was
+gathered to test this hypothesis versus the alternative hypothesis
+stating that there is a sudden transition from slow and accurate
+responding to fast responding at chance level.  At each trial of the
+experiment, the participant is shown the current setting of the
+relative reward for speed versus accuracy.  The bottom panel of
+figure~ref{fig:speed} shows the values of this variable.  The main
+question in this experiment was what would happen when this reward
+variable would change from reward for accuracy only to reward for
+speed only.  The \code{speed} data that we analyse here are from
+participant A in Experiment 1 in \citet{Dutilh2009}, who provide a
+complete description of the experiment and the relevant theoretical
+background.
+
+The central hypothesis about this data is whether indeed it is best 
+described by two modes of responding rather than a single mode of 
+responding with a continuous trade-off between speed and accuracy. 
+The hallmark of having a discontinuity between slow versus speeded 
+responding is that the switching between the two modes is assymetric
+\citep[see e.g.][for a theoretical underpinning of this 
+claim]{Maas1992}. The \code{fit} help page of \pkg{depmixS4} provides 
+a number of examples in which the assymetricity of the switching 
+process is tested; those examples are discussed at length in 
+\citet{Visser2009b}.
 
@@
 \code{depmix}-function as follows:
-
 <<>>=
@@
 set.seed(1)
-mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
-@
-
-The \code{depmix} function returns an object of class \code{depmix}
-which contains the model specification (and not a fitted model!).
+mod <- depmix(response=rt~1, data=speed, nstates=2, trstart=runif(4))
+@
+
+The first line of code loads the \pkg{depmixS4} package and 
+\code{data(speed)} loads the \code{speed} data set. The line
+\code{set.seed(1)} is necessary to get starting values that will 
+result in the right model, see more on starting values below.
+
+The call to \code{depmix} specifies the model with a number of
+arguments.  The \code{response} argument is used to specify the
+response variable, possibly with covariates, in the familiar format
+using formulae such as in \code{lm} or \code{glm} models.  The second
+argument, \code{data=speed}, provides the \code{data.frame} in which
+the variables from \code{response} are interpreted.  Third, the number
+of states of the model is given by \code{nstates=2}.
+
+\subsubsection{Starting values}
+
 Note also that start values for the transition parameters are provided
 in this call using the \code{trstart} argument.  At this time, the
-package does not provide automatic starting values.
+package does not provide automatic starting values.  Starting values
+for parameters can be provided using three arguments: \code{instart}
+for the parameters relating to the prior probabilities,
+\coce{trstart}, relating the transition models, and \code{respstart}
+for the parameters of the response models.  Note that the starting
+values for the initial and transition models as well as of the
+multinomial logit response models are interpreted as {\em probabilities},
+and internally converted to multinomial logit parameters.
+
+\subsubsection{Fitting the model and printing results}
+
+The \code{depmix} function returns an object of class \code{depmix}
+which contains the model specification, and not a fitted model. Hence,
+the model needs to be fitted by calling \code{fit}:
 
@@
 parameter estimates, first for the prior probabilities model, second
-for the transition model, and third for the response models.
+for the transition models, and third for the response models.
 <<>>=
@@
 
-Note that, since no further arguments were specified, the initial
-state, state transition and response distributions were set to their
-defaults (multinomial distributions for the first two, and Gaussian
-distributions for the response distributions).
-
-\subsection{Covariates on transition parameters}
+Since no further arguments were specified, the initial state, state
+transition and response distributions were set to their defaults
+(multinomial distributions for the first two, and a Gaussian
+distribution for the response).
+
+\subsection{Covariates on transition parameters\label{sec:trans}}
 
@@
 
-Applying \code{summary} to the \code{fit}ted object gives (only transition models
-printed here): 
+Applying \code{summary} to the \code{fit}ted object gives (only
+transition models printed here):
 <<>>=
@@
 
-\subsection{Adding covariates on the prior probabilities}
+\subsection{Adding covariates on the prior probabilities\label{sec:prior}}
 
@@
 data consists of 4 binary items (correct-incorrect) on a balance scale
-task \citet{Siegler1981}.  The data form a subset of the data
-published in \citet{Jansen2002}.
+task \citep{Siegler1981}.  The data form a subset of the data
+published in \citet{Jansen2002}. Before specifying specifying a model 
+for these data, we briefly describe them.
+
+The balance scale taks is a famous task for testing cognitive
+strategies developed by Piaget \citep{Siegler1981}.
+Figure~\ref{fig:balance} provides an example of a balance scale item.
+Participants' task is to say to which side the balance will tip when
+released, or alternatively, whether it will stay in balance.  The item
+shown in Figure~\ref{fig:balance} is a so-callled distance item: the
+number of weights placed on each side is equal, and only the distance
+of the weights to the fulcrum differs between each side.
+
+\begin{figure}[htbp]
+\begin{center}
+	\includegraphic{graphs:baldist.pdf}
+	\caption{Balance scale item; this is a so-called distance item.}
+	\label{fig:balance}  
+\end{center}
+\end{figure}
+
+Children in the lower grades of primary school are known to ignore the
+distance dimension, and base their answer only on the number of
+weights on each side.  Hence, they would typically provide the wrong
+answer to these distance items.  Slightly older children do take
+distance into account when responding to balance scale items, but they
+only do so when the number of weights is equal on each side.  These
+two strategies that children employ are known a Rule I and Rule II.
+Other strategies can be teased apart by administering different items.
+The \code{balance} data set that we analyse here consists of 4
+distance items on a balance scale task administered to 779
+participants ranging from 5 to 19 years of age.  The full set of items
+consisted of 25 items; other items in the test are used to detect
+other strategies that children (and young adults) employ in solving
+balance scale items \citep[see][for details]{Jansen2002}.
 

### 2010-02-25 papers/jss/dpx4Rev.Rnw
Minor updates in jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@372 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 a number of examples in which the assymetricity of the switching
-process is tested; those examples are discussed at length in 
-\citet{Visser2009b}.
+process is tested; those examples and other candidate models are
+discussed at length in \citet{Visser2009b}.
 
@@
 
-\subsubsection{Starting values}
-
-Note also that start values for the transition parameters are provided
-in this call using the \code{trstart} argument.  At this time, the
-package does not provide automatic starting values.  Starting values
-for parameters can be provided using three arguments: \code{instart}
-for the parameters relating to the prior probabilities,
-\coce{trstart}, relating the transition models, and \code{respstart}
-for the parameters of the response models.  Note that the starting
-values for the initial and transition models as well as of the
-multinomial logit response models are interpreted as {\em probabilities},
-and internally converted to multinomial logit parameters.
-
-\subsubsection{Fitting the model and printing results}
-
-The \code{depmix} function returns an object of class \code{depmix}
-which contains the model specification, and not a fitted model. Hence,
-the model needs to be fitted by calling \code{fit}:
-
+\paragraph{Starting values} Note also that start values for the
+transition parameters are provided in this call using the
+\code{trstart} argument.  At this time, the package does not provide
+automatic starting values.  Starting values for parameters can be
+provided using three arguments: \code{instart} for the parameters
+relating to the prior probabilities, \coce{trstart}, relating the
+transition models, and \code{respstart} for the parameters of the
+response models.  Note that the starting values for the initial and
+transition models as well as of the multinomial logit response models
+are interpreted as {\em probabilities}, and internally converted to
+multinomial logit parameters.
+
+\paragraph{Fitting the model and printing results} The \code{depmix}
+function returns an object of class \code{depmix} which contains the
+model specification, and not a fitted model.  Hence, the model needs
+to be fitted by calling \code{fit}:
 <<>>=
@@
 @
+
 The \code{fit}-function returns an object of class
@@
 convergence information (and information about constraints if these
-were applied, see below).  The \code{print} method provides summary
-information on convergence, the log-likelihood and the AIC and BIC
-values:
+were applied, see below).  The class has \code{print} and
+\code{summary} methods to see the results.  The \code{print} method
+provides summary information on convergence, the log-likelihood and
+the AIC and BIC values:
 <<>>=
@@
 <<>>=
-summary(fm) 
+summary(fm, which="transition") 
 @
@@
 <<>>=
-summary(fm)
+summary(fm,which="response")
 @
@@
 <<>>=
-summary(fm)
-@
+summary(fm,which="prior")
+@
+
+%
+% add more explanation about the parameters, possibly add a plot of 
+% the class membership as function of age
+%
 
@@
 
-Using package \pkg{Rdonlp2} by \citet{Tamura2009}, parameters may be
-fitted subject to general linear (in-)equality constraints.
-Constraining and fixing parameters is done using the \code{conpat}
-argument to the \code{fit}-function, which specifies for each
-parameter in the model whether it's fixed (0) or free (1 or higher).
-Equality constraints can be imposed by having two parameters have the
-same number in the \code{conpat} vector.  When only fixed values are
-required, the \code{fixed} argument can be used instead of
-\code{conpat}, with zeroes for fixed parameters and other values (ones
-e.g.) for non-fixed parameters.  Fitting the models subject to these
-constraints is handled by the optimization routine \code{donlp2}.  To
-be able to construct the \code{conpat} and/or \code{fixed} vectors one
-needs the correct ordering of parameters which is briefly discussed
-next before proceeding with an example.
+Using package \pkg{Rdonlp2} by \citet{Tamura2009} or \pkg{Rsolnp}
+\citet{ZZZZ???}, parameters may be fitted subject to general linear
+(in-)equality constraints.  Constraining and fixing parameters is done
+using the \code{conpat} argument to the \code{fit}-function, which
+specifies for each parameter in the model whether it's fixed (0) or
+free (1 or higher).  Equality constraints can be imposed by having two
+parameters have the same number in the \code{conpat} vector.  When
+only fixed values are required, the \code{fixed} argument can be used
+instead of \code{conpat}, with zeroes for fixed parameters and other
+values (ones e.g.) for non-fixed parameters.  Fitting the models
+subject to these constraints is handled by the optimization routine
+\code{donlp2}.  To be able to construct the \code{conpat} and/or
+\code{fixed} vectors one needs the correct ordering of parameters
+which is briefly discussed next before proceeding with an example.
 
@@
 )
+
 @
@@
     function(object,log=FALSE) {
-        dexGAUS(object@y, mu = predict(object), sigma = exp(object@parameters$sigma), nu = exp(object@parameters$nu), log = log)
+    	dexGAUS(object@y, mu = predict(object), 
+	sigma = exp(object@parameters$sigma), 
+	nu = exp(object@parameters$nu), 
+	log = log)
     }

### 2010-02-25 papers/jss/dpx4Rev.tex
Various changes in response to reviewers for jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@362 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Minor updates in jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@372 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{center}
-\begin{Schunk}
-\begin{Soutput}
-##
-## Markov Chain Monte Carlo Package (MCMCpack)
-## Copyright (C) 2003-2010 Andrew D. Martin, Kevin M. Quinn, and Jong Hee Park
-##
-## Support provided by the U.S. National Science Foundation
-## (Grants SES-0350646 and SES-0350613)
-##
-\end{Soutput}
-\end{Schunk}
 \includegraphics{dpx4Rev-001}
@@
 already indicated in the Introduction, it consists of three time
-series with three variables: response time, accuracy, and a covariate
-Pacc which defines the relative pay-off for speeded versus accurate
-responding.  The participant in this experiment switches between fast
-responding at chance level and relatively slower responding at a high
-level of accuracy.  Interesting hypotheses to test are: is the
-switching regime symmetric?  Is there evidence for two states or does
-one state suffice?  Is the guessing state actually a guessing state,
-i.e., is the probability of a correct response at chance level (0.5)?
+series with three variables: response time, accuracy, and a covariate,
+Pacc, which defines the relative pay-off for speeded versus accurate
+responding.  Before describing some of the models that are fitted
+to these data, we provide a brief sketch of the reasons for gathering
+these data in the first place.
+
+Response times are a very common dependent variable in psychological
+experiments and hence form the basis for infernce about many
+psychological processes.  A potential threat to such inference based
+on response times is formed by the speed-accuracy trade-off: different
+participants in an experiment may respond differently to typical
+instructions to `respond as fast and accurate as possible'.  A popular
+model which takes the speed-accuracy trade-off into account is the
+diffusion model \citep{Ratcliff1978}, which has proven to provide
+accurate descriptions of response times in many different settings.
+
+One potential problem with the diffusion model is that it predicts a
+continuous trade-off between speed and accuracy of responding, i.e.,
+when participants are pressed to respond faster and faster, the
+diffusion model predicts that this would lead to a gradual decrease in
+accuracy.  The \code{speed} data set that we analyze below was
+gathered to test this hypothesis versus the alternative hypothesis
+stating that there is a sudden transition from slow and accurate
+responding to fast responding at chance level.  At each trial of the
+experiment, the participant is shown the current setting of the
+relative reward for speed versus accuracy.  The bottom panel of
+figure~ref{fig:speed} shows the values of this variable.  The main
+question in this experiment was what would happen when this reward
+variable would change from reward for accuracy only to reward for
+speed only.  The \code{speed} data that we analyse here are from
+participant A in Experiment 1 in \citet{Dutilh2009}, who provide a
+complete description of the experiment and the relevant theoretical
+background.
+
+The central hypothesis about this data is whether indeed it is best
+described by two modes of responding rather than a single mode of
+responding with a continuous trade-off between speed and accuracy.
+The hallmark of having a discontinuity between slow versus speeded
+responding is that the switching between the two modes is assymetric
+\citep[see e.g.][for a theoretical underpinning of this
+claim]{Maas1992}.  The \code{fit} help page of \pkg{depmixS4} provides
+a number of examples in which the assymetricity of the switching
+process is tested; those examples and other candidate models are
+discussed at length in \citet{Visser2009b}.
 
@@
 \code{depmix}-function as follows:
-
-The \code{depmix} function returns an object of class \code{depmix}
-which contains the model specification (and not a fitted model!).
-Note also that start values for the transition parameters are provided
-in this call using the \code{trstart} argument.  At this time, the
-package does not provide automatic starting values.
-
-\begin{Schunk}
+\begin{Schunk}
+\begin{Sinput}
+> library(depmixS4)
+> data(speed)
+> set.seed(1)
+> mod <- depmix(response = rt ~ 1, data = speed, nstates = 2, trstart = runif(4))
+\end{Sinput}
+\end{Schunk}
+
+The first line of code loads the \pkg{depmixS4} package and 
+\code{data(speed)} loads the \code{speed} data set. The line
+\code{set.seed(1)} is necessary to get starting values that will 
+result in the right model, see more on starting values below.
+
+The call to \code{depmix} specifies the model with a number of
+arguments.  The \code{response} argument is used to specify the
+response variable, possibly with covariates, in the familiar format
+using formulae such as in \code{lm} or \code{glm} models.  The second
+argument, \code{data=speed}, provides the \code{data.frame} in which
+the variables from \code{response} are interpreted.  Third, the number
+of states of the model is given by \code{nstates=2}.
+
+\paragraph{Starting values} Note also that start values for the
+transition parameters are provided in this call using the
+\code{trstart} argument.  At this time, the package does not provide
+automatic starting values.  Starting values for parameters can be
+provided using three arguments: \code{instart} for the parameters
+relating to the prior probabilities, \coce{trstart}, relating the
+transition models, and \code{respstart} for the parameters of the
+response models.  Note that the starting values for the initial and
+transition models as well as of the multinomial logit response models
+are interpreted as {\em probabilities}, and internally converted to
+multinomial logit parameters.
+
+\paragraph{Fitting the model and printing results} The \code{depmix}
+function returns an object of class \code{depmix} which contains the
+model specification, and not a fitted model.  Hence, the model needs
+to be fitted by calling \code{fit}:
+\begin{Schunk}
+\begin{Sinput}
+> fm <- fit(mod)
+\end{Sinput}
 \begin{Soutput}
@@
 \end{Schunk}
+
 The \code{fit}-function returns an object of class
@@
 convergence information (and information about constraints if these
-were applied, see below).  The \code{print} method provides summary
-information on convergence, the log-likelihood and the AIC and BIC
-values:
-\begin{Schunk}
-\begin{Soutput}
-Convergence info: Log likelihood converged to within tol. 
+were applied, see below).  The class has \code{print} and
+\code{summary} methods to see the results.  The \code{print} method
+provides summary information on convergence, the log-likelihood and
+the AIC and BIC values:
+\begin{Schunk}
+\begin{Sinput}
+> fm
+\end{Sinput}
+\begin{Soutput}
+Convergence info: Log likelihood converged to within tol. (relative change crit.) 
 'log Lik.' -84.34175 (df=7)
@@
 parameter estimates, first for the prior probabilities model, second
-for the transition model, and third for the response models.
-\begin{Schunk}
+for the transition models, and third for the response models.
+\begin{Schunk}
+\begin{Sinput}
+> summary(fm)
+\end{Sinput}
 \begin{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> set.seed(1)
+> mod <- depmix(rt ~ 1, data = speed, nstates = 2, family = gaussian(), 
++     transition = ~scale(Pacc), instart = runif(2))
+> fm <- fit(mod)
+\end{Sinput}
 \begin{Soutput}
@@
 
-Applying \code{summary} to the \code{fit}ted object gives (only transition models
-printed here): 
-\begin{Schunk}
-\begin{Soutput}
-Initial state probabilties model 
-Model of type multinomial, formula: ~1
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 10.71779
-Probalities at zero values of the covariates.
-2.214681e-05 0.9999779
-
+Applying \code{summary} to the \code{fit}ted object gives (only
+transition models printed here):
+\begin{Schunk}
+\begin{Sinput}
+> summary(fm, which = "transition")
+\end{Sinput}
+\begin{Soutput}
 Transition model for state (component) 1 
@@
 0.07788458 0.9221154 
-
-Response model(s) for state 1
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 5.512179
-sd  0.1921098
-
-Response model(s) for state 2
-
-Response model for response 1 
-Model of type gaussian, formula: rt ~ 1
-Coefficients: 
-[1] 6.3885
-sd  0.2402693 
 \end{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> set.seed(1)
+> mod <- depmix(list(rt ~ 1, corr ~ 1), data = speed, nstates = 2, 
++     family = list(gaussian(), multinomial()), transition = ~scale(Pacc), 
++     instart = runif(2))
+> fm <- fit(mod)
+\end{Sinput}
 \begin{Soutput}
@@
 iteration 30 logLik: -248.9723 
-iteration 33 logLik: -248.9723 
+iteration 31 logLik: -248.9723 
 \end{Soutput}
@@
 \begin{Schunk}
-\begin{Soutput}
-Initial state probabilties model 
-Model of type multinomial, formula: ~1
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 10.24230
-Probalities at zero values of the covariates.
-3.562961e-05 0.9999644
-
-Transition model for state (component) 1 
-Model of type multinomial, formula: ~scale(Pacc)
-Coefficients: 
-     [,1]       [,2]
-[1,]    0 -0.8924638
-[2,]    0  2.1293542
-Probalities at zero values of the covariates.
-0.7093984 0.2906016
-
-Transition model for state (component) 2 
-Model of type multinomial, formula: ~scale(Pacc)
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 2.390396
-[2,]    0 3.690785
-Probalities at zero values of the covariates.
-0.08390802 0.916092
-
+\begin{Sinput}
+> summary(fm, which = "response")
+\end{Sinput}
+\begin{Soutput}
 Response model(s) for state 1
@@
 Coefficients: 
-[1] 5.52169
-sd  0.2028857
+[1] 5.521695
+sd  0.2028917
 
@@
      [,1]      [,2]
-[1,]    0 0.1030554
-Probalities at zero values of the covariates.
-0.4742589 0.5257411
+[1,]    0 0.1030452
+Probalities at zero values of the covariates.
+0.4742615 0.5257385
 
@@
 Coefficients: 
-[1] 6.39369
-sd  0.2373650
+[1] 6.393693
+sd  0.2373635
 
@@
      [,1]     [,2]
-[1,]    0 2.245514
-Probalities at zero values of the covariates.
-0.09573715 0.9042629 
+[1,]    0 2.245562
+Probalities at zero values of the covariates.
+0.09573297 0.904267 
 \end{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> data(balance)
+> balance$age <- balance$age - 5
+> set.seed(1)
+> mod <- mix(list(d1 ~ 1, d2 ~ 1, d3 ~ 1, d4 ~ 1), data = balance, 
++     nstates = 2, family = list(multinomial(), multinomial(), 
++         multinomial(), multinomial()), respstart = c(rep(c(0.9, 
++         0.1), 4), rep(c(0.1, 0.9), 4)), prior = ~age, initdata = balance)
+> fm <- fit(mod)
+\end{Sinput}
 \begin{Soutput}
@@
 iteration 5 logLik: -951.2919 
-iteration 10 logLik: -951.2918 
-iteration 10 logLik: -951.2918 
+iteration 8 logLik: -951.2918 
 \end{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> summary(fm, which = "prior")
+\end{Sinput}
 \begin{Soutput}
@@
      [,1]       [,2]
-[1,]    0 -2.5182573
-[2,]    0  0.5512996
-Probalities at zero values of the covariates.
-0.9254119 0.07458815
-
-Response model(s) for state 1
-
-Response model for response 1 
-Model of type multinomial, formula: d1 ~ 1
-Coefficients: 
-     [,1]      [,2]
-[1,]    0 -3.390018
-Probalities at zero values of the covariates.
-0.9673911 0.03260889
-
-Response model for response 2 
-Model of type multinomial, formula: d2 ~ 1
-Coefficients: 
-     [,1]      [,2]
-[1,]    0 -2.311851
-Probalities at zero values of the covariates.
-0.9098538 0.0901462
-
-Response model for response 3 
-Model of type multinomial, formula: d3 ~ 1
-Coefficients: 
-     [,1]      [,2]
-[1,]    0 -3.415948
-Probalities at zero values of the covariates.
-0.9681992 0.03180076
-
-Response model for response 4 
-Model of type multinomial, formula: d4 ~ 1
-Coefficients: 
-     [,1]      [,2]
-[1,]    0 -2.247889
-Probalities at zero values of the covariates.
-0.9044683 0.0955317
-
-Response model(s) for state 2
-
-Response model for response 1 
-Model of type multinomial, formula: d1 ~ 1
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 2.950105
-Probalities at zero values of the covariates.
-0.04973154 0.9502685
-
-Response model for response 2 
-Model of type multinomial, formula: d2 ~ 1
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 3.065500
-Probalities at zero values of the covariates.
-0.04455298 0.955447
-
-Response model for response 3 
-Model of type multinomial, formula: d3 ~ 1
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 2.808197
-Probalities at zero values of the covariates.
-0.05688286 0.9431171
-
-Response model for response 4 
-Model of type multinomial, formula: d4 ~ 1
-Coefficients: 
-     [,1]     [,2]
-[1,]    0 3.350205
-Probalities at zero values of the covariates.
-0.03388844 0.9661116 
-\end{Soutput}
-\end{Schunk}
+[1,]    0 -2.5182590
+[2,]    0  0.5512897
+Probalities at zero values of the covariates.
+0.925412 0.07458803 
+\end{Soutput}
+\end{Schunk}
+
+%
+% add more explanation about the parameters, possibly add a plot of 
+% the class membership as function of age
+%
 
@@
 \begin{Schunk}
+\begin{Sinput}
+> setpars(mod, value = 1:npar(mod))
+\end{Sinput}
 \begin{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> setpars(mod, getpars(mod, which = "fixed"))
+\end{Sinput}
 \begin{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> trst <- c(0.9, 0.1, 0, 0, 0.1, 0.9, 0, 0)
+> mod <- depmix(list(rt ~ 1, corr ~ 1), data = speed, transition = ~Pacc, 
++     nstates = 2, family = list(gaussian(), multinomial()), trstart = trst, 
++     inst = c(0.999, 0.001))
+> fm1 <- fit(mod)
+\end{Sinput}
 \begin{Soutput}
@@
 iteration 20 logLik: -248.9723 
-iteration 23 logLik: -248.9723 
+iteration 22 logLik: -248.9723 
 \end{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> conpat <- c(0, 1, rep(c(0, 1), 4), 1, 1, 0, 1, 1, 1, 0, 1)
+> conpat[6] <- conpat[10] <- 2
+> fm2 <- fit(fm1, equal = conpat)
+\end{Sinput}
 \begin{Soutput}
@@
 
-Loading required package: Rsolnp (version 0.3)
     1 fx=   0.000000e+00 upsi=  6.7e+00 b2n= -1.0e+00 umi=  0.0e+00 nr   1 si-1
-    2 fx=   2.640247e+02 upsi=  0.0e+00 b2n=  2.1e+01 umi=  0.0e+00 nr   1 si-1
-    3 fx=   2.636495e+02 upsi=  0.0e+00 b2n=  5.4e+01 umi=  0.0e+00 nr   1 si-1
-    4 fx=   2.587358e+02 upsi=  0.0e+00 b2n=  2.7e+02 umi=  0.0e+00 nr   1 si-1
-    5 fx=   2.528309e+02 upsi=  0.0e+00 b2n=  1.2e+02 umi=  0.0e+00 nr   1 si-1
-    6 fx=   2.523880e+02 upsi=  0.0e+00 b2n=  9.5e+01 umi=  0.0e+00 nr   1 si-1
-    7 fx=   2.523802e+02 upsi=  0.0e+00 b2n=  9.4e+01 umi=  0.0e+00 nr   1 si-1
-    8 fx=   2.521969e+02 upsi=  0.0e+00 b2n=  9.2e+01 umi=  0.0e+00 nr   1 si-1
-    9 fx=   2.515470e+02 upsi=  0.0e+00 b2n=  5.7e+01 umi=  0.0e+00 nr   1 si-1
-   10 fx=   2.510761e+02 upsi=  0.0e+00 b2n=  5.1e+01 umi=  0.0e+00 nr   1 si-1
-   11 fx=   2.507338e+02 upsi=  0.0e+00 b2n=  7.1e+01 umi=  0.0e+00 nr   1 si-1
-   12 fx=   2.503019e+02 upsi=  0.0e+00 b2n=  1.9e+01 umi=  0.0e+00 nr   1 si-1
-   13 fx=   2.502725e+02 upsi=  0.0e+00 b2n=  9.5e-01 umi=  0.0e+00 nr   1 si-1
+    2 fx=   2.640133e+02 upsi=  0.0e+00 b2n=  2.1e+01 umi=  0.0e+00 nr   1 si-1
+    3 fx=   2.636388e+02 upsi=  0.0e+00 b2n=  5.4e+01 umi=  0.0e+00 nr   1 si-1
+    4 fx=   2.587645e+02 upsi=  0.0e+00 b2n=  2.7e+02 umi=  0.0e+00 nr   1 si-1
+    5 fx=   2.528471e+02 upsi=  0.0e+00 b2n=  1.2e+02 umi=  0.0e+00 nr   1 si-1
+    6 fx=   2.523984e+02 upsi=  0.0e+00 b2n=  9.6e+01 umi=  0.0e+00 nr   1 si-1
+    7 fx=   2.523907e+02 upsi=  0.0e+00 b2n=  9.5e+01 umi=  0.0e+00 nr   1 si-1
+    8 fx=   2.522057e+02 upsi=  0.0e+00 b2n=  9.2e+01 umi=  0.0e+00 nr   1 si-1
+    9 fx=   2.515501e+02 upsi=  0.0e+00 b2n=  5.8e+01 umi=  0.0e+00 nr   1 si-1
+   10 fx=   2.510856e+02 upsi=  0.0e+00 b2n=  5.2e+01 umi=  0.0e+00 nr   1 si-1
+   11 fx=   2.507418e+02 upsi=  0.0e+00 b2n=  7.2e+01 umi=  0.0e+00 nr   1 si-1
+   12 fx=   2.503033e+02 upsi=  0.0e+00 b2n=  1.9e+01 umi=  0.0e+00 nr   1 si-1
+   13 fx=   2.502725e+02 upsi=  0.0e+00 b2n=  9.8e-01 umi=  0.0e+00 nr   1 si-1
    14 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.5e-01 umi=  0.0e+00 nr   1 si-1
-   15 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.0e-02 umi=  0.0e+00 nr   1 si-1
-   16 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.5e-02 umi=  0.0e+00 nr   1 si-1
-   17 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  6.7e-02 umi=  0.0e+00 nr   1 si-1
-   18 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.1e-02 umi=  0.0e+00 nr   1 si-1
-   19 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.3e-03 umi=  0.0e+00 nr   1 si-1
+   15 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.1e-02 umi=  0.0e+00 nr   1 si-1
+   16 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.1e-02 umi=  0.0e+00 nr   1 si-1
+   17 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  6.3e-02 umi=  0.0e+00 nr   1 si-1
+   18 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.9e-02 umi=  0.0e+00 nr   1 si-1
+   19 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.1e-03 umi=  0.0e+00 nr   1 si-1
    20 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.4e-04 umi=  0.0e+00 nr   1 si-1
-   21 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.0e-03 umi=  0.0e+00 nr   1 si-1
-   22 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-01 umi=  0.0e+00 nr   1 si-1
-   23 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  7.1e-02 umi=  0.0e+00 nr   1 si-1
-   24 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.8e-02 umi=  0.0e+00 nr   1 si-1
-   25 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.9e-02 umi=  0.0e+00 nr   1 si-1
-   26 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.3e-02 umi=  0.0e+00 nr   1 si-1
-   27 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.6e-02 umi=  0.0e+00 nr   1 si-1
-   28 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  5.3e-02 umi=  0.0e+00 nr   1 si-1
-   29 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  9.0e-02 umi=  0.0e+00 nr   1 si-1
-   30 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
-   31 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
-   32 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
-   33 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
-   34 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  8.8e-02 umi=  0.0e+00 nr   1 si-1
-   35 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  3.5e-02 umi=  0.0e+00 nr   1 si-1
-   36 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.3e-02 umi=  0.0e+00 nr   1 si-1
-   37 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.1e-02 umi=  0.0e+00 nr   1 si-1
-   38 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-02 umi=  0.0e+00 nr   1 si-1
-   39 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.9e-04 umi=  0.0e+00 nr   1 si-1
+   21 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.3e-03 umi=  0.0e+00 nr   1 si-1
+   22 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  9.4e-02 umi=  0.0e+00 nr   1 si-1
+   23 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-01 umi=  0.0e+00 nr   1 si-1
+   24 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-01 umi=  0.0e+00 nr   1 si-1
+   25 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.7e-02 umi=  0.0e+00 nr   1 si-1
+   26 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.6e-02 umi=  0.0e+00 nr   1 si-1
+   27 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.0e-03 umi=  0.0e+00 nr   1 si-1
+   28 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  2.0e-02 umi=  0.0e+00 nr   1 si-1
+   29 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  4.9e-02 umi=  0.0e+00 nr   1 si-1
+   30 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.2e-01 umi=  0.0e+00 nr   1 si-1
+   31 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.2e-01 umi=  0.0e+00 nr   1 si-1
+   32 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.2e-01 umi=  0.0e+00 nr   1 si-1
+   33 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.2e-01 umi=  0.0e+00 nr   1 si-1
+   34 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.2e-01 umi=  0.0e+00 nr   1 si-1
+   35 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.2e-01 umi=  0.0e+00 nr   1 si-1
+   36 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-01 umi=  0.0e+00 nr   1 si-1
+   37 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.1e-01 umi=  0.0e+00 nr   1 si-1
+   38 fx=   2.502723e+02 upsi=  0.0e+00 b2n=  1.5e-02 umi=  0.0e+00 nr   1 si-1
 
@@
 startvalue
-  1.1073511e+01  -4.2221899e+00   9.1331180e+00  -3.3737739e+00   1.5804725e+01 
-  5.5217096e+00   2.0290950e-01   1.0308039e-01   6.3936995e+00   2.3736013e-01 
-  2.2456377e+00
+  1.1073464e+01  -4.2226665e+00   9.1333225e+00  -3.3734642e+00   1.5804664e+01 
+  5.5217147e+00   2.0291526e-01   1.0305078e-01   6.3937023e+00   2.3735863e-01 
+  2.2456760e+00
 
@@
  KT-conditions satisfied, no further correction computed
- evaluations of f                          142
- evaluations of grad f                      39
- evaluations of constraints                143
+ evaluations of f                          149
+ evaluations of grad f                      38
+ evaluations of constraints                150
  evaluations of grads of constraints         0
- final scaling of objective           2.577477e+00
- norm of grad(f)                      5.469761e-01
- lagrangian violation                 1.047009e-05
+ final scaling of objective           2.578752e+00
+ norm of grad(f)                      5.469760e-01
+ lagrangian violation                 3.850303e-06
  feasibility violation                0.000000e+00
  dual feasibility violation           0.000000e+00
- optimizer runtime sec's              1.057179e+02
-
- optimal value of f =   2.50272256993535e+02
+ optimizer runtime sec's              1.093723e+02
+
+ optimal value of f =   2.50272256745793e+02
 
  optimal solution  x =
-  1.57776829062875e+01 -5.33950896009860e+00  1.26660714788775e+01
- -2.42709038921118e+00  1.26660714788775e+01  5.52688670442125e+00
-  2.09034923400015e-01  1.03019934348728e-01  6.39615051757101e+00
-  2.36133700896234e-01  2.27764349273245e+00
+  1.66634775085811e+01 -5.33950908658836e+00  1.26660713921942e+01
+ -2.42709046868253e+00  1.26660713921942e+01  5.52688670523495e+00
+  2.09034923220784e-01  1.03019938087600e-01  6.39615051769729e+00
+  2.36133701083779e-01  2.27764355098866e+00
 
@@
    22   1.0000000e+20    0.0000000e+00 
-   23   0.0000000e+00    3.8677045e-01    1.4142136e+00
+   23   0.0000000e+00    3.8677046e-01    1.4142136e+00
    24   0.0000000e+00    0.0000000e+00
@@
  evaluations of restrictions and their gradients
- (   143,     0)
+ (   150,     0)
  last estimate of cond.nr. of active gradients   1.000e+00
 
- last estimate of cond.nr. of approx.  hessian   3.607e+05
-iterative steps total              39
+ last estimate of cond.nr. of approx.  hessian   3.187e+06
+iterative steps total              38
 # of restarts                       1
-# of full regular updates          37
-# of updates                       37
+# of full regular updates          36
+# of updates                       36
 # of regularized full SQP-steps     0
@@
 \begin{Schunk}
+\begin{Sinput}
+> setClass("exgaus", contains = "response")
+\end{Sinput}
 \begin{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, 
++     ...) standardGeneric("exgaus"))
+\end{Sinput}
 \begin{Soutput}
@@
 \end{Soutput}
-\begin{Soutput}
- **********   GAMLSS Version 3.0-1 ********** 
+\begin{Sinput}
+> require(gamlss)
+\end{Sinput}
+\begin{Soutput}
+ **********   GAMLSS Version 3.1-0 ********** 
 For more on GAMLSS look at http://www.gamlss.com/ 
@@
 \end{Soutput}
+\begin{Sinput}
+> require(gamlss.dist)
+> setMethod("exgaus", signature(y = "ANY"), function(y, pstart = NULL, 
++     fixed = NULL, ...) {
++     y <- matrix(y, length(y))
++     x <- matrix(1)
++     parameters <- list()
++     npar <- 3
++     if (is.null(fixed)) 
++         fixed <- as.logical(rep(0, npar))
++     if (!is.null(pstart)) {
++         if (length(pstart) != npar) 
++             stop("length of 'pstart' must be ", npar)
++         parameters$mu <- pstart[1]
++         parameters$sigma <- log(pstart[2])
++         parameters$nu <- log(pstart[3])
++     }
++     mod <- new("exgaus", parameters = parameters, fixed = fixed, 
++         x = x, y = y, npar = npar)
++     mod
++ })
+\end{Sinput}
 \begin{Soutput}
@@
 \begin{Schunk}
+\begin{Sinput}
+> setMethod("fit", "exgaus", function(object, w) {
++     if (missing(w)) 
++         w <- NULL
++     y <- object@y
++     fit <- gamlss(y ~ 1, weights = w, family = exGAUS(), control = gamlss.control(n.cyc = 100, 
++         trace = FALSE), mu.start = object@parameters$mu, sigma.start = exp(object@parameters$sigma), 
++         nu.start = exp(object@parameters$nu))
++     pars <- c(fit$mu.coefficients, fit$sigma.coefficients, fit$nu.coefficients)
++     object <- setpars(object, pars)
++     object
++ })
+\end{Sinput}
 \begin{Soutput}
@@
 We first create all the response models that we need as a double list: 
+\begin{Schunk}
+\begin{Sinput}
+> rModels <- list(list(exgaus(speed$rt, pstart = c(5, 0.1, 0.1)), 
++     GLMresponse(formula = corr ~ 1, data = speed, family = multinomial(), 
++         pstart = c(0.5, 0.5))), list(exgaus(speed$rt, pstart = c(6, 
++     0.1, 0.1)), GLMresponse(formula = corr ~ 1, data = speed, 
++     family = multinomial(), pstart = c(0.1, 0.9))))
+\end{Sinput}
+\end{Schunk}
 
@@
 the response class): 
+\begin{Schunk}
+\begin{Sinput}
+> trstart = c(0.9, 0.1, 0.1, 0.9)
+> transition <- list()
+> transition[[1]] <- transInit(~Pacc, nst = 2, data = speed, pstart = c(0.9, 
++     0.1, 0, 0))
+> transition[[2]] <- transInit(~Pacc, nst = 2, data = speed, pstart = c(0.1, 
++     0.9, 0, 0))
+> inMod <- transInit(~1, ns = 2, pstart = c(0.1, 0.9), data = data.frame(1))
+\end{Sinput}
+\end{Schunk}
 
@@
 \begin{Schunk}
+\begin{Sinput}
+> mod <- makeDepmix(response = rModels, transition = transition, 
++     prior = inMod, stat = FALSE)
+> fm <- fit(mod)
+\end{Sinput}
 \begin{Soutput}
@@
 iteration 40 logLik: -232.2526 
-iteration 45 logLik: -232.2526 
-iteration 50 logLik: -232.2526 
-iteration 52 logLik: -232.2526 
+iteration 43 logLik: -232.2526 
 \end{Soutput}

### 2010-03-03 papers/jss/dpx4Rev.Rnw
Jss paper updates, removed .bib files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@374 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 provided using three arguments: \code{instart} for the parameters
-relating to the prior probabilities, \coce{trstart}, relating the
+relating to the prior probabilities, \code{trstart}, relating the
 transition models, and \code{respstart} for the parameters of the
@@
 
-\subsection{Covariates on transition parameters\label{sec:trans}}
+\subsection{Covariates on transition parameters}
 
@@
 
-\subsection{Adding covariates on the prior probabilities\label{sec:prior}}
+\subsection{Adding covariates on the prior probabilities}
 
@@
 The balance scale taks is a famous task for testing cognitive
-strategies developed by Piaget \citep{Siegler1981}.
+strategies developed by Jean Piaget \citep[see]{Siegler1981}.
 Figure~\ref{fig:balance} provides an example of a balance scale item.
@@
 \begin{center}
-	\includegraphic{graphs:baldist.pdf}
+	\includegraphics{graphs/baldist.pdf}
 	\caption{Balance scale item; this is a so-called distance item.}
@@
 only do so when the number of weights is equal on each side.  These
-two strategies that children employ are known a Rule I and Rule II.
+two strategies that children employ are known as Rule I and Rule II.
 Other strategies can be teased apart by administering different items.
@@
 consisted of 25 items; other items in the test are used to detect
-other strategies that children (and young adults) employ in solving
+other strategies that children and young adults employ in solving
 balance scale items \citep[see][for details]{Jansen2002}.
 
-Similarly to the transition matrix, covariates on the prior
-probabilities of the latent states (or classes in this case), are
-defined by using a one-sided formula:
+In the following model, age is included as covariate on class
+membership to test whether, with age, children apply more complex
+rules in solving balance scale items.  Similarly to the transition
+matrix, covariates on the prior probabilities of the latent states (or
+classes in this case), are defined by using a one-sided formula
+\code{prior=~age}: 
 <<>>= 
 data(balance)
-balance$age <- balance$age-5
 set.seed(1)
-mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
-    family=list(multinomial(), multinomial(), multinomial(),
-    multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
-    prior=~age, initdata=balance)
+mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=3,
+	family=list(multinomial("identity"), multinomial("identity"),
+	multinomial("identity"), multinomial("identity")),
+	respstart=runif(24), prior=~age, initdata=balance)
 fm <- fit(mod) 
+fm
 @
@@
 
+As can be seen from the print of the fitted model above, the BIC for 
+this model equals 1941.6. The similarly defined 2-class model for 
+these data has a BIC of 1969.2, and the 4-class model has BIC equal to
+1950.4. Hence, the 3-class seems to be adequate for describing these 
+data.
+
 The summary of the \code{fit}ted model gives the following (only the
@@
 
-%
-% add more explanation about the parameters, possibly add a plot of 
-% the class membership as function of age
-%
-
-Hence at young ages, children have a high probability of incorrect 
-responding in class~1, whereas the prior probability for class~2 
-increases with age.
+The intercept values of the multinomial logistic parameters provide
+the prior probabilities of each class when the covariate has value
+zero (note that in this case the value zero does not have much
+significance, scaling and/or centering of covariates may be useful in
+such cases).  The summary function prints these values.  As can be
+seen from those values, at age zero, the prior probability is
+overwhelmingly at the second class.  Inspection of the response
+parameters reveals that class~2 is associated with incorrect
+responding, whereas class~1 is associated with correct responding;
+class~3 is an intermediate class with guessing behavior.
+Figure~\ref{fig:balprior} depicts the prior class probabilities as
+function of age based on the fitted parameters
+
+\begin{figure}[htbp]
+\begin{center}
+<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=
+x <- mlogit(base=1)
+coeff <- coefficients(fm@prior@parameters)
+
+pr1 <- function(y) {sapply(y, function(z) {x$linkinv(c(t(coeff)%*%c(1,z)), base=1)[1]})}
+pr2 <- function(y) {sapply(y, function(z) {x$linkinv(c(t(coeff)%*%c(1,z)), base=1)[2]})}
+pr3 <- function(y) {sapply(y, function(z) {x$linkinv(c(t(coeff)%*%c(1,z)), base=1)[3]})}
+
+plot(pr1,min(balance$age),max(balance$age),lty=1,ylim=c(0,1),
+main="Prior probabilities by age, balance scale data", xlab="age", ylab="Pr")
+plot(pr2,min(balance$age),max(balance$age),add=T,lty=2)
+plot(pr3,min(balance$age),max(balance$age),add=T,lty=3)
+
+legend("right",legend=c("Class 1 (correct)","Class 2 (incorrect)","Class 3 (guess)"),lty=1:3,inset=c(0.1,0))
+@
+	\caption{Class probabilities as a function of age.}
+	\label{fig:balprior}  
+\end{center}
+\end{figure}
+
+As can be seen from Figure~\ref{fig:balprior}, at younger ages 
+children predominantly apply Rule I, the wrong strategy for these 
+items. According to the model, approximately 90 \% of children at age
+5 apply Rule I. The remaining 10 \% are evenly split among the 2 
+other classes. At age 19, almost all participants belong to class~1. 
+Interestingly, prior probability of the 'guess' class 2, first 
+increases with age, and then decreases again. This suggests that 
+children pass through a phase in which they are uncertain or possibly 
+switch between applying different strategies.
 
@@
 setpars(mod, getpars(mod,which="fixed"))
-
 @
@@
 After this, we use the fitted values from this model to constrain the
-regression coefficients on the transition matrix (parameters numbers~6
+regression coefficients on the transition matrix (parameters number~6
 and~10):

### 2010-03-03 papers/jss/dpx4Rev.tex
Jss paper updates, removed .bib files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@374 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{Soutput}
-iteration 0 logLik: -305.3302 
-iteration 5 logLik: -305.3251 
-iteration 10 logLik: -305.3133 
-iteration 15 logLik: -305.2824 
-iteration 20 logLik: -305.1948 
-iteration 25 logLik: -304.8879 
-iteration 30 logLik: -302.2484 
-iteration 35 logLik: -109.1782 
-iteration 40 logLik: -84.3722 
-iteration 45 logLik: -84.34183 
-iteration 50 logLik: -84.34175 
-iteration 50 logLik: -84.34175 
+iteration 0 logLik: -305.3323 
+iteration 5 logLik: -305.3291 
+iteration 10 logLik: -305.3234 
+iteration 15 logLik: -305.3093 
+iteration 20 logLik: -305.2689 
+iteration 25 logLik: -305.1367 
+iteration 30 logLik: -304.5522 
+iteration 35 logLik: -291.1151 
+iteration 40 logLik: -87.53257 
+iteration 45 logLik: -84.34731 
+iteration 50 logLik: -84.34178 
+iteration 54 logLik: -84.34176 
 \end{Soutput}
@@
 Convergence info: Log likelihood converged to within tol. (relative change crit.) 
-'log Lik.' -84.34175 (df=7)
+'log Lik.' -84.34176 (df=7)
 AIC:  182.6835 
@@
      [,1]      [,2]
-[1,]    0 -11.25688
-Probalities at zero values of the covariates.
-0.999987 1.291797e-05
+[1,]    0 -11.08968
+Probalities at zero values of the covariates.
+0.9999847 1.526887e-05
 
@@
 Coefficients: 
-[1]  0.000000 -2.392455
-Probalities at zero values of the covariates.
-0.9162501 0.08374986
+[1] 0.9162498 0.0837502
 
@@
 Coefficients: 
-[1] 0.000000 2.139255
-Probalities at zero values of the covariates.
-0.1053396 0.8946604
+[1] 0.1053408 0.8946592
 
@@
 Coefficients: 
-[1] 6.385492
-sd  0.2439376
+[1] 6.385491
+sd  0.2439384
 
@@
 Coefficients: 
-[1] 5.511151
-sd  0.1926063 
+[1] 5.511149
+sd  0.1926048 
 \end{Soutput}
@@
 
+Loading required package: Rsolnp (version 0.3)
     1 fx=   0.000000e+00 upsi=  6.7e+00 b2n= -1.0e+00 umi=  0.0e+00 nr   1 si-1
@@
  dual feasibility violation           0.000000e+00
- optimizer runtime sec's              1.093723e+02
+ optimizer runtime sec's              1.020294e+02
 
@@
 \begin{Soutput}
- **********   GAMLSS Version 3.1-0 ********** 
+ **********   GAMLSS Version 3.0-1 ********** 
 For more on GAMLSS look at http://www.gamlss.com/ 

### 2010-03-03 papers/jss/dpx4Rev.Rnw
Cleaned up jss directory, added new bib files
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@375 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	estimating dependent mixture models in the \proglang{R}
-	programming language \citep{R2009}.  This includes standard Markov
+	programming language \citep{R2010}.  This includes standard Markov
 	models, latent/hidden Markov models, and latent class and finite
@@
 Using package \pkg{Rdonlp2} by \citet{Tamura2009} or \pkg{Rsolnp}
-\citet{ZZZZ???}, parameters may be fitted subject to general linear
+\citet{Ghalanos2010}, parameters may be fitted subject to general linear
 (in-)equality constraints.  Constraining and fixing parameters is done

### 2010-03-03 papers/jss/dpx4Rev.Rnw
Paper and bib file updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@376 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2010-03-03 papers/jss/dpx4Rev.Rnw
@@
 %% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
-
-% Refs to check: Jansen 2002, Dutilh 2009
 
@@
 Figure~\ref{fig:speed}).  These data are more fully described in
-\citet{Dutilh2009}.
+\citet{Dutilh2010}.
 
@@
 speed only.  The \code{speed} data that we analyse here are from
-participant A in Experiment 1 in \citet{Dutilh2009}, who provide a
+participant A in Experiment 1 in \citet{Dutilh2010}, who provide a
 complete description of the experiment and the relevant theoretical
@@
     transition=~scale(Pacc), instart=runif(2))
-fm <- fit(mod)
+fm <- fit(mod,verbose=FALSE)
 @
@@
     transition=~scale(Pacc),instart=runif(2))
-fm <- fit(mod)
+fm <- fit(mod,verbose=FALSE)
 @
@@
 	respstart=runif(24), prior=~age, initdata=balance)
-fm <- fit(mod) 
+fm <- fit(mod,verbose=FALSE) 
 fm
@@
 To see the ordering of parameters use the following:
-<<>>=
+<<term=FALSE>>=
 setpars(mod, value=1:npar(mod))
@@
 initial state probabilities model):
-<<>>=
+<<term=FALSE>>=
 setpars(mod, getpars(mod,which="fixed"))
@@
 	  trstart=trst,inst=c(.999,0.001))
-fm1 <- fit(mod)
+fm1 <- fit(mod,verbose=FALSE)
 @
@@
 and~10):
-<<>>=
+<<term=FALSE>>=
 # start with fixed and free parameters
@@
 defined as follows:
-<<>>=
+<<term=FALSE>>=
 setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
@@
     prior=inMod,stat=FALSE)
-fm <- fit(mod)
+fm <- fit(mod,verbose=FALSE)
 @
@@
 Economic Learning and Social Evolution (ELSE).  Han van der Maas
-provided the speed-accuracy data \cite{Dutilh2009} and thereby
+provided the speed-accuracy data \cite{Dutilh2010} and thereby
 neccessitated implementing models with time-dependent covariates.

### 2010-03-04 papers/jss/dpx4Rev.Rnw
- identity link now default in init models - minor changes to the jss article
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@377 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 function for the response time variable, and a Bernoulli distribution
-for the accuracy variable.  In the models we are considering here,
+for the accuracy variable.  In the models considered here,
 both the transition probability functions $a_{ij}$ and the initial
@@
 Response times are a very common dependent variable in psychological
-experiments and hence form the basis for infernce about many
+experiments and hence form the basis for inference about many
 psychological processes.  A potential threat to such inference based
@@
 relative reward for speed versus accuracy.  The bottom panel of
-figure~ref{fig:speed} shows the values of this variable.  The main
-question in this experiment was what would happen when this reward
-variable would change from reward for accuracy only to reward for
+figure~\ref{fig:speed} shows the values of this variable.  The 
+experiment was designed to investigate what would happen when this reward
+variable changes from reward for accuracy only to reward for
 speed only.  The \code{speed} data that we analyse here are from
@@
 
-The central hypothesis about this data is whether indeed it is best
+The central question regarding this data is whether it is indeed best
 described by two modes of responding rather than a single mode of
 responding with a continuous trade-off between speed and accuracy.
-The hallmark of having a discontinuity between slow versus speeded
-responding is that the switching between the two modes is assymetric
+The hallmark of a discontinuity between slow versus speeded
+responding is that switching between the two modes is asymmetric
 \citep[see e.g.][for a theoretical underpinning of this
 claim]{Maas1992}.  The \code{fit} help page of \pkg{depmixS4} provides
-a number of examples in which the assymetricity of the switching
+a number of examples in which the asymmetry of the switching
 process is tested; those examples and other candidate models are
@@
 set.seed(1)
-mod <- depmix(response=rt~1, data=speed, nstates=2, trstart=runif(4))
+mod <- depmix(response=rt~1, data=speed, nstates=2, 
+    trstart=runif(4))
 @
@@
 \paragraph{Starting values} Note also that start values for the
-transition parameters are provided in this call using the
+transition parameters are provided in this call by the
 \code{trstart} argument.  At this time, the package does not provide
@@
 response models.  Note that the starting values for the initial and
-transition models as well as of the multinomial logit response models
+transition models as well as multinomial logit response models
 are interpreted as {\em probabilities}, and internally converted to
-multinomial logit parameters.
+multinomial logit parameters (if a logit link function is used).
 
@@
 these data, i.e. assuming there is no mixture, has a log-likelihood of
--305.33, and 614.66, and 622.83 for the AIC and BIC respectively.
+$-305.33$, and 614.66, and 622.83 for the AIC and BIC respectively.
 Hence, the 2-state model fits the data much better than the 1-state
@@
 By default, the transition probabilities and the initial state
-probabilities are parameterized using the multinomial logistic model.
-More precisely, each row of the transition matrix is parameterized by
+probabilities are parameterized using a multinomial model with an 
+identity link function.
+Using a multinomial logistic model allows us to include covariates on the
+initial state and transition probabilities. In this case, each row of the 
+transition matrix is parameterized by
 a baseline category logistic multinomial, meaning that the parameter
@@
 state probability model would have three parameters of which the first
-is fixed at zero and the other two are freely estimated.
-
-The multinomial logistic model allows us to include covariates on the
-initial state and transition probabilities.  \citet{Chung2007} discuss
+is fixed at zero and the other two are freely estimated. \citet{Chung2007} discuss
 a related latent transition model for repeated measurement data
@@
 responses.  In above examples we have only used the response times
-which were modelled with the Gaussian distribution.  The accuracy
+which were modelled as a Gaussian distribution.  The accuracy
 variable in the \code{speed} data can be modelled with a multinomial
@@
 The balance scale taks is a famous task for testing cognitive
-strategies developed by Jean Piaget \citep[see]{Siegler1981}.
+strategies developed by Jean Piaget \citep[see][]{Siegler1981}.
 Figure~\ref{fig:balance} provides an example of a balance scale item.
@@
 Note here that we define a \code{mix} model instead of a \code{depmix}
-models as these data form independent observations.  More formally,
+model as these data form independent observations.  More formally,
 \code{depmix} models extend the class of \code{mix} models by adding
-the transition models.  As for fitting \code{mix} models: as can be
-seen in equation~\ref{eq:Q}, the EM algorithm can be applied by simply
+transition models.  As for fitting \code{mix} models: as can be
+seen in Equation~\ref{eq:Q}, the EM algorithm can be applied by simply
 dropping the second summand containing the transition parameters, and 
@@
 specifies for each parameter in the model whether it's fixed (0) or
-free (1 or higher).  Equality constraints can be imposed by having two
-parameters have the same number in the \code{conpat} vector.  When
+free (1 or higher).  Equality constraints can be imposed by giving two
+parameters the same number in the \code{conpat} vector.  When
 only fixed values are required, the \code{fixed} argument can be used
@@
 \code{fixed} arguments, complete parameter vectors should be supplied,
-i.e., these vectors should have length of the number of parameters of
+i.e., these vectors should have length equal to the number of parameters of
 the model, which can be obtained by calling \code{npar(object)}.  Note
@@
 	\item paramaters: a named list with the coefficients and possibly 
-	other parameters, e.g., the standard deviation in the Gaussian 
-	response model
+	other parameters (e.g., the standard deviation in the Gaussian 
+	response model)
 	\item fixed: a vector of logicals indicating whether parameters are 
@@
 After defining all the necessary methods for the new response model, 
-we can  now define the dependent mixture model using this reponse model. 
+we can  now define the dependent mixture model using this response model. 
 The function \code{makeDepmix} is included in \pkg{depmixS4} to have 
@@
 describes the data much better than the same model in which the
-response times are modeled as gaussian.
+response times are modelled as Gaussian.
 
@@
 this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
-intended for modeling of (individual) time series data with the aim of
+intended for modelling of (individual) time series data with the aim of
 characterizing the transition processes underlying the data.  The
@@
 Ingmar Visser was supported by an EC Framework 6 grant, project 516542
-(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
-Economic Learning and Social Evolution (ELSE).  Han van der Maas
-provided the speed-accuracy data \cite{Dutilh2010} and thereby
+(NEST).  Maarten Speekenbrink was supported by ESRC grant RES-062-23-1511 
+and the ESRC Centre for Economic Learning and Social Evolution (ELSE).  
+Han van der Maas provided the speed-accuracy data \cite{Dutilh2010} and thereby
 neccessitated implementing models with time-dependent covariates.

### 2010-03-04 papers/jss/dpx4Rev.Rnw
Yet more paper updates ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@378 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	for latent class and finite mixture models, see example below on adding
-	covariates to prior probabilities in section~\ref{sec:prior})
+	covariates to prior probabilities)
 

### 2010-03-04 papers/jss/dpx4Rev.Rnw
Minor changes in paper and getdf for transInit models (not done yet ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@382 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	parameters, by direct numerical optimization with the
-	\pkg{Rdonlp2} routine.}
-
-\Keywords{hidden Markov model, dependent mixture model, mixture model}
-
-\Plainkeywords{hidden Markov model, dependent mixture model, mixture model} %% without formatting
+	\pkg{Rdonlp2} or \pkg{Rsolnp} routines.}
+
+\Keywords{hidden Markov model, dependent mixture model, mixture 
+model, constraints}
+
+\Plainkeywords{hidden Markov model, dependent mixture model, mixture 
+model, constraints} %% without formatting
 %% at least one keyword must be supplied
@@
 of responses generated by a single participant in a psychological
-response time experiment.  The data consists of three variables,
+response time experiment.  The data consists of three variables:
 response time, response accuracy, and a covariate which is a pay-off
@@
 responding.  These variables are measured on 168, 134 and 137
-occasions respectively (the first part of this series is plotted in
+occasions respectively (the first series of 168 trials is plotted in
 Figure~\ref{fig:speed}).  These data are more fully described in
-\citet{Dutilh2010}.
+\citet{Dutilh2010}, and in the next section a number of example models
+for these data is described.
 
@@
 \begin{center}
+
 	<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=
@@
 	@
-	 \caption{Response times (rt), accuracy (corr) and pay-off values (Pacc) for
-	  the first series of responses in dataset \code{speed}.}
+
+	\caption{Response times (rt), accuracy (corr) and pay-off 
+	values (Pacc) for the first series of responses in dataset 
+	\code{speed}.}
+
 	\label{fig:speed}
@@
 univariate time series \citep[][chapter~1]{Cappe2005}.  We use the
-term ``dependent mixture model'' because one of the authors (Ingmar
+term "dependent mixture model" because one of the authors (Ingmar
 Visser) thought it was time for a new name to relate these
@@
 \end{enumerate}
-
-%In the next paragraphs, the likelihood
-%equation and estimation procedures for the dependent mixture model are
-%described for data of the above form.
-
-%The likelihood of the dependent mixture model conditional on the
-%unknown (or hidden) state sequence $\vc{S}_{1:T} = (S_1,\ldots,S_T)$ 
-%and the model is given
-%by:
-%\begin{equation}
-%	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
-%	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
-%\end{equation}
-%where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
-%at the marginal likelihood of the data for parameter vector
-%$\greekv{\theta}$, i.e.\ unconditional on the state sequence, we need to sum
-%above likelihood over all possible state sequences. This likelihood 
-%is written as: 
-%\begin{equation}
-%	\Prob(\vc{O}_{1:T}|\greekv{\theta}) =  
-%	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
-%	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
-%\end{equation}
-%where we have the following elements:
-%\begin{enumerate}
-%	
-%	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
-%	of $n$ latent classes or states; we write for short $S_{i}$ to
-%	denote $S_{t}=i$.
-%	
-%	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
-%	giving the probability of class/state $i$ at time $t=1$ with
-%	covariate $\vc{z}_1$.
-%	
-%	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
-%	provides the probability of a transition from state $i$ to state
-%	$j$ with covariate $\vc{z}_t$,
-%	
-%	\item $\vc{b}_{j}$ is a vector of observation densities
-%	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
-%	conditional densities of observations $O_{t}^k$ associated with
-%	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
-%	$k=1, \ldots, m$. 
-%\end{enumerate}
-
-% The dependent mixture model is defined by the following equations: 
-% %\begin{align}
-% %	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
-% %	O_{t} &= b(O_{t}|S_{t}), 
-% %\end{align}
-% \begin{align}
-% 	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
-% 	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
-% \end{align}
-%where $S_{t}$ is a sequence of hidden states, $A$ is a transition
-%matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
-%density function for $O_{t}$ conditional on the hidden state $S_{t}$.
 
@@
 consisting of three subvectors with parameters for the prior model,
-transition model, and response model respectively.  The joint
-log-likelihood can be written as
+transition model, and response models respectively.  The joint
+log-likelihood can be written as:
 \begin{equation}
@@
 Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
-(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
+(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta})),
 \end{equation}
-can be written as
+can be written as:
 \begin{multline}
@@
 
-%The maximising values of $a_{jk}$ are \cite<e.g.,>{Rabiner89}
-%\begin{equation}
-%\label{eq:MLA}
-%\hat{a}_{jk} = \frac{1}{N(T-1)} \sum_{i=1}^N \sum_{t=2}^T  \frac{\xi^i_t(j,k)}{\gamma_{t-1}^i(j)}.
-%\end{equation}
-
-%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does
-%not affect the estimation of the other elements.  When elements
-%$a_{jk}$ are assumed identical, we simply extend the summation in
-%(\ref{eq:MLA}) to include all those elements, changing the
-%denominator $N(T-1)$ accordingly.
-
-%To estimate $\greekv{\lambda}$, we note that the term containing this
-%parameter has the form of a weighted likelihood, where
-%$\gamma_t^i(j)$ can be interpreted as the number of replications of
-%$r^i_t$.  Hence, we can rely on standard ML estimates of the logistic
-%regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$
-%as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To
-%be more specific, we replicate each observation $r^i_t$ a total of 6
-%times, once for each of the states besides the random state (which
-%offers no information regarding $\lambda_i$).  For the $j$-th
-%replication (corresponding to the $j$-th state), we used
-%$v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a
-%case weight.  All these replications were used to obtain the maximum
-%likelihood estimate of $\lambda$ from a single GLM, using the
-%``glm.fit'' function in R \cite{R}.}.  The resulting parameter
-%estimates are then used in a new Expectation step, followed by
-%another Maximisation step, until convergence.  Under mild regularity
-%conditions, it can be shown that the EM algorithm converges to a
-%(local) maximum of the likelihood \cite{EM}.
-
 The EM algorithm however has some drawbacks.  First, it can be slow to
@@
 direct optimization using package \pkg{Rdonlp2}
-\citep{Tamura2009,Spellucci2002}, or package \pkg{Rsolnp}.  Both
-packages can handle general linear (in)equality constraints, and
-optionally also non-linear constraints.
-
-%Need some more on EM and how/why it is justified to do separate weighted
-%fits of the response models and transition and prior models.
-
-%Also mention use of glm, nnet and possibly other packages that we use.
+\citep{Tamura2009,Spellucci2002}, or package \pkg{Rsolnp}
+\citep{Ghalanos2010}.  Both packages can handle general linear
+(in)equality constraints (and optionally also non-linear constraints).
 
@@
 Pacc, which defines the relative pay-off for speeded versus accurate
-responding.  Before describing some of the models that are fitted
-to these data, we provide a brief sketch of the reasons for gathering
+responding.  Before describing some of the models that are fitted to
+these data, we provide a brief sketch of the reasons for gathering
 these data in the first place.
@@
 figure~\ref{fig:speed} shows the values of this variable.  The
-experiment was designed to investigate what would happen when this reward
-variable changes from reward for accuracy only to reward for
+experiment was designed to investigate what would happen when this
+reward variable changes from reward for accuracy only to reward for
 speed only.  The \code{speed} data that we analyse here are from
@@
 responding with a continuous trade-off between speed and accuracy.
-The hallmark of a discontinuity between slow versus speeded
-responding is that switching between the two modes is asymmetric
-\citep[see e.g.][for a theoretical underpinning of this
-claim]{Maas1992}.  The \code{fit} help page of \pkg{depmixS4} provides
-a number of examples in which the asymmetry of the switching
-process is tested; those examples and other candidate models are
-discussed at length in \citet{Visser2009b}.
+The hallmark of a discontinuity between slow versus speeded responding
+is that switching between the two modes is asymmetric \citep[see
+e.g.][for a theoretical underpinning of this claim]{Maas1992}.  The
+\code{fit} help page of \pkg{depmixS4} provides a number of examples
+in which the asymmetry of the switching process is tested; those
+examples and other candidate models are discussed at length in
+\citet{Visser2009b}.
 

### 2010-03-10 papers/jss/dpx4Rev.Rnw
Final changes to jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@407 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 	multivariate data with distributions from the \code{glm} family,
-	the logistic multinomial, or the multivariate normal distribution.
+	the (logistic) multinomial, or the multivariate normal distribution.
 	Other distributions can be added easily, and an example is
@@
 	parameters, by direct numerical optimization with the
-	\pkg{Rdonlp2} or \pkg{Rsolnp} routines.}
+	\pkg{Rsolnp} or \pkg{Rdonlp2} routines.}
 
@@
 \begin{center}
-
 <<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
+.Options$digits<-3
 library(depmixS4)
@@
 @
-
 	\caption{Response times (rt), accuracy (corr) and pay-off 
@@
 	\code{speed}.}
-
 	\label{fig:speed}
-
 \end{center}
@@
 otherwise, direct optimization is used.  Two options are available for
-direct optimization using package \pkg{Rdonlp2}
-\citep{Tamura2009,Spellucci2002}, or package \pkg{Rsolnp}
-\citep{Ghalanos2010}.  Both packages can handle general linear
-(in)equality constraints (and optionally also non-linear constraints).
+direct optimization using package \pkg{Rsolnp}
+\citep{Ghalanos2010,Ye1987}, or \pkg{Rdonlp2}
+\citep{Tamura2009,Spellucci2002}.  Both packages can handle general
+linear (in)equality constraints (and optionally also non-linear
+constraints).
 
@@
 mod <- depmix(response=rt~1, data=speed, nstates=2, 
-    trstart=runif(4))
+    trst=runif(4))
 @
@@
 \paragraph{Starting values} Note also that start values for the
-transition parameters are provided in this call by the
-\code{trstart} argument.  At this time, the package does not provide
-automatic starting values.  Starting values for parameters can be
-provided using three arguments: \code{instart} for the parameters
-relating to the prior probabilities, \code{trstart}, relating the
-transition models, and \code{respstart} for the parameters of the
-response models.  Note that the starting values for the initial and
-transition models as well as multinomial logit response models
-are interpreted as {\em probabilities}, and internally converted to
-multinomial logit parameters (if a logit link function is used).
+transition parameters are provided in this call by the \code{trstart}
+argument.  Starting values for parameters can be provided using three
+arguments: \code{instart} for the parameters relating to the prior
+probabilities, \code{trstart}, relating the transition models, and
+\code{respstart} for the parameters of the response models.  Note that
+the starting values for the initial and transition models as well as
+multinomial logit response models are interpreted as {\em
+probabilities}, and internally converted to multinomial logit
+parameters (if a logit link function is used).  Start values can also
+be generated randomly within the EM algorithm by generating random
+uniform values for the values of $\gamma_{t}$ in (\ref{eq:Q}) at
+iteration 0.  Automatic generation of starting values is not available
+for constrained models (see below).
 
@@
 \code{summary} methods to see the results.  The \code{print} method
-provides summary information on convergence, the log-likelihood and
+provides information on convergence, the log-likelihood and
 the AIC and BIC values:
@@
 (multinomial distributions for the first two, and a Gaussian
-distribution for the response).
+distribution for the response).  The resulting model indicates two
+well-separated states, one with slow and the second with fast
+responses.  The transition probabilities indicate rather stable
+states, i.e., the probability in either of the states is around 0.9.
+The initial state probability estimates indicate that state 1 is the
+starting for the process, with a negligible probability of starting in
+state 2.
 
@@
 probabilities are parameterized using a multinomial model with an
-identity link function.
-Using a multinomial logistic model allows us to include covariates on the
-initial state and transition probabilities. In this case, each row of the 
-transition matrix is parameterized by
-a baseline category logistic multinomial, meaning that the parameter
-for the base category is fixed at zero \citep[see][p.\ 267 ff., for
-multinomial logistic models and various
+identity link function.  Using a multinomial logistic model allows us
+to include covariates on the initial state and transition
+probabilities.  In this case, each row of the transition matrix is
+parameterized by a baseline category logistic multinomial, meaning
+that the parameter for the base category is fixed at zero
+\citep[see][p.\ 267 ff., for multinomial logistic models and various
 parameterizations]{Agresti2002}.  The default baseline category is the
@@
 state probability model would have three parameters of which the first
-is fixed at zero and the other two are freely estimated. \citet{Chung2007} discuss
-a related latent transition model for repeated measurement data
-($T=2$) using logistic regression on the transition parameters; they
-rely on Bayesian methods of estimation.  Covariates on the transition
-probabilities can be specified using a one-sided formula as in the
-following example:
+is fixed at zero and the other two are freely estimated.
+\citet{Chung2007} discuss a related latent transition model for
+repeated measurement data ($T=2$) using logistic regression on the
+transition parameters; they rely on Bayesian methods of estimation.
+Covariates on the transition probabilities can be specified using a
+one-sided formula as in the following example: 
 <<>>=
@@
 	   transition=~scale(Pacc), instart=runif(2)) 
-fm <- fit(mod,verbose=FALSE)
-@
-
-Applying \code{summary} to the \code{fit}ted object gives (only
-transition models printed here):
+fm <-fit(mod,verbose=FALSE) 
+@
+
+Note the use of \code{verbose=FALSE} to suppress printing of
+information on the iterations of the fitting process.  Applying
+\code{summary} to the \code{fit}ted object gives (only transition
+models printed here by using argument \code{which}): 
 <<>>= 
@@
 variables.
+
+\subsection{Fixing and constraining parameters}
+
+Using package \pkg{Rsolnp} \citet{Ghalanos2010} or \pkg{Rdonlp2} by
+\citet{Tamura2009}, parameters may be fitted subject to general linear
+(in-)equality constraints.  Constraining and fixing parameters is done
+using the \code{conpat} argument to the \code{fit}-function, which
+specifies for each parameter in the model whether it's fixed (0) or
+free (1 or higher).  Equality constraints can be imposed by giving two
+parameters the same number in the \code{conpat} vector.  When only
+fixed values are required, the \code{fixed} argument can be used
+instead of \code{conpat}, with zeroes for fixed parameters and other
+values (ones e.g.) for non-fixed parameters.  Fitting the models
+subject to these constraints is handled by the optimization routine
+\code{donlp2}.  To be able to construct the \code{conpat} and/or
+\code{fixed} vectors one needs the correct ordering of parameters
+which is briefly discussed next before proceeding with an example.
+
+\paragraph{Parameter numbering} When using the \code{conpat} and
+\code{fixed} arguments, complete parameter vectors should be supplied,
+i.e., these vectors should have length equal to the number of parameters of
+the model, which can be obtained by calling \code{npar(object)}.  Note
+that this is not the same as the degrees of freedom used e.g.\ in the
+\code{logLik} function because \code{npar} also counts the baseline
+category zeroes from the multinomial logistic models.  Parameters are
+numbered in the following order:
+\begin{enumerate}
+	\item  the prior model parameters
+	\item  the parameters for the transition models
+	\item  the response model parameters per state (and subsequently
+	per response in the case of multivariate time series)
+\end{enumerate}
+
+To see the ordering of parameters use the following:
+<<term=FALSE>>=
+setpars(mod, value=1:npar(mod))
+@
+
+To see which parameters are fixed (by default only baseline parameters
+in the multinomial logistic models for the transition models and the
+initial state probabilities model):
+<<term=FALSE>>=
+setpars(mod, getpars(mod,which="fixed"))
+@
+When fitting constraints it is useful to have good starting values 
+for the parameters and hence we first fit the following model without
+constraints:
+<<>>=
+trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
+mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
+	  nstates=2,family=list(gaussian(),multinomial()),
+	  trstart=trst,inst=c(.999,0.001))
+fm1 <- fit(mod,verbose=FALSE)
+@
+
+After this, we use the fitted values from this model to constrain the
+regression coefficients on the transition matrix (parameters number~6
+and~10):
+<<term=FALSE>>=
+pars <- c(unlist(getpars(fm1)))
+pars[6] <- pars[10] <- 11
+fm1 <- setpars(fm1,pars)
+# start with fixed and free parameters
+conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
+# constrain the beta's on the transition parameters to be equal
+conpat[6] <- conpat[10] <- 2
+fm2 <- fit(fm1,equal=conpat)
+@
+
+Using \code{summary} on the fitted model shows that the regression
+coefficients are now estimated at the same value of 12.667.  The
+function \code{llratio} computes the likelihood ratio
+$\chi^2$-statistic and the associated $p$-value with appropriate
+degrees of freedom for testing the tenability of constraints
+\citep{Dannemann2007}.  Note that these arguments (i.e. \code{conpat}
+and \code{conrows}) provide the possibility for arbitrary constraints,
+also between, e.g., a multinomial regression coefficient for the
+transition matrix and the mean of a Gaussian response model.  Whether
+such constraints make sense is hence the responsibility of the user.
 
@@
 switch between applying different strategies.
-
-\subsection{Fixing and constraining parameters}
-
-Using package \pkg{Rdonlp2} by \citet{Tamura2009} or \pkg{Rsolnp}
-\citet{Ghalanos2010}, parameters may be fitted subject to general linear
-(in-)equality constraints.  Constraining and fixing parameters is done
-using the \code{conpat} argument to the \code{fit}-function, which
-specifies for each parameter in the model whether it's fixed (0) or
-free (1 or higher).  Equality constraints can be imposed by giving two
-parameters the same number in the \code{conpat} vector.  When
-only fixed values are required, the \code{fixed} argument can be used
-instead of \code{conpat}, with zeroes for fixed parameters and other
-values (ones e.g.) for non-fixed parameters.  Fitting the models
-subject to these constraints is handled by the optimization routine
-\code{donlp2}.  To be able to construct the \code{conpat} and/or
-\code{fixed} vectors one needs the correct ordering of parameters
-which is briefly discussed next before proceeding with an example.
-
-\paragraph{Parameter numbering} When using the \code{conpat} and
-\code{fixed} arguments, complete parameter vectors should be supplied,
-i.e., these vectors should have length equal to the number of parameters of
-the model, which can be obtained by calling \code{npar(object)}.  Note
-that this is not the same as the degrees of freedom used e.g.\ in the
-\code{logLik} function because \code{npar} also counts the baseline
-category zeroes from the multinomial logistic models.  Parameters are
-numbered in the following order:
-\begin{enumerate}
-	\item  the prior model parameters
-	\item  the parameters for the transition models
-	\item  the response model parameters per state (and subsequently
-	per response in the case of multivariate time series)
-\end{enumerate}
-
-To see the ordering of parameters use the following:
-<<term=FALSE>>=
-setpars(mod, value=1:npar(mod))
-@
-
-To see which parameters are fixed (by default only baseline parameters
-in the multinomial logistic models for the transition models and the
-initial state probabilities model):
-<<term=FALSE>>=
-setpars(mod, getpars(mod,which="fixed"))
-@
-When fitting constraints it is useful to have good starting values 
-for the parameters and hence we first fit the following model without
-constraints:
-<<>>=
-trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
-mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
-	  nstates=2,family=list(gaussian(),multinomial()),
-	  trstart=trst,inst=c(.999,0.001))
-fm1 <- fit(mod,verbose=FALSE)
-@
-
-After this, we use the fitted values from this model to constrain the
-regression coefficients on the transition matrix (parameters number~6
-and~10):
-<<term=FALSE>>=
-# start with fixed and free parameters
-conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
-# constrain the beta's on the transition parameters to be equal
-conpat[6] <- conpat[10] <- 2
-fm2 <- fit(fm1,equal=conpat)
-@
-
-Using \code{summary} on the fitted model shows that the regression 
-coefficients are now estimated at the same value of 12.66. The function
-\code{llratio} computes the likelihood ratio $\chi^2$-statistic and the
-associated $p$-value with appropriate degrees of freedom for testing the
-tenability of constraints \citep{Dannemann2007}. Note that these arguments 
-provide the possibility for arbitrary 
-constraints, also between, e.g., a multinomial regression coefficient 
-for the transition matrix and the mean of a Gaussian response model. 
-Whether such constraints make sense is hence the responsibility of 
-the user.
 
@@
 
-\bibliography{all,ingmar}
+\bibliography{dpx4Rev}
 

### 2010-06-30 papers/jss/dpx4Rev.Rnw
Minor edits to conform to jss standards
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@423 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\title{\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models}
+\title{\pkg{depmixS4} : An \proglang{R}-package for Hidden Markov Models}
 
@@
 \Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
-\Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting
+\Plaintitle{depmixS4: An R-package for Hidden Markov Models} %% without formatting
 
@@
 	multivariate data with distributions from the \code{glm} family,
-	the (logistic) multinomial, or the multivariate normal distribution.
-	Other distributions can be added easily, and an example is
-	provided with the exgaus distribution.  Parameters are estimated by
-	the EM algorithm or, when (linear) constraints are imposed on the
-	parameters, by direct numerical optimization with the
-	\pkg{Rsolnp} or \pkg{Rdonlp2} routines.}
+	the (logistic) multinomial, or the multivariate normal
+	distribution.  Other distributions can be added easily, and an
+	example is provided with the {\em exgaus} distribution.
+	Parameters are estimated by the EM algorithm or, when (linear)
+	constraints are imposed on the parameters, by direct numerical
+	optimization with the \pkg{Rsolnp} or \pkg{Rdonlp2} routines.
+
+}
 
@@
 
-%\usepackage{Sweave}
+%% \usepackage{Sweave}
 
@@
 
+<<echo=FALSE,results=hide>>=	
+.Options$digits<-3
+options(continue="+	",width=70,useFancyQuotes=FALSE)
+library(depmixS4)
+@
+
 %% include your article here, just as usual
 %% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
-
-\begin{center}
-\bf{DRAFT: DO NOT QUOTE WITHOUT CONTACTING AUTHOR}
-\end{center}
 
@@
 models with extensions.  Further examples of applications can be found
-in e.g.\ \citet[][chapter~1]{Cappe2005}.
+in e.g.\ \citet[][chapter~1]{Cappe2005}.  A more gentle introduction
+into hidden Markov models with applications is the book by
+\citet{Zucchini2009}.
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[!Htbp]
 \begin{center}
 <<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
-.Options$digits<-3
-library(depmixS4)
 data(speed)
@@
 probabilities can all depend on covariates $\vc{z}_t$.
-
-%transition probability functions $a_{ij}$ and the initial state
-%probability functions $\greekv{\pi}$ may depend on covariates as well
-%as the response distributions $\vc{b_{j}^{k}}$.  See for example
-%\citet{Fruhwirth2006} for an overview of hidden Markov models with
-%extensions.
 
@@
 through the use of a general Newton-Raphson optimizer.  In the EM
-algorithm, parameters are estimated by iteratively maximising the
+algorithm, parameters are estimated by iteratively maximizing the
 expected joint log-likelihood of the parameters given the observations and
@@
 computed effectively by the forward-backward algorithm \citep[see
-e.g.,][]{Rabiner1989}.  The Maximisation step consists of the
-maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
-side of (\ref{eq:Q}) consists of three separate parts, we can maximise
+e.g.,][]{Rabiner1989}.  The Maximization step consists of the
+maximization of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
+side of (\ref{eq:Q}) consists of three separate parts, we can maximize
 separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
-$\greekv{\theta}_3$.  In common models, maximisation for
+$\greekv{\theta}_3$.  In common models, maximization for
 $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
 \code{nnet.default} routine in the \pkg{nnet} package
-\citep{Venables2002}, and maximisation for $\greekv{\theta}_3$ by the
-standard \code{glm} routine.  Note that for the latter maximisation,
+\citep{Venables2002}, and maximization for $\greekv{\theta}_3$ by the
+standard \code{glm} routine.  Note that for the latter maximization,
 the expected values $\gamma_t(j)$ are used as prior weights of the
@@
 \begin{enumerate}
-	\item model specification with function \code{depmix} (or with \code{mix}
-	for latent class and finite mixture models, see example below on adding
-	covariates to prior probabilities)
+
+	\item model specification with function \code{depmix} (or with
+	\code{mix} for latent class and finite mixture models, see example
+	below on adding covariates to prior probabilities)
 
@@
 responses.  The transition probabilities indicate rather stable
-states, i.e., the probability in either of the states is around 0.9.
-The initial state probability estimates indicate that state 1 is the
-starting for the process, with a negligible probability of starting in
-state 2.
+states, i.e., the probability of remaining in either of the states is
+around 0.9.  The initial state probability estimates indicate that
+state 1 is the starting state for the process, with a negligible
+probability of starting in state 2.
 
@@
 probabilities are parameterized using a multinomial model with an
-identity link function.  Using a multinomial logistic model allows us
+identity link function.  Using a multinomial logistic model allows one
 to include covariates on the initial state and transition
@@
 mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
-    family=list(gaussian(),multinomial()),
+    family=list(gaussian(), multinomial("identity")),
 		transition=~scale(Pacc),instart=runif(2))
@@
 a single element in the list.  Effectively, this means specifying a
-multivariate response model.  The only multivariate response model
-currently implemented in \pkg{depmixS4} is for multivariate normal
+multivariate response model.  Currently, \pkg{depmixS4} has one 
+multivariate response model which is for multivariate normal
 variables.
@@
 subject to these constraints is handled by the optimization routine
-\code{donlp2}.  To be able to construct the \code{conpat} and/or
-\code{fixed} vectors one needs the correct ordering of parameters
-which is briefly discussed next before proceeding with an example.
+\code{solnp} or, optionally, by \code{donlp2}.  To be able to
+construct the \code{conpat} and/or \code{fixed} vectors one needs the
+correct ordering of parameters which is briefly discussed next before
+proceeding with an example.
 
@@
 mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
-	  nstates=2,family=list(gaussian(),multinomial()),
-	  trstart=trst,inst=c(.999,0.001))
+	  nstates=2,family=list(gaussian(),multinomial("identity")),
+	  trstart=trst,inst=c(0.99,0.01))
 fm1 <- fit(mod,verbose=FALSE)
@@
 pars[6] <- pars[10] <- 11
-fm1 <- setpars(fm1,pars)
+pars[1]=0
+pars[2]=1
+pars[13] <- pars[14] <- 0.5
+fm1 <- setpars(mod,pars)
 # start with fixed and free parameters
-conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
+conpat <- c(0,0,rep(c(0,1),4),1,1,0,0,1,1,1,1)
 # constrain the beta's on the transition parameters to be equal
@@
 Using \code{summary} on the fitted model shows that the regression
-coefficients are now estimated at the same value of 12.667.  The
-function \code{llratio} computes the likelihood ratio
-$\chi^2$-statistic and the associated $p$-value with appropriate
-degrees of freedom for testing the tenability of constraints
-\citep{Dannemann2007}.  Note that these arguments (i.e. \code{conpat}
-and \code{conrows}) provide the possibility for arbitrary constraints,
-also between, e.g., a multinomial regression coefficient for the
-transition matrix and the mean of a Gaussian response model.  Whether
-such constraints make sense is hence the responsibility of the user.
+coefficients are now estimated at the same value of 12.66.  Setting
+elements 13 and 14 of \code{conpat} to zero resulted in fixing those
+parameters at their starting values of 0.5.  This means that state 1
+can now be interpreted as a 'guessing' state which is associated with
+comparatively fast responses.  Similarly for elements 1 and 2,
+resulting in fixed initial probabilities.  The function \code{llratio}
+computes the likelihood ratio $\chi^2$-statistic and the associated
+$p$-value with appropriate degrees of freedom for testing the
+tenability of constraints \citep{Dannemann2007}.  Note that these
+arguments (i.e. \code{conpat} and \code{conrows}) provide the
+possibility for arbitrary constraints, also between, e.g., a
+multinomial regression coefficient for the transition matrix and the
+mean of a Gaussian response model.  Whether such constraints make
+sense is hence the responsibility of the user.
 
@@
 
-The balance scale taks is a famous task for testing cognitive
+The balance scale task is a famous task for testing cognitive
 strategies developed by Jean Piaget \citep[see][]{Siegler1981}.
@@
 released, or alternatively, whether it will stay in balance.  The item
-shown in Figure~\ref{fig:balance} is a so-callled distance item: the
+shown in Figure~\ref{fig:balance} is a so-called distance item: the
 number of weights placed on each side is equal, and only the distance
@@
 	\includegraphics{graphs/baldist.pdf}
-	\caption{Balance scale item; this is a so-called distance item.}
+	\caption{Balance scale item; this is a distance item (see the text 
+	for details).}
 	\label{fig:balance}  
@@
 	\item x: the design matrix, possibly only an intercept
-	\item paramaters: a named list with the coefficients and possibly 
+	\item parameters: a named list with the coefficients and possibly 
 	other parameters (e.g., the standard deviation in the Gaussian 
@@
 
-<<echo=FALSE>>=
+<<echo=FALSE, term=FALSE>>=
 setMethod("dens","exgaus",
@@
 use of \code{makeDepmix} allows the possibility of, say, fitting a
-Gaussian in one state and an exgaus distribution in another state.
+gaussian in one state and an exgaus distribution in another state.
 Note also that according to the AIC and BIC, the model with the exgaus
 describes the data much better than the same model in which the
-response times are modelled as Gaussian.
+response times are modelled as gaussian.
 
@@
 and transition models with two goals in mind.  First, to speed up
-(constrained) parameter optimization using \pkg{Rdonlp2}.  Second,
-analytic gradients are useful in computing the Hessian of the
-estimated parameters so as to arrive at standard errors for those.  We
-are also planning to implement goodness-of-fit statistics
-\citep{Titman2008}, and automatic generation of starting values.
+(constrained) parameter optimization using \pkg{Rdonlp2} or
+\pkg{Rsolnp}.  Second, analytic gradients are useful in computing the
+Hessian of the estimated parameters so as to arrive at standard errors
+for those.  We are also planning to implement goodness-of-fit
+statistics \citep{Titman2008}.
 
@@
 Ingmar Visser was supported by an EC Framework 6 grant, project 516542
-(NEST).  Maarten Speekenbrink was supported by ESRC grant RES-062-23-1511 
-and the ESRC Centre for Economic Learning and Social Evolution (ELSE).  
-Han van der Maas provided the speed-accuracy data \cite{Dutilh2010} and thereby
-neccessitated implementing models with time-dependent covariates.
-Brenda Jansen provided the balance scale data set \citep{Jansen2002}
-which was the perfect opportunity to test the covariates on the prior
-model parameters.  The examples in the help files use both of these
-data sets.
+(NEST).  Maarten Speekenbrink was supported by ESRC grant
+RES-062-23-1511 and the ESRC Centre for Economic Learning and Social
+Evolution (ELSE).  Han van der Maas provided the speed-accuracy data
+\citep{Dutilh2010} and thereby necessitated implementing models with
+time-dependent covariates.  Brenda Jansen provided the balance scale
+data set \citep{Jansen2002} which was the perfect opportunity to test
+the covariates on the prior model parameters.  The examples in the
+help files use both of these data sets.
 

### 2010-07-01 papers/jss/dpx4Rev.Rnw
Updates to bib file.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@425 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
More reference and code chunk edits ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@427 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 %set width of figures produced by Sweave
-\setkeys{Gin}{width=0.9\textwidth}
+\setkeys{Gin}{width=0.8\textwidth}
 
@@
 <<echo=FALSE,results=hide>>=	
-.Options$digits<-3
-options(continue="+	",width=70,useFancyQuotes=FALSE)
+options(prompt="R> ", continue="+	",width=70, useFancyQuotes=FALSE,
+digits=3)
 library(depmixS4)
@@
 sciences, in different areas and applications.  In psychology, they
-are used for modelling learning processes (see \citealp{Wickens1982},
-for an overview, and e.g. \citealp{Schmittmann2006}, for a recent
-application).  In economics, latent Markov models are so-called regime
+are used for modelling learning processes; see \citet{Wickens1982},
+for an overview, and e.g. \citet{Schmittmann2006}, for a recent
+application.  In economics, latent Markov models are so-called regime
 switching models (see e.g., \citealp{Kim1994} and
@@
 set.seed(1)
-mod <- depmix(response=rt~1, data=speed, nstates=2, 
-    trst=runif(4))
+mod <- depmix(response=rt~1, data=speed, nstates=2, trstart=runif(4))
 @
@@
 Hence, the 2-state model fits the data much better than the 1-state
-model.  Note that the 1-state model can be specified using \code{mod <-
-depmix(rt~1, data=speed, nstates=1)}, although this model is trivial
-as it will simply return the mean and the sd of the rt variable.
+model.  Note that the 1-state model can be specified using \code{mod
+<- depmix(rt~1, data=speed, nstates=1)}, although this model is
+trivial as it will simply return the mean and standard deviation of
+the \code{rt} variable.
 
@@
 	  nstates=2,family=list(gaussian(),multinomial("identity")),
-	  trstart=trst,inst=c(0.99,0.01))
+	  trstart=trst,instart=c(0.99,0.01))
 fm1 <- fit(mod,verbose=FALSE)
@@
 conpat[6] <- conpat[10] <- 2
+@
+<<echo=TRUE,results=hide>>=
 fm2 <- fit(fm1,equal=conpat)
@@
 \begin{enumerate}
-	\item y: the response variable
-	\item x: the design matrix, possibly only an intercept
-	\item parameters: a named list with the coefficients and possibly 
-	other parameters (e.g., the standard deviation in the Gaussian 
+	\item \code{y}: the response variable
+	\item \code{x}: the design matrix, possibly only an intercept
+	\item \code{parameters}: a named list with the coefficients and possibly 
+	other parameters (e.g., the standard deviation in the normal 
 	response model)
-	\item fixed: a vector of logicals indicating whether parameters are 
+	\item \code{fixed}: a vector of logicals indicating whether parameters are 
 	fixed
-	\item npar: numerical indicating the number of parameters of the model
+	\item \code{npar}: numerical indicating the number of parameters of the model
 \end{enumerate}
@@
 	with starting values
-	\item show method: to print the model to the terminal
-	\item dens: the function that computes the density of the responses
-	\item getpars and setpars: to get and set parameters 
-	\item predict: to generate predict'ed values 
-	\item fit: function to fit the model using posterior weights (used 
+	\item \code{show}: to print the model to the terminal
+	\item \code{dens}: the function that computes the density of the responses
+	\item \code{getpars} and \code{setpars}: to get and set parameters 
+	\item \code{predict}: to generate predict'ed values 
+	\item \code{fit}: function to fit the model using posterior weights (used 
 	by the EM algorithm)
@@
 
-Only the constructor and the fit methods are provided here; the
+Only the constructor and the \code{fit} methods are provided here; the
 complete code can be found in the help file of the \code{makeDepmix}
-function.  The \code{exgaus} example uses the \pkg{gamlss} and
-\pkg{gamlss.distr} packages
-\citep{Stasinopoulos2009a,Stasinopoulos2009b} for computing the
-\code{dens}ity and for \code{fit}ting the parameters.
+function.  The example with the exgaus distribution uses the
+\pkg{gamlss} and \pkg{gamlss.distr} packages \citep{Rigby2005,
+Stasinopoulos2007,Stasinopoulos2009a,Stasinopoulos2009b} for computing
+the \code{dens}ity and for \code{fit}ting the parameters.
 
@@
 defined as follows:
+<<results=hide>>=
+require(gamlss)
+require(gamlss.dist)
+@
 <<term=FALSE>>=
@@
 
-require(gamlss)
-require(gamlss.dist)
-
 setMethod("exgaus", 
   signature(y="ANY"), 
-  function(y,pstart=NULL,fixed=NULL, ...) {
+  function(y, pstart=NULL, fixed=NULL, ...) {
     y <- matrix(y,length(y))
@@
 )
-
 @
@@
 setMethod("fit","exgaus",
-  function(object,w) {
+  function(object, w) {
     if(missing(w)) w <- NULL
     y <- object@y
+
 	fit <- gamlss(y~1,weights=w,family=exGAUS(),
       control=gamlss.control(n.cyc=100,trace=FALSE),
-      mu.start=object@parameters$mu, 
-      sigma.start=exp(object@parameters$sigma), 
-      nu.start=exp(object@parameters$nu))
-    pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
+      mu.start=object@parameters$mu, sigma.start=exp(object@parameters$sigma), 
+      nu.start=exp(object@parameters$nu)
+	)
+
+	pars <- c(fit$mu.coefficients, fit$sigma.coefficients, 
+		fit$nu.coefficients)
     object <- setpars(object,pars)
@@
 <<>>=
-rModels <- list( list( 
-		exgaus(speed$rt,pstart=c(5,.1,.1)),	
-		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))),
-	list(
-		exgaus(speed$rt,pstart=c(6,.1,.1)),
-		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))))
+rModels <- list()
+rModels[[1]] <- list()
+rModels[[1]][[1]] <- exgaus(speed$rt, pstart=c(5,.1,.1))
+rModels[[1]][[2]] <- GLMresponse(formula=corr~1, data=speed, 
+	family=multinomial(), pstart=c(0.5,0.5))
+
+rModels[[2]] <- list()
+rModels[[2]][[1]] <- exgaus(speed$rt, pstart=c(6,.1,.1))
+rModels[[2]][[2]] <- GLMresponse(formula=corr~1, data=speed, 
+	family=multinomial(), pstart=c(.1,.9))
 @
@@
 transition <- list()
-transition[[1]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.9,0.1,0,0))
-transition[[2]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.1,0.9,0,0))
-inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
+transition[[1]] <- transInit(~Pacc,nst=2, data=speed, 
+	pstart=c(0.9,0.1,0,0))
+transition[[2]] <- transInit(~Pacc,nst=2, data=speed, 
+	pstart=c(0.1,0.9,0,0))
+inMod <- transInit(~1, ns=2, pstart=c(0.1,0.9), data=data.frame(1))
 @
@@
 <<>>=
-mod <- makeDepmix(response=rModels,transition=transition,
-    prior=inMod,stat=FALSE)
-fm <- fit(mod,verbose=FALSE)
+mod <- makeDepmix(response=rModels, transition=transition,
+    prior=inMod, stat=FALSE)
+fm <- fit(mod, verbose=FALSE)
 @
@@
 this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
-intended for modelling of (individual) time series data with the aim of
+intended for modelling (individual) time series data with the aim of
 characterizing the transition processes underlying the data.  The

### 2010-07-02 papers/jss/dpx4Rev.Rnw
updates to formating ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@429 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 options(prompt="R> ", continue="+	",width=70, useFancyQuotes=FALSE,
-digits=3)
+digits=4)
 library(depmixS4)
@@
 The \code{fit} method is defined as follows: 
+<<echo=FALSE>>=
+options(width=60)
+@
 <<>>=
@@
       control=gamlss.control(n.cyc=100,trace=FALSE),
-      mu.start=object@parameters$mu, sigma.start=exp(object@parameters$sigma), 
-      nu.start=exp(object@parameters$nu)
-	)
+      mu.start=object@parameters$mu, 
+	  sigma.start=exp(object@parameters$sigma), 
+      nu.start=exp(object@parameters$nu))
 
@@
 We first create all the response models that we need as a double list: 
+<<echo=FALSE>>=
+options(width=80)
+@
 <<>>=
@@
 the response class): 
+<<echo=FALSE>>=
+options(width=70)
+@
 <<>>=

### 2010-08-09 papers/jss/V2Revision/dpx4Rev.Rnw
Cleaned up papers directory
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@437 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2010-06-30 papers/jss/dpx4Rev.Rnw
@@
 
-\title{\pkg{depmixS4} : An \proglang{R}-package for Hidden Markov Models}
+\title{\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models}
 
@@
 \Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
-\Plaintitle{depmixS4: An R-package for Hidden Markov Models} %% without formatting
+\Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting
 
@@
 	estimating dependent mixture models in the \proglang{R}
-	programming language \citep{R2010}.  This includes standard Markov
+	programming language \citep{R2009}.  This includes standard Markov
 	models, latent/hidden Markov models, and latent class and finite
@@
 	multivariate data with distributions from the \code{glm} family,
-	the (logistic) multinomial, or the multivariate normal
-	distribution.  Other distributions can be added easily, and an
-	example is provided with the {\em exgaus} distribution.
-	Parameters are estimated by the EM algorithm or, when (linear)
-	constraints are imposed on the parameters, by direct numerical
-	optimization with the \pkg{Rsolnp} or \pkg{Rdonlp2} routines.
-
-}
+	the (logistic) multinomial, or the multivariate normal distribution.
+	Other distributions can be added easily, and an example is
+	provided with the exgaus distribution.  Parameters are estimated by
+	the EM algorithm or, when (linear) constraints are imposed on the
+	parameters, by direct numerical optimization with the
+	\pkg{Rsolnp} or \pkg{Rdonlp2} routines.}
 
@@
 
-%% \usepackage{Sweave}
+%\usepackage{Sweave}
 
@@
 
-<<echo=FALSE,results=hide>>=	
-.Options$digits<-3
-options(continue="+	",width=70,useFancyQuotes=FALSE)
-library(depmixS4)
-@
-
 %% include your article here, just as usual
 %% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
+
+\begin{center}
+\bf{DRAFT: DO NOT QUOTE WITHOUT CONTACTING AUTHOR}
+\end{center}
 
@@
 models with extensions.  Further examples of applications can be found
-in e.g.\ \citet[][chapter~1]{Cappe2005}.  A more gentle introduction
-into hidden Markov models with applications is the book by
-\citet{Zucchini2009}.
+in e.g.\ \citet[][chapter~1]{Cappe2005}.
 
@@
 
-\begin{figure}[!Htbp]
+\begin{figure}[htbp]
 \begin{center}
 <<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
+.Options$digits<-3
+library(depmixS4)
 data(speed)
@@
 probabilities can all depend on covariates $\vc{z}_t$.
+
+%transition probability functions $a_{ij}$ and the initial state
+%probability functions $\greekv{\pi}$ may depend on covariates as well
+%as the response distributions $\vc{b_{j}^{k}}$.  See for example
+%\citet{Fruhwirth2006} for an overview of hidden Markov models with
+%extensions.
 
@@
 through the use of a general Newton-Raphson optimizer.  In the EM
-algorithm, parameters are estimated by iteratively maximizing the
+algorithm, parameters are estimated by iteratively maximising the
 expected joint log-likelihood of the parameters given the observations and
@@
 computed effectively by the forward-backward algorithm \citep[see
-e.g.,][]{Rabiner1989}.  The Maximization step consists of the
-maximization of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
-side of (\ref{eq:Q}) consists of three separate parts, we can maximize
+e.g.,][]{Rabiner1989}.  The Maximisation step consists of the
+maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
+side of (\ref{eq:Q}) consists of three separate parts, we can maximise
 separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
-$\greekv{\theta}_3$.  In common models, maximization for
+$\greekv{\theta}_3$.  In common models, maximisation for
 $\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
 \code{nnet.default} routine in the \pkg{nnet} package
-\citep{Venables2002}, and maximization for $\greekv{\theta}_3$ by the
-standard \code{glm} routine.  Note that for the latter maximization,
+\citep{Venables2002}, and maximisation for $\greekv{\theta}_3$ by the
+standard \code{glm} routine.  Note that for the latter maximisation,
 the expected values $\gamma_t(j)$ are used as prior weights of the
@@
 \begin{enumerate}
-
-	\item model specification with function \code{depmix} (or with
-	\code{mix} for latent class and finite mixture models, see example
-	below on adding covariates to prior probabilities)
+	\item model specification with function \code{depmix} (or with \code{mix}
+	for latent class and finite mixture models, see example below on adding
+	covariates to prior probabilities)
 
@@
 responses.  The transition probabilities indicate rather stable
-states, i.e., the probability of remaining in either of the states is
-around 0.9.  The initial state probability estimates indicate that
-state 1 is the starting state for the process, with a negligible
-probability of starting in state 2.
+states, i.e., the probability of remaining in either of the states is around 0.9.
+The initial state probability estimates indicate that state 1 is the
+starting state for the process, with a negligible probability of starting in
+state 2.
 
@@
 mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
-    family=list(gaussian(), multinomial("identity")),
+    family=list(gaussian(),multinomial("identity")),
     transition=~scale(Pacc),instart=runif(2))
@@
 subject to these constraints is handled by the optimization routine
-\code{solnp} or, optionally, by \code{donlp2}.  To be able to
-construct the \code{conpat} and/or \code{fixed} vectors one needs the
-correct ordering of parameters which is briefly discussed next before
-proceeding with an example.
+\code{solnp} or, optionally, by \code{donlp2}.  To be able to construct the \code{conpat} and/or
+\code{fixed} vectors one needs the correct ordering of parameters
+which is briefly discussed next before proceeding with an example.
 
@@
 coefficients are now estimated at the same value of 12.66.  Setting 
-elements 13 and 14 of \code{conpat} to zero resulted in fixing those
-parameters at their starting values of 0.5.  This means that state 1
-can now be interpreted as a 'guessing' state which is associated with
-comparatively fast responses.  Similarly for elements 1 and 2,
-resulting in fixed initial probabilities.  The function \code{llratio}
-computes the likelihood ratio $\chi^2$-statistic and the associated
-$p$-value with appropriate degrees of freedom for testing the
-tenability of constraints \citep{Dannemann2007}.  Note that these
-arguments (i.e. \code{conpat} and \code{conrows}) provide the
-possibility for arbitrary constraints, also between, e.g., a
-multinomial regression coefficient for the transition matrix and the
-mean of a Gaussian response model.  Whether such constraints make
-sense is hence the responsibility of the user.
+elements 13 and 14 of \code{conpat} to zero resulted in fixing those parameters
+at their starting values of 0.5.  This means that state 1 can now be interpreted
+as a 'guessing' state which is associated with comparatively fast responses.
+Similarly for elements 1 and 2, resulting in fixed initial probabilities. The
+function \code{llratio} computes the likelihood ratio
+$\chi^2$-statistic and the associated $p$-value with appropriate
+degrees of freedom for testing the tenability of constraints
+\citep{Dannemann2007}.  Note that these arguments (i.e. \code{conpat}
+and \code{conrows}) provide the possibility for arbitrary constraints,
+also between, e.g., a multinomial regression coefficient for the
+transition matrix and the mean of a Gaussian response model.  Whether
+such constraints make sense is hence the responsibility of the user.
 
@@
 
-The balance scale task is a famous task for testing cognitive
+The balance scale taks is a famous task for testing cognitive
 strategies developed by Jean Piaget \citep[see][]{Siegler1981}.
@@
 released, or alternatively, whether it will stay in balance.  The item
-shown in Figure~\ref{fig:balance} is a so-called distance item: the
+shown in Figure~\ref{fig:balance} is a so-callled distance item: the
 number of weights placed on each side is equal, and only the distance
@@
 	\item x: the design matrix, possibly only an intercept
-	\item parameters: a named list with the coefficients and possibly 
+	\item paramaters: a named list with the coefficients and possibly 
 	other parameters (e.g., the standard deviation in the Gaussian 
@@
 and transition models with two goals in mind.  First, to speed up
-(constrained) parameter optimization using \pkg{Rdonlp2} or
-\pkg{Rsolnp}.  Second, analytic gradients are useful in computing the
-Hessian of the estimated parameters so as to arrive at standard errors
-for those.  We are also planning to implement goodness-of-fit
-statistics \citep{Titman2008}.
+(constrained) parameter optimization using \pkg{Rdonlp2}.  Second,
+analytic gradients are useful in computing the Hessian of the
+estimated parameters so as to arrive at standard errors for those.  We
+are also planning to implement goodness-of-fit statistics
+\citep{Titman2008}.
 
@@
 Ingmar Visser was supported by an EC Framework 6 grant, project 516542
-(NEST).  Maarten Speekenbrink was supported by ESRC grant
-RES-062-23-1511 and the ESRC Centre for Economic Learning and Social
-Evolution (ELSE).  Han van der Maas provided the speed-accuracy data
-\citep{Dutilh2010} and thereby necessitated implementing models with
-time-dependent covariates.  Brenda Jansen provided the balance scale
-data set \citep{Jansen2002} which was the perfect opportunity to test
-the covariates on the prior model parameters.  The examples in the
-help files use both of these data sets.
+(NEST).  Maarten Speekenbrink was supported by ESRC grant RES-062-23-1511 
+and the ESRC Centre for Economic Learning and Social Evolution (ELSE).  
+Han van der Maas provided the speed-accuracy data \cite{Dutilh2010} and thereby
+neccessitated implementing models with time-dependent covariates.
+Brenda Jansen provided the balance scale data set \citep{Jansen2002}
+which was the perfect opportunity to test the covariates on the prior
+model parameters.  The examples in the help files use both of these
+data sets.
 

### 2010-10-18 papers/jss/depmixS4.Rnw
Minor typo's corrected.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@440 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2010-07-01 papers/jss/dpx4Rev.Rnw
@@
-% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
-
-\documentclass[article]{jss}
-\usepackage{amsmath}
-
-%\usepackage[]{amsmath, amsfonts, amstext, amsthm} 
-%\usepackage{amssymb}
-%\usepackage[]{graphics} 
-%\usepackage{epsfig}
-%\usepackage{epstopdf}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% almost as usual
+\documentclass[article,nojss]{jss}
+\usepackage{amsmath,thumbpdf}
+\graphicspath{{Figures/}}
+
+\newcommand{\vc}{\mathbf}
+\newcommand{\mat}{\mathbf}
+\newcommand{\greekv}[1]{\mbox{\boldmath$\mathrm{#1}$}}
+\newcommand{\greekm}[1]{\mbox{\boldmath$#1$}}
+
+%% \usepackage{Sweave}
+\SweaveOpts{engine = R, echo = TRUE, keep.source = TRUE, eps = FALSE}
+
+%\VignetteIndexEntry{depmixS4: An R Package for Hidden Markov Models}
+%\VignetteDepends{depmixS4,gamlss,gamlss.data}
+%\VignetteKeywords{hidden Markov model, dependent mixture model, mixture model, constraints}
+%\VignettePackage{depmixS4}
+
 \author{Ingmar Visser\\University of Amsterdam \And 
         Maarten Speekenbrink\\University College London}
-
-\title{\pkg{depmixS4} : An \proglang{R}-package for Hidden Markov Models}
-
-%% for pretty printing and a nice hypersummary also set:
-\Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
-\Plaintitle{depmixS4: An R-package for Hidden Markov Models} %% without formatting
-
-\Shorttitle{depmixS4: Hidden Markov Models} %% a short title (if necessary)
-
-%% an abstract and keywords
+\Plainauthor{Ingmar Visser, Maarten Speekenbrink}
+
+\title{\pkg{depmixS4}: An \proglang{R} Package for Hidden Markov Models}
+\Plaintitle{depmixS4: An R Package for Hidden Markov Models}
+
 \Abstract{	
+  This introduction to the \proglang{R} package \pkg{depmixS4} is a (slightly)
+  modified version of \cite{Visser+Speekenbrink:2010}, published in the
+  \emph{Journal of Statistical Software}.
 
@@
   estimating dependent mixture models in the \proglang{R}
-	programming language \citep{R2010}.  This includes standard Markov
+  programming language.  This includes standard Markov
   models, latent/hidden Markov models, and latent class and finite
@@
   example is provided with the {\em exgaus} distribution.
-	Parameters are estimated by the EM algorithm or, when (linear)
+  Parameters are estimated by the expectation-maximization (EM) algorithm or, when (linear)
   constraints are imposed on the parameters, by direct numerical
   optimization with the \pkg{Rsolnp} or \pkg{Rdonlp2} routines.   
-
 }
 
-\Keywords{hidden Markov model, dependent mixture model, mixture 
-model, constraints}
-
-\Plainkeywords{hidden Markov model, dependent mixture model, mixture 
-model, constraints} %% without formatting
-%% at least one keyword must be supplied
-
-%% publication information
-%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
-%% \Volume{13}
-%% \Issue{9}
-%% \Month{September}
-%% \Year{2004}
-%% \Submitdate{2004-09-29}
-%% \Acceptdate{2004-09-29}
-
-%% The address of (at least) one author should be given
-%% in the following format:
+\Keywords{hidden Markov model, dependent mixture model, mixture model, constraints}
+
+\Volume{36}
+\Issue{7}
+\Month{August}
+\Year{2010}
+\Submitdate{2009-08-19}
+\Acceptdate{2010-06-21}
+
 \Address{
@@
   Roetersstraat 15\\
-1018 WB, Amsterdam\\
-The Netherlands\\
+  1018 WB, Amsterdam, The Netherlands\\
   E-mail: \email{i.visser@uva.nl} \\
@@
 
-%% It is also possible to add a telephone and fax number
-%% before the e-mail in the following format:
-%% Telephone: +43/1/31336-5053
-%% Fax: +43/1/31336-734
-
-\newcommand{\vc}{\mathbf}
-\newcommand{\mat}{\mathbf}
-\newcommand{\greekv}[1]{\mbox{\boldmath$\mathrm{#1}$}}
-\newcommand{\greekm}[1]{\mbox{\boldmath$#1$}}
-
-%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\batchmode
-
-\SweaveOpts{echo=TRUE}
-\usepackage{a4wide}
-
-%% \usepackage{Sweave}
-
 \begin{document}
 
-%set width of figures produced by Sweave
-\setkeys{Gin}{width=0.8\textwidth}
-
-\maketitle
-
 <<echo=FALSE,results=hide>>=	
-options(prompt="R> ", continue="+	",width=70, useFancyQuotes=FALSE,
-digits=3)
-library(depmixS4)
-@
-
-%% include your article here, just as usual
-%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
+options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE, digits = 4)
+library("depmixS4")
+@
 
 \section{Introduction}
-%% Note: If there is markup in \(sub)section, then it has to be escape as above.
 
@@
 are used for modelling learning processes; see \citet{Wickens1982},
-for an overview, and e.g. \citet{Schmittmann2006}, for a recent
+for an overview, and e.g., \citet{Schmittmann2006}, for a recent
 application.  In economics, latent Markov models are so-called regime
@@
 models with extensions.  Further examples of applications can be found
-in e.g.\ \citet[][chapter~1]{Cappe2005}.  A more gentle introduction
+in e.g., \citet[][Chapter~1]{Cappe2005}.  A more gentle introduction
 into hidden Markov models with applications is the book by
@@
 package was available for fitting such models.  Existing software for
-estimating Markovian models include Panmark \citep{Pol1996}, and for
-latent class models Latent Gold \citep{Vermunt2003}.  These programs
+estimating Markovian models include \pkg{Panmark} \citep{Pol1996}, and for
+latent class models \pkg{Latent Gold} \citep{Vermunt2003}.  These programs
 lack a number of important features, besides not being freely
@@
 \begin{enumerate}	
-
 	\item to be able to estimate parameters subject to general
-	linear (in)equality constraints
+	linear (in)equality constraints;
 
 	\item to be able to fit transition models with covariates, i.e.,
-	to have time-dependent transition matrices
+	to have time-dependent transition matrices;
 
 	\item to be able to include covariates in the prior or initial
-	state probabilities
+	state probabilities;
 
@@
 	easily add new uni- or multivariate response distributions and
-	new transition models, e.g., continuous time observation models
+	new transition models, e.g., continuous time observation models.
 
@@
 effects of covariates on prior and transition probabilities is a
-distinguishing feature of \pkg{depmixS4}.  In the next section, we
-provide an outline of the model and likelihood equations.
-
-\section{The dependent mixture model}
+distinguishing feature of \pkg{depmixS4}.
+In Section~\ref{sec:2}, we provide an outline of the model and likelihood equations.
+
+The \pkg{depmixS4} package is implemented using the \proglang{R} system
+for statistical computing \citep{R2010} and is available from the
+Comprehensive \proglang{R} Archive Network at \url{http://CRAN.R-project.org/package=depmixS4}.
+
+\section{The dependent mixture model} \label{sec:2}
 
@@
 
-\begin{figure}[!Htbp]
+\setkeys{Gin}{width=0.8\textwidth}
+\begin{figure}[t!]
 \begin{center}
 <<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=	
-data(speed)
-plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
+data("speed")
+plot(as.ts(speed[1:168,]), main = "Speed-accuracy trade-off")
 @
@@
 contrast, the hidden Markov model is typically only used for `long'
-univariate time series \citep[][chapter~1]{Cappe2005}.  We use the
-term "dependent mixture model" because one of the authors (Ingmar
+univariate time series \citep[][Chapter~1]{Cappe2005}.  We use the
+term ``dependent mixture model" because one of the authors (Ingmar
 Visser) thought it was time for a new name to relate these
-models\footnote{Only later did I find out that \citet{Leroux1992}
+models\footnote{Only later he found out that \citet{Leroux1992}
 already coined the term dependent mixture models in an application
@@
 	provides the probability of a transition from state $i$ to state
-	$j$ with covariate $\vc{z}_t$,
+	$j$ with covariate $\vc{z}_t$.
 
@@
 where $\Prob(\vc{O}_{1}|\vc{O}_{0}):=\Prob(\vc{O}_{1})$. Note that for a 
-simple, i.e.\ observed, Markov chain these probabilities reduce to 
+simple, i.e., observed, Markov chain these probabilities reduce to 
 $\Prob(\vc{O}_{t}|\vc{O}_{1},\ldots, 
@@
 
-Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
+Parameters are estimated in \pkg{depmixS4} using the expectation-maximization (EM) algorithm or
 through the use of a general Newton-Raphson optimizer.  In the EM
@@
 log-likelihood can be written as:
-\begin{equation}
+\begin{multline}
 \log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
 \Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
-+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
++ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) \\
 + \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
-\end{equation}
+\end{multline}
 This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
@@
 	\code{mix} for latent class and finite mixture models, see example
-	below on adding covariates to prior probabilities)
-
-	\item  model fitting with function \code{fit}
+	below on adding covariates to prior probabilities);
+
+	\item  model fitting with function \code{fit}.
 \end{enumerate}
@@
 Throughout this article a data set called \code{speed} is used.  As
-already indicated in the Introduction, it consists of three time
-series with three variables: response time, accuracy, and a covariate,
-Pacc, which defines the relative pay-off for speeded versus accurate
-responding.  Before describing some of the models that are fitted to
-these data, we provide a brief sketch of the reasons for gathering
-these data in the first place.
+already indicated in the introduction, it consists of three time
+series with three variables: response time \code{rt}, accuracy
+\code{corr}, and a covariate, \code{Pacc}, which defines the relative
+pay-off for speeded versus accurate responding.  Before describing
+some of the models that are fitted to these data, we provide a brief
+sketch of the reasons for gathering these data in the first place.
 
@@
 relative reward for speed versus accuracy.  The bottom panel of
-figure~\ref{fig:speed} shows the values of this variable.  The
+Figure~\ref{fig:speed} shows the values of this variable.  The
 experiment was designed to investigate what would happen when this
@@
 A dependent mixture model can be created with the
-\code{depmix}-function as follows:
-<<>>=
-library(depmixS4)
-data(speed)
+\code{depmix} function as follows:
+<<>>=
+library("depmixS4")
+data("speed")
 set.seed(1)
-mod <- depmix(response=rt~1, data=speed, nstates=2, trstart=runif(4))
+mod <- depmix(response = rt ~ 1, data = speed, nstates = 2,
+  trstart = runif(4))
 @
@@
 using formulae such as in \code{lm} or \code{glm} models.  The second
-argument, \code{data=speed}, provides the \code{data.frame} in which
+argument, \code{data = speed}, provides the \code{data.frame} in which
 the variables from \code{response} are interpreted.  Third, the number
-of states of the model is given by \code{nstates=2}.
-
-\paragraph{Starting values} Note also that start values for the
+of states of the model is given by \code{nstates = 2}.
+
+\paragraph{Starting values.} Note also that start values for the
 transition parameters are provided in this call by the \code{trstart}
@@
 
-\paragraph{Fitting the model and printing results} The \code{depmix}
-function returns an object of class \code{depmix} which contains the
+\paragraph{Fitting the model and printing results.} The \code{depmix}
+function returns an object of class `\code{depmix}' which contains the
 model specification, and not a fitted model.  Hence, the model needs
@@
 
-The \code{fit}-function returns an object of class
-\code{depmix.fitted} which extends the \code{depmix} class, adding
+The \code{fit} function returns an object of class
+`\code{depmix.fitted}' which extends the `\code{depmix}' class, adding
 convergence information (and information about constraints if these
@@
 and \code{BIC}, respectively.  By comparison, a 1-state model for
-these data, i.e. assuming there is no mixture, has a log-likelihood of
+these data, i.e., assuming there is no mixture, has a log-likelihood of
 $-305.33$, and 614.66, and 622.83 for the AIC and BIC respectively.
 Hence, the 2-state model fits the data much better than the 1-state
-model.  Note that the 1-state model can be specified using \code{mod
-<- depmix(rt~1, data=speed, nstates=1)}, although this model is
+model.  Note that the 1-state model can be specified using
+\code{mod <- depmix(rt ~ 1, data = speed, nstates = 1)}, although this model is
 trivial as it will simply return the mean and standard deviation of
@@
 that the parameter for the base category is fixed at zero
-\citep[see][p.\ 267 ff., for multinomial logistic models and various
+\citep[see][p.~267~ff., for multinomial logistic models and various
 parameterizations]{Agresti2002}.  The default baseline category is the
@@
 set.seed(1)
-mod <- depmix(rt~1, data=speed, nstates=2, family=gaussian(),
-	   transition=~scale(Pacc), instart=runif(2)) 
-fm <-fit(mod,verbose=FALSE) 
-@
-
-Note the use of \code{verbose=FALSE} to suppress printing of
+mod <- depmix(rt ~ 1, data = speed, nstates = 2, family = gaussian(),
+  transition = ~ scale(Pacc), instart = runif(2)) 
+fm <- fit(mod, verbose = FALSE) 
+@
+
+Note the use of \code{verbose = FALSE} to suppress printing of
 information on the iterations of the fitting process.  Applying
@@
 <<>>= 
-summary(fm, which="transition") 
+summary(fm, which = "transition") 
 @
@@
 set.seed(1)
-mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
-    family=list(gaussian(), multinomial("identity")),
-		transition=~scale(Pacc),instart=runif(2))
-fm <- fit(mod,verbose=FALSE)
+mod <- depmix(list(rt ~ 1,corr ~ 1), data = speed, nstates = 2, 
+  family = list(gaussian(), multinomial("identity")),
+  transition = ~ scale(Pacc), instart = runif(2))
+fm <- fit(mod,verbose = FALSE)
 @
@@
 <<>>=
-summary(fm,which="response")
+summary(fm, which = "response")
 @
@@
 
-Using package \pkg{Rsolnp} \citet{Ghalanos2010} or \pkg{Rdonlp2} by
-\citet{Tamura2009}, parameters may be fitted subject to general linear
+Using package \pkg{Rsolnp} \citep{Ghalanos2010} or \pkg{Rdonlp2}
+\citep{Tamura2009}, parameters may be fitted subject to general linear
 (in-)equality constraints.  Constraining and fixing parameters is done
-using the \code{conpat} argument to the \code{fit}-function, which
+using the \code{conpat} argument to the \code{fit} function, which
 specifies for each parameter in the model whether it's fixed (0) or
@@
 instead of \code{conpat}, with zeroes for fixed parameters and other
-values (ones e.g.) for non-fixed parameters.  Fitting the models
+values (e.g., ones) for non-fixed parameters.  Fitting the models
 subject to these constraints is handled by the optimization routine
@@
 
-\paragraph{Parameter numbering} When using the \code{conpat} and
+\paragraph{Parameter numbering.} When using the \code{conpat} and
 \code{fixed} arguments, complete parameter vectors should be supplied,
@@
 the model, which can be obtained by calling \code{npar(object)}.  Note
-that this is not the same as the degrees of freedom used e.g.\ in the
+that this is not the same as the degrees of freedom used e.g., in the
 \code{logLik} function because \code{npar} also counts the baseline
@@
 \begin{enumerate}
-	\item  the prior model parameters
-	\item  the parameters for the transition models
+	\item  the prior model parameters;
+	\item  the parameters for the transition models;
 	\item  the response model parameters per state (and subsequently
-	per response in the case of multivariate time series)
+	per response in the case of multivariate time series).
 \end{enumerate}
@@
 <<term=FALSE>>=
-setpars(mod, value=1:npar(mod))
+setpars(mod, value = 1:npar(mod))
 @
@@
 <<term=FALSE>>=
-setpars(mod, getpars(mod,which="fixed"))
+setpars(mod, getpars(mod, which = "fixed"))
 @
@@
 <<>>=
-trst <- c(0.9,0.1,0,0,0.1,0.9,0,0)
-mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
-	  nstates=2,family=list(gaussian(),multinomial("identity")),
-	  trstart=trst,instart=c(0.99,0.01))
-fm1 <- fit(mod,verbose=FALSE)
+trst <- c(0.9, 0.1, 0, 0, 0.1, 0.9, 0, 0)
+mod <- depmix(list(rt ~ 1,corr ~ 1), data = speed, transition = ~ Pacc,
+  nstates = 2, family = list(gaussian(), multinomial("identity")),
+  trstart = trst, instart = c(0.99, 0.01))
+fm1 <- fit(mod,verbose = FALSE)
 @
@@
 and~10):
-<<term=FALSE>>=
+<<term=FALSE, echo=TRUE,results=hide>>=
 pars <- c(unlist(getpars(fm1)))
 pars[6] <- pars[10] <- 11
-pars[1]=0
-pars[2]=1
+pars[1] <- 0
+pars[2] <- 1
 pars[13] <- pars[14] <- 0.5
-fm1 <- setpars(mod,pars)
-# start with fixed and free parameters
-conpat <- c(0,0,rep(c(0,1),4),1,1,0,0,1,1,1,1)
-# constrain the beta's on the transition parameters to be equal
+fm1 <- setpars(mod, pars)
+conpat <- c(0, 0, rep(c(0, 1), 4), 1, 1, 0, 0, 1, 1, 1, 1)
 conpat[6] <- conpat[10] <- 2
-@
-<<echo=TRUE,results=hide>>=
-fm2 <- fit(fm1,equal=conpat)
+fm2 <- fit(fm1, equal = conpat)
 @
@@
 tenability of constraints \citep{Dannemann2007}.  Note that these
-arguments (i.e. \code{conpat} and \code{conrows}) provide the
+arguments (i.e., \code{conpat} and \code{conrows}) provide the
 possibility for arbitrary constraints, also between, e.g., a
@@
 
-\begin{figure}[htbp]
+\setkeys{Gin}{width=0.7\textwidth}
+\begin{figure}[t!]
 \begin{center}
-	\includegraphics{graphs/baldist.pdf}
+	\includegraphics{baldist.pdf}
 	\caption{Balance scale item; this is a distance item (see the text 
@@
 only do so when the number of weights is equal on each side.  These
-two strategies that children employ are known as Rule I and Rule II.
+two strategies that children employ are known as Rule~I and Rule~II.
 Other strategies can be teased apart by administering different items.
@@
 classes in this case), are defined by using a one-sided formula
-\code{prior=~age}: 
-<<>>= 
-data(balance)
+\code{prior = ~ age}: 
+<<>>= 
+data("balance")
 set.seed(1)
-mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=3,
-	family=list(multinomial("identity"), multinomial("identity"),
-	multinomial("identity"), multinomial("identity")),
-	respstart=runif(24), prior=~age, initdata=balance)
-fm <- fit(mod,verbose=FALSE) 
+mod <- mix(list(d1 ~ 1, d2 ~ 1, d3 ~ 1, d4 ~ 1), data = balance,
+  nstates = 3, family = list(multinomial("identity"),
+  multinomial("identity"), multinomial("identity"),
+  multinomial("identity")), respstart = runif(24), prior = ~ age,
+  initdata = balance)
+fm <- fit(mod, verbose = FALSE) 
 fm
 @
-
 Note here that we define a \code{mix} model instead of a \code{depmix}
 model as these data form independent observations.  More formally,
-\code{depmix} models extend the class of \code{mix} models by adding
+\code{depmix} models extend the class of `\code{mix}' models by adding
 transition models.  As for fitting \code{mix} models: as can be
@@
 <<>>=
-summary(fm,which="prior")
+summary(fm, which = "prior")
 @
@@
 Figure~\ref{fig:balprior} depicts the prior class probabilities as
-function of age based on the fitted parameters
+function of age based on the fitted parameters.
 
@@
 
-User-defined distributions should extend the \code{response}-class and
+User-defined distributions should extend the `\code{response}' class and
 have the following slots:
 \begin{enumerate}
-	\item \code{y}: the response variable
-	\item \code{x}: the design matrix, possibly only an intercept
-	\item \code{parameters}: a named list with the coefficients and possibly 
+	\item \code{y}: The response variable.
+	\item \code{x}: The design matrix, possibly only an intercept.
+	\item \code{parameters}: A named list with the coefficients and possibly 
 	other parameters (e.g., the standard deviation in the normal 
-	response model)
-	\item \code{fixed}: a vector of logicals indicating whether parameters are 
-	fixed
-	\item \code{npar}: numerical indicating the number of parameters of the model
+	response model).
+	\item \code{fixed}: A vector of logicals indicating whether parameters are 
+	fixed.
+	\item \code{npar}: Numerical indicating the number of parameters of the model.
 \end{enumerate}
@@
 the response times with an exgaus rather than a Gaussian distribution.
-To do so, we first define an exgaus-class extending the
-response-class:
+To do so, we first define an `\code{exgaus}' class extending the
+`\code{response}' class:
 <<>>=
@@
 \begin{enumerate}
-	\item constructor: function to create instances of the class 
-	with starting values
-	\item \code{show}: to print the model to the terminal
-	\item \code{dens}: the function that computes the density of the responses
-	\item \code{getpars} and \code{setpars}: to get and set parameters 
-	\item \code{predict}: to generate predict'ed values 
-	\item \code{fit}: function to fit the model using posterior weights (used 
-	by the EM algorithm)
+	\item constructor: Function to create instances of the class 
+	with starting values.
+	\item \code{show}: To print the model to the terminal.
+	\item \code{dens}: The function that computes the density of the responses.
+	\item \code{getpars} and \code{setpars}: To get and set parameters .
+	\item \code{predict}: To generate predicted values.
+	\item \code{fit}: Function to fit the model using posterior weights (used 
+	by the EM algorithm).
 \end{enumerate}
@@
 function.  The example with the exgaus distribution uses the
-\pkg{gamlss} and \pkg{gamlss.distr} packages \citep{Rigby2005,
+\pkg{gamlss} and \pkg{gamlss.dist} packages \citep{Rigby2005,
 Stasinopoulos2007,Stasinopoulos2009a,Stasinopoulos2009b} for computing
@@
 
-The constructor method return an object of class \code{exgaus}, and is
+The constructor method return an object of class `\code{exgaus}', and is
 defined as follows:
-<<results=hide>>=
-require(gamlss)
-require(gamlss.dist)
-@
-<<term=FALSE>>=
+<<results=hide, term=FALSE>>=
+library("gamlss")
+library("gamlss.dist")
 setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
@@
 setMethod("exgaus", 
-  signature(y="ANY"), 
-  function(y, pstart=NULL, fixed=NULL, ...) {
-    y <- matrix(y,length(y))
+  signature(y = "ANY"), 
+  function(y, pstart = NULL, fixed = NULL, ...) {
+    y <- matrix(y, length(y))
     x <- matrix(1) 
@@
     npar <- 3
-    if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
+    if(is.null(fixed)) fixed <- as.logical(rep(0, npar))
     if(!is.null(pstart)) {
-      if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
+      if(length(pstart) != npar) stop("length of 'pstart' must be ", npar)
       parameters$mu <- pstart[1]
@@
     }
-    mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
+    mod <- new("exgaus", parameters = parameters, fixed = fixed,
+      x = x, y = y, npar = npar)
     mod
@@
 <<>>=
-setMethod("fit","exgaus",
+setMethod("fit", "exgaus",
   function(object, w) {
@@
     y <- object@y
-
-	fit <- gamlss(y~1,weights=w,family=exGAUS(),
-      control=gamlss.control(n.cyc=100,trace=FALSE),
-      mu.start=object@parameters$mu, sigma.start=exp(object@parameters$sigma), 
-      nu.start=exp(object@parameters$nu)
-	)
-
+    fit <- gamlss(y ~ 1, weights = w, family = exGAUS(),
+      control = gamlss.control(n.cyc = 100, trace = FALSE),
+      mu.start = object@parameters$mu,
+      sigma.start = exp(object@parameters$sigma),
+      nu.start = exp(object@parameters$nu))    
     pars <- c(fit$mu.coefficients, fit$sigma.coefficients, 
@@
 rModels[[1]] <- list()
-rModels[[1]][[1]] <- exgaus(speed$rt, pstart=c(5,.1,.1))
-rModels[[1]][[2]] <- GLMresponse(formula=corr~1, data=speed, 
-	family=multinomial(), pstart=c(0.5,0.5))
+rModels[[1]][[1]] <- exgaus(speed$rt, pstart = c(5, 0.1, 0.1))
+rModels[[1]][[2]] <- GLMresponse(formula = corr ~ 1, data = speed,
+  family = multinomial(), pstart = c(0.5, 0.5))
 
 rModels[[2]] <- list()
-rModels[[2]][[1]] <- exgaus(speed$rt, pstart=c(6,.1,.1))
-rModels[[2]][[2]] <- GLMresponse(formula=corr~1, data=speed, 
-	family=multinomial(), pstart=c(.1,.9))
+rModels[[2]][[1]] <- exgaus(speed$rt, pstart = c(6, 0.1, 0.1))
+rModels[[2]][[2]] <- GLMresponse(formula = corr ~ 1, data = speed, 
+  family = multinomial(), pstart = c(0.1, 0.9))
 @
@@
 \code{transInit} function (which produces a transInit model, which also extends 
-the response class): 
-<<>>=
-trstart=c(0.9,0.1,0.1,0.9)
+the `\code{response}' class): 
+<<>>=
+trstart <- c(0.9, 0.1, 0.1, 0.9)
 transition <- list()
-transition[[1]] <- transInit(~Pacc,nst=2, data=speed, 
-	pstart=c(0.9,0.1,0,0))
-transition[[2]] <- transInit(~Pacc,nst=2, data=speed, 
-	pstart=c(0.1,0.9,0,0))
-inMod <- transInit(~1, ns=2, pstart=c(0.1,0.9), data=data.frame(1))
+transition[[1]] <- transInit(~ Pacc, nst = 2, data = speed, 
+  pstart = c(0.9, 0.1, 0, 0))
+transition[[2]] <- transInit(~ Pacc, nst = 2, data = speed, 
+  pstart = c(0.1, 0.9, 0, 0))
+inMod <- transInit(~ 1, ns = 2, pstart = c(0.1, 0.9),
+  data = data.frame(1))
 @
@@
 <<>>=
-mod <- makeDepmix(response=rModels, transition=transition,
-    prior=inMod, stat=FALSE)
-fm <- fit(mod, verbose=FALSE)
+mod <- makeDepmix(response = rModels, transition = transition,
+  prior = inMod, stat = FALSE)
+fm <- fit(mod, verbose = FALSE)
 @
@@
 
-\section[Conclusions and future work]{Conclusions \& future work}
+\section{Conclusions and future work}
 
@@
 it should be noted that more specialized packages are available for
-this such as \pkg{FlexMix} \citep{Leisch2004}.  The package is
+this such as \pkg{flexmix} \citep{Leisch2004,GruenLeisch2008}.  The package is
 intended for modelling (individual) time series data with the aim of
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 
@@
 
-\bibliography{dpx4Rev}
+\bibliography{depmixS4}
 

### 2011-09-16 tags/depmix-0.9.8/man/.Rd2dvi1150/Rd2.tex
Tag for depmix 0.9.8
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@490 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\nonstopmode{}
\documentclass[a4paper]{book}
\usepackage[times,hyper]{Rd}
\usepackage{makeidx}
\makeindex{}
\begin{document}
\chapter*{}
\begin{center}
{\textbf{\huge \R{} documentation}} \par\bigskip{{\Large of \file{-o=depmix.dvi} etc.}}
\par\bigskip{\large \today}
\end{center}
\Rdcontents{\R{} topics documented:}
\HeaderA{depmix}{Fitting Dependent Mixture Models}{depmix}
\aliasA{bootstrap}{depmix}{bootstrap}
\aliasA{fitdmm}{depmix}{fitdmm}
\aliasA{loglike}{depmix}{loglike}
\aliasA{oneliner}{depmix}{oneliner}
\aliasA{posterior}{depmix}{posterior}
\aliasA{summary.fit}{depmix}{summary.fit}
\keyword{models}{depmix}
\begin{Description}\relax
\item[fitdmm] \code{fitdmm} fits mixtures of hidden/latent Markov
models on arbitrary length time series fitdmm mixed categorical and
continuous data.  This includes latent class models (for time series of
length 1).

\item[\code{posterior}] \code{posterior} computes the most likely
latent state sequence for a given dataset and model.
\end{Description}
\begin{Usage}
\begin{verbatim}

        fitdmm(dat, dmm, printlevel = 1, post = TRUE, tdcov = 0, ses
                 = TRUE, method = "nlm", der = 1, vfactor = 15,
                 iterlim = 300, accuracy = "standard", kmst = !dmm$st,
                 postst = TRUE)
        loglike(dat, dmm, tdcov = 0, grad = FALSE, hess = FALSE, set
                 = TRUE, grInd = 0, sca = 1, printlevel = 1)
        posterior(dat,dmm,tdcov=0,printlevel=1)
        bootstrap(object,dat,samples=100, pvalonly=0,...)
        summary.fit(object,precision=3,fd=1,...)
        oneliner(object,precision=3)

\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{dat}] An object (or list of objects) of class \code{md}, see
markovdata.  If dat is a list of objects of class \code{md} a
multigroup model is fitted on these data sets.
\item[\code{dmm}] An object (or a list of objects) of class \code{dmm}, see
dmm.  If dmm is a list of objects of class \code{dmm}, these are taken
to components of a mixture of dmm's model and will be coerced to class
\code{mixdmm}.  In any case, the model that is fitted a multigroup
mixture of dmm's with default ngroups=1 and number of components=1.
\item[\code{printlevel}] \code{printlevel} controls the output provided by the
C-routines that are called to optimize the parameters.  The default of
1 provides minmal output: just the initial and final loglikelihood of
the model.  Setting higher values will provide more output on the
progress the iterations.
\item[\code{post}] By default posteriors are computed, the result of which
can be found in fit\$post.
\item[\code{method}] This is the optimization algorithm that is used. NLM is the 
default method. There is further support for optim and NPSOL.
\item[\code{der}] Specifies whether derivatives are to be used in optimization.
\item[\code{vfactor}] vfactor controls optimization in optim and nlm. Since in 
those routines there is no possibility for enforcing constraints, 
constraints are enforced by adding a penalty term to the loglikelihood. 
The penalty term is printed at the end of optimization if it is not close
enough to zero. This may have several reasons. When parameters are 
estimated at bounds for example. This can be solved by fixing those 
parameters on their boundary values. When this is not acceptable vfactor 
may be increased such that the penalty is larger and the probability that 
they actually hold in the fitted model is correspondingly higher.
\item[\code{tdcov}] Logical, when set to TRUE, given that the model and data have 
covariates, the corresponding parameters will be estimated.
\item[\code{ses}] Logical, determines whether standard errors are computed after
optimization.
\item[\code{iterlim}] The iteration limit for npsol, defaults to 100, which may
be too low for large models.
\item[\code{accuracy}] This argument can be used to set accuracy of optimization 
when using \code{nlm} as optimizer. It can take values "standard" (the 
default), "high" and "best" for increasing levels of accuracy.
\item[\code{grad}] logical; if TRUE the gradients are returned.
\item[\code{hess}] logical; if TRUE the hessian is returned.
\item[\code{set}] Whith the default value TRUE, the data and models parameters
are sent to the C/C++ routines before computing the loglikelihood. When set
is FALSE, this is not done. If an incorrect model was set earlier in the 
C-routines this may cause serious errors and/or crashes.
\item[\code{sca}] If set to -1.0 the negative loglikelihood, gradients and hessian 
are returned.
\item[\code{object}] An object of class \code{fit}, ie the return value of 
fitdmm.
\item[\code{kmst,postst}] These arguments control the generation of starting
values by kmeans and posterior estimates respectively.
\item[\code{grInd}] Logical argument; if TRUE, individual contributions of
each independent realization to the gradient vector will be returned.
\item[\code{fd}] Print the finite difference based standard errors in the summary
if both those and bootstrapped standard errors are available.
\item[\code{samples}] The number of samples to be used in bootstrapping.
\item[\code{pvalonly}] Logical, if 1 only a bootstrapped pvalue is returned and not 
fitted paramaters to compute standard errors, optimization is truncated
when the loglikelihood is better than the original loglikelihood.
\item[\code{precision}] Precision sets the number of digits to be printed in the
summary functions.
\item[\code{...}] Used in summary.
\end{ldescription}
\end{Arguments}
\begin{Details}\relax
The function \code{fitdmm} optimizes the parameters of a mixture of
\code{dmm}s using a general purpose optimization routine subject to linear
and nonlinear constraints on the parameters.
\end{Details}
\begin{Value}
\code{fitdmm} returns an object of class \code{fit} which has a summary
method that prints the summary of the fitted model, and the following fields:

\begin{ldescription}
\item[\code{date,timeUsed,totMem}] The date that the model was fitted, the
time it took to so and the memory usage.
\item[\code{loglike}] The loglikelihood of the fitted model.
\item[\code{aic}] The AIC of the fitted model.
\item[\code{bic}] The BIC of the fitted model.
\item[\code{mod}] The fitted model.
\item[\code{post}] See function posterior for details.

\code{loglike} returns a list of the following:

\item[logl] The loglikelihood.

\item[gr,grset] \code{gr} contains the gradients. \code{grset} is a logical vector
giving information as to which gradients are set, currently all gradients are set 
except the gradients for the mixing proportions.

\item[hs,hsset] \code{hs} contains the hessian. \code{hsset} is a logical giving 
information as to which elements are computed.

\code{posterior} returns lists of the following:

\item[states] A matrix of dimension 2+sum(nstates) by sum(length(ntimes)) containing 
in the first column the a posteriori component, in the second column the a posteriori 
state and in the remaining column the posterior probabilities of all states.

\item[comp] Contains the posterior component number for each independent
realization; all ones for a single component model.

\code{bootstrap} returns an object of class \code{fit} with three extra fields,
the bootstrapped standard errors, bse, a matrix  with goodness-of-fit measures
of the bootstrap samples, ie logl, AIC and BIC and pbetter, which is the 
proportion of bootstrap samples that resulted in better fits than the original
model.

\code{summary.fit} pretty-prints the above fields.

\code{oneliner} returns a vector of loglike, aic, bic, mod$npars,
mod$freepars, date.
\end{ldescription}
\end{Value}
\begin{Note}\relax
The \code{\LinkA{repeated}{repeated}} library
by Jim Lindsey fits hidden markov models.  \code{fitdmm} fits time series of
arbitrary length and mixtures of \code{dmm}s, where, to the best of my
knowledge, other packages are limited due to the different optimization
routines that are commonly used for these types of models (this is
certainly so for categorical data models).
\end{Note}
\begin{Author}\relax
Ingmar Visser
\end{Author}
\begin{References}\relax
Articles
\end{References}
\begin{SeeAlso}\relax
\code{\LinkA{dmm}{dmm}},\code{\LinkA{markovdata}{markovdata}},\code{\LinkA{repeated}{repeated}}
\end{SeeAlso}
\begin{Examples}
\begin{ExampleCode}

# COMBINED RT AND CORRECT/INCORRECT SCORES from a 'switching' experiment

data(rudy)
mod <- dmm(nsta=2,itemt=c(1,2)) # gaussian and binary items
fit1 <- fitdmm(dat=rudy,dmm=mod)
summary(fit1)

# add some constraints using conpat
conpat=rep(1,15)
conpat[1]=0
conpat[14:15]=0
conpat[8:9]=0
# use starting values from the previous model fit, except for the guessing 
# parameters which should really be 0.5
stv=c(1,.896,.104,.084,.916,5.52,.20,.5,.5,6.39,.24,.098,.90,0,1)
mod=dmm(nstates=2,itemt=c("n",2),stval=stv,conpat=conpat)

fit2 <- fitdmm(dat=rudy,dmm=mod)
summary(fit2)

# add covariates to the model to incorporate the fact the accuracy pay off changes per trial
# 2-state model with covariates + other constraints
conpat=rep(1,15)
conpat[1]=0
conpat[8:9]=0
conpat[14:15]=0
conpat[2]=2
conpat[5]=2
stv=c(1,0.9,0.1,0.1,0.9,5.5,0.2,0.5,0.5,6.4,0.25,0.9,0.1,0,1)
tdfix=rep(0,15)
tdfix[2:5]=1
stcov=rep(0,15)
stcov[2:5]=c(-0.4,0.4,0.15,-0.15)

mod<-dmm(nstates=2,itemt=c("n",2),stval=stv,conpat=conpat,tdfix=tdfix,tdst=stcov,modname="twoboth+cov")

fit3 <- fitdmm(dat=rudy,dmm=mod,tdcov=1,der=0,ses=0,vfa=80,accu="best")
summary(fit3)

# split the data into three time series
data(rudy)
r1=markovdata(dat=rudy[1:168,],item=itemtypes(rudy))
r2=markovdata(dat=rudy[169:302,],item=itemtypes(rudy))
r3=markovdata(dat=rudy[303:439,],item=itemtypes(rudy))

# define 2-state model with constraints
conpat=rep(1,15)
conpat[1]=0
conpat[8:9]=0
conpat[14:15]=0
stv=c(1,0.9,0.1,0.1,0.9,5.5,0.2,0.5,0.5,6.4,0.25,0.9,0.1,0.5,0.5)
mod<-dmm(nstates=2,itemt=c("n",2),stval=stv,conpat=conpat)

# define 3-group model with equal transition parameters, and no 
# equalities between the obser parameters
mgr <-mgdmm(dmm=mod,ng=3,trans=TRUE,obser=FALSE)

fitmg <- fitdmm(dat=list(r1,r2,r3),dmm=mgr)
summary(fitmg)

# LEARNING DATA AND MODELS (with absorbing states)

data(mdslow)
summary(mdslow)

# all or none model with error prob in the learned state
fixed = c(0,0,0,1,1,1,1,0,0,0,0)
stv = c(1,1,0,0.07,0.93,0.9,0.1,0.5,0.5,0,1)
allor <- dmm(nstates=2,itemtypes=2,fixed=fixed,stval=stv,modname="All-or-none")
fit4 <- fitdmm(dat=mdslow,dmm=allor)
summary(fit4)

# Concept identification model: learning only after an error

data(mdrat)
summary(mdrat)

# Concept identification model: learning only after an error
st=c(1,1,0,0,0,0.5,0.5,0.5,0.25,0.25,0.8,0.2,1,0,0,1,0.25,0.375,0.375)
# fix some parameters
fx=rep(0,19)
fx[8:12]=1
fx[17:19]=1
# add a couple of constraints
conr1 <- rep(0,19)
conr1[9]=1
conr1[10]=-1
conr2 <- rep(0,19)
conr2[18]=1
conr2[19]=-1
conr3 <- rep(0,19)
conr3[8]=1
conr3[17]=-2
conr=c(conr1,conr2,conr3)
cim <- dmm(nstates=3,itemtypes=2,fixed=fx,conrows=conr,stval=st,modname="CIM")

fit5 <- fitdmm(dat=mdrat,dmm=cim)
summary(fit5)

# define a mixture of the above models ...
mix <- mixdmm(dmm=list(allor,cim),modname="MixAllCim")

data(mdall)

# ... and fit it on the combined data mdall
fit6 <- fitdmm(dat=mdall,dmm=mix)
summary(fit6)

\end{ExampleCode}
\end{Examples}

\printindex{}
\end{document}
